---
title: Team 3 - Lab 1
subtitle: Forecasting with ARIMA and ETS models
author: "Ben Makhlouf"
date: April XX, 2025
output: 
  html_document:
    code-folding: true
    toc: true
    toc_float: true
---

##### Questions to address 

# Broad Goal : Compare forecast accuracy of ARIMA using yearly vs Monthly data. 

# Steps: 
1. Load the data 
   - Filter the data to only be Washington + Oregon
   
2. Test stationarity 
   - Show a quick ACF and PACF plots.
3. Fit an ARIMA model using Autoarima to monthly and yearly training data (80%)
4. Forecast using the fitted ARIMA model (with plot)
5. Run diagnostics 
6. Forecast using accuracy()
7. Compare accuracy of the two models using RMSE, MAE, MAPE, and MASE

# First, load the data 

```{r message=FALSE, warning=FALSE}
# if you need to install
# install.packages('atsalibrary', repos = c('https://atsa-es.r-universe.dev', 'https://cloud.r-project.org'))
library(atsalibrary) # This is ALL the data 

# This is the monthly data 
chinook.month <- chinook.month %>%
  mutate(Date = as.Date(paste(Year, Month, "01", sep = "-"), format = "%Y-%b-%d"))

chinook.monthWA<- chinook.month %>%
  filter(State %in% c("WA"))

# Plot to check we only have Washington and Oregon. Some missing values, thats ok
ggplot(chinook.monthWA, aes(x = Date, y = log.metric.tons)) +
  geom_line() +
  facet_wrap(~ State, ncol = 1) +  # 3 panels vertically
  labs(title = "Monthly Chinook Salmon Landings",
       x = "Date",
       y = "Log(Metric Tons)") +
  theme_minimal()


# This is the yearly data 
# for some reason they type out the full name. Weird. Ayways... 
chinook.year <- chinook.year

chinook.yearWA<- chinook.year %>%
  filter(State %in% c("Washington"))

ggplot(chinook.yearWA, aes(x = Year, y = log.metric.tons)) +
  geom_line() +
  facet_wrap(~ State, scales = "free_y") +
  labs(
    title = "Chinook Landings by State",
    x = "Year",
    y = "Log(Metric Tons)"
  ) +
  theme_minimal()

```


```{r}
### Create a forecast using ARIMA and ETS models for Washington using both monthly and yearly data 

# Forecast package
library(forecast)

# Convert yearly Washington data to time series, both starting at 1990
wa_yearly <- chinook.yearWA %>% 
  arrange(Year) %>%
  pull(log.metric.tons)

# what is the first date 
firstyear<- chinook.yearWA %>%
  arrange(Year) %>%
  pull(Year) %>%
  min()

print(firstyear)

wa_yearly_ts <- ts(wa_yearly, start = 1950, frequency = 1)


# Create time series for monthly Washington data

firstyearmonth<- chinook.monthWA %>%
  arrange(Date) %>%
  pull(Date) %>%
  min()

print(firstyearmonth)

wa_monthly <- chinook.monthWA %>%
  arrange(Date) %>%
  pull(log.metric.tons)
wa_monthly_ts <- ts(wa_monthly, start = c(1990,1), frequency = 12)

plot(wa_monthly_ts)
plot(wa_yearly_ts)

# lets trim the data so they both start 1990 and end in 2015 
wa_monthly_ts <- window(wa_monthly_ts, start = c(1990, 1), end = c(2015, 12))
wa_yearly_ts <- window(wa_yearly_ts, start = 1990, end = 2015)

#Plot to verify 
plot(wa_monthly_ts, main = "Monthly Chinook Salmon Landings in Washington")
plot(wa_yearly_ts, main = "Yearly Chinook Salmon Landings in Washington")

## Split the data into testing and training data. Testing is up to 2010 
wa_monthly_train <- window(wa_monthly_ts, end = c(2010, 12))
wa_monthly_test <- window(wa_monthly_ts, start = c(2011, 1))

wa_yearly_train <- window(wa_yearly_ts, end = 2010)
wa_yearly_test <- window(wa_yearly_ts, start = 2011)

plot(wa_monthly_test, main = "Monthly Chinook Salmon Landings in Washington - Training Data")
plot(wa_yearly_test, main = "Yearly Chinook Salmon Landings in Washington - Training Data")

####### Now, we're ready to forecast 

```

```{r}

# use autoarima to fit arima 
monthfit<- auto.arima(wa_monthly_train)
yearfit<- auto.arima(wa_yearly_train)

# look at the models 
summary(monthfit)
## ARIMA(1,0,0)(0,1,2)[12]
# non seasonal component: Arima(1,0,0) # AR Model with no differencing 
# seasonal component: Arima (0,1,2)[12] - 1 seasonal difference and 2 seasonal moving average terms. 
# 12 month seasonality 
# Drift term (trend)
# slight long-term decline 


summary(yearfit)
# ARIMA(1,0,0 with non-zero mean)
# non seasonal AR 1 model 
# No seasonal component 
# Stronger correlation with the previous years data (.669)
# Also suggests slightly downward trend (-0.0037)

# fit Holt to year 
fitYearHolt<- forecast::holt(wa_yearly_train)


```

#Diagnostics 
```{r}

forecast::ndiffs(wa_monthly_ts, test = "kpss")
forecast::nsdiffs(wa_monthly_ts, test = "ads")

```




Great, model is trained up. Lets forecast 

```{r}
YearFit<- forecast::auto.arima(wa_yearly_train)
MonthFit<- forecast::auto.arima(wa_monthly_train)

yearforecast<- forecast::forecast(YearFit)
monthforecast<- forecast::forecast(MonthFit, h = 12) 

plot(yearforecast, main = "Yearly Chinook Salmon Landings in Washington")
#points(wa_yearly_test) not working... 

plot(monthforecast, main = "Monthly Chinook Salmon Landings in Washington")
#points(wa_monthly_test) not working.. 

# Use Holt and Holt-winters to test ETS 

# Holt-Winters for monthly data
#fitMonthHoltWint<- forecast::holt(wa_monthly_train)
#MonthHWForecast<- forecast(fitMonthHoltWint, h = 12)

# fit holt to yearly 
fitYearHolt<- forecast::holt(wa_yearly_train)
YearHoltForecast<- forecast(fitYearHolt)

plot(YearHoltForecast, main = "Yearly Chinook Salmon Landings in Washington - Holt")

```
Testing Forecast Fit 
```{r}

########### 
########### ARIMA 
### 

# Month
accuracy(monthforecast, wa_monthly_test)

#  ME      RMSE       MAE      MPE     MAPE      MASE        ACF1 Theil's U
# Training set -0.02184778 0.9954557 0.6554148     -Inf      Inf 0.9092448 -0.07884648        NA
# Test set      0.14756882 0.9190212 0.6894202 22.77132 43.16707 0.9564199  0.22153911 0.6305218

# Year
accuracy(yearforecast, wa_yearly_test)

#                      ME      RMSE       MAE        MPE     MAPE      MASE         ACF1 Theil's U
# Training set -0.0247061 0.2505295 0.2007125 -0.4458757 2.735247 0.9375834 -0.007278968        NA
# Test set      0.3564054 0.4265469 0.3564054  4.4324755 4.432476 1.6648677  0.469532636  2.401485

# Month 
#NA, NOT WORKING 

# Year 
accuracy(YearHoltForecast, wa_yearly_test)

#                      ME      RMSE       MAE        MPE     MAPE      MASE         ACF1 Theil's U
# Training set -0.0247061 0.2505295 0.2007125 -0.4458757 2.735247 0.9375834 -0.007278968        NA
# Test set      0.3564054 0.4265469 0.3564054  4.4324755 4.432476 1.6648677  0.469532636  2.401485


```








# Create ARIMA models
wa_yearly_arima <- auto.arima(wa_yearly_ts)
wa_monthly_arima <- auto.arima(wa_monthly_ts, seasonal = TRUE)

# Create ETS models
wa_yearly_ets <- ets(wa_yearly_ts)
wa_monthly_ets <- ets(wa_monthly_ts)

# Generate fo






```




# Data

Discuss what parts of the data sets you will work with.

# Question your team will address

What will your team do with the data? You can do many things with the data. The only constraints are that you

* fit ARIMA and ETS models. You can fit other models in addition to these if you want (lm, glm, gam, etc)
* do diagnostics for ARIMA and ETS models
* make forecasts
* test how good your forecasts or compare forecasts (many options here)

Example, "Our team decided to compare the accuracy of forecasts using best fit ARIMA models for WA, OR and CA commercial yearly landings. Our question is whether forecast accuracy is different for simple (1st order) versus 2nd to 4th order ARIMA models."

# Method you will use

## Initial plan

Describe what you plan to do to address your question.

Example, "We will fit ARIMA models to 1960-1980 data and make 1981-1985  forecasts (5 year). We will compare the accuracy of the forecasts to the actual values. We will measure accuracy with mean squared error. We will use the forecast package to fit the ARIMA models with different AR orders (1st up to 4th order)."

## What you actually did

Example, "We were able to do our plan fairly quickly, so after writing the basic code. We compared the accuracy of forecasts for different time periods using 20-years for training and 5-year forecasts each time. We compared the RMSE, MAE, and MAPE accuracy measures."

# Diagnostics and preliminary exploration

## Plot the data

Plot the data and discuss any obvious problems with stationarity from your visual test.


## Use ACF and PACF

Use the ACF and PACF plots to explore the time series. What did you learn? Also try decomposition.

## Test for stationarity

Run tests using `forecast::ndiffs` and discuss any stationarity issues that are revealed.

# Results

# Discussion

# Description of each team member's contributions

Example: "All team members helped decide on the goal and ran the analyses for the individual regions. Team members 2 & 3 wrote most of the code for the analysis of the regions. Team member 4 researched approaches for measuring accuracy of forecasts in [Hyndman & Athanasopoulos[OTexts.com/fpp2] and team member 2 added code for that to the methods. Team member 4 also researched tests for stationarity and worked with team member 2 to code that up. Team member 1 worked on the plotting section of the report using and adapting code that team member 3 wrote. All team members helped edit the report and wrote the discussion together."

