---
title: Lab 1 Forecasting with ARIMA and ETS models
author: Eli Holmes
output: 
  html_document:
    code-folding: true
    toc: true
    toc_float: true
---

```{r include=FALSE}
library(tidyverse)
options(dplyr.summarise.inform = FALSE)
knitr::opts_chunk$set(warning = FALSE)
# library(forecast)
```

#We decided to look at just Washington data for this lab. To do so, we isolate yearly and monthly data from just the Washington subset. 

```{r}
# Seperate the monthly and yearly data 
library(atsalibrary) # This is ALL the data 

# This is the monthly data 
chinook.month <- chinook.month %>%
  mutate(Date = as.Date(paste(Year, Month, "01", sep = "-"), format = "%Y-%b-%d"))

chinook.month <- chinook.month %>%
  filter(State %in% c("WA"))

chinook.year <- chinook.year 

chinook.year_filtered <- chinook.year %>%
  filter(State == "Washington")

ggplot(chinook.year_filtered, aes(x = Year, y = log.metric.tons)) +
  geom_line() +
  facet_wrap(~ State, scales = "free_y") +
  labs(
    title = "Chinook Landings by State",
    x = "Year",
    y = "Log(Metric Tons)"
  ) +
  theme_minimal()


# Quick plot of year 

ggplot(chinook.year_filtered, aes(x = Year, y = log.metric.tons)) +
  geom_line() +
  facet_wrap(~ State, ncol = 1) +  # 3 panels vertically
  labs(title = "Monthly Chinook Salmon Landings",
       x = "Date",
       y = "Log(Metric Tons)") +
  theme_minimal()

# Quick plot of month 

ggplot(chinook.month, aes(x = Date, y = log.metric.tons)) +
  geom_line() +
  facet_wrap(~ State, ncol = 1) +  # 3 panels vertically
  labs(title = "Monthly Chinook Salmon Landings",
       x = "Date",
       y = "Log(Metric Tons)") +
  theme_minimal()

```


# Step2,  Check stationarity and differencing 

```{r}

yearly <- chinook.year_filtered %>% 
  arrange(Year) %>%
  pull(log.metric.tons)

yearly_ts <- ts(yearly, start = 1950, end = 2016, frequency = 1)


monthly<- chinook.month %>%
  #change log metric tons inf to NA
  mutate(log.metric.tons = ifelse(log.metric.tons == Inf, NA, log.metric.tons)) %>%
  mutate(log.metric.tons = ifelse(log.metric.tons == -Inf, NA, log.metric.tons)) %>%
  filter(!is.na(log.metric.tons))

monthly <- monthly %>%
  arrange(Date) %>%
  pull(log.metric.tons)

# Are there NAs in monthly 
# remove NA

monthly_ts <- ts(monthly, start = c(1990,1), frequency = 12)


##############################

# Check the need to difference the data 

adf.test(yearly_ts)

# Augmented Dickey-Fuller Test
# 
# data:  yearly_ts
# Dickey-Fuller = -2.5627, Lag order = 4, p-value = 0.3467
# alternative hypothesis: stationary

# Not stationary, p value greater than .01 


adf.test(monthly_ts)

# Augmented Dickey-Fuller Test
# 
# data:  monthly_ts
# Dickey-Fuller = -9.6198, Lag order = 7, p-value = 0.01
# alternative hypothesis: stationary

# These data ARE Stationary 

#############################

# wE can run the ndiff to see how many differences we would need to get to stationarity 

ndiffs(yearly_ts)

# [1], we need to difference ONCE, we expect the middle number in autoarima to be 1 

```
### Step 3, explore correlation and autocorrelation using ACF, PACF 

```{r}
### ACF PACF HERE 
```


```{r}

# Create training and test sets (80% training, 20% test)
yearly_train_end <- floor(0.8 * length(yearly_ts))
yearly_train <- window(yearly_ts, end = time(yearly_ts)[yearly_train_end])
yearly_test <- window(yearly_ts, start = time(yearly_ts)[yearly_train_end + 1])

monthly_train_end <- floor(0.8 * length(monthly_ts))
monthly_train <- window(monthly_ts, end = time(monthly_ts)[monthly_train_end])
monthly_test <- window(monthly_ts, start = time(monthly_ts)[monthly_train_end + 1])

# Delete monthly_train_end
rm(monthly_train_end)
# Delete yearly_train_end
rm(yearly_train_end)

# Fit ARIMA model to training data
yearly_arima <- auto.arima(yearly_train)
# Series: yearly_train 
# ARIMA(0,1,0) 
# 
# sigma^2 = 0.04824:  log likelihood = 5.04
# AIC=-8.07   AICc=-7.99   BIC=-6.12

monthly_arima <- auto.arima(monthly_train)

# Series: monthly_train 
# ARIMA(2,0,2) with non-zero mean 
# 
# Coefficients:
#          ar1      ar2      ma1     ma2    mean
#       1.6522  -0.9335  -1.1833  0.4651  3.2705
# s.e.  0.0423   0.0444   0.1831  0.1728  0.0830
# 
# sigma^2 = 2.048:  log likelihood = -526.49
# AIC=1064.98   AICc=1065.27   BIC=1087.14


# Fit ETS models
yearly_ets <- ets(yearly_train)

# ETS(A,N,N) 
# 
# Call:
# ets(y = yearly_train)
# 
#   Smoothing parameters:
#     alpha = 0.9999 
# 
#   Initial states:
#     l = 8.2942 
# 
#   sigma:  0.2218
# 
#      AIC     AICc      BIC 
# 54.74317 55.23297 60.65405 

monthly_ets <- ets(monthly_train)

# ETS(A,N,N) 
# 
# Call:
# ets(y = monthly_train)
# 
#   Smoothing parameters:
#     alpha = 0.9999 
# 
#   Initial states:
#     l = 3.2657 
# 
#   sigma:  1.8451
# 
#      AIC     AICc      BIC 
# 2058.871 2058.953 2069.952 

```

Forecast
```{r}

# Generate forecasts for the test period
yearly_arima_forecast <- forecast(yearly_arima, h = length(yearly_test))
yearly_ets_forecast <- forecast(yearly_ets, h = length(yearly_test))

monthly_arima_forecast <- forecast(monthly_arima, h = length(monthly_test))
monthly_ets_forecast <- forecast(monthly_ets, h = length(monthly_test))

```


## Accuracy metrics 
```{r}
# Evaluate forecast accuracy
yearly_arima_accuracy <- accuracy(yearly_arima_forecast, yearly_test)
yearly_ets_accuracy <- accuracy(yearly_ets_forecast, yearly_test)

monthly_arima_accuracy <- accuracy(monthly_arima_forecast, monthly_test)
monthly_ets_accuracy <- accuracy(monthly_ets_forecast, monthly_test)

# Plot forecasts
autoplot(yearly_arima_forecast)
autoplot(yearly_ets_forecast)

autoplot(monthly_arima_forecast)
autoplot(monthly_ets_forecast)

```




# What we need still 
######## initial plan (EMMA)
####### What we actually did (EMMA)
####### Questions we address (EMMA)

###### Use ACF and PACF (TINA) 

###  Results and Discussion (TINA: FIT RESULTS , 
                             Ben: Accuracy Results)
                             
### Add in and comment out the accuracy results (BEN)

# Team members contributions 


