---
title: Lab 1 Forecasting with ARIMA and ETS models
author: Eli Holmes
output: 
  html_document:
    code-folding: true
    toc: true
    toc_float: true
---

```{r include=FALSE}
library(tidyverse)
options(dplyr.summarise.inform = FALSE)
knitr::opts_chunk$set(warning = FALSE)
# library(forecast)
```

#We decided to look at just Washington data for this lab. To do so, we isolate yearly and monthly data from just the Washington subset. 

# Step1,  load data and filter to Washington with basic plots

```{r}
# Seperate the monthly and yearly data 
library(atsalibrary) # This is ALL the data 

# This is the monthly data 
chinook.month <- chinook.month %>%
  mutate(Date = as.Date(paste(Year, Month, "01", sep = "-"), format = "%Y-%b-%d"))

chinook.month <- chinook.month %>%
  filter(State %in% c("WA"))

chinook.year <- chinook.year 

chinook.year_filtered <- chinook.year %>%
  filter(State == "Washington")

# Quick plot of year 

ggplot(chinook.year_filtered, aes(x = Year, y = log.metric.tons)) +
  geom_line() +
  facet_wrap(~ State, ncol = 1) +  # 3 panels vertically
  labs(title = "Monthly Chinook Salmon Landings",
       x = "Date",
       y = "Log(Metric Tons)") +
  theme_minimal()

# Quick plot of month 

ggplot(chinook.month, aes(x = Date, y = log.metric.tons)) +
  geom_line() +
  facet_wrap(~ State, ncol = 1) +  # 3 panels vertically
  labs(title = "Monthly Chinook Salmon Landings",
       x = "Date",
       y = "Log(Metric Tons)") +
  theme_minimal()

```


# Step2,  check stationarity and differencing 

```{r}

yearly <- chinook.year_filtered %>% 
  arrange(Year) %>%
  pull(log.metric.tons)

yearly_ts <- ts(yearly, start = 1950, end = 2016, frequency = 1)


monthly<- chinook.month %>%
  #change log metric tons inf to NA
  mutate(log.metric.tons = ifelse(log.metric.tons == Inf, NA, log.metric.tons)) %>%
  mutate(log.metric.tons = ifelse(log.metric.tons == -Inf, NA, log.metric.tons)) %>%
  filter(!is.na(log.metric.tons))

monthly <- monthly %>%
  arrange(Date) %>%
  pull(log.metric.tons)

# Are there NAs in monthly 
# remove NA

monthly_ts <- ts(monthly, start = c(1990,1), frequency = 12)


##############################

# Check the need to difference the data 

adf.test(yearly_ts)

# Augmented Dickey-Fuller Test
# 
# data:  yearly_ts
# Dickey-Fuller = -2.5627, Lag order = 4, p-value = 0.3467
# alternative hypothesis: stationary

# Not stationary, p value greater than .01 


adf.test(monthly_ts)

# Augmented Dickey-Fuller Test
# 
# data:  monthly_ts
# Dickey-Fuller = -9.6198, Lag order = 7, p-value = 0.01
# alternative hypothesis: stationary

# These data ARE Stationary 

#############################

# wE can run the ndiff to see how many differences we would need to get to stationarity 

ndiffs(yearly_ts)

# [1], we need to difference ONCE, we expect the middle number in autoarima to be 1 

```
### Step 3, explore correlation and autocorrelation using ACF and PACF 

```{r}
### ACF PACF HERE
acf(yearly_ts, lag.max = 20, main = "ACF of Yearly Chinook Salmon Landings in Washington")
# The ACF plot shows a tails off structure
pacf(yearly_ts, lag.max = 20, main = "PACF of Yearly Chinook Salmon Landings in Washington")
# The PACF plot show a cuts off structure after lag=1
# These are AR(1) model characteristics
# Combining with results from the above stationary test, we expect an ARIMA(1,1,0) model to fit the yearly Chinook Salmon Landings in Washington

acf(monthly_ts, lag.max = 36, main = "ACF of Monthly Chinook Salmon Landings in Washington")
# We observe a clear deterministic seasonal pattern and expect a big negative coefficient for AR component
pacf(monthly_ts, lag.max = 36, main = "PACF of Monthly Chinook Salmon Landings in Washington")
# Several correlation patterns are detected and there is a strong correlation with lag=2
# Choosing the optimal p and q value may need to be done by comparing AIC and BIC values, which is expected to be conducted in auto.arima function
```
### Step 4, fit ARIMA and ETS models

```{r}

# Create training and test sets (80% training, 20% test)
yearly_train_end <- floor(0.8 * length(yearly_ts))
yearly_train <- window(yearly_ts, end = time(yearly_ts)[yearly_train_end])
yearly_test <- window(yearly_ts, start = time(yearly_ts)[yearly_train_end + 1])

monthly_train_end <- floor(0.8 * length(monthly_ts))
monthly_train <- window(monthly_ts, end = time(monthly_ts)[monthly_train_end])
monthly_test <- window(monthly_ts, start = time(monthly_ts)[monthly_train_end + 1])

# Delete monthly_train_end
rm(monthly_train_end)
# Delete yearly_train_end
rm(yearly_train_end)

# Fit ARIMA model to training data
yearly_arima <- auto.arima(yearly_train)
# Series: yearly_train 
# The model result suggests an random walk ARIMA(0,1,0) instead of ARIMA(1,1,0), because the AR(1) coefficient might be small or not statistically significant
# 
# sigma^2 = 0.04824:  log likelihood = 5.04
# AIC=-8.07   AICc=-7.99   BIC=-6.12

monthly_arima <- auto.arima(monthly_train)

# Series: monthly_train 
# ARIMA(2,0,2) with non-zero mean 
# 
# Coefficients:
#          ar1      ar2      ma1     ma2    mean
#       1.6522  -0.9335  -1.1833  0.4651  3.2705
# s.e.  0.0423   0.0444   0.1831  0.1728  0.0830
# 
# sigma^2 = 2.048:  log likelihood = -526.49
# AIC=1064.98   AICc=1065.27   BIC=1087.14


# Fit ETS models
yearly_ets <- ets(yearly_train)

# ETS(A,N,N) 
# 
# Call:
# ets(y = yearly_train)
# 
#   Smoothing parameters:
#     alpha = 0.9999 
# 
#   Initial states:
#     l = 8.2942 
# 
#   sigma:  0.2218
# 
#      AIC     AICc      BIC 
# 54.74317 55.23297 60.65405 

monthly_ets <- ets(monthly_train)

# ETS(A,N,N) 
# 
# Call:
# ets(y = monthly_train)
# 
#   Smoothing parameters:
#     alpha = 0.9999 
# 
#   Initial states:
#     l = 3.2657 
# 
#   sigma:  1.8451
# 
#      AIC     AICc      BIC 
# 2058.871 2058.953 2069.952 

```
### Step 5, model performance residual diagnostics
```{r}
checkresiduals(yearly_arima)
# ACF of residuals shows no significant autocorrelation and looks like a normal distribution, indicating a very good fit
checkresiduals(monthly_arima)
# ACF of residuals have two autocorrelation spikes but could be random weird thing and looks like a normal distribution, indicating a relative good fit
checkresiduals(yearly_ets)
# ACF of residuals shows no significant autocorrelation and looks like a normal distribution, indicating a good fit
checkresiduals(monthly_ets)
# ACF of residuals shows several significant autocorrelations though looks like a normal distribution, indicating a poor fit

## In general, the ARIMA models seem to fit better than the ETS models for both yearly and monthly data.
```


### Step 6, forecast
```{r}

# Generate forecasts for the test period
yearly_arima_forecast <- forecast(yearly_arima, h = length(yearly_test))
yearly_ets_forecast <- forecast(yearly_ets, h = length(yearly_test))

monthly_arima_forecast <- forecast(monthly_arima, h = length(monthly_test))
monthly_ets_forecast <- forecast(monthly_ets, h = length(monthly_test))

# Plot forecasts
plot1<-autoplot(yearly_arima_forecast)
plot2<-autoplot(yearly_ets_forecast)

plot3<-autoplot(monthly_arima_forecast)
plot4<-autoplot(monthly_ets_forecast)

# Put into a 4 plot panel 
library(gridExtra)
library(grid)
library(ggplot2)

grid.arrange(plot1, plot2, plot3, plot4, ncol = 2)



```


## Accuracy metrics 
```{r}
# Evaluate forecast accuracy
yearly_arima_accuracy <- accuracy(yearly_arima_forecast, yearly_test)

#                     ME      RMSE       MAE        MPE     MAPE      MASE      ACF1 Theil's U
# Training set -0.01046748 0.2175570 0.1732156 -0.1712576 2.225795 0.9820193 0.0292464        NA
# Test set     -0.07686666 0.2753066 0.2363580 -1.1236585 3.107135 1.3399954 0.5697271  1.139735

yearly_ets_accuracy <- accuracy(yearly_ets_forecast, yearly_test)

#                     ME      RMSE       MAE        MPE     MAPE      MASE       ACF1
# Training set -0.01062423 0.2175548 0.1730609 -0.1731526 2.223930 0.9811418 0.02860653
# Test set     -0.07683952 0.2752990 0.2363503 -1.1233035 3.107024 1.3399514 0.56972710
             
monthly_arima_accuracy <- accuracy(monthly_arima_forecast, monthly_test)

#                     ME     RMSE       MAE  MPE MAPE      MASE         ACF1 Theil's U
# Training set -0.0620450 1.314323 0.9435214 -Inf  Inf 0.8521606 -0.001970859        NA
# Test set     -0.9640865 3.266402 2.5277314 -Inf  Inf 2.2829721  0.737652808         0

monthly_ets_accuracy <- accuracy(monthly_ets_forecast, monthly_test)

#                       ME     RMSE      MAE  MPE MAPE     MASE      ACF1 Theil's U
# Training set  0.0008512003 1.653080 1.088517 -Inf  Inf 0.983116 0.3694314        NA
# Test set     -1.8938374286 3.914736 2.998623 -Inf  Inf 2.708267 0.7424152         0

```

The accuracy metrics show that for yearly data, ARIMA and ETS models performed nearly the same with a MAPE value of about 3.1%, indicating reasonably accurate forecasts. For the monthly data, the ARIMA model outperformed ETS with lower RMSE (3.27 vs 3.91) and MAE (2.53 vs 3.00)

Visually, the plots suggest better monthly performance but the numerical metrics indicate higher absolute errors for monthly forecasts compared to yearly ones. It would be interesting to compare how each of these predictions preform on test data from the other dataset (i.e. do predictions made from monthly data preform better for yearly data)


################ 
################ Team Member Contributions 

BM, EC, and TT all contributed code and analysis to the final analysis. EC wrote details about the initial plan and questions we addressed. TT added details about checking the residuals and the ACF/PACF plots, and wrote details about the results and discussion. BM wrote and organized final code and added details about Accuracy results. 





# What we need still 
######## initial plan (EMMA)
####### What we actually did (EMMA)
####### Questions we address (EMMA)

###### Use ACF and PACF (TINA) 

###  Results and Discussion (TINA: FIT RESULTS , 
                             Ben: Accuracy Results) DONE 
                             
### Add in and comment out the accuracy results (BEN) DONE

# Team members contributions Done-ish 




