---
title: Lab 1 Forecasting with ARIMA and ETS models
author: Eli Holmes
output: 
  html_document:
    code-folding: true
    toc: true
    toc_float: true
---

```{r include=FALSE}
library(tidyverse)
options(dplyr.summarise.inform = FALSE)
knitr::opts_chunk$set(warning = FALSE)
# library(forecast)
```

#Initial Plan and What Actually Happened
We will compare the accuracy of ARIMA models in yearly vs. monthly data, answering the question: Is the accuracy of monthly or yearly data better? To do this we will look only at the Washington data.
To begin, we will test the stationarity of the yearly data and the monthly data.
If either are not stationary, we will use differencing to make them stationary for the purposes of this analysis. After the data are stationary, we will fit an ARIMA and ETS model to the monthly data using AutoARIMA on the earliest 80% of the data.
This will allow use to compare the accuracy of the ARIMA model in predicting the last 20% of the data to the actual data. After this is done, we will do the same to the yearly data. Then, we can use the accuracy function to compare the two methods. 

We loaded the data and wrangled it, filtering it to only include the Washington data. This allowed us to compare between both the yearly and monthly data. From there we used the auto.arima() function, which auto-checked for stationarity for us. We needed to make the yearly data stationary by differencing it once. Then, we fit both ETS and arima models to both the monthly and yearly catch data from Washington. After fitting the models, we concluded that the monthly data forecast was more accurate than the yearly data forecast.`


#Question Addressed
1) Is an ARIMA model and/or an ETS model adequate for using monthly data and yearly data?
2) Is the accuracy of forecasting with monthly data or yearly data "better"?


```{r}
# Seperate the monthly and yearly data 
library(atsalibrary) # This is ALL the data 

# This is the monthly data 
chinook.month <- chinook.month %>%
  mutate(Date = as.Date(paste(Year, Month, "01", sep = "-"), format = "%Y-%b-%d"))

chinook.month <- chinook.month %>%
  filter(State %in% c("WA"))

chinook.year <- chinook.year 

chinook.year_filtered <- chinook.year %>%
  filter(State == "Washington")

# Quick plot of year 

ggplot(chinook.year_filtered, aes(x = Year, y = log.metric.tons)) +
  geom_line() +
  facet_wrap(~ State, ncol = 1) +  # 3 panels vertically
  labs(title = "Monthly Chinook Salmon Landings",
       x = "Date",
       y = "Log(Metric Tons)") +
  theme_minimal()

# Quick plot of month 

ggplot(chinook.month, aes(x = Date, y = log.metric.tons)) +
  geom_line() +
  facet_wrap(~ State, ncol = 1) +  # 3 panels vertically
  labs(title = "Monthly Chinook Salmon Landings",
       x = "Date",
       y = "Log(Metric Tons)") +
  theme_minimal()

```


# Step2,  Check stationarity and differencing 

```{r}

yearly <- chinook.year_filtered %>% 
  arrange(Year) %>%
  pull(log.metric.tons)

yearly_ts <- ts(yearly, start = 1950, end = 2016, frequency = 1)


monthly<- chinook.month %>%
  #change log metric tons inf to NA
  mutate(log.metric.tons = ifelse(log.metric.tons == Inf, NA, log.metric.tons)) %>%
  mutate(log.metric.tons = ifelse(log.metric.tons == -Inf, NA, log.metric.tons)) %>%
  filter(!is.na(log.metric.tons))

monthly <- monthly %>%
  arrange(Date) %>%
  pull(log.metric.tons)

# Are there NAs in monthly 
# remove NA

monthly_ts <- ts(monthly, start = c(1990,1), frequency = 12)


##############################

# Check the need to difference the data 

adf.test(yearly_ts)

# Augmented Dickey-Fuller Test
# 
# data:  yearly_ts
# Dickey-Fuller = -2.5627, Lag order = 4, p-value = 0.3467
# alternative hypothesis: stationary

# Not stationary, p value greater than .01 


adf.test(monthly_ts)

# Augmented Dickey-Fuller Test
# 
# data:  monthly_ts
# Dickey-Fuller = -9.6198, Lag order = 7, p-value = 0.01
# alternative hypothesis: stationary

# These data ARE Stationary 

#############################

# wE can run the ndiff to see how many differences we would need to get to stationarity 

ndiffs(yearly_ts)

# [1], we need to difference ONCE, we expect the middle number in autoarima to be 1 

```
### Step 3, explore correlation and autocorrelation using ACF, PACF 

```{r}
### ACF PACF HERE 
```


```{r}

# Create training and test sets (80% training, 20% test)
yearly_train_end <- floor(0.8 * length(yearly_ts))
yearly_train <- window(yearly_ts, end = time(yearly_ts)[yearly_train_end])
yearly_test <- window(yearly_ts, start = time(yearly_ts)[yearly_train_end + 1])

monthly_train_end <- floor(0.8 * length(monthly_ts))
monthly_train <- window(monthly_ts, end = time(monthly_ts)[monthly_train_end])
monthly_test <- window(monthly_ts, start = time(monthly_ts)[monthly_train_end + 1])

# Delete monthly_train_end
rm(monthly_train_end)
# Delete yearly_train_end
rm(yearly_train_end)

# Fit ARIMA model to training data
yearly_arima <- auto.arima(yearly_train)
# Series: yearly_train 
# ARIMA(0,1,0) 
# 
# sigma^2 = 0.04824:  log likelihood = 5.04
# AIC=-8.07   AICc=-7.99   BIC=-6.12

monthly_arima <- auto.arima(monthly_train)

# Series: monthly_train 
# ARIMA(2,0,2) with non-zero mean 
# 
# Coefficients:
#          ar1      ar2      ma1     ma2    mean
#       1.6522  -0.9335  -1.1833  0.4651  3.2705
# s.e.  0.0423   0.0444   0.1831  0.1728  0.0830
# 
# sigma^2 = 2.048:  log likelihood = -526.49
# AIC=1064.98   AICc=1065.27   BIC=1087.14


# Fit ETS models
yearly_ets <- ets(yearly_train)

# ETS(A,N,N) 
# 
# Call:
# ets(y = yearly_train)
# 
#   Smoothing parameters:
#     alpha = 0.9999 
# 
#   Initial states:
#     l = 8.2942 
# 
#   sigma:  0.2218
# 
#      AIC     AICc      BIC 
# 54.74317 55.23297 60.65405 

monthly_ets <- ets(monthly_train)

# ETS(A,N,N) 
# 
# Call:
# ets(y = monthly_train)
# 
#   Smoothing parameters:
#     alpha = 0.9999 
# 
#   Initial states:
#     l = 3.2657 
# 
#   sigma:  1.8451
# 
#      AIC     AICc      BIC 
# 2058.871 2058.953 2069.952 

```

Forecast
```{r}

# Generate forecasts for the test period
yearly_arima_forecast <- forecast(yearly_arima, h = length(yearly_test))
yearly_ets_forecast <- forecast(yearly_ets, h = length(yearly_test))

monthly_arima_forecast <- forecast(monthly_arima, h = length(monthly_test))
monthly_ets_forecast <- forecast(monthly_ets, h = length(monthly_test))

# Plot forecasts
plot1<-autoplot(yearly_arima_forecast)
plot2<-autoplot(yearly_ets_forecast)

plot3<-autoplot(monthly_arima_forecast)
plot4<-autoplot(monthly_ets_forecast)

# Put into a 4 plot panel 
library(gridExtra)
library(grid)
library(ggplot2)

grid.arrange(plot1, plot2, plot3, plot4, ncol = 2)



```


## Accuracy metrics 
```{r}
# Evaluate forecast accuracy
yearly_arima_accuracy <- accuracy(yearly_arima_forecast, yearly_test)

#                     ME      RMSE       MAE        MPE     MAPE      MASE      ACF1 Theil's U
# Training set -0.01046748 0.2175570 0.1732156 -0.1712576 2.225795 0.9820193 0.0292464        NA
# Test set     -0.07686666 0.2753066 0.2363580 -1.1236585 3.107135 1.3399954 0.5697271  1.139735

yearly_ets_accuracy <- accuracy(yearly_ets_forecast, yearly_test)

#                     ME      RMSE       MAE        MPE     MAPE      MASE       ACF1
# Training set -0.01062423 0.2175548 0.1730609 -0.1731526 2.223930 0.9811418 0.02860653
# Test set     -0.07683952 0.2752990 0.2363503 -1.1233035 3.107024 1.3399514 0.56972710
             
monthly_arima_accuracy <- accuracy(monthly_arima_forecast, monthly_test)

#                     ME     RMSE       MAE  MPE MAPE      MASE         ACF1 Theil's U
# Training set -0.0620450 1.314323 0.9435214 -Inf  Inf 0.8521606 -0.001970859        NA
# Test set     -0.9640865 3.266402 2.5277314 -Inf  Inf 2.2829721  0.737652808         0

monthly_ets_accuracy <- accuracy(monthly_ets_forecast, monthly_test)

#                       ME     RMSE      MAE  MPE MAPE     MASE      ACF1 Theil's U
# Training set  0.0008512003 1.653080 1.088517 -Inf  Inf 0.983116 0.3694314        NA
# Test set     -1.8938374286 3.914736 2.998623 -Inf  Inf 2.708267 0.7424152         0

```

The accuracy metrics show that for yearly data, ARIMA and ETS models performed nearly the same with a MAPE value of about 3.1%, indicating reasonably accurate forecasts. For the monthly data, the ARIMA model outperformed ETS with lower RMSE (3.27 vs 3.91) and MAE (2.53 vs 3.00)

Visually, the plots suggest better monthly performance but the numerical metrics indicate higher absolute errors for monthly forecasts compared to yearly ones. It would be interesting to compare how each of these predictions preform on test data from the other dataset (i.e. do predictions made from monthly data preform better for yearly data)


################ 
################ Team Member Contributions 

BM, EC, and TT all contributed code and analysis to the final analysis. EC wrote details about the initial plan and questions we addressed. TT added details about checking the residuals and the ACF/PACF plots, and wrote details about the results and discussion. BM wrote and organized final code and added details about Accuracy results. 









