---
title: Team 2 - Lab 4
subtitle: Hidden Markov Models
author: "Name 1, Name 2, Name 3"
date: May XX, 2025
output: 
  html_document:
    code-folding: true
    toc: true
    toc_float: true
---



```{r setup, include = FALSE}
options(dplyr.summarise.inform = FALSE)
library(dplyr)
# install.packages("rsoi")
# install.packages("depmixS4")
library(depmixS4)
library(ggplot2)
```

# Part 1. Hidden Markov Models
For the first part of the homework, we’ll use data from the Pacific Decadal Oscillation (PDO) to ask questions about identifying regimes. This dataset can be accessed via the `rsoi` package. We will look at the winter PDO only. We need to shift the year for Oct-Dec by 1 since Oct-Feb spans 2 calendar years.
```{r}
pdo <- rsoi::download_pdo()
pdo$Year[which(pdo$Month%in%c("Oct","Nov","Dec"))] <- pdo$Year[which(pdo$Month%in%c("Oct","Nov","Dec"))] + 1
pdo <- dplyr::group_by(pdo, Year) %>%
  dplyr::summarize(winter_pdo = mean(PDO[which(Month %in% c("Oct","Nov","Dec","Jan","Feb"))])) %>% 
  dplyr::select(winter_pdo, Year)
# The first year will be missing Oct-Dec
pdo <- pdo[-1,]
```

## 1.1 Fit HMM model
Fit a 2-state HMM to the annual indices of winter PDO. Assume Gaussian errors (default).
```{r}
# Prepare data for depmixS4
winter_pdo <- data.frame(winter_pdo = pdo$winter_pdo)

# Define HMM with 2 hidden states and Gaussian distribution
hmm_model <- depmix(response = winter_pdo ~ 1, 
                    data = winter_pdo, 
                    nstates = 2, 
                    family = gaussian())

# Fit the model
hmm_fit <- fit(hmm_model, verbose = TRUE)

# Summary of the fitted HMM
summary(hmm_fit)
```

## 1.2 Check Stability
Try fitting the model 10-20 times. Does the likelihood seem reasonably stable? (Note `logLik()` gets you the log-likelihood from model fits in R).

```{r}
# Fit HMM 15 times and store log-likelihoods
set.seed(550) 

n_runs <- 15
logLik_values <- numeric(n_runs)

for (i in 1:n_runs) {
  hmm_model <- depmix(response = winter_pdo ~ 1, 
                      data = winter_pdo, 
                      nstates = 2, 
                      family = gaussian())
  
  hmm_fit <- fit(hmm_model, verbose = FALSE)
  
  logLik_values[i] <- logLik(hmm_fit)
}

# Print the log-likelihoods
print(logLik_values)
```

Every run converged to the exact same log-likelihood value: -96.16431. The number of iterations varied (e.g., from 43 to 81), but the final log-likelihood was consistent. Therefore, we can conclude that the log-likelihood values are stable.

## 1.3 Transition Matrix and Persistence Probabilities
What is the transition matrix for the best model? What are the persistence probabilities (e.g. probabilities of staying in the same state from step $t$ to $t+1$)?

Based on the previous output of `summary(hmm_fit)`, the transition matrix can be obtained as follows:

> Transition matrix <br> 
> &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; toS1  toS2 <br>
> fromS1 0.707 0.293 <br>
> fromS2 0.168 0.832 <br>

- From State 1:
  - Probability of staying in State 1: 0.707 (persistence probability)
  - Probability of switching to State 2: 0.293
- From State 2:
  - Probability of staying in State 2: 0.832 (persistence probability)
  - Probability of switching to State 1: 0.168

## 1.4 Plot the predicted values versus year
```{r}
# Combine everything into one data frame
pdo_plot <- pdo
pdo_plot$est_state <- factor(posterior(hmm_fit)$state, 
                             levels = c(1, 2), 
                             labels = c("State1", "State2"))

pdo_plot <- pdo_plot %>%
  mutate(State = if_else(est_state == "State1", 0, 1))

# Plot
ggplot(pdo_plot, aes(x = Year, y = State, color = est_state)) +
  geom_point(size = 3) +
  labs(title = "HMM Predicted States Over Time",
       x = "Year",
       y = "State",
       color = "Estimated State") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## 1.5 Plot the posterior probability
Plot the posterior probability of being in the various states from your best model (e.g. probability of being in state 1 over time)
```{r}
# Extract posterior probabilities and combine with year for plotting
pdo_prob_plot <- data.frame(
  Year = pdo$Year,
  prob_state1 = posterior(hmm_fit)$S1,  # S1 = probability of being in state 1
  prob_state2 = posterior(hmm_fit)$S2   # S2 = probability of being in state 2
)

# Plot probability of being in State 1 over time
p1 <- ggplot(pdo_prob_plot, aes(x = Year, y = prob_state1)) +
  geom_line(color = "blue") +
  geom_point(color = "blue", size = 2) +
  labs(title = "Posterior Probability of Being in State 1 Over Time",
       x = "Year",
       y = "P(State 1)") +
  theme_minimal()

# Plot probability of being in State 2 over time
p2 <- ggplot(pdo_prob_plot, aes(x = Year, y = prob_state2)) +
  geom_line(color = "red") +
  geom_point(color = "red", size = 2) +
  labs(title = "Posterior Probability of Being in State 2 Over Time",
       x = "Year",
       y = "P(State 2)") +
  theme_minimal()

ggpubr::ggarrange(p1, p2, nrow = 2)
```

## 1.6 Long-run probability
What is the long-run probability that the PDO is in state 1 versus state 2? 
You can calculate this from the transition matrix. There is
an analytical solution for this (a bit of googling will find it). Or
you can run a `for` loop to find it. Let $p_1$ be the probability
that PDO is in state 1 and $p_2$ be the probability that PDO is in
state 2. Note $p_1 + p_2 = 1$. If $P$ is the transition matrix (in
Q3),

$$\begin{bmatrix}p_1&p_2\end{bmatrix}_n = \begin{bmatrix}p_1&p_2\end{bmatrix}_{n-1} P$$
Note this is a 1x2 matrix times a 2x2 matrix on the right. Start with
$p_1=1$ and $p_2=0$, say. Run a `for` loop until

$$\begin{bmatrix}p_1&p_2\end{bmatrix}_n \approx \begin{bmatrix}p_1&p_2\end{bmatrix}_{n-1}$$
That $\begin{bmatrix}p_1&p_2\end{bmatrix}_n$ is the long-run probability
in each state.

```{r}
# Extract the transition matrix from your best model
n_states <- 2
trans_mat <- matrix(NA, nrow = n_states, ncol = n_states)

for (i in 1:n_states) {
  # Access transition model for state i
  trans_model <- hmm_fit@transition[[i]]
  
  # Get parameters (probabilities to all states from state i)
  trans_mat[i, ] <- getpars(trans_model)
}

# Initialize probabilities
p <- c(1, 0)  # Start: 100% in state 1

# Iterate until convergence
threshold <- 1e-8
max_iter <- 1000

for (i in 1:max_iter) {
  p_new <- p %*% trans_mat  # Matrix multiplication
  if (sum(abs(p_new - p)) < threshold) {
    break
  }
  p <- p_new
}

# Display result
colnames(p) <- c("State 1", "State 2")
p
```

In the long run, the PDO is expected to be:

- In State 1 about 36.5% of the time
- In State 2 about 63.5% of the time


# Part 2. Fitting Multivariate HMMs

## Data

As part of the California Current Integrated Ecosystem Report, NOAA
scientists do annual updates of [stoplight charts for ecosystem
indicators](https://www.fisheries.noaa.gov/west-coast/science-data/ocean-conditions-indicators-trends).

We have included the `stoplight.csv` dataset for this week. One of the
columns divides indicators into groups (e.g. Local Physical, Local
Biological, etc). Please pick a type of indicators, and develop a 2- or
3-state multivariate HMM. A few tips:

Describe what plankton groups and covariates you used and indicate any temporal 
subsetting you chose. Also describe any standardization, centering or scaling 
that was used. 

## Load the data

One of the
columns divides indicators into groups (e.g. Local Physical, Local
Biological, etc). 

```{r load_data}
stoplight <- read.csv(here::here("Lab-4", "stoplight.csv"))
```

## Wrangle the data

```{r wrangle_data}
library(dplyr)
#Read in the data
stoplight <- read_csv("stoplight.csv")

#Going to model Nearshore ichthyoplankton and juvenile chinook as two biological indicators
#Subset the data to only include rows 13 and 15

dat <- stoplight %>%  filter(!row_number() %in% c(1:12,14, 16:23))

#Need to convert from wide to long 
library(tidyr)
library(atsalibrary)
library(ggplot2)
#install.packages("hmmTMB")
library(hmmTMB)
#install.packages("depmixS4")
library(depmixS4)
dat_long <- dat %>% 
  pivot_longer(cols = -c("Ecosystem Indicators", "Type"), names_to = "year", values_to = "vals")

#Remove the first digit in the year column
dat_long$year <- gsub("X", "", dat_long$year)

#Convert to the correct wide
dat_wide <- dat_long %>%
  dplyr::select(-Type) %>%
  tidyr::pivot_wider(names_from = `Ecosystem Indicators`, values_from = vals)

dat_wide$time <- dat_wide$year
dat_wide$time <- as.numeric(dat_wide$time)
dat_wide$year <- NULL 
dat_wide$Nearshore_Ichthyoplankton <- as.numeric(dat_wide$Nearshore_Ichthyoplankton)
dat_wide$Chinook_salmon_juv <- as.numeric(dat_wide$Chinook_salmon_juv)

```


# General tasks

Each group has the same general tasks, but you will adapt them as you work on the data.

Please pick a type of indicators, and develop a 2- or 3-state multivariate HMM. 

We have decided to use the juvenile chinook data and the nearshore icthyoplankton data. This should be an interesting comparison because of the potential links as predator/prey.

# Methods

A few tips:

-   Assume all responses are Gaussian.

-   You're welcome to include covariates (year? Climate variables?) --
    but fitting a simple model without covariates is also totally fine

```{r fit_hmm_model}

# Fit a 2-state model to the icthyoplankton and juvenile chinook data
set.seed(123)
hidden <- MarkovChain$new(data = dat_wide, n_states = 2)
dists <- list(Nearshore_Ichthyoplankton = "norm", Chinook_salmon_juv = "norm") 
par0 <- list(Nearshore_Ichthyoplankton = list(mean = c(0,3), sd = c(0.1, 0.1)),
             Chinook_salmon_juv = list(mean = c(0,3), sd = c(0.1, 0.1)))
obs_model <- Observation$new(data = dat_wide,
                             dists = dists,
                             n_states = 2,
                             par = par0)
hmm <- HMM$new(obs = obs_model, 
               hid = hidden)
hmm$fit(silent=TRUE)

#Check the fit
states <- hmm$viterbi()
dat_wide$state <- states  # attach to data


```


# Results

Summarize the model you've created. Specifically,

-   Does it converge?

-   How many states seem to be most supported?

-   Plot the time series of estimated states. What does this mean?

-   What are the transition probabilities?

-   If included, what are the covariate effects? What do these mean?

-   Anything else interesting that you've discovered?

# Discussion


# Team contributions

Example: "All team members helped decide on the goal and ran the analyses for the individual regions. Team members 2 & 3 wrote most of the code for the analysis. Team member 1 worked on the plotting section of the report using and adapting code that team member 3 wrote. All team members helped edit the report and wrote the discussion together."

