---
title: Team 1 - Lab 4
subtitle: Hidden Markov Models
author: "Name 1, Name 2, Name 3"
date: May XX, 2025
output: 
  html_document:
    code-folding: true
    toc: true
    toc_float: true
---



```{r setup, include = FALSE}
options(dplyr.summarise.inform = FALSE)
```

Make sure to label your final write-up with "final" in the title so we know which one is the final one.

# Data

As part of the California Current Integrated Ecosystem Report, NOAA
scientists do annual updates of [stoplight charts for ecosystem
indicators](https://www.fisheries.noaa.gov/west-coast/science-data/ocean-conditions-indicators-trends).

We have included the `stoplight.csv` dataset for this week. One of the
columns divides indicators into groups (e.g. Local Physical, Local
Biological, etc). Please pick a type of indicators, and develop a 2- or
3-state multivariate HMM. A few tips:

Describe what plankton groups and covariates you used and indicate any temporal 
subsetting you chose. Also describe any standardization, centering or scaling 
that was used. 

## Load the data

One of the
columns divides indicators into groups (e.g. Local Physical, Local
Biological, etc). 

```{r load_data}
stoplight <- read.csv(here::here("Lab-4", "stoplight.csv"))
library(dplyr)
```

## Wrangle the data

```{r wrangle_data}
library(dplyr)
#Read in the data
stoplight <- read.csv("stoplight.csv")

#Going to model Nearshore icthyoplankton and juvenile chinook as two biological indicators
#Subset the data to only include rows 13 and 15

dat <- stoplight %>%  filter(!row_number() %in% c(1:12,14, 16:23))

#Need to convert from wide to long 
library(tidyr)
library(atsalibrary)
library(ggplot2)
#install.packages("hmmTMB")
library(hmmTMB)
#install.packages("depmixS4")
library(depmixS4)
dat_long <- dat %>% 
  pivot_longer(cols = -c(Ecosystem.Indicators, Type), names_to = "year", values_to = "vals")

#Remove the first digit in the year column
dat_long$year <- gsub("X", "", dat_long$year)

#Convert to the correct wide
dat_wide <- dat_long %>%
  select(-Type) %>%  # If 'Type' is not needed; keep it if relevant
  pivot_wider(names_from = Ecosystem.Indicators, values_from = vals)

dat_wide$time <- dat_wide$year
dat_wide$time <- as.numeric(dat_wide$time)
dat_wide$year <- NULL 
dat_wide$Nearshore_Ichthyoplankton <- as.numeric(dat_wide$Nearshore_Ichthyoplankton)
dat_wide$Chinook_salmon_juv <- as.numeric(dat_wide$Chinook_salmon_juv)

```


# General tasks

Each group has the same general tasks, but you will adapt them as you work on the data.

Please pick a type of indicators, and develop a 2- or 3-state multivariate HMM. 


We have decided to use the juvenile chinook data and the nearshore icthyoplankton data. This should be an interesting comparison because of the potential links as predator/prey.

# Methods

A few tips:

-   Assume all responses are Gaussian.

-   You're welcome to include covariates (year? Climate variables?) --
    but fitting a simple model without covariates is also totally fine


```{r fit_hmm_model}

# Fit a 2-state model to the icthyoplankton and juvenile chinook data
set.seed(123)
hidden <- MarkovChain$new(data = dat_wide, n_states = 2)
dists <- list(Nearshore_Ichthyoplankton = "norm", Chinook_salmon_juv = "norm") 
par0 <- list(Nearshore_Ichthyoplankton = list(mean = c(0,3), sd = c(0.1, 0.1)),
             Chinook_salmon_juv = list(mean = c(0,3), sd = c(0.1, 0.1)))
obs_model <- Observation$new(data = dat_wide,
                             dists = dists,
                             n_states = 2,
                             par = par0)
hmm <- HMM$new(obs = obs_model, 
               hid = hidden)
hmm$fit(silent=TRUE)

#Check the fit
states <- hmm$viterbi()
dat_wide$state <- states  # attach to data

ggplot(dat_wide, aes(x = time, y = Nearshore_Ichthyoplankton, color = factor(state))) +
  geom_line() +
  labs(color = "State")

ggplot(dat_wide, aes(x = time, y = Chinook_salmon_juv, color = factor(state))) +
  geom_line() +
  labs(color = "State")

#plot both states on the same plot
ggplot(dat_wide, aes(x = time)) +
  geom_line(aes(y = Nearshore_Ichthyoplankton, color = "Nearshore Ichthyoplankton")) +
  geom_line(aes(y = Chinook_salmon_juv, color = "Chinook Salmon Juv")) +
  labs(color = "State") +
  scale_color_manual(values = c("blue", "red")) +
  theme(legend.position = "bottom") +
  ylab("Values") +
  xlab("Year")



```

# Results

Summarize the model you've created. Specifically,

-   Does it converge?

-   How many states seem to be most supported?

-   Plot the time series of estimated states. What does this mean?

-   What are the transition probabilities?

-   If included, what are the covariate effects? What do these mean?

-   Anything else interesting that you've discovered?

# Discussion


# Team contributions

Example: "All team members helped decide on the goal and ran the analyses for the individual regions. Team members 2 & 3 wrote most of the code for the analysis. Team member 1 worked on the plotting section of the report using and adapting code that team member 3 wrote. All team members helped edit the report and wrote the discussion together."

