---
title: Team 1 - Lab 1
subtitle: Dynamic Factor Analysis (DFA) - Common trends in plankton data
author: "Matheus de Barros, Tina Tao, Brian McGreal"
date: May 6, 2025
output: 
  html_document:
    code-folding: true
    toc: true
    toc_float: true
---

```{r setup, message=FALSE, warning=FALSE, echo=FALSE}
options(dplyr.summarise.inform = FALSE)
```

# Data

We selected the five primary producer from the phytoplankton group, which include

* `Cryptomonas`: small brown or green algae (edible)  
* `Diatoms`: small algae rich in silica (edible)  
* `Greens`: general class of algae (edible)  
* `Unicells`: very small algae (edible)  
* `Other.algae`: generic catch-all for atypical species (edible)

because there are too many missing values in `Bluegreens`. Likewise, the group `Neomysis` is excluded from the zooplankton data due to the high volume of missing observations, leaving the following classifications of zooplankton:

* `Conochilus`: colonial form of rotifer (grazer)  
* `Cyclops`: cyclopoid copepod (grazer) 
* `Daphnia`: cladoceran (grazer) 
* `Diaptomus`: calanoid copepod (grazer) 
* `Epischura`: very large calanoid copepod (predator)
* `Leptodora`: very large cladoceran (predator)  
* `Non.daphnid.cladocerans`: catch-all for other cladocerans (grazers)
* `Non.colonial.rotifers`: free-floating rotifers (grazers)

For covariates, We selected environmental factors including
* `Temp`: water temperature in degrees C  
* `TP`: total phosphorous concentration in mg m<sup>-3</sup>  
* `pH`: pH.

Our analysis includes data from the year 1985 onward through the end of the dataset (December, 1994). These data are 'cleaned', meaning they are log-transformed and scaled.

```{r load_data, echo=FALSE, warning=FALSE, message=FALSE}
## load MARSS for data and analyses
library(here)
library(MARSS)
library(panelr)
library(tidyverse)
library(zoo)

here::i_am("Lab-3/Team-2/Lab-3-TT.Rmd")

## load the raw data (there are 3 datasets contained here)
data(lakeWAplankton, package = "MARSS")

## we want `lakeWAplanktonTrans`, which has been transformed
## so the 0's are replaced with NA's and the data z-scored
all_dat <- lakeWAplanktonTrans
```

```{r wrangle_data, echo=FALSE, warning=FALSE, message=FALSE}
## use 10 years from 1985-1994
yr_frst <- 1985
yr_last <- 1994
plank_dat <- all_dat[all_dat[, "Year"] >= yr_frst & 
                     all_dat[, "Year"] <= yr_last,]

## create vector of phytoplankton group names
phytoplankton <- c("Cryptomonas", "Diatoms", "Greens",
                   "Unicells", "Other.algae")

## get only the phytoplankton
dat_1985 <- plank_dat[, phytoplankton]

## transpose data so time goes across columns
dat_1985 <- t(dat_1985)

## get number of time series
N_ts <- dim(dat_1985)[1]

## get length of time series
TT <- dim(dat_1985)[2]

## mean of each taxon
y_bar <- apply(dat_1985, 1, mean, na.rm = TRUE)

## de-mean the data - subtract the means
dat <- dat_1985 - y_bar

## assign new column names
spp <- rownames(dat_1985)
rownames(dat) <- spp
```

```{r plot_data, echo=FALSE, warning=FALSE, message=FALSE}
## set plot colors for six groups
clr <- c("blue", "darkgreen", "darkred", "purple", "orange")

## initialize a counter
cnt <- 1

## set up plotting space for six time series
par(mfrow = c(N_ts, 1), mai = c(0.5, 0.7, 0.1, 0.1), omi = c(0, 0, 0, 0))

## loop through each phytoplankton group
for(i in spp){
  plot(dat[i,], bty = "L", xaxt = "n", pch = 16,
       xlab = "",
       ylab = "Abundance index", col = clr[cnt], type = "b")
  axis(1, 12 * (0:dim(dat_1985)[2]) + 1, yr_frst + 0:dim(dat_1985)[2])
  title(i)
  cnt <- cnt + 1
}
```

# Methods

## Phytoplankton abundance

We examined models that consider phytoplankton abundance being driven by three underlying state options: one, two, or three states. We also test three observation error matrices, diagonal and equal, diagonal and unequal, and equal variance and covariance. Our aim is to consider all combinations of three environmental covariates and how they relate to phytoplankton abundance.

```{r model_params, echo=FALSE, warning=FALSE, message=FALSE}
model.states <- 1:3
names(model.states) <- c("one_state", "two_states","three_states")

# testing different observation error matrices
R.models <- c("diagonal and unequal","equalvarcov", "diagonal and equal")

# covariates must be z-scored if the response variables are demeaned, and subset covariates to same time window and transpose
cov_temp <- t(zscore(plank_dat[,"Temp", drop = FALSE]))
cov_TP   <- t(zscore(plank_dat[,"TP", drop = FALSE]))
cov_pH   <- t(zscore(plank_dat[,"pH", drop = FALSE]))
cov_temp_TP   <- t(zscore(plank_dat[, c("Temp", "TP")]))
cov_TP_pH     <- t(zscore(plank_dat[, c("TP", "pH")]))
cov_temp_pH   <- t(zscore(plank_dat[, c("Temp", "pH")]))
cov_all       <- t(zscore(plank_dat[, c("Temp", "TP", "pH")]))

d.models <- list(
  Temp            = cov_temp,
  TP              = cov_TP,
  pH              = cov_pH,
  "Temp and TP"   = cov_temp_TP,
  "TP and pH"     = cov_TP_pH,
  "Temp and pH"   = cov_temp_pH,
  "Temp,TP, and pH" = cov_all
)

# define model list 
mod.list = list(
  A = "zero", # this is set to zero because it is the offset/scaling
  Q = "identity", # Q is set to the default value, the identify matrix, which can also be "diagonal and equal" or "diagonal and unequal"
  x0 = "zero" # x0 can be also be set to "unconstrained" or "unequal" it might be useful to try these later
)

```



```{r MARSS_models_fitting, echo=FALSE, warning=FALSE, message=FALSE}
## Create a results list to store model fits and AICs
model_fits <- list()
model_AICs <- data.frame(
  ModelID = character(),
  NumStates = integer(),
  R_Model = character(),
  Covariates = character(),
  AICc = numeric(),
  stringsAsFactors = FALSE
)

# Check and handle NAs in covariates (assuming plank_dat is already subsetted)
# Apply linear interpolation to covariates to fill NAs
library(zoo)
plank_df <- as.data.frame(plank_dat)
plank_df$Temp <- na.approx(plank_df$Temp, na.rm = FALSE)
plank_df$TP <- na.approx(plank_df$TP, na.rm = FALSE)
plank_df$pH <- na.approx(plank_df$pH, na.rm = FALSE)

# Recreate covariates with scaling after handling NAs
cov_temp <- t(scale(plank_df[,"Temp", drop = FALSE]))
cov_TP   <- t(scale(plank_df[,"TP", drop = FALSE]))
cov_pH   <- t(scale(plank_df[,"pH", drop = FALSE]))
cov_temp_TP   <- t(scale(plank_df[, c("Temp", "TP")]))
cov_TP_pH     <- t(scale(plank_df[, c("TP", "pH")]))
cov_temp_pH   <- t(scale(plank_df[, c("Temp", "pH")]))
cov_all       <- t(scale(plank_df[, c("Temp", "TP", "pH")]))

# Update d.models with corrected covariates
d.models <- list(
  Temp            = cov_temp,
  TP              = cov_TP,
  pH              = cov_pH,
  "Temp and TP"   = cov_temp_TP,
  "TP and pH"     = cov_TP_pH,
  "Temp and pH"   = cov_temp_pH,
  "Temp,TP, and pH" = cov_all
)

model_id <- 1
for(n_state in model.states){
  for(R_mod in R.models){
    for(d_name in names(d.models)){
      
      # Define the Z matrix with correct dimensions
      if(n_state == 1){
        Z.model <- matrix("z", N_ts, n_state)  # All species load on single trend
      } else {
        Z.model <- matrix(0, N_ts, n_state)  # Start with zeros
        param_count <- 1
        for(i in 1:N_ts){
          for(j in 1:n_state){
            if(i < N_ts && j > i){
              # upper-triangle for first m-1 rows stays zero
              next
            } else {
              if(i == j){
                Z.model[i, j] <- paste0("z", param_count, "+")  # diagonal: positive
              } else {
                Z.model[i, j] <- paste0("z", param_count)
              }
              param_count <- param_count + 1
            }
          }
        }
      }
      
      # Define model list with updated Z and correct dimensions
      mod.list <- list(
        Z = Z.model,
        R = R_mod,
        A = "zero",
        Q = "identity",
        x0 = "zero",
        D = "unconstrained",
        d = d.models[[d_name]]
      )
      
      # Fit the model
      fit <- tryCatch({
        MARSS(dat, model = mod.list, silent = TRUE, control = list(maxit = 5000))
      }, error = function(e) NULL)
      
      # Store results if successful
      if(!is.null(fit) && !is.null(fit$AICc)){
        model_fits[[paste0("model", model_id)]] <- fit
        model_AICs <- rbind(model_AICs, data.frame(
          ModelID = paste0("model", model_id),
          NumStates = n_state,
          R_Model = R_mod,
          Covariates = d_name,
          AICc = fit$AICc
        ))
      } else {
        cat("Model", model_id, "failed or AICc missing\n")
      }
      
      model_id <- model_id + 1
    }
  }
}

```

```{r MARSS_models_comparison, echo=FALSE, warning=FALSE, message=FALSE}
library(dplyr)
model_AICs %>% 
  arrange(AICc) %>% 
  head(10)

# Find the best model ID
best_model_id <- model_AICs$ModelID[which.min(model_AICs$AICc)]

# Get the best model object
best_model <- model_fits[[best_model_id]]

# View model summary
summary(best_model)

# View estimated parameters
coef(best_model, type = "matrix")$Z  # Loadings
coef(best_model, type = "matrix")$D  # Covariate effects
```

```{r best_model_diagnostics, echo=FALSE, warning=FALSE, message=FALSE}
# Extract residuals from the best model
resids <- MARSSresiduals(best_model)

# Time series plot of model residuals (observation errors)
model_res <- resids$model.residuals
ts.plot(t(model_res), main = "Time Series of Model Residuals", ylab = "Residuals", xlab = "Time", col = 1:nrow(model_res))
legend("topright", legend = rownames(model_res), col = 1:nrow(model_res), lty = 1)

# Autocorrelation check for each time series
par(mfrow = c(ceiling(nrow(model_res)/2), 2), mar = c(4, 4, 2, 1))  # layout for ACF plots
for(i in 1:nrow(model_res)){
  acf(model_res[i, ], na.action = na.pass, main = paste("ACF -", rownames(model_res)[i]))
}
par(mfrow = c(1,1))  # reset layout
```

```{r best_model_visualization, echo=FALSE, warning=FALSE, message=FALSE}
# Plot the estimated trends
library(ggplot2)
trends <- data.frame(
  Time = 1:TT,
  Trend = as.vector(best_model$states),
  SE = as.vector(best_model$states.se)
)

ggplot(trends, aes(x = Time, y = Trend)) +
  geom_line() +
  geom_ribbon(aes(ymin = Trend - 1.96*SE, ymax = Trend + 1.96*SE), alpha = 0.2) +
  labs(title = "Estimated Common Trends")

# Get fitted values
fitted_vals <- fitted(best_model)

# Plot for each species
ggplot(fitted_vals, aes(x = t, y = .fitted)) +
  geom_line(color = "blue") +
  geom_point(aes(y = y), color = "black", size = 1) +
  facet_wrap(~.rownames, scales = "free_y") +
  labs(x = "Time", y = "Abundance", title = "Model Fits vs Observations")

# Get covariate coefficients
covar_effects <- coef(best_model, type = "matrix")$D

# Plot effects
data.frame(
  Species = rownames(dat),
  covar_effects
) %>% 
  tidyr::pivot_longer(-Species, names_to = "Covariate", values_to = "Effect") %>% 
  ggplot(aes(x = Covariate, y = Effect, color = Species)) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  coord_flip() +
  theme_minimal()
```

## Zooplankton abundance

This analysis includes model comparison to determine the optimal number of hidden states that drive zooplankton abundance. 

Loading and wrangling data:
```{r}
library(MARSS)
library(ggplot2)
library(dplyr)
library(ggpubr)

data(lakeWAplankton, package = "MARSS")

data = lakeWAplanktonTrans

zooplankton = c("Conochilus", "Cyclops", "Daphnia", "Diaptomus", "Epischura",
                "Leptodora", "Non.daphnid.cladocerans", "Non.colonial.rotifers") # I'm excluding 'Neomysis' since they dont have lots of data

env_covars = c("Temp", "TP", "pH")

time_vars = c("Year", "Month")

combined = unique(c(zooplankton, env_covars, time_vars))

# filter data
data = lakeWAplanktonTrans[, colnames(lakeWAplanktonTrans) %in% combined, drop = FALSE]

# select zooplankton only and de-mean data
zoop_data = data[, colnames(data) %in% zooplankton, drop = FALSE]
y_mean    = apply(zoop_data, 2, mean, na.rm = TRUE)
zoop_data = zoop_data - y_mean

# select last 10 years only
zoop_data = zoop_data[120:dim(zoop_data)[2],]

# transpose so time is columns 
zoop_data = t(zoop_data)
N_ts      = nrow(zoop_data)
TT        = ncol(zoop_data)
```

Plotting:
```{r}
cnt = 1
par(mfrow = c(4, 2), mai = c(0.5, 0.7, 0.15, 0.5), omi = c(0, 0, 0, 0))

for (i in zooplankton) {
  plot(zoop_data[cnt, ], type = "l", xlab = "Time", ylab = i,
       main = paste("Time series of", i), ylim = c(-2, 2))
  cnt = cnt + 1
}
```
#### Model 1: 1 Hidden State

##### The observation model

$$
\begin{bmatrix}
Y_{1t} \\
Y_{2t} \\
Y_{3t} \\
Y_{4t} \\
Y_{5t} \\
Y_{6t} \\
Y_{7t} \\
Y_{8t}
\end{bmatrix}
=
\begin{bmatrix}
z_{11} \\
z_{21} \\
z_{31} \\
z_{41} \\
z_{51} \\
z_{61} \\
z_{71} \\
z_{81}
\end{bmatrix}
X_t
+
\begin{bmatrix}
0 \\
0 \\
0 \\
0 \\
0 \\
0 \\
0 \\
0
\end{bmatrix}
+
\begin{bmatrix}
0 \\
0 \\
0 \\
0 \\
0 \\
0 \\
0 \\
0
\end{bmatrix}
+
\varepsilon_t
$$
Where:
- \( \varepsilon_t \sim N(0, R) \)
- The variance-covariance matrix \( R \) is:
$$
R = 
\begin{bmatrix}
\sigma_1^2 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
0 & \sigma_2^2 & 0 & 0 & 0 & 0 & 0 & 0 \\
0 & 0 & \sigma_3^2 & 0 & 0 & 0 & 0 & 0 \\
0 & 0 & 0 & \sigma_4^2 & 0 & 0 & 0 & 0 \\
0 & 0 & 0 & 0 & \sigma_5^2 & 0 & 0 & 0 \\
0 & 0 & 0 & 0 & 0 & \sigma_6^2 & 0 & 0 \\
0 & 0 & 0 & 0 & 0 & 0 & \sigma_7^2 & 0 \\
0 & 0 & 0 & 0 & 0 & 0 & 0 & \sigma_8^2
\end{bmatrix}
$$

$$
X_t = 
\begin{bmatrix}
1
\end{bmatrix}
X_{t-1} +
\begin{bmatrix}
0
\end{bmatrix}
+
\begin{bmatrix}
0
\end{bmatrix}
+
\eta_t
$$

Where:
- \( \eta_t \sim N(0, Q) \)
- The variance-covariance matrix \( Q \) is:
$$
Q = 
\begin{bmatrix}
1
\end{bmatrix}
$$


```{r}
###### Model 1: 1 Hidden State ######

## 1. Observation model
Z_vals_1 = list(  "z11"  ,  
                  "z21",
                  "z31",
                  "z41",
                  "z51",
                  "z61",
                  "z71",
                  "z81" )

ZZ_1 = matrix(Z_vals_1, nrow = N_ts, ncol = 1, byrow = TRUE)
aa_1 = "zero"
DD_1 = "zero"                 # no covariates
dd_1 = "zero"
RR_1 = "diagonal and unequal" # observation error variance-covariance matrix

## 2. Process model
mm_1 = 1           # number of hidden states
BB_1 = "identity"  # diag(mm)
uu_1 = "zero"      # process bias (i.e. growth rate)
CC_1 = "zero"      # no covariates
cc_1 = "zero"
QQ_1 = "identity"  # process error variance-covariance matrix

mod_list_1 = list(Z = ZZ_1, A = aa_1, D = DD_1, d = dd_1, R = RR_1,
                  B = BB_1, U = uu_1, C = CC_1, c = cc_1, Q = QQ_1)

init_list_1 = list(x0 = 0)

con_list_1   = list(maxit = 1000, allow.degen = TRUE)
dfa_1_hidden = MARSS(y = zoop_data, model = mod_list_1, inits = init_list_1, control = con_list_1)

## Plot fits against data
model_1_fits    = dfa_1_hidden$ytT
model_1_fits_se = dfa_1_hidden$ytT.se

par(mfrow = c(4, 2), mai = c(0.5, 0.7, 0.15, 0.5), omi = c(0, 0, 0, 0))
cnt = 1
for (i in zooplankton) {
  plot(zoop_data[cnt, ], type = "p", xlab = "Time", ylab = i,
       main = paste("Time series of", i), ylim = c(-2, 2))
  lines(model_1_fits[cnt, ], col = "red")
  lines(model_1_fits[cnt, ] + model_1_fits_se[cnt, ], col = "blue", lty = 2)
  lines(model_1_fits[cnt, ] - model_1_fits_se[cnt, ], col = "blue", lty = 2)
  cnt = cnt + 1
}

## Plot states and loadings
Z_est_1 = coef(dfa_1_hidden, type = "matrix")$Z
H_inv_1 = varimax(Z_est_1)

states_1 = as.vector(dfa_1_hidden$states) 
states_1 = data.frame(1:TT, states_1); names(states_1) = c("Time", "State")

states_plot_1 = states_1 %>%
  ggplot(aes(x = Time, y = State)) +
  geom_line() +
  theme_minimal()

loadings_1 = as.vector(H_inv_1[,1])
loadings_1 = data.frame(loadings_1, zooplankton)

loadings_plot_1 = loadings_1 %>%
  ggplot(aes(x = zooplankton, y = loadings_1)) +
  geom_bar(stat = "identity", fill = "white", color = "black") +
  theme_minimal() +
  coord_flip() +
  ylab("") +
  xlab("Loadings")

ggarrange(states_plot_1, loadings_plot_1)
```

#### Model 2: 2 Hidden States

##### The observation model

$$
\begin{bmatrix}
Y_{1t} \\
Y_{2t} \\
Y_{3t} \\
Y_{4t} \\
Y_{5t} \\
Y_{6t} \\
Y_{7t} \\
Y_{8t}
\end{bmatrix}
=
\begin{bmatrix}
z_{11} & 0 \\
z_{21} & z_{22} \\
z_{31} & z_{32} \\
z_{41} & z_{42} \\
z_{51} & z_{52} \\
z_{61} & z_{62} \\
z_{71} & z_{72} \\
z_{81} & z_{82}
\end{bmatrix}
\begin{bmatrix}
X_{1t} \\
X_{2t}
\end{bmatrix}
+
\begin{bmatrix}
0 \\
0 \\
0 \\
0 \\
0 \\
0 \\
0 \\
0
\end{bmatrix}
+
\begin{bmatrix}
0 \\
0 \\
0 \\
0 \\
0 \\
0 \\
0 \\
0
\end{bmatrix}
+
\varepsilon_t
$$

Where:
- \( \varepsilon_t \sim N(0, R) \)
- The variance-covariance matrix \( R \) is:
$$
R = 
\begin{bmatrix}
\sigma_1^2 & 0 & \cdots & 0 \\
0 & \sigma_2^2 & \cdots & 0 \\
\vdots & \vdots & \ddots & \vdots \\
0 & 0 & \cdots & \sigma_8^2
\end{bmatrix}
$$


##### The process model

$$
\begin{bmatrix}
X_{1t} \\
X_{2t}
\end{bmatrix}
=
\begin{bmatrix}
1 & 0 \\
0 & 1
\end{bmatrix}
\begin{bmatrix}
X_{1, t-1} \\
X_{2, t-1}
\end{bmatrix}
+
\begin{bmatrix}
0 \\
0
\end{bmatrix}
+
\begin{bmatrix}
0 \\
0
\end{bmatrix}
+
\eta_t
$$

Where:
- \( \eta_t \sim N(0, Q) \)
- The variance-covariance matrix \( Q \) is:
$$
Q = 
\begin{bmatrix}
1 & 0 \\
0 & 1
\end{bmatrix}
$$

```{r}

###### Model 2: 2 Hidden States ######

## 1. Observation model
Z_vals_2 = list(  0  , 0  ,
                "z21", 0  ,
                "z31","z32",
                "z41","z42",
                "z51","z52",
                "z61","z62",
                "z71","z72",
                "z81","z82")
ZZ_2 = matrix(Z_vals_2, nrow = N_ts, ncol = 2, byrow = TRUE)
aa_2 = "zero"
DD_2 = "zero"                 # no covariates
dd_2 = "zero"
RR_2 = "diagonal and unequal" # observation error variance-covariance matrix

## 2. Process model
mm_2 = 2           # number of hidden states
BB_2 = "identity"  # diag(mm)
uu_2 = "zero"      # process bias (i.e. growth rate)
CC_2 = "zero"      # no covariates
cc_2 = "zero"
QQ_2 = "identity"  # process error variance-covariance matrix

mod_list_2 = list(Z = ZZ_2, A = aa_2, D = DD_2, d = dd_2, R = RR_2,
                  B = BB_2, U = uu_2, C = CC_2, c = cc_2, Q = QQ_2)

init_list_2 = list(x0 = matrix(rep(0, mm_2), mm_2, 1))

con_list_2 = list(maxit = 1000, allow.degen = TRUE)
dfa_2_hidden = MARSS(y = zoop_data, model = mod_list_2, inits = init_list_2, control = con_list_2)

## Plot fits against data
model_2_fits    = dfa_2_hidden$ytT
model_2_fits_se = dfa_2_hidden$ytT.se

par(mfrow = c(4, 2), mai = c(0.5, 0.7, 0.15, 0.5), omi = c(0, 0, 0, 0))
cnt = 1
for (i in zooplankton) {
  plot(zoop_data[cnt, ], type = "p", xlab = "Time", ylab = i,
       main = paste("Time series of", i), ylim = c(-2, 2))
  lines(model_2_fits[cnt, ], col = "red")
  lines(model_2_fits[cnt, ] + model_2_fits_se[cnt, ], col = "blue", lty = 2)
  lines(model_2_fits[cnt, ] - model_2_fits_se[cnt, ], col = "blue", lty = 2)
  cnt = cnt + 1
}

## Plot states and loadings
Z_est_2 = coef(dfa_2_hidden, type = "matrix")$Z
#H_inv_2 = varimax(Z_est_2)$rotmat

states_2 = as.data.frame(t(dfa_2_hidden$states))
states_2 = data.frame(1:TT, states_2); names(states_2) = c("Time", "State 1", "State 2")

states_plot_2 = states_2 %>%
  ggplot(aes(x = Time)) +
  geom_line(aes(y = `State 1`, color = "State 1")) +
  geom_line(aes(y = `State 2`, color = "State 2")) +
  theme_minimal() +
  ylab("States") +
  scale_color_manual(values = c("blue", "red")) +
  theme(legend.title = element_blank(),
        legend.position = "top")

loadings_2 = as.data.frame(Z_est_2)
loadings_2 = data.frame(loadings_2, zooplankton)
loadings_plot_2 = loadings_2 %>%
  ggplot(aes(x = zooplankton)) +
  geom_bar(aes(y = X1), stat = "identity", fill = "blue", color = "black") +
  geom_bar(aes(y = X2), stat = "identity", fill = "red", color = "black") +
  theme_minimal() +
  coord_flip() +
  ylab("") +
  xlab("Loadings")

ggarrange(states_plot_2, loadings_plot_2)

```

#### Model 3: 3 Hidden States

##### The observation model
$$
\begin{bmatrix}
Y_{1t} \\
Y_{2t} \\
Y_{3t} \\
Y_{4t} \\
Y_{5t} \\
Y_{6t} \\
Y_{7t} \\
Y_{8t}
\end{bmatrix}
=
\begin{bmatrix}
z_{11} & 0 & 0 \\
z_{21} & z_{22} & 0 \\
z_{31} & z_{32} & z_{33} \\
z_{41} & z_{42} & z_{43} \\
z_{51} & z_{52} & z_{53} \\
z_{61} & z_{62} & z_{63} \\
z_{71} & z_{72} & z_{73} \\
z_{81} & z_{82} & z_{83}
\end{bmatrix}
\begin{bmatrix}
X_{1t} \\
X_{2t} \\
X_{3t}
\end{bmatrix}
+
\begin{bmatrix}
0 \\
0 \\
0 \\
0 \\
0 \\
0 \\
0 \\
0
\end{bmatrix}
+
\begin{bmatrix}
0 \\
0 \\
0 \\
0 \\
0 \\
0 \\
0 \\
0
\end{bmatrix}
+
\varepsilon_t
$$

Where:
- \( \varepsilon_t \sim N(0, R) \)
- The variance-covariance matrix \( R \) is:
$$
R = 
\begin{bmatrix}
\sigma_1^2 & 0 & \cdots & 0 \\
0 & \sigma_2^2 & \cdots & 0 \\
\vdots & \vdots & \ddots & \vdots \\
0 & 0 & \cdots & \sigma_8^2
\end{bmatrix}
$$
##### The process model
$$
\begin{bmatrix}
X_{1t} \\
X_{2t} \\
X_{3t}
\end{bmatrix}
=
\begin{bmatrix}
1 & 0 & 0 \\
0 & 1 & 0 \\
0 & 0 & 1
\end{bmatrix}
\begin{bmatrix}
X_{1, t-1} \\
X_{2, t-1} \\
X_{3, t-1}
\end{bmatrix}
+
\begin{bmatrix}
0 \\
0 \\
0
\end{bmatrix}
+
\begin{bmatrix}
0 \\
0 \\
0
\end{bmatrix}
+
\eta_t
$$

Where:
- \( \eta_t \sim N(0, Q) \)
- The variance-covariance matrix \( Q \) is:
$$
Q = 
\begin{bmatrix}
1 & 0 & 0 \\
0 & 1 & 0 \\
0 & 0 & 1
\end{bmatrix}
$$

```{r}
###### Model 3: 3 Hidden States ######

## 1. Observation model
Z_vals_3 = list("z11",  0  ,  0  ,
                "z21","z22",  0  ,
                "z31","z32","z33",
                "z41","z42","z43",
                "z51","z52","z53",
                "z61","z62","z63",
                "z71","z72","z73",
                "z81","z82","z83")
ZZ_3 = matrix(Z_vals_3, nrow = N_ts, ncol = 3, byrow = TRUE)
aa_3 = "zero"
DD_3 = "zero"                 # no covariates
dd_3 = "zero"
RR_3 = "diagonal and unequal" # observation error variance-covariance matrix

## 2. Process model
mm_3 = 3           # number of hidden states
BB_3 = "identity"  # diag(mm)
uu_3 = "zero"      # process bias (i.e. growth rate)
CC_3 = "zero"      # no covariates
cc_3 = "zero"
QQ_3 = "identity"  # process error variance-covariance matrix

mod_list_3 = list(Z = ZZ_3, A = aa_3, D = DD_3, d = dd_3, R = RR_3,
                  B = BB_3, U = uu_3, C = CC_3, c = cc_3, Q = QQ_3)

init_list_3 = list(x0 = matrix(rep(0, mm_3), mm_3, 1))

con_list_3 = list(maxit = 1000, allow.degen = TRUE)
dfa_3_hidden = MARSS(y = zoop_data, model = mod_list_3, inits = init_list_3, control = con_list_3)

## Plot fits against data
model_3_fits    = dfa_3_hidden$ytT
model_3_fits_se = dfa_3_hidden$ytT.se
par(mfrow = c(4, 2), mai = c(0.5, 0.7, 0.15, 0.5), omi = c(0, 0, 0, 0))
cnt = 1
for (i in zooplankton) {
  plot(zoop_data[cnt, ], type = "p", xlab = "Time", ylab = i,
       main = paste("Time series of", i), ylim = c(-2, 2))
  lines(model_3_fits[cnt, ], col = "red")
  lines(model_3_fits[cnt, ] + model_3_fits_se[cnt, ], col = "blue", lty = 2)
  lines(model_3_fits[cnt, ] - model_3_fits_se[cnt, ], col = "blue", lty = 2)
  cnt = cnt + 1
}

## Plot states and loadings
Z_est_3 = coef(dfa_3_hidden, type = "matrix")$Z
H_inv_3 = varimax(Z_est_3)$rotmat

states_3 = as.data.frame(t(dfa_3_hidden$states))
states_3 = data.frame(1:TT, states_3); names(states_3) = c("Time", "State 1", "State 2", "State 3")
states_plot_3 = states_3 %>%
  ggplot(aes(x = Time)) +
  geom_line(aes(y = `State 1`, color = "State 1")) +
  geom_line(aes(y = `State 2`, color = "State 2")) +
  geom_line(aes(y = `State 3`, color = "State 3")) +
  theme_minimal() +
  ylab("States") +
  scale_color_manual(values = c("blue", "red", "green4")) +
  theme(legend.title = element_blank(),
        legend.position = "top")

loadings_3 = as.data.frame(Z_est_3)
loadings_3 = data.frame(loadings_3, zooplankton)
loadings_plot_3 = loadings_3 %>%
  ggplot(aes(x = zooplankton)) +
  geom_bar(aes(y = X1), stat = "identity", fill = "blue", color = "black") +
  geom_bar(aes(y = X2), stat = "identity", fill = "red", color = "black") +
  geom_bar(aes(y = X3), stat = "identity", fill = "green4", color = "black") +
  theme_minimal() +
  coord_flip() +
  ylab("") +
  xlab("Loadings")

ggarrange(states_plot_3, loadings_plot_3)

```

##### Comparing models

According to the AICc values, the model with 3 hidden states is the best fit for the data, which showed a slightly better fit compared to the model with 2 hidden states. The model with 1 hidden state was the worst fit.

```{r}
### Make a model comparison table
model_comparison = data.frame(
  Model = c("Model 1", "Model 2", "Model 3"),
  AIC = c(dfa_1_hidden$AIC, dfa_2_hidden$AIC, dfa_3_hidden$AIC)
)
model_comparison = model_comparison %>%
  mutate(Delta_AIC = AIC - min(AIC),
         Weight = exp(-0.5 * Delta_AIC) / sum(exp(-0.5 * Delta_AIC)))

model_comparison
```

## Grazer abundance, phytoplankton covariates

This final section of our analysis focuses on modeling grazing zooplankton, treating phytoplankton abundance as a covariate as grazers feed on phytoplankton. The dataset contains time series abundance data on six species of grazing zooplankton: conochilus, cyclops, daphnia, diaptomus, non-daphnid cladocerans, and non- colonial rotifers. The first stage of this process involves modeling phytoplankton using a traditional MARSS model and recording model estimates of the underlying state of phytoplankton in Lake Washington. The next stage utilizes a dynamica factor analysis framework to consider zooplankton grazer abundance.

#### Modeling phytoplankton

Modeling phytoplankton abundance with traditional MARSS models is necessary for two reasons. One, the dynamic factor analysis of zooplankton grazers will include phytoplankton abundance as covariates. Covariates included in DFA models must have no missing data, and the observed phytoplankton data contains some gaps in the time series. Using MARSS models, the underlying state of phytoplankton abundance can be estimated with no missing data points. These estimates will then be passed into the DFA model. The second reason is that treating time series of five distinct phytoplankton species as representative samples of one underlying state of phytoplankton abundance is likely all that is required to observe the impact of phytoplankton abundance on the abundance of grazing zooplankton. Unless grazers are very selective in their diets, the overall level of phytoplankton in a given time period will likely be more important than the abundances of five individual phtoplankton groupings. 

```{r zp_datamanipulation, echo=FALSE, warning=FALSE, message=FALSE}
## load the data (there are 3 datasets contained here)
data(lakeWAplankton, package = "MARSS")
all_dat <- as.data.frame(lakeWAplanktonTrans)

# transform date
all_dat$Date <- as.yearmon(paste(all_dat$Year, all_dat$Month), "%Y %m")

# plot data
# for ( i in seq(3,length( all_dat )-1,1) ) plot(all_dat[,i], x=all_dat$Date, ylab=names(all_dat[i]),type="l")

# dropping years pre-1980 and critters with poor data
yr_frst <- 1985
data <- subset(all_dat, Year  >= yr_frst)
data <- data[-c(9, 18)]

# different lists of variables
phytoclimate <- c("Temp", 
                  "TP", 
                  "pH")
zooclimate <- c("Temp", 
                "pH")
phytoplankton <- c("Cryptomonas", 
                   "Diatoms", 
                   "Greens",
                   "Unicells", 
                   "Other.algae")
zooplankton <- c("Conochilus", 
                 "Cyclops", 
                 "Daphnia",
                 "Diaptomus", 
                 "Non.daphnid.cladocerans",
                 "Non.colonial.rotifers")
  # only including grazers

# get only the phytoplankton
dates <- data$Date
data_phyto <- t(data[, phytoplankton])
colnames(data_phyto) <- dates

# get phytoplankton climate covariates
data_phytoclimate <- t(data[, phytoclimate])
colnames(data_phytoclimate) <- dates

# monthly factor covariate matrix
month_cov <- matrix(0, 12, 120)
month.abb <- unique(data$Month)
monrow <- match(data$Month, month.abb)[1:120]
month_cov[cbind(monrow,1:120)] <- 1

# stack up the old covariate matrix
phyto.c <- rbind(month_cov, data_phytoclimate)
```

This analysis begins by considering two models with different covariate structures: one which takes into account Lake Washington's temperature, pH level, and total phosphorous measures, and another which includes a matrix of ones and zeroes corresponding to each month in order captrue seasonal trends. Each model treats each category of phytoplankton as its own underlying state.

```{r zp_modelbuild-phyto, echo=FALSE, warning=FALSE, message=FALSE}
# build model - seasonal
phyto_month.list <- list(
  B = "identity",
  U = "zero",
  Q = "diagonal and unequal",
  Z = "identity",
  A = "zero",
  R = "diagonal and equal",
  c = month_cov,
  C= "unconstrained",
  x0 = "unequal",
  V0 = "zero",
  tinitx = 0
)

# build model - climate
phyto_climate.list <- list(
  B = "identity",
  U = "zero",
  Q = "diagonal and unequal",
  Z = "identity",
  A = "zero",
  R = "diagonal and equal",
  c = data_phytoclimate,
  C= "unconstrained",
  x0 = "unequal",
  V0 = "zero",
  tinitx = 0
)

# model control
con_list <- list(maxit = 3000, allow.degen = TRUE)

# modeling - seasonal
if(!file.exists(here::here("Lab-3", "Team-2", "data", "phyto_month.rds"))){
  ssm.phyto_month <- MARSS(data_phyto, model = phyto_month.list, method = "kem", control = con_list)
  saveRDS(ssm.phyto_month, file=here::here("Lab-3", "Team-2", "data", "phyto_month.rds"))
}
# load in ssm.phyto_month
ssm.phyto_month <- readRDS(file=here::here("Lab-3", "Team-2", "data", "phyto_month.rds"))

# modeling - climate
if(!file.exists(here::here("Lab-3", "Team-2", "data", "phyto_climate.rds"))){
  ssm.phyto_climate <- MARSS(data_phyto, model = phyto_climate.list, method = "kem", control = con_list)
  saveRDS(ssm.phyto_climate, file=here::here("Lab-3", "Team-2", "data", "phyto_climate.rds"))
}
# load in ssm.phyto_climate
ssm.phyto_climate <- readRDS(file=here::here("Lab-3", "Team-2", "data", "phyto_climate.rds"))

# model comparison
# an empty list to collect outputs
out_list <- list()

# Add each model's name and AICc 
out_list[[1]] <- data.frame(Model = "seasonal", AICc = ssm.phyto_month$AICc)
out_list[[2]] <- data.frame(Model = "climate", AICc = ssm.phyto_climate$AICc)

# Combine all into one dataframe
out <- do.call(rbind, out_list)
out
  # seasonal model fits better than climate

ssm.phyto <- ssm.phyto_month
```

Per AICc, the model including seasonal covariates (as opposed to lake temperature and chemical compsotion) is a better fit for phytoplankton data. Abundances are described in the figure below.

```{r zp_plot-phyto, echo=FALSE, warning=FALSE, message=FALSE}
# grabbing model estimates
states <- ssm.phyto$states
statesSE <- ssm.phyto$states.se

states <- as.data.frame(states)
statesSE <- as.data.frame(statesSE)

states_long <- states %>% 
  pivot_longer("V1":"V120", names_to="Date", values_to="fitted")
states_long <- states_long %>% 
  mutate(Date = rep(dates, 5))

states_long <- states_long %>% 
  mutate(state = c(rep(phytoplankton[1], 120),
                   rep(phytoplankton[2], 120),
                   rep(phytoplankton[3], 120),
                   rep(phytoplankton[4], 120),
                   rep(phytoplankton[5], 120)
  ))

statesSE_long <- statesSE %>% 
  pivot_longer("V1":"V120", names_to="Date", values_to="se")

states_long <- states_long %>% 
  mutate(se = statesSE_long$se)

states_long$lb <- states_long$fitted - states_long$se
states_long$ub <- states_long$fitted + states_long$se

# save long state estimates
save(states_long, file=here::here("Lab-3", "Team-2", "data", "ssm.phyto_long.Rda"))

plot <- ggplot(data = states_long, aes(x = Date, y = fitted, group = state)) +
  geom_line() +
  labs(x = "", 
       y="State Estimate",
       title='State-space estimates of phyotplankton abundance in Lake Washington') +
  geom_ribbon(aes(ymin=lb, ymax=ub, fill=state), alpha=0.35, linetype=0) +
  theme_classic() +
  theme(legend.position="bottom") + 
  scale_x_discrete(breaks = c(1980, 1983, 1986, 1989, 1992, 1995))
plot
```

Next we consider models which treat all five time series of phytoplankton groups as representative samples from one larger underlying state of phytoplankton abundance in Lake Washington. Once again, the two covariate structures described above will be evaluated for best fit. 

```{r zp_modelbuild-phyto1s, echo=FALSE, warning=FALSE, message=FALSE}
# build model - seasonal
phyto_month.list <- list(
  B = "identity",
  U = "zero",
  Q = "diagonal and unequal",
  Z = matrix(rep(1, nrow(data_phyto))),
  A = "zero",
  R = "diagonal and equal",
  c = month_cov,
  C= "unconstrained",
  x0 = "unequal",
  V0 = "zero",
  tinitx = 0
)

# build model - climate
phyto_climate.list <- list(
  B = "identity",
  U = "zero",
  Q = "diagonal and unequal",
  Z = matrix(rep(1, nrow(data_phyto))),
  A = "zero",
  R = "diagonal and equal",
  c = data_phytoclimate,
  C= "unconstrained",
  x0 = "unequal",
  V0 = "zero",
  tinitx = 0
)

# modeling - seasonal
if(!file.exists(here::here("Lab-3", "Team-2", "data", "phyto_month_1s.rds"))){
  ssm.phyto_month <- MARSS(data_phyto, model = phyto_month.list, method = "kem", control = con_list)
  saveRDS(ssm.phyto_month, file="Lab-3/Team-2/data/phyto_month_1s.rds")
}
# load in ssm.phyto_month
ssm.phyto_month <- readRDS(file=here::here("Lab-3", "Team-2", "data", "phyto_month_1s.rds"))

# modeling - climate
if(!file.exists(here::here("Lab-3", "Team-2", "data", "phyto_climate_1s.rds"))){
  ssm.phyto_climate <- MARSS(data_phyto, model = phyto_climate.list, method = "kem", control = con_list)
  saveRDS(ssm.phyto_climate, file=here::here("Lab-3", "Team-2", "data", "phyto_climate_1s.rds"))
}
# load in ssm.phyto_climate
ssm.phyto_climate <- readRDS(file=here::here("Lab-3", "Team-2", "data", "phyto_climate_1s.rds"))

# model comparison
# an empty list to collect outputs
out_list <- list()

# Add each model's name and AICc 
out_list[[1]] <- data.frame(Model = "seasonal", AICc = ssm.phyto_month$AICc)
out_list[[2]] <- data.frame(Model = "climate", AICc = ssm.phyto_climate$AICc)

# Combine all into one dataframe
out <- do.call(rbind, out_list)
out

ssm.phyto <- ssm.phyto_month
```

As before, AICc finds that the model including seasonal covariates is a better fit for phytoplankton data. This overall abundance of phytoplankton is described in the figure below.

```{r zp_plot-phyto1s, echo=FALSE, warning=FALSE, message=FALSE}
# grabbing model estimates
states <- ssm.phyto$states
statesSE <- ssm.phyto$states.se

states <- as.data.frame(states)
statesSE <- as.data.frame(statesSE)

states_long <- states %>% 
  pivot_longer("V1":"V120", names_to="Date", values_to="fitted")
states_long <- states_long %>% 
  mutate(Date = dates)

states_long <- states_long %>% 
  mutate(state = c(rep(phytoplankton[1], 120)
  ))

statesSE_long <- statesSE %>% 
  pivot_longer("V1":"V120", names_to="Date", values_to="se")

states_long <- states_long %>% 
  mutate(se = statesSE_long$se)

states_long$lb <- states_long$fitted - states_long$se
states_long$ub <- states_long$fitted + states_long$se

# save long state estimates
save(states_long, file=here::here("Lab-3", "Team-2", "data", "ssm.phyto_long_1s.Rda"))

plot <- ggplot(data = states_long, aes(x = Date, y = fitted, group = state)) +
  geom_line() +
  labs(x = "", 
       y="State Estimate",
       title='State-space estimates of phyotplankton abundance in Lake Washington') +
  geom_ribbon(aes(ymin=lb, ymax=ub, fill=state), alpha=0.35, linetype=0) +
  theme_classic() +
  theme(legend.position="bottom") + 
  scale_x_continuous(breaks = c(1980, 1983, 1986, 1989, 1992, 1995))
plot
```

This model of phytoplankton abundance as one underlying state will be used going forward in the dynamic factor analysis of zooplankton grazers. 

#### Modeling grazing zooplankton

With phytoplankton abundance estimates in hand, we now move on to the evaluation of zooplankton grazers using dynamic factor analysis. We begin by considering which number of underlying will best fit the model. We evealutate models with 1 to 5 underlying factors (as there are six time series of zooplankton grazers it is appropriate at most to consider one fewer factor). These initial models will include covariates describing temperature, pH, and phytoplankton abundance. One the best fit is determined in terms of underlying factors, these covariates will be additionally evaluated to determine whether they are all necessaty. 

```{r zp_modelbuild-zoo, echo=FALSE, warning=FALSE, message=FALSE}
# get only zooplankton
data_zoo <- t(data[, zooplankton])
colnames(data_zoo) <- dates

# get zooplankton climate covariates
data_zooclimate <- t(data[, zooclimate])

# stack up the old covariate matrix
zoo.d <- rbind(states_long$fitted, data_zooclimate)
rownames(zoo.d) <- c("Phyto", "Temp", "pH")
# print(cor(t(zoo.d)))
  # not super highly correlated

# dynamic factor anaylsis - zooplankton

# preparing to loop
## empty lists
# loadings <- list()
aicc_list <- list()

# looping
for(i in 1:(nrow(data_zoo)-1)){
  
  ## loading matrices
  NN <- nrow(data_zoo)
  MM <- i
  ZZ <- matrix(paste0("z", seq(NN*MM)), NN, MM)
  ZZ[upper.tri(ZZ)] <- 0
  # loadings[[i]] <- assign(paste0("ZZ", i), ZZ)


  ## build model
  zoo.list <- list(
    B = "identity",
    U = "zero",
    Q = "identity",
    # Z = loadings[[i]],
    Z = ZZ,
    A = "zero",
    R = "diagonal and unequal",
    d = zoo.d,
    D = "unconstrained",
    x0 = "equal",
    V0 = "zero",
    tinitx = 0
  )

  ## modeling
  if(!file.exists(here::here("Lab-3", "Team-2", "data", paste("zoo_", i, ".rds", sep = "")))){
    ssm <- MARSS(data_zoo, model = zoo.list, method = "kem", control = con_list)
    saveRDS(ssm, file=here::here("Lab-3", "Team-2", "data", paste("zoo_", i, ".rds", sep = "")))
  }
  # load in ssm.phyto_climate
  ssm <- readRDS(file=here::here("Lab-3", "Team-2", "data", paste("zoo_", i, ".rds", sep = "")))
  
  ## model comparison
  # Add each model's name and AICc 
  aicc_list[[i]] <- data.frame(Model = paste("zoo_", i, sep = ""), AICc = ssm$AICc)
}

# Combine all into one dataframe
aicc <- do.call(rbind, aicc_list)
aicc
  # per AICc, the model with 4 underlying factors displays the best fit

ssm <- readRDS(file=here::here("Lab-3", "Team-2", "data", "zoo_4.rds"))
```

AICc determines that the model with four underlying factors is the best fit per the data. WE now move on to evaluating which set of covariates are best suited to improve model fit. All possible combinations of water temperature, pH levels, and phytoplankton abundance will be considered. Also, a seasonal covariate will be introduced and considered both on it's own and as part of other covariate combinations, yielding a total of 15 covariate structures to be considered.

```{r zp_modelcov-zoo, echo=FALSE, warning=FALSE, message=FALSE}
# seasonality as sine/cosine waves
cos_t <- cos(2 * pi * seq(120) / 12)
sin_t <- sin(2 * pi * seq(120) / 12)

# checking different levels of covariates
d.all <- rbind(zoo.d, cos_t, sin_t)
d.phyto <- t(as.matrix(zoo.d[1,]))
d.temp <- t(as.matrix(zoo.d[2,]))
d.ph <- t(as.matrix(zoo.d[3,]))
d.phytemp <- as.matrix(zoo.d[1:2,])
d.phyph <- as.matrix(zoo.d[c(1,3),])
d.tempph <- as.matrix(zoo.d[2:3,])
d.cov <- zoo.d
d.S <- as.matrix(d.all[c(4:5),])
d.phytoS <- as.matrix(d.all[c(1, 4:5),])
d.tempS <- as.matrix(d.all[c(2, 4:5),])
d.phS <- as.matrix(d.all[c(3, 4:5),])
d.phytempS <- as.matrix(d.all[c(1:2, 4:5),])
d.phyphS <- as.matrix(d.all[c(1,3:5),])
d.tempphS <- as.matrix(d.all[2:5,])

# lil lists
covaicc_list <- list()
cov_list <- list(d.all, d.phyto, d.temp, d.ph, d.phytemp, d.phyph, d.tempph, d.cov, d.S, d.phytoS, d.tempS, d.phS, d.phytempS, d.phyphS, d.tempphS)

mod_list = list(m = 4, R = "diagonal and unequal")
for(i in 1:length(cov_list)){
  if(!file.exists(here("Lab-3", "Team-2", "data", paste("cov_", i, ".rds", sep = "")))){
    dfa <- MARSS(data_zoo, model = mod_list, form = "dfa", z.score = FALSE,
                    control = con_list, covariates = cov_list[[i]])
    saveRDS(dfa, file=here("Lab-3", "Team-2", "data", paste("cov_", i, ".rds", sep = "")))
  }
  # load in ssm.phyto_climate
  dfa <- readRDS(file=here("Lab-3", "Team-2", "data", paste("cov_", i, ".rds", sep = "")))
  covaicc_list[[i]] <- data.frame(Model = paste("cov_", i, sep = ""), AICc = dfa$AICc)
}

# Combine all into one dataframe
covaicc <- do.call(rbind, covaicc_list)
covaicc
  # per AICc, the model including phyoplankton abundance, pH levels, and a seasonal component with 4 underlying factors displays the best fit
ssm <- readRDS(file=here("Lab-3", "Team-2", "data", "cov_14.rds"))

covar_effects <- coef(ssm, type = "matrix")$D[, 1:2]
colnames(covar_effects) <- list("Phytoplankton", "pH")
covar_effects
```

AICc finds that a model which allows for four underlying factors and includes phytoplankton abundance and pH levels as well as a seasonal component best describes the zooplankton grazer abundance data. The effects of phytoplankton abundance and pH levels on each species of grazer are reported in the table above.

# Results

## Phytoplankton Abundance

## Zooplankton Abundance

## Grazer abundance, phytoplankton covariates

With a best fit model in hand, the next step is to determine a rotation matrix H that maximizes the difference between loadings in Z, to better understand what grazer zooplankton species abundances are being driven by which underlying factors.

```{r zp_hmatrix, echo=FALSE, warning=FALSE, message=FALSE}
# the ro-to
Z_est <- coef(ssm, type = "matrix")$Z
H_inv <- varimax(Z_est)$rotmat
Z_rot <- Z_est %*% H_inv
proc_rot <- solve(H_inv) %*% ssm$states
H_inv
Z_est
Z_rot
```


```{r zp_factor loadings, echo=FALSE, warning=FALSE, message=FALSE}
# plotting estimated factors and loadings (?)
## plot labels
ylbl <- zooplankton
w_ts <- seq(dim(data_zoo)[2])
minZ <- 0
N_ts <- dim(data_zoo)[1]
clr <- c("brown2", "blue4", "darkgreen", "darkred", "purple", "gold", "orange3", "cyan3")

## set up plot area
layout(matrix(c(1:10), nrow(proc_rot), 2), widths = c(2,1))
par(mai = c(0.5, 0.5, 0.5, 0.1), omi = c(0, 0, 0, 0))

## plot the processes
for(i in 1:nrow(proc_rot)) {
  ylm <- c(-1, 1) * max(abs(proc_rot[i,]))
  ## set up plot area
  plot(w_ts,proc_rot[i,], type = "n", bty = "L",
       ylim = ylm, xlab = "", ylab = "", xaxt = "n")
  ## draw zero-line
  abline(h = 0, col = "gray")
  ## plot trend line
  lines(w_ts, proc_rot[i,], lwd = 2)
  lines(w_ts, proc_rot[i,], lwd = 2)
  ## add panel labels
  mtext(paste("State",i), side = 3, line = 0.5)
  axis(1, 12 * (0:dim(data_zoo)[2]) + 1, yr_frst + 0:dim(data_zoo)[2])
}

## plot the loadings
ylm <- c(-1, 1) * max(abs(Z_rot))
for(i in 1:nrow(proc_rot)) {
  plot(x = c(1:N_ts)[abs(Z_rot[,i])>minZ],
       y = as.vector(Z_rot[abs(Z_rot[,i])>minZ,i]),
       type = "h",
       lwd = 2, xlab = "", ylab = "", xaxt = "n", ylim = ylm,
       xlim = c(0.5, N_ts + 0.5), col = clr)
  for(j in 1:N_ts) {
    if(Z_rot[j,i] > minZ) {text(j, -0.03, ylbl[j], srt = 90, adj = 1, cex = 1.2, col = clr[j])}
    if(Z_rot[j,i] < -minZ) {text(j, 0.03, ylbl[j], srt = 90, adj = 0, cex = 1.2, col = clr[j])}
    abline(h = 0, lwd = 1.5, col = "gray")
  } 
  mtext(paste("Factor loadings on state", i), side = 3, line = 0.5)
}
```

The above plot shows no clear pattern in species being driven by one trend or another. Perhaps cyclops and daphnia exhibit similar drivers and likewise with non-daphnid cladocerans and non-colonial rotifers. Also, despite including covariates to account for seasonal trends, there still appears to be some seasonality in states 1, 2, & 4.

```{r zp_model fit, echo=FALSE, warning=FALSE, message=FALSE}
# plotting data and model fit
get_DFA_fits <- function(MLEobj, dd = NULL, alpha = 0.05) {
  ## empty list for results
  fits <- list()
  ## extra stuff for var() calcs
  Ey <- MARSS:::MARSShatyt(MLEobj)
  ## model params
  ZZ <- coef(MLEobj, type="matrix")$Z
  ## number of obs ts
  nn <- dim(Ey$ytT)[1]
  ## number of time steps
  TT <- dim(Ey$ytT)[2]
  ## get the inverse of the rotation matrix
  H_inv <- varimax(ZZ)$rotmat
  ## check for covars
  if(!is.null(dd)) {
    DD <- coef(MLEobj, type = "matrix")$D
    ## model expectation
    fits$ex <- ZZ %*% H_inv %*% MLEobj$states + DD %*% dd
  } else {
    ## model expectation
    fits$ex <- ZZ %*% H_inv %*% MLEobj$states
  }
  ## Var in model fits
  VtT <- MARSSkfss(MLEobj)$VtT
  VV <- NULL
  for(tt in 1:TT) {
    RZVZ <- coef(MLEobj, type = "matrix")$R - ZZ %*% VtT[,,tt] %*% t(ZZ)
    SS <- Ey$yxtT[,,tt] - Ey$ytT[,tt,drop = FALSE] %*% t(MLEobj$states[,tt,drop = FALSE])
    VV <- cbind(VV, diag(RZVZ + SS %*% t(ZZ) + ZZ %*% t(SS)))
  }
  SE <- sqrt(VV)
  ## upper & lower (1-alpha)% CI
  fits$up <- qnorm(1-alpha/2)*SE + fits$ex
  fits$lo <- qnorm(alpha/2)*SE + fits$ex
  return(fits)
}
## get model fits & CI's
mod_fit <- get_DFA_fits(ssm)

## set up plotting area
par(mfrow = c(N_ts, 1), mai = c(0.5, 0.7, 0.1, 0.1), omi = c(0, 0, 0, 0))

## plot the fits
for(i in 1:N_ts) {
  up <- mod_fit$up[i,]
  mn <- mod_fit$ex[i,]
  lo <- mod_fit$lo[i,]
  plot(w_ts, mn, type = "n", ylim = c(min(lo), max(up)),
       cex.lab = 1.2,
       xlab = "", ylab = ylbl[i], xaxt = "n")
  axis(1, 12 * (0:dim(data_zoo)[2]) + 1, yr_frst + 0:dim(data_zoo)[2])
  points(w_ts,data_zoo[i,], pch = 16, col = clr[i])
  lines(w_ts, up, col = "darkgray")
  lines(w_ts, mn, col = "black", lwd = 2)
  lines(w_ts, lo, col = "darkgray")
}
```

Finally, we compare model fit to the actual data for all six groupings of grazer zooplankton. Generally fit is good, with the model proving better at estimating abundances of certain species of zooplankton than others. Diaptomus, non-colonial rotifers, and conochilus estimates fit the data fairly well, while daphnia and to (maybe) a lesser extent non-daphnid cladocerans and cyclops estimates are less accurate predictions of the data observed.

# Discussion

Including phytoplankton abundance as a covariate in models describing the abundance of various grazing zooplaknton proved fruitful. A dynamic factor analysis revealed four underlying factors driving/predicting grazer abundance in Lake Washington. Phytoplaknton abundances were included in the model of best fit, along with pH levels and a seasonal component. While some species were modeled very effectively with this structure, others were not estimated as efficiently. A potential next step would be modeling each species of grazer separately in order to better capture the effects of phytoplankton abundance on each.

# Team contributions

Each team member contributed to a specific analysis. Tina modeled phytoplankton, Matheus modeled zooplankton, and Brian modeled zooplankton grazers with phytoplankton as covariates.

