---
title: Team 1 - Lab 1
subtitle: Dynamic Factor Analysis (DFA) - Common trends in plankton data
author: "Name 1, Name 2, Name 3"
date: April XX, 2025
output: 
  html_document:
    code-folding: true
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
options(dplyr.summarise.inform = FALSE)
```

# Data

Describe what plankton groups and covariates you used and indicate any temporal subsetting you chose.

I selected the five primary producer from the phytoplankton group, which include
* `Cryptomonas`: small brown or green algae (edible)  
* `Diatoms`: small algae rich in silica (edible)  
* `Greens`: general class of algae (edible)  
* `Unicells`: very small algae (edible)  
* `Other.algae`: generic catch-all for atypical species (edible)

because there are too many missing values in `Bluegreens`.  

For the covariates, I selected environmental factors including
* `Temp`: water temperature in degrees C  
* `TP`: total phosphorous concentration in mg m<sup>-3</sup>  
* `pH`: pH.


## Load the data

```{r load_data}
## load MARSS for data and analyses
library(MARSS)

## load the raw data (there are 3 datasets contained here)
data(lakeWAplankton, package = "MARSS")

## we want `lakeWAplanktonTrans`, which has been transformed
## so the 0's are replaced with NA's and the data z-scored
all_dat <- lakeWAplanktonTrans
```

## Wrangle the data

* You can choose any taxa to include in your analyses, but make sure there are at least 5 groups. You can mix phytoplankton and zooplankton, if you'd like, but justify any choices you make. 

* You will want to subset the data to a time period with adequate observations for the taxa of interest. Your time window should include at lest 5 years (60 months).

```{r wrangle_data}
## use 10 years from 1985-1994
yr_frst <- 1985
yr_last <- 1994
plank_dat <- all_dat[all_dat[, "Year"] >= yr_frst & 
                     all_dat[, "Year"] <= yr_last,]

## create vector of phytoplankton group names
phytoplankton <- c("Cryptomonas", "Diatoms", "Greens",
                   "Unicells", "Other.algae")

## get only the phytoplankton
dat_1985 <- plank_dat[, phytoplankton]

## transpose data so time goes across columns
dat_1985 <- t(dat_1985)

## get number of time series
N_ts <- dim(dat_1985)[1]

## get length of time series
TT <- dim(dat_1985)[2]

## mean of each taxon
y_bar <- apply(dat_1985, 1, mean, na.rm = TRUE)

## de-mean the data - subtract the means
dat <- dat_1985 - y_bar

## assign new column names
spp <- rownames(dat_1985)
rownames(dat) <- spp
```

```{r plot_data}
## set plot colors for six groups
clr <- c("blue", "darkgreen", "darkred", "purple", "orange")

## initialize a counter
cnt <- 1

## set up plotting space for six time series
par(mfrow = c(N_ts, 1), mai = c(0.5, 0.7, 0.1, 0.1), omi = c(0, 0, 0, 0))

## loop through each phytoplankton group
for(i in spp){
  plot(dat[i,], bty = "L", xaxt = "n", pch = 16,
       xlab = "",
       ylab = "Abundance index", col = clr[cnt], type = "b")
  axis(1, 12 * (0:dim(dat_1980)[2]) + 1, yr_frst + 0:dim(dat_1980)[2])
  title(i)
  cnt <- cnt + 1
}

```

I plan to examine models that considered three underlying state options: one, two,
and three states. I will also test three observation error matrices,
diagonal and equal, diagonal and unequal, and equal variance and covariance.
I aim to consider all combinations of three environmental covariates.

```{r model_params}
model.states <- 1:3
names(model.states) <- c("one_state", "two_states","three_states")

# testing different observation error matrices
R.models <- c("diagonal and unequal","equalvarcov", "diagonal and equal")

# covariates must be z-scored if the response variables are demeaned, and subset covariates to same time window and transpose
cov_temp <- t(zscore(plank_dat[,"Temp", drop = FALSE]))
cov_TP   <- t(zscore(plank_dat[,"TP", drop = FALSE]))
cov_pH   <- t(zscore(plank_dat[,"pH", drop = FALSE]))
cov_temp_TP   <- t(zscore(plank_dat[, c("Temp", "TP")]))
cov_TP_pH     <- t(zscore(plank_dat[, c("TP", "pH")]))
cov_temp_pH   <- t(zscore(plank_dat[, c("Temp", "pH")]))
cov_all       <- t(zscore(plank_dat[, c("Temp", "TP", "pH")]))

d.models <- list(
  Temp            = cov_temp,
  TP              = cov_TP,
  pH              = cov_pH,
  "Temp and TP"   = cov_temp_TP,
  "TP and pH"     = cov_TP_pH,
  "Temp and pH"   = cov_temp_pH,
  "Temp,TP, and pH" = cov_all
)

# define model list 
mod.list = list(
  A = "zero", # this is set to zero because it is the offset/scaling
  Q = "identity", # Q is set to the default value, the identify matrix, which can also be "diagonal and equal" or "diagonal and unequal"
  x0 = "zero" # x0 can be also be set to "unconstrained" or "unequal" it might be useful to try these later
)

```



```{r MARSS_models_fitting}
## Create a results list to store model fits and AICs
model_fits <- list()
model_AICs <- data.frame(
  ModelID = character(),
  NumStates = integer(),
  R_Model = character(),
  Covariates = character(),
  AICc = numeric(),
  stringsAsFactors = FALSE
)

# Check and handle NAs in covariates (assuming plank_dat is already subsetted)
# Apply linear interpolation to covariates to fill NAs
library(zoo)
plank_df <- as.data.frame(plank_dat)
plank_df$Temp <- na.approx(plank_df$Temp, na.rm = FALSE)
plank_df$TP <- na.approx(plank_df$TP, na.rm = FALSE)
plank_df$pH <- na.approx(plank_df$pH, na.rm = FALSE)

# Recreate covariates with scaling after handling NAs
cov_temp <- t(scale(plank_df[,"Temp", drop = FALSE]))
cov_TP   <- t(scale(plank_df[,"TP", drop = FALSE]))
cov_pH   <- t(scale(plank_df[,"pH", drop = FALSE]))
cov_temp_TP   <- t(scale(plank_df[, c("Temp", "TP")]))
cov_TP_pH     <- t(scale(plank_df[, c("TP", "pH")]))
cov_temp_pH   <- t(scale(plank_df[, c("Temp", "pH")]))
cov_all       <- t(scale(plank_df[, c("Temp", "TP", "pH")]))

# Update d.models with corrected covariates
d.models <- list(
  Temp            = cov_temp,
  TP              = cov_TP,
  pH              = cov_pH,
  "Temp and TP"   = cov_temp_TP,
  "TP and pH"     = cov_TP_pH,
  "Temp and pH"   = cov_temp_pH,
  "Temp,TP, and pH" = cov_all
)

model_id <- 1
for(n_state in model.states){
  for(R_mod in R.models){
    for(d_name in names(d.models)){
      
      # Define the Z matrix with correct dimensions
      if(n_state == 1){
        Z.model <- matrix("z", N_ts, n_state)  # All species load on single trend
      } else {
        # Create unique loadings for each species and state
        Z.model <- matrix(paste0("z", 1:(N_ts * n_state)), N_ts, n_state)
      }
      
      # Define model list with updated Z and correct dimensions
      mod.list <- list(
        Z = Z.model,
        R = R_mod,
        A = "zero",
        Q = "identity",
        x0 = "zero",
        D = "unconstrained",
        d = d.models[[d_name]]
      )
      
      # Fit the model
      fit <- tryCatch({
        MARSS(dat, model = mod.list, silent = TRUE, control = list(maxit = 5000))
      }, error = function(e) NULL)
      
      # Store results if successful
      if(!is.null(fit) && !is.null(fit$AICc)){
        model_fits[[paste0("model", model_id)]] <- fit
        model_AICs <- rbind(model_AICs, data.frame(
          ModelID = paste0("model", model_id),
          NumStates = n_state,
          R_Model = R_mod,
          Covariates = d_name,
          AICc = fit$AICc
        ))
      } else {
        cat("Model", model_id, "failed or AICc missing\n")
      }
      
      model_id <- model_id + 1
    }
  }
}

```

```{r MARSS_models_comparison}
library(dplyr)
model_AICs %>% 
  arrange(AICc) %>% 
  head(10)

# Find the best model ID
best_model_id <- model_AICs$ModelID[which.min(model_AICs$AICc)]

# Get the best model object
best_model <- model_fits[[best_model_id]]

# View model summary
summary(best_model)

# View estimated parameters
coef(best_model, type = "matrix")$Z  # Loadings
coef(best_model, type = "matrix")$D  # Covariate effects
```
```{r best_model_visualization}
# Plot the estimated trends
library(ggplot2)
trends <- data.frame(
  Time = 1:TT,
  Trend = as.vector(best_model$states),
  SE = as.vector(best_model$states.se)
)

ggplot(trends, aes(x = Time, y = Trend)) +
  geom_line() +
  geom_ribbon(aes(ymin = Trend - 1.96*SE, ymax = Trend + 1.96*SE), alpha = 0.2) +
  labs(title = "Estimated Common Trends")

# Get fitted values
fitted_vals <- fitted(best_model)

# Plot for each species
ggplot(fitted_vals, aes(x = t, y = .fitted)) +
  geom_line(color = "blue") +
  geom_point(aes(y = y), color = "black", size = 1) +
  facet_wrap(~.rownames, scales = "free_y") +
  labs(x = "Time", y = "Abundance", title = "Model Fits vs Observations")

# Get covariate coefficients
covar_effects <- coef(best_model, type = "matrix")$D

# Plot effects
data.frame(
  Species = rownames(dat),
  covar_effects
) %>% 
  tidyr::pivot_longer(-Species, names_to = "Covariate", values_to = "Effect") %>% 
  ggplot(aes(x = Covariate, y = Effect, color = Species)) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  coord_flip() +
  theme_minimal()
```

```{r best_model_diagnostics}
# Residuals analysis
resids <- residuals(best_model)
plot(resids)

# Autocorrelation check
acf(resids$model.residuals, na.action = na.pass)
```

# General tasks

Each group has the same general tasks, but you will adapt them as you work on the data.

1) Find the most parsimonious model among a set that examines the effects of environmental covariates and/or an indicator of seasonality, varying numbers of trends, and different forms of variance-covariance matrices for the observation errors.

2) Plot trends and individual loadings for the model identified in task (1) above.

3) Plot model fits and uncertainty intervals for the model identified in task (1) above.

4) Describe the effects of environmental or dummy variables on (possibly seasonal) patterns in the data.






# Methods

Please address the following in your methods:

* Which plankton taxa did you choose and how did you choose them?
   
* What time period(s) did you examine and why?

* What environmental or dummy variables did you include and why?

* What forms of models did you fit (ie, write them out in matrix form)?

* What sort of model diagnostics did you use to examine model assumptions?


# Results


# Discussion


# Team contributions

Example: "All team members helped decide on the goal and ran the analyses for the individual regions. Team members 2 & 3 wrote most of the code for the analysis. Team member 1 worked on the plotting section of the report using and adapting code that team member 3 wrote. All team members helped edit the report and wrote the discussion together."

