---
title: Climate Variables
author: "E Holmes"
date: "April 22, 2025"
output:
  html_document:
    code-folding: yes
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

```{r message=FALSE}
library(tidyverse)
options(dplyr.summarise.inform = FALSE)
```

## Climate Indices

See also <https://psl.noaa.gov/data/atmoswrit/corr/timeseries/> and <https://psl.noaa.gov/data/climateindices/list/>

| Index                            | Region                            | Influences                                         | Notes                                                                 | Data Source |
|----------------------------------|------------------------------------|----------------------------------------------------|-----------------------------------------------------------------------|-------------|
| **PDO** (Pacific Decadal Osc.)  | North Pacific                      | Ocean temps, productivity, stratification          | Warm phase → poor survival in Alaska, good in CA (and vice versa)    | [PDO CSV](https://www.ncei.noaa.gov/pub/data/cmb/ersst/v5/index/ersst.v5.pdo.dat) |
| **ONI** (Oceanic Niño Index) | Equatorial Pacific (global effects) | Upwelling, SST, marine food web                    | El Niño often harmful for juvenile salmon on West Coast              | [ONI TXT](https://www.cpc.ncep.noaa.gov/data/indices/oni.ascii.txt) |
| **NPGO** ([North Pacific Gyre Osc](https://www.o3d.org/npgo/)) | North Pacific                    | Salinity, nutrient flux, productivity              | Positively correlated with salmon survival (esp. CA Current)         | [NPGO TXT](https://www.o3d.org/npgo/data/NPGO.txt) |
| Cold Pool Index | Bering Sea shelf  | Bottom temperature, habitat compression            | Cold pool extent influences migration, prey availability, and predation | [coldpool r package](https://github.com/afsc-gap-products/coldpool) |
| **NEP SST** (NE Pacific SST)     | Coastal Alaska, BC, WA, OR, CA     | Marine heatwaves, thermal stress | associated with reduced survival  |  |
| **Upwelling Index** (Bakun)      | California/Oregon coast            | Nutrient delivery, primary productivity            | Drives ocean growth conditions for smolts                            | [NOAA Upwelling](https://mjacox.com/upwelling-indices/) |
| **PNA** (Pacific-North American) | North America/western U.S.         | Winter temps, river flow, snowpack                 | Influences freshwater survival, stream temperatures [info](https://www.climate.gov/news-features/understanding-climate/climate-variability-pacific-north-american-pattern)               | [PNA TXT](https://www.ncei.noaa.gov/access/monitoring/pna/) |
| **MEI / MEIv2** (Multivariate ENSO Index) | ENSO-linked               | Broad atmospheric/ocean composite                  | More integrated than ONI; includes wind, pressure, SST               | [MEIv2 TXT](https://psl.noaa.gov/enso/mei/) |
| West Coast Stream Flow |   |    |    | [Daily West Coast Stream Flow](https://noaa-nwfsc.github.io/Streamflow-Means-NWM/) |


## Loading data

This can be painful. Share your code in the class repo so others can use your work! [Post here](https://github.com/atsa-es/fish550-2025/discussions/4)

```{r}
pdo_url <- "https://www.ncei.noaa.gov/pub/data/cmb/ersst/v5/index/ersst.v5.pdo.dat"
pdo <- read.table(pdo_url, skip = 2, header = FALSE, check.names = FALSE)
pdo[pdo==99.99] <- NA
pdo_vec <- as.numeric(t(pdo[, 2:13]))
pdots <- ts(pdo_vec, freq=12, start=c(1854, 1))
plot(pdots)
```

## Prepping your covariates for `MARSS()`

You need a matrix:

* Columns = time and number of columns must match your data
* Rows = 1 row for each covariate

**Example for PDO**

Let's say we want to use the April-June average PDO. 

```{r}
# col 1 is year
years <- which(pdo[,1] %in% 2000:2025)
spring_pdo <- rowMeans(pdo[years, 5:7], na.rm = TRUE)
cmat <- matrix(spring_pdo, nrow=1)
cmat
```
This is now $\mathbf{c}_t$ in 
$$
\mathbf{x}_t = \mathbf{x}_{t-1} + \mathbf{u} + \mathbf{C}\mathbf{c}_t + \mathbf{w}_t
$$

## Use PDO to predict steelhead returns

Idea is that pdo ($p_t$) is correlated with the fluctuations in log spawner returns ($x_t$).

$$
x_t = x_{t-1} + u + (\alpha p_t + w_t)
$$
But we have 4 steelhead time series for this example. We can write this as:
$$
\begin{bmatrix}x_{1,t}\\x_{2,t}\\x_{3,t}\\x_{4,t}\end{bmatrix}=
\begin{bmatrix}x_{1,t-1}\\x_{2,t-1}\\x_{3,t-1}\\x_{4,t-1}\end{bmatrix} + 
\begin{bmatrix}u_1\\u_2\\u_3\\u_4\end{bmatrix}+ 
\begin{bmatrix}\alpha_1\\\alpha_2\\\alpha_3\\\alpha_4\end{bmatrix} p_t +
\begin{bmatrix}w_1\\w_2\\w_3\\w_4\end{bmatrix}
$$

What about the $w_t$'s? How are they related? Let's say they are independent with unequal variances (the $x_t$ are diagonal and unequal). What about our observations? Let's say each $y_t$ measures one $x_t$ with independent and equal error (R is diagonal and equal). Let's fit this model with MARSS.

Wrangle the raw spawner data to get the $x_t$.
```{r}
library(tidyverse)
load("Data_Images/esa-salmon.rda")
esu <- unique(esa.salmon$esu_dps)
esunum <- which(esu == "Steelhead (Upper Columbia River DPS)")
esuname <- esu[esunum]

dat <- esa.salmon %>% 
  subset(esu_dps == esuname) %>% # get only this ESU
  mutate(log.spawner = log(value)) %>% # create a column called log.spawner
  select(esapopname, spawningyear, log.spawner) %>% # get just the columns that I need
  pivot_wider(names_from = "esapopname", values_from = "log.spawner") %>% 
  column_to_rownames(var = "spawningyear") %>% # make the years rownames
  as.matrix() %>% # turn into a matrix with year down the rows
  t() # make time across the columns
# MARSS complains if I don't do this
dat[is.na(dat)] <- NA
# Clean up rownames
tmp <- rownames(dat)
tmp <- stringr::str_replace(tmp, "Steelhead [(]Upper Columbia River DPS[)]", "")
tmp <- stringr::str_replace(tmp, "River - summer", "")
tmp <- stringr::str_trim(tmp)
rownames(dat) <- tmp
```

Get the PDO for the years in our data:
```{r}
# col 1 is year
years <- which(pdo[,1] %in% colnames(dat))
spring_pdo <- rowMeans(pdo[years, 5:7], na.rm = TRUE)
cmat <- matrix(spring_pdo, nrow=1)
```

Check that `cmat` and `dat` have the same number of columns.
```{r}
dim(cmat)
dim(dat)
```

Specify a model
```{r}
# pdo
mod.list1 <- list(
  U = "unequal",
  R = "diagonal and equal",
  Q = "diagonal and unequal",
  c = cmat,
  C = "unequal"
)

# no pdo
mod.list2 <- list(
  U = "unequal",
  R = "diagonal and equal",
  Q = "diagonal and unequal")
```

```{r}
library(MARSS)
fit1 <- MARSS(dat, model=mod.list1)
fit2 <- MARSS(dat, model=mod.list2)
```

Which one fits better based on AIC? The model without PDO is better (lower AICc).

```{r}
fit1$AICc
fit2$AICc
```

## Resources

Chapter 7 MARSS models. ATSA Lab Book. <https://atsa-es.github.io/atsa-labs/chap-mss.html>

Chapter 8 MARSS models with covariates. ATSA Lab Book. <https://atsa-es.github.io/atsa-labs/chap-msscov.html>

Chapter 16 Modeling cyclic sockeye <https://atsa-es.github.io/atsa-labs/chap-cyclic-sockeye.html>
