---
title: Climate Variables
author: "E Holmes"
date: "April 22, 2025"
output:
  html_document:
    code-folding: yes
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

```{r message=FALSE}
library(tidyverse)
options(dplyr.summarise.inform = FALSE)
```

## Climate Indices

See also <https://psl.noaa.gov/data/atmoswrit/corr/timeseries/> and <https://psl.noaa.gov/data/climateindices/list/>.

I have created a dataset of seasonal climate variables from the PSL page. 

Seasons are defined as:

- **Winter**: January, February, March
- **Spring**: April, May, June
- **Summer**: May, June, July, August
- **Fall**: September, October, November
- **Annual**: whole year

Variables:

| Index                                | Region                              | Influences on Salmon                                  | Notes                                                      |
|--------------------------------------|-------------------------------------|-------------------------------------------------------|------------------------------------------------------------|
| **PDO** (Pacific Decadal Oscillation) | North Pacific                      | Ocean productivity, survival of juveniles             | Warm phase reduces survival in Alaska, improves in CA.    |
| **ONI** (Oceanic Niño Index)          | Equatorial Pacific (global effects) | Marine heatwaves, prey availability                   | El Niño years generally reduce survival along West Coast. |
| **MEIv2** (Multivariate ENSO Index)   | Equatorial Pacific/global          | Similar to ONI but broader; marine habitat stress     | Integrated index of ENSO variability; affects productivity.|
| **PNA** (Pacific-North American Pattern) | North America and North Pacific  | River flows, snowpack, freshwater survival            | Positive PNA linked to warmer/drier winters, lower survival. |
| **EPO** (East Pacific Oscillation)    | Eastern North Pacific              | Winter ocean conditions                              | Negative EPO can improve early marine survival conditions. |
| **WP** (Western Pacific Pattern)      | Western Pacific, Eastern Asia      | Atmospheric circulation affecting ocean entry timing  | Alters timing and success of smolt outmigration indirectly.|
| **CENSO** (Central ENSO Index)         | Central Pacific                    | Central Pacific ENSO variability                     | May modulate effects of traditional El Niño on survival.  |
| **WHWP** (Western Hemisphere Warm Pool) | Atlantic and Eastern Pacific      | Hurricane activity (Atlantic)                        | Not directly linked to salmon survival in the North Pacific. |
| **IPOTPI** (Indo-Pacific Ocean Temperature Pattern Index) | Indo-Pacific | Tropical SST variability patterns                     | Emerging index; indirect possible link to broad-scale climate affecting salmon. |
| **NP** (North Pacific Index)          | North Pacific                      | Aleutian Low strength, ocean upwelling                | Strong Aleutian Low (lower NP) → better salmon survival.  |
| **PACWARM** (Pacific Warm Pool Index) | Equatorial Pacific                 | Warm pool expansion influencing ENSO events           | Can modulate strength of El Niño impacts on marine survival.|
| **GMSST** (Global Mean Sea Surface Temperature) | Global Ocean                     | Baseline warming trend                               | Higher GMSST linked to long-term habitat stress.           |


## Loading data

I have created a data frame of North Pacific climate variables from NOAA PSL. Update your atsalibrary to version 1.6 to get the climate data or load `np_climate.RData` from the `Data_Images` directory.

```{r message=FALSE, warning=FALSE}
library(atsalibrary)
data("np_climate")
# this has seasonal and annual averages
# use ?np_climate for info
colnames(np_climate_month)
```

```{r}
pdots <- ts(np_climate_month$pdo, freq=12, start=c(np_climate_month$year[1], 1))
plot(pdots)
```

```{r message=FALSE, warning=FALSE}
library(atsalibrary)
data("np_climate")
# this has seasonal and annual averages
# use ?np_climate for info
colnames(np_climate_seasonal)
```


## Prepping your covariates for `MARSS()`

You need a matrix:

* Columns = time and number of columns must match your data
* Rows = 1 row for each covariate

**Example for PDO**

Let's say we want to use the April-June average PDO. We would construct a matrix with ncols being the years and 1 row for spring PDO.

```{r}
# col 1 is year
years <- which(np_climate_seasonal$year %in% 2000:2024)
spring_pdo <- np_climate_seasonal$pdo.spring[years]
cmat <- matrix(spring_pdo, nrow=1)
cmat
```
This is now $\mathbf{c}_t$ in 
$$
\mathbf{x}_t = \mathbf{x}_{t-1} + \mathbf{u} + \mathbf{C}\mathbf{c}_t + \mathbf{w}_t
$$

## Use PDO to predict steelhead returns

Idea is that pdo ($p_t$) is correlated with the fluctuations in log spawner returns ($x_t$).

$$
x_t = x_{t-1} + u + (\alpha p_t + w_t)
$$
But we have 4 steelhead time series for this example. We can write this as:
$$
\begin{bmatrix}x_{1,t}\\x_{2,t}\\x_{3,t}\\x_{4,t}\end{bmatrix}=
\begin{bmatrix}x_{1,t-1}\\x_{2,t-1}\\x_{3,t-1}\\x_{4,t-1}\end{bmatrix} + 
\begin{bmatrix}u_1\\u_2\\u_3\\u_4\end{bmatrix}+ 
\begin{bmatrix}\alpha_1\\\alpha_2\\\alpha_3\\\alpha_4\end{bmatrix} p_t +
\begin{bmatrix}w_1\\w_2\\w_3\\w_4\end{bmatrix}
$$

What about the $w_t$'s? How are they related? Let's say they are independent with unequal variances (the $x_t$ are diagonal and unequal). What about our observations? Let's say each $y_t$ measures one $x_t$ with independent and equal error (R is diagonal and equal). Let's fit this model with MARSS.

Wrangle the raw spawner data to get the $y_t$ for a MARSS model.
```{r salmondat, message=FALSE, warning=FALSE}
library(tidyverse)
load("Data_Images/esa-salmon.rda")
esu <- unique(esa.salmon$esu_dps)
esunum <- which(esu == "Steelhead (Upper Columbia River DPS)")
esuname <- esu[esunum]

dat <- esa.salmon %>% 
  subset(esu_dps == esuname) %>% # get only this ESU
  mutate(log.spawner = log(value)) %>% # create a column called log.spawner
  dplyr::select(esapopname, spawningyear, log.spawner) %>% # get just the columns that I need
  pivot_wider(names_from = "esapopname", values_from = "log.spawner") %>% 
  column_to_rownames(var = "spawningyear") %>% # make the years rownames
  as.matrix() %>% # turn into a matrix with year down the rows
  t() # make time across the columns
# MARSS complains if I don't do this
dat[is.na(dat)] <- NA
# Clean up rownames
tmp <- rownames(dat)
tmp <- stringr::str_replace(tmp, "Steelhead [(]Upper Columbia River DPS[)]", "")
tmp <- stringr::str_replace(tmp, "River - summer", "")
tmp <- stringr::str_trim(tmp)
rownames(dat) <- tmp
```

Update your atsalibrary to version 1.6 to get the climate data or load `np_climate.RData` from the `Data_Images` directory.
```
install.packages('atsalibrary', repos = c('https://atsa-es.r-universe.dev', 'https://cloud.r-project.org'))
```

We will use spring PDO (Apr-Jun).
```{r}
library(atsalibrary)
data("np_climate")
years <- which(np_climate_seasonal$year %in% colnames(dat))
spring_pdo <- np_climate_seasonal$pdo.spring[years]
cmat <- matrix(spring_pdo, nrow=1)
# No NAs allowed. Interpolate if needed
any(is.na(cmat))
```

Check that `cmat` and `dat` have the same number of columns (years).
```{r}
dim(cmat)
dim(dat)
```

Specify a model
```{r}
# pdo
mod.list1 <- list(
  U = "unequal",
  R = "diagonal and equal",
  Q = "diagonal and unequal",
  c = cmat,
  C = "unequal"
)

# no pdo
mod.list2 <- list(
  U = "unequal",
  R = "diagonal and equal",
  Q = "diagonal and unequal")
```

```{r}
library(MARSS)
fit1 <- MARSS(dat, model=mod.list1)
fit2 <- MARSS(dat, model=mod.list2)
```

Which one fits better based on AIC? The model without PDO is better (lower AICc).

```{r}
fit1$AICc
fit2$AICc
```

## Resources

Chapter 7 MARSS models. ATSA Lab Book. <https://atsa-es.github.io/atsa-labs/chap-mss.html>

Chapter 8 MARSS models with covariates. ATSA Lab Book. <https://atsa-es.github.io/atsa-labs/chap-msscov.html>

Chapter 16 Modeling cyclic sockeye <https://atsa-es.github.io/atsa-labs/chap-cyclic-sockeye.html>
