---
title: "Team 1 - Lab 2"
author: "Nicole, Mary, Tina"
date: "April 22, 2025"
output:
  html_document:
    code-folding: yes
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
subtitle: MARSS models
---

# Data

Our lab team looked at the Oregon Coast coho salmon ESU. We subset this data from the esa-salmon.rda into one matrix 'coho' that contained time series data for 21 sub-populatins within this ESU. We also created a reference matrix called 'pop.groups' that assigned the 21 sub-populations into five assumed groups.

```{r, include=FALSE}
load(here::here("Lab-2", "Data_Images", "esa-salmon.rda"))
library(MARSS)
library(ggplot2)
library(dplyr)
library(forecast)
library(tidyverse)
library(zoo)
library(corrplot)
```

```{r, echo=FALSE}
### Data exploration
esu <- unique(esa.salmon$esu_dps)
esunum <- which(esu == esu[7])
esuname <- esu[esunum]

## ---- new-plotesu -----------------------------------------------------------
plotesu <- function(esuname, keep_set = NULL) {
  df <- esa.salmon %>%
    filter(esu_dps %in% esuname) %>%
    { if (!is.null(keep_set)) filter(., esapopname %in% keep_set) else . }

  ggplot(df, aes(x = spawningyear,
                 y = log(value),
                 colour = majorpopgroup)) +
    geom_point(size = 0.2, na.rm = TRUE) +
    facet_wrap(~esapopname) +
    theme(strip.text.x = element_text(size = 3),
          axis.text.x  = element_text(size = 5, angle = 90)) +
    ggtitle(paste0(esuname, collapse = "\n"))
}
```

```{r, echo=FALSE}
## ---- pick-a-handful --------------------------------------------------------
keep <- head(unique(esa.salmon$esapopname[esa.salmon$esu_dps == esu[7]]), 6)

## ---- plot-the-handful ------------------------------------------------------
plotesu(esu[7], keep)

## Create data frames for our ESU
esa_filtered <- esa.salmon %>%
  filter(esu_dps == esuname, esapopname != "") %>%
  mutate(log.spawner = log(value)) %>%
  mutate(log.spawner = replace(log.spawner, is.infinite(log.spawner), NA))


#Lookup data frame for major pop groups
pop.groups <- esa_filtered %>%
  select(esapopname, majorpopgroup) %>%
  distinct()%>%
  as.matrix()
pop.groups<-pop.groups[-c(1,7,15, 20, 26),]
# Time series matrix for MARSS
coho<- esa_filtered %>%
  select(esapopname, spawningyear, log.spawner) %>%
  pivot_wider(
    names_from = "esapopname",
    values_from = "log.spawner",
    values_fn = mean
  ) %>%
  column_to_rownames(var = "spawningyear") %>%
  as.matrix() %>%
  t()
coho[is.na(coho)] <- NA

#subset fall runs (don't know what "Stratum" rows are)
coho<-coho[-c(1,7,15, 20, 26),]
nrow(coho)
#clean up row names- our 26 sub-pops
tmp <- rownames(coho)
tmp <- stringr::str_replace(tmp, "Salmon, coho [(]Oregon Coast ESU[)]", "")
tmp <- stringr::str_replace(tmp, "- fall", "")
tmp <- stringr::str_trim(tmp)
rownames(coho) <- tmp
rownames(pop.groups)<-tmp
#subset colnames in case we need them later
cohocolnames<-colnames(coho)


```

# General Questions

1. Create estimates of spawner abundance for all missing years and provide estimates of the decline from the historical abundance.

2. Evaluate support for the major population groups. Are the populations in the groups more correlated than outside the groups?

3. Evaluate the evidence of cycling in the data. *We will talk about how to do this on the Tuesday after lab.*

# Methods
### Question 1
To create estimates of spawner abundance, we fit an unbiased random walk model using the MARSS function to create estimates of the "states" or x values. For this first model, we assumed R and Q are diagonal and equal, and U is equal to zero, thus no bias term will be included in this estimate. Below is a plot of sub-populations with the state estimates and observed spawner abundances.
```{r, include=FALSE}
#Part 1: Getting spawner abundance estiamtes
# `R="diagonal and equal"` and `A="scaling"
mod.list<-list(
  R="diagonal and equal",
  Q="diagonal and equal",
  A="scaling", 
  U="zero"
)
obj<-MARSS(coho, mod.list)
cohofit<-obj$states
rownames(cohofit)<-tmp
colnames(cohofit)<-cohocolnames
#cohofit is the estimated "states" from an unbiased random walk
```

```{r, echo=FALSE}
autoplot(obj, plot.type="fitted.ytT")
autoplot(obj, plot.type="residuals")
```
To estimate the population trends over time, we fit biased random walk models to each subpopulation's data. In one model, we assumed the U term is equal across all sub-populations. This model estimated U to be 0.0349, which indicates that the Oregon Coast coho ESU is generally stable and experiencing a slight increase from historical abundances. When we change this assumption of U being equal to U being unequal, the model estimated a population trend for each of the 21 sub-populations in the ESU. 

``` {r, include=FALSE}
## Part 2: Estimates of decrease from historical abundance
#Fit with biased random walk
mod.list<-list(
  R="diagonal and equal",
  Q="diagonal and equal",
  A="scaling", 
  U="equal"
)
cohofit.biased<-MARSS(coho, mod.list)
```
```{r, echo=FALSE}
U<-matrix(coef(cohofit.biased)$U)
rownames(U)<-c("U=")
colnames(U)<-"U Estimate for total ESU"
print(U)
```
Oregon Coast ESU sub-population U estimates from biased random walk with unequal U terms
```{r, echo=FALSE}
#Unequal U biased random walk
#Are sub-groups experiencing same trends over time?
mod.list<-list(
  R="diagonal and equal",
  Q="diagonal and equal",
  A="scaling", 
  U="unequal"
)
coho.unequal<-MARSS(coho, mod.list)
cohoU<-coef(coho.unequal, type="matrix")$U
colnames(cohoU)<-c("U Estimate")
print(cohoU)
mintrend<-min(cohoU)
maxtrend<-max(cohoU)
```

### Question 2
The Oregon Coast Coho ESU is divided into five “major population groups” (MPGs), based on biological or geographical similarities. In this question, we test whether these groups reflect real ecological patterns—specifically, whether populations within the same MPG show more similar (i.e., correlated) abundance trends than populations in different groups.

We evaluate this by fitting three MARSS models with different assumptions about grouping and comparing them using AICc. The model with the lowest AICc provides the best support for its grouping hypothesis. For this first analysis, we assumed unequal drift, which aligns with the varying population trends we calculated in question 1, and hidden states allowed to correlate (Q="equalvarcov"). We also fit these models with an assumption of equal drift, and zero drift. These both could also be reasonable assumptions because the population trends amongst the sub-populations in this ESU did not vary much, with a range in U values of -0.047 to 0.087. These are also close to zero, and so we could test the assumption that there is no underlying population trend. *Add matrix form for the three models, then the other two 5 group models with U=zero and U=equal*

```{r, echo=FALSE}
# Helper function: converts a grouping vector into a Z matrix
make_Z <- function(group_vec) {
  model.matrix(~ 0 + factor(group_vec))
}

# Extract the major population group (MPG) label for each population
gvec <- pop.groups[rownames(coho), "majorpopgroup"]

# Z matrix for the 5-group MPG model
Z.grp <- make_Z(gvec)

# Z matrix for the one-population model (all sites share a state)
Z.one <- matrix(1, nrow(coho), 1)

# Z matrix for the site-level model (each site has its own state)
Z.site <- diag(1, nrow(coho))

```

```{r, echo=FALSE}
# Base model components (shared across all groupings)
base.list <- list(
  B = "identity",            # random walk
  U = "unequal",                # unequal drift
  Q = "equalvarcov",         # hidden states allowed to correlate
  R = "diagonal and unequal",# each site has its own observation error
  A = "scaling"              # site-specific intercepts
)
```

```{r, include=FALSE}
# Fit model with 5 major population groups
fit.grp <- MARSS(coho, c(base.list, list(Z = Z.grp)), silent = TRUE)
# Fit model with one shared hidden state
fit.one <- MARSS(coho, c(base.list, list(Z = Z.one)), silent = TRUE)

# Fit model with one hidden state per site
fit.site <- MARSS(coho, c(base.list, list(Z = Z.site)), silent = TRUE)
```
AICc results when unequal drift is assumed:
```{r, echo=FALSE}
AICc.tab <- tibble(
  model = c("5-group MPG", "1-population", "Site-level"),
  num_states = c(ncol(Z.grp), ncol(Z.one), ncol(Z.site)),
  AICc = c(fit.grp$AICc, fit.one$AICc, fit.site$AICc)
) %>%
  arrange(AICc)

AICc.tab

```

```{r, include=FALSE}
# Change assumption to zero U
base.list.zero <- list(
  B = "identity",            # random walk
  U = "zero",                # unequal drift
  Q = "equalvarcov",         # hidden states allowed to correlate
  R = "diagonal and unequal",# each site has its own observation error
  A = "scaling"              # site-specific intercepts
)

# Fit model with 5 major population groups
fit.grp.zero <- MARSS(coho, c(base.list.zero, list(Z = Z.grp)), silent = TRUE)
# Fit model with one shared hidden state
fit.one.zero <- MARSS(coho, c(base.list.zero, list(Z = Z.one)), silent = TRUE)

# Fit model with one hidden state per site
fit.site.zero <- MARSS(coho, c(base.list.zero, list(Z = Z.site)), silent = TRUE)
```
AICc results when zero drift is assumed:
```{r, echo=FALSE}
AICc.tab.zero <- tibble(
  model = c("5-group MPG", "1-population", "Site-level"),
  num_states = c(ncol(Z.grp), ncol(Z.one), ncol(Z.site)),
  AICc = c(fit.grp.zero$AICc, fit.one.zero$AICc, fit.site.zero$AICc)
) %>%
  arrange(AICc)

AICc.tab.zero
```

```{r, include=FALSE}
# Change assumption to equal U
base.list.equal <- list(
  B = "identity",            # random walk
  U = "equal",                # unequal drift
  Q = "equalvarcov",         # hidden states allowed to correlate
  R = "diagonal and unequal",# each site has its own observation error
  A = "scaling"              # site-specific intercepts
)

# Fit model with 5 major population groups
fit.grp.equal <- MARSS(coho, c(base.list.equal, list(Z = Z.grp)), silent = TRUE)
# Fit model with one shared hidden state
fit.one.equal <- MARSS(coho, c(base.list.equal, list(Z = Z.one)), silent = TRUE)

# Fit model with one hidden state per site
fit.site.equal <- MARSS(coho, c(base.list.equal, list(Z = Z.site)), silent = TRUE)
```
AICc results when equal drift is assumed:
```{r, echo=FALSE}
AICc.tab.equal <- tibble(
  model = c("5-group MPG", "1-population", "Site-level"),
  num_states = c(ncol(Z.grp), ncol(Z.one), ncol(Z.site)),
  AICc = c(fit.grp.equal$AICc, fit.one.equal$AICc, fit.site.equal$AICc)
) %>%
  arrange(AICc)

AICc.tab.equal
```
The comparison of these models changes when the assumption of population trend is changed. When equal or zero drift is assumed, the 5-group MPG performs best. When unequal drift is assumed, the site-level model performs best. This could suggest there is an alternative grouping that could explain the variation in population trends observed in the Oregon Coast ESU. Also of note, when we compare the 5-group model AICc values to each other when the population trend assumptions are changed, the unequal drift model performs best. This supports the assumption of unequal population trends (U-term). One hypothesis is that these population trends can be explained by changing the assumptions of groupings for these ESU sub-populations. To test this hypothesis, we fit several alternative models with different grouping structures to the Oregon Coast ESU sub-populations. We maintain the assumption of hidden states that are allowed to correlate (models fit with this assumption performed better than other assumptions about Q; see AICc results below) 

```{r, include=FALSE}
unequal.drift.Q.equal <- list(
  B = "identity",            # random walk
  U = "unequal",                # no drift
  Q = "diagonal and equal",         # hidden states allowed to correlate
  R = "diagonal and unequal",# each site has its own observation error
  A = "scaling"              # site-specific intercepts
)
fit.grp.Q.equal <- MARSS(coho, c(unequal.drift.Q.equal, list(Z = Z.grp)), silent = TRUE)

unequal.drift.Q.unequal <- list(
  B = "identity",            # random walk
  U = "unequal",                # no drift
  Q = "diagonal and unequal",         # hidden states allowed to correlate
  R = "diagonal and unequal",# each site has its own observation error
  A = "scaling"              # site-specific intercepts
)
fit.grp.Q.unequal <- MARSS(coho, c(unequal.drift.Q.unequal, list(Z = Z.grp)), silent = TRUE)
```

```{r, echo=FALSE}
Q.tab<-tibble(
  model = c("Q=equalvarcov", "Q=diagonal and equal", "Q= diagonal and unequal"),
  num_states = c(ncol(Z.grp), ncol(Z.grp), ncol(Z.grp)),
  AICc = c(fit.grp$AICc, fit.grp.Q.equal$AICc, fit.grp.Q.unequal$AICc)
) %>%
  arrange(AICc)
Q.tab
```

First we fit alternative groupings that divided the ESU into 6 groups. One model separated Beaver Creek and Salmon River into a sixth group, since they geographically distant from the other sub-populations (more inland). The other 6-group model separated Floras Creek and Sixes River into a sixth group. These were the two populations experiencing declining population trends, and are further south than most of the other sites, which could mean they form a unique group. We also fit a model with 7 groupings, with groups 6 and 7 being the sub-groups described previously.
```{r, include=FALSE}
mod.list<-list(
  B = "identity",            # random walk
  U = "unequal",                # no drift
  Q = "equalvarcov",         # hidden states allowed to correlate
  R = "diagonal and unequal",# each site has its own observation error
  A = "scaling"              # site-specific intercepts
)
### 6 groups separate beaver creek and salmon river 
pop.groups.6.1<-pop.groups
pop.groups.6.1[c("Beaver Creek", "Salmon River"),2]<-"Group 6"
gvec <- pop.groups.6.1[rownames(coho), "majorpopgroup"]
#Z matrix for the first 6-group model
Z.grp.6.1 <- make_Z(gvec)
#Fit model
fit.grp.6.1 <- MARSS(coho, c(mod.list, list(Z = Z.grp.6.1)), silent = TRUE)

# 6 groups separate floras creek and sixes river
pop.groups.6.2<-pop.groups
pop.groups.6.2[c("Floras Creek/New River", "Sixes River"),2]<-"Group 6"
gvec <- pop.groups.6.2[rownames(coho), "majorpopgroup"]
#Z matrix for the first 6-group model
Z.grp.6.2 <- make_Z(gvec)
#Fit model
fit.grp.6.2 <- MARSS(coho, c(mod.list, list(Z = Z.grp.6.2)), silent = TRUE)

# 7 group model
pop.groups.6.2[c("Beaver Creek", "Salmon River"),2]<-"Group 7"
gvec <- pop.groups.6.2[rownames(coho), "majorpopgroup"]
#Z matrix for the first 6-group model
Z.grp.7 <- make_Z(gvec)
#Fit model
fit.grp.7 <- MARSS(coho, c(mod.list, list(Z = Z.grp.6.2)), silent = TRUE)
```
AICc results for splitting the groups into 6 and 7 groups:
```{r, echo=FALSE}
groups6.7<-tibble(
  model = c("Beaver and Salmon group", "Flores and Sixes Group", "5-Group MPG", "7 Groups"),
  num_states = c(ncol(Z.grp.6.1), ncol(Z.grp.6.2), ncol(Z.grp),ncol(Z.grp.7)),
  AICc = c(fit.grp.6.1$AICc, fit.grp.6.2$AICc, fit.grp$AICc, fit.grp.7$AICc)
) %>%
  arrange(AICc)

groups6.7
```
None of the 6- and 7-groups models were supported when comparing AICc values to the original 5-Group MPG model. Thus, we decided not to further divide the groupings and tried fitting models with the same number of groups but re-arranged.

The first alternative 5 group model divided the sub-populations up equally along a north-south gradient, and the second model set Beaver Creek and Salmon River, and Floras Creek and Sixes River into their own groups, and combined the coastal groups into one.
```{r, include=FALSE}
# 5 groups- divide evenly north to south so each group has 4 or 5 pops in it
#does not treat lakes and umpqua as a unique group
#"A" is furthest north, "E" is furthest south
pop.groups.5<-pop.groups
geolist<-c("C", "C", "D", "C", "B", "A", "B", "C", "B", "E", "E", "E", "E", "A", "A", "B",
           "A", "D", "D", "D", "D")
pop.groups.5[, "majorpopgroup"]<-geolist

gvec <- pop.groups.5[rownames(coho), "majorpopgroup"]
#Z matrix for the first 6-group model
Z.grp.5.2 <- make_Z(gvec)
#Fit model
fit.grp.5.2 <- MARSS(coho, c(mod.list, list(Z = Z.grp.5.2)), silent = TRUE)

# 5 groups
#Beaver and salmon group, sixes and floras group, lakes, coast, umpqua
pop.groups.6.2 <- as.data.frame(pop.groups.6.2)
pop.groups.6.2[pop.groups.6.2$majorpopgroup %in% c("Mid-Coast", "Mid-South Coast", "North Coast"), "majorpopgroup"] <- "Coast"
gvec <- pop.groups.6.2[rownames(coho), "majorpopgroup"]
Z.grp.5.3 <- make_Z(gvec)
#Fit model
fit.grp.5.3 <- MARSS(coho, c(mod.list, list(Z = Z.grp.5.3)), silent = TRUE)
```

None of these performed better than the original 5-group model.
```{r, echo=FALSE}
groups5<-tibble(
  model = c("North-South groups", "Combined Coastal Group", "5-Group MPG"),
  num_states = c(ncol(Z.grp), ncol(Z.grp), ncol(Z.grp)),
  AICc = c(fit.grp.5.2$AICc, fit.grp.5.3$AICc, fit.grp$AICc)
) %>%
  arrange(AICc)

groups5
```
Next we tried fitting models with fewer than 5 groups: one there is no lake group (merged with mid-Coast), two where there is no umpqua group (one merged with mid-south coast, in the other one it is merged with the mid-coast group), one in which there three groups- coastal, lake, and upmqua, and one in which there are only three coastal groups.

```{r, include=FALSE}
#4 groups- no lake, all lakes in mid-Coast
pop.groups.4<-as.data.frame(pop.groups)
pop.groups.4[pop.groups.4$majorpopgroup %in% c("Lakes"), "majorpopgroup"]<-"Mid-Coast"

gvec <- pop.groups.4[rownames(coho), "majorpopgroup"]
#Z matrix for the first 6-group model
Z.grp.4.1 <- make_Z(gvec)
#Fit model
fit.grp.4.1 <- MARSS(coho, c(mod.list, list(Z = Z.grp.4.1)), silent = TRUE)

# 4 groups- no umpqua- umpqua is in mid-south coast
pop.groups.4.2<-as.data.frame(pop.groups)
pop.groups.4.2[pop.groups.4.2$"majorpopgroup"%in%c("Umpqua"), "majorpopgroup"]<-"Mid-South Coast"

gvec <- pop.groups.4.2[rownames(coho), "majorpopgroup"]
#Z matrix for the first 6-group model
Z.grp.4.2 <- make_Z(gvec)
#Fit model
fit.grp.4.2 <- MARSS(coho, c(mod.list, list(Z = Z.grp.4.2)), silent = TRUE)

# 4 groups- no umpua- umpqua is in mid-coast
pop.groups.4.3<-as.data.frame(pop.groups)
pop.groups.4.3[pop.groups.4.3$"majorpopgroup"%in%c("Umpqua"), "majorpopgroup"]<-"Mid-Coast"

gvec <- pop.groups.4.3[rownames(coho), "majorpopgroup"]
#Z matrix for the first 6-group model
Z.grp.4.3 <- make_Z(gvec)
#Fit model
fit.grp.4.3 <- MARSS(coho, c(mod.list, list(Z = Z.grp.4.3)), silent = TRUE)
fit.grp.4.3$AICc
fit.grp.4.2$AICc

# 3 groups- lake, coast, umpqua
pop.groups.3.1<-as.data.frame(pop.groups)
pop.groups.3.1[pop.groups.3.1$"majorpopgroup"%in%c("Mid-Coast", "Mid-South Coast", "North Coast"), "majorpopgroup"]<-"Coast"

gvec <- pop.groups.3.1[rownames(coho), "majorpopgroup"]
#Z matrix for the first 6-group model
Z.grp.3.1 <- make_Z(gvec)
#Fit model
fit.grp.3.1 <- MARSS(coho, c(mod.list, list(Z = Z.grp.3.1)), silent = TRUE)

# 3 groups all coastal
pop.groups.4.2[pop.groups.4.2$majorpopgroup %in% c("Lakes"), "majorpopgroup"]<-"Mid-Coast"
gvec <- pop.groups.4.2[rownames(coho), "majorpopgroup"]
#Z matrix for the first 6-group model
Z.grp.3.2 <- make_Z(gvec)
#Fit model
fit.grp.3.2 <- MARSS(coho, c(mod.list, list(Z = Z.grp.3.2)), silent = TRUE)
```
Again none of these performed better than the original 5-group model. We can see some interesting results in comparing these models is the Umpqua populations appear to be more similar to the Mid-South Coast than Mid-Coast populations, and the lake populations do appear to be distinct (much higher AICs values when they are grouped in with the coastal groups). Because no other patterns for more effective groupings were found in our hypothesis testing, we have determined the best grouping assumption is the one laid out in the original data set.
```{r, echo=FALSE}
groups4.3<-tibble(
  model = c("No Lakes", "Umpqua in Mid-South Coast", "Umpqua in Mid-Coast",
           "Lake, Coast, Umpqua", "3 Coastal", "5-Group MPG"),
  num_states = c(ncol(Z.grp.4.1), ncol(Z.grp.4.2), ncol(Z.grp.4.3), ncol(Z.grp.3.1),
                 ncol(Z.grp.3.2), ncol(Z.grp)),
  AICc = c(fit.grp.4.1$AICc, fit.grp.4.2$AICc, fit.grp.4.3$AICc,
           fit.grp.3.1$AICc, fit.grp.3.2$AICc,fit.grp$AICc)
) %>%
  arrange(AICc)
groups4.3
```
### Question 3
To further explore variation in population trends over time for the Oregon Coast ESU, we evaluate the evidence of cycling in Oregon Coast coho spawner abundance. We compared three structural approaches using MARSS models: a polynomial trend, a Fourier series (cyclic component), and a null model (the best performed from above analysis). We hypothesized that coho populations might exhibit 4-year cycles due to biological factors.

```{r, echo=FALSE}
# Create time covariates
years <- as.numeric(colnames(coho))
t <- years - min(years) + 1  # Time index starting at 1

# Polynomial trend model (linear + quadratic)
time <- t
time_sq <- t^2
cov_poly <- rbind(time, time_sq)

mod.poly <- list(
  B = "identity", U = "zero", C = "unconstrained",
  c = cov_poly, Q = "diagonal and equal",
  Z = "identity", R = "diagonal and equal", A = "scaling"
)

# Fourier model (4-year cycle)
sin4 <- sin(2 * pi * t / 4)
cos4 <- cos(2 * pi * t / 4)
cov_fourier <- rbind(sin4, cos4)

mod.fourier <- list(
  B = "identity", U = "zero", C = "unconstrained",
  c = cov_fourier, Q = "diagonal and equal",
  Z = "identity", R = "diagonal and equal", A = "scaling"
)

# Fit all models
fit.null <- MARSS(coho, c(base.list, list(Z = Z.site)), silent = TRUE)
fit.poly <- MARSS(coho, model = mod.poly, silent = TRUE)
fit.fourier <- MARSS(coho, model = mod.fourier, silent = TRUE)

# Compare AICc
aicc.tab <- tibble(
  Model = c("Null", "Polynomial", "Fourier (4-year)"),
  AICc = c(fit.null$AICc, fit.poly$AICc, fit.fourier$AICc),
  Delta_AICc = c(0, fit.poly$AICc - fit.null$AICc, 
                fit.fourier$AICc - fit.null$AICc)
) %>% arrange(AICc)

print(aicc.tab)
```

It seems that still the null model performs the best, but we want to systematically test different cycle periods that might be biologically plausible for coho salmon (2-10 years).

```{r, echo=FALSE}
test_cycles <- function(data, periods = 2:10) {
  years <- as.numeric(colnames(data))
  t <- years - min(years) + 1
  
  # Null model for comparison
  fit.null <- MARSS(coho, c(base.list, list(Z = Z.site)), silent = TRUE)
  
  # Test each period
  results <- map_df(periods, function(p) {
    # Create Fourier terms
    sin_term <- sin(2 * pi * t / p)
    cos_term <- cos(2 * pi * t / p)
    covariates <- rbind(sin_term, cos_term)
    
    # Fourier model
    mod.fourier <- list(
      B = "identity", U = "zero", C = "unconstrained",
      c = covariates, Q = "diagonal and equal",
      Z = "identity", R = "diagonal and equal", A = "scaling"
    )
    
    fit <- MARSS(data, model = mod.fourier, silent = TRUE)
    
    tibble(
      Period = p,
      AICc = fit$AICc,
      Delta_AICc = fit$AICc - fit.null$AICc,
      Converged = fit$convergence == 0
    )
  })
  
  arrange(results, AICc)
}

cycle_results <- test_cycles(coho, periods = 2:10)
```


# Results
### Estimating Population Trends

Both of the models we fit to estimate population trends for the Oregon Coast ESU populations found they are relatively stable, and have experienced slight increases since 1990. Below is a histogram of the distribution of estimated population trends for each sub-population in this ESU. There is some variation in the population trends among the sub-populations in this ESU, but the Floras Creek/New River and Sixes River fall spawners are the only groups that are experiencing a declining trend.
```{r, echo=FALSE}
hist(cohoU, main="Coho Population Trends", xlab="U coefficient")
```

### Fitting state-space models
To evaluate whether the five major population groups (MPGs) are meaningful in explaining Coho population dynamics, we fit three MARSS models with different grouping assumptions: one with a hidden state for each MPG, one treating all populations as a single group, and one giving each site its own independent state. We also fit these three models with different assumptions on population trend- zero drift, equal drift, or unequal drift. When zero or equal drift is assumed, the model with the lowest AICc was the **5-group MPG model (AICc = 1457)**, with a substantial improvement over both the **site-level model (AICc = 1688)** and the **1-population model (AICc = 1706)**. These results support the biological relevance of the MPG structure: populations within a group show similar dynamics, while those in different groups behave differently.This provides ecological justification for modeling these populations as grouped rather than treating all sites as independent or pooled together.

### Visualization of the correlation results:
To better understand the relationships between major population groups, we visualized the correlation matrix (Q) from the MARSS model that assigned one hidden state to each group. The Q matrix reflects how similarly the groups behave over time.

In the heatmap below, stronger red indicates higher positive correlation (groups move together), while blue indicates negative correlation (groups move in opposite directions), and white indicates weak or no correlation.

This visual allows us to assess whether some MPGs are tightly linked or behaving independently, even within the group structure.
*Something happened here, I copied your code but the output does not look the same. I then tried copying the code from the Lab2-MARSS.rmd and am getting the same results, did I do something wrong while fitting the model?*

```{r}
Q <- coef(fit.site, type="matrix")$Q
corrmat <- diag(1/sqrt(diag(Q))) %*% Q %*% diag(1/sqrt(diag(Q)))
corrplot(corrmat, main="Site model")

Q <- coef(fit.grp, type="matrix")$Q
corrmat <- diag(1/sqrt(diag(Q))) %*% Q %*% diag(1/sqrt(diag(Q)))
corrplot(corrmat, main="5 Group model")

```

However, when unequal drift is assumed, the model that treats each sub-population as its own group performs the best. The 5-group MPG model that assumes unequal drift also performs better than the same model fit with equal or zero drift, which further supports the idea that there might be differences among the populations in this ESU resulting in different population trends. 

To further investigate this, we changed our assumptions of the groupings and fit alternative models to the Oregon Coast ESU sub-populations. However, none of the combinations of 7, 6, 5 (different from the original groups), 4, or 3 groupings that we fit to the data performed better. The original 5 groupings considered in the Oregon Coast ESU did seem to be the most appropriate for the data. While larger groups or considering each population as its own group might produce better model results, the goal of creating population groups to explain the data is achieved with the original five. The model that treated each sub-population did not perform extremely better relative to the 5 group model, and so there is merit in that grouping assumption. 

### Question 3
All tested cycle periods (2–10 years) yielded higher AICc values than the null model (ΔAICc > 0), indicating no statistical evidence for cyclic dynamics in Oregon Coast coho spawner abundance (see AICc comparisons below). 
```{r, echo=FASLSE}
print(aicc.tab)
print(cycle_results)
```

# Discussion
In this lab, we explored population structure, trends, and cyclic dynamics of the Oregon Coast coho salmon ESU. This ESU consists of 21 sub-populations of coho, for which there is time series data of adult returns from 1990 to 2023. We fit MARSS models to the data to investigate the underlying population dynamics of this ESU and found the following:

There is variation in the population trends experienced by the 21 sub-populations; on average, the sub-populations in this ESU have experienced slightly positive trends from their historical abundances in the early 1990's. Two populations- the Floras Creek and Sixed River- have experienced slight declines. 

We then tested several hypotheses about the ESU population groups to explore the variation amongst these twenty-one sub-populations. We did this by changing the assumptions of population groupings in the Z-matrix of our model equations based on possible underlying population structures (all sub-populations are one homogenous group, or each is their own individual group, population groupings based on latitude, etc.). Comparing the models we fit to the data strongly supported the original five sub-groups included in the original data set. When we change the model assumption to allow unequal drift (population trends) the best-performing model was the one that considered each sub-population as its own group. When the drift was assumed to be equal or zero, the original 5-group model performed best. This is a reasonable assumption since the population trends observed for each sub-population are centered around zero and very small. And so this grouping assumption is reasonable, and is supported by the data. 

An alternative hypothesis to population structure driving variation within this ESU is the hypothesis of cyclic dynamics within the time series data. To investigate this, we fit several models with varying cycle periods to the data.

While cyclic dynamics are theoretically plausible for salmonids, this analysis found no empirical support for cyclic trends in Oregon Coast coho spawner abundance at any tested period (2–10 years). The null model’s superiority suggests that population trends are likely driven by non-cyclic factors, and management should prioritize addressing persistent stressors rather than anticipating cyclic rebounds.

The lack of cyclicity suggests that no dominant environmental driver synchronizes populations at these timescales. Populations may be buffered against cyclic pressures by life-history diversity and local adaptation to heterogeneous habitats. 

# Description of each team member's contributions
Nicole addressed question 1 and fit alternative grouping-assumption models in question 2. Mary provided the code for all data visualization and for fitting MARSS models in question 2. Tina addressed question 3. All group members contributed to the write-up and discussion sections of this report.
