---
title: Team X - Lab 2
subtitle: MARSS models
author: "Nicole, Mary, Tina"
date: April 22, 2025
output: 
  html_document:
    code-folding: true
    toc: true
    toc_float: true
---

# Data

Our lab team looked at the Oregon Coast coho salmon ESU. We subset this data from the esa-salmon.rda into one matrix 'coho' that contained time series data for 26 sub-populatins within this ESU. We also created a reference matrix called 'pop.groups' that assigned the 26 sub-populations into five assumed groups.

Load the data.
```{r}
load(here::here("Lab-2", "Data_Images", "esa-salmon.rda"))
```

Wrangle the data.

```{r, echo=FALSE}
### Data exploration
esu <- unique(esa.salmon$esu_dps)
esunum <- which(esu == esu[7])
esuname <- esu[esunum]

#Data visualization
plotesu <- function(esuname){
  df <- esa.salmon %>% subset(esu_dps %in% esuname)
  ggplot(df, aes(x=spawningyear, y=log(value), color=majorpopgroup)) + 
    geom_point(size=0.2, na.rm = TRUE) + 
    theme(strip.text.x = element_text(size = 3)) +
    theme(axis.text.x = element_text(size = 5, angle = 90)) +
    facet_wrap(~esapopname) +
    ggtitle(paste0(esuname, collapse="\n"))
}

#Plot of our ESU sub-pops
plotesu(esu[7])

## Create data frames for our ESU
esa_filtered <- esa.salmon %>%
  filter(esu_dps == esuname, esapopname != "") %>%
  mutate(log.spawner = log(value)) %>%
  mutate(log.spawner = replace(log.spawner, is.infinite(log.spawner), NA))


#Lookup data frame for major pop groups
pop.groups <- esa_filtered %>%
  select(esapopname, majorpopgroup) %>%
  distinct()%>%
  as.matrix()

# Time series matrix for MARSS
coho<- esa_filtered %>%
  select(esapopname, spawningyear, log.spawner) %>%
  pivot_wider(
    names_from = "esapopname",
    values_from = "log.spawner",
    values_fn = mean
  ) %>%
  column_to_rownames(var = "spawningyear") %>%
  as.matrix() %>%
  t()
coho[is.na(coho)] <- NA
head(coho)

#clean up row names- our 26 sub-pops
tmp <- rownames(coho)
tmp <- stringr::str_replace(tmp, "Steelhead [(]Upper Columbia River DPS[)]", "")
tmp <- stringr::str_replace(tmp, "River - summer", "")
tmp <- stringr::str_trim(tmp)
rownames(coho) <- tmp
rownames(pop.groups)<-tmp
#subset colnames in case we need them later
cohocolnames<-colnames(coho)
```

# General Questions

1. Create estimates of spawner abundance for all missing years and provide estimates of the decline from the historical abundance.

2. Evaluate support for the major population groups. Are the populations in the groups more correlated than outside the groups?

3. Evaluate the evidence of cycling in the data. *We will talk about how to do this on the Tuesday after lab.*


# Methods

Address the following in your methods

* Describe your assumptions about the x and how the data time series are related to x.

   - How are the x and y (data) related? 1 x for 1 y or will you assume 1 x for all y or 1 x for each major population group? How will you choose? 
   - What will you assume about the U for the x's?
   - What will you assume about the Q matrix?
   
* Write out your assumptions as different models **in matrix form**, fit each and then compare these with AIC or AICc.

* Do your estimates differ depending on the assumptions you make about the structure of the data, i.e. you assumptions about the x's, Q, and U.

### Tips

**Simplify**

If your ESU has many populations, start with a smaller set of 4-7 populations.

**Assumptions**

You can assume that `R="diagonal and equal"` and `A="scaling"`. Assume that "historical" means the earliest years available for your group.

**States**

Your abundance estimate is the "x" or "state" estimates. You can get this from
```
fit$states
```
or 
```
tsSmooth(fit)
```
where `fit` is from `fit <- MARSS()`

**plotting**

Estimate of the mean of the spawner counts based on your x model.
```
autoplot(fit, plot.type="fitted.ytT")
```

**diagnostics**
```
autoplot(fit, plot.type="residuals")
```

# Results

# Discussion

# Description of each team member's contributions

Example: "All team members helped decide on the goal and ran the analyses for the individual regions. Team members 2 & 3 wrote most of the code for the analysis of the regions. Team member 4 researched approaches for measuring accuracy of forecasts in [Hyndman & Athanasopoulos[OTexts.com/fpp2] and team member 2 added code for that to the methods. Team member 4 also researched tests for stationarity and worked with team member 2 to code that up. Team member 1 worked on the plotting section of the report using and adapting code that team member 3 wrote. All team members helped edit the report and wrote the discussion together."

