---
title: "Team 1 - Lab 2"
subtitle: "MARSS models"
author: "Mary Orrand, Nicole, Tina"
date: "April 22, 2025"
output: 
  html_document:
    code-folding: true
    toc: true
    toc_float: true
    theme: flatly
    number_sections: true
---


# Data

In this lab, we use multivariate autoregressive state-space (MARSS) models to estimate spawner abundance over time for the Oregon Coast Coho ESU. The goals of the lab are to estimate missing values, evaluate trends in abundance, and explore whether grouping populations improves model fit. Team 1 will be working with the Eastern Or coho salmon data, "1. Salmon, coho (Oregon Coast ESU) `esu[7]`".

Question 1: Create estimates of spawner abundance for all missing years and provide estimates of the decline from the historical abundance.

Load the data.
```{r}
load(here::here("Lab-2", "Data_Images", "esa-salmon.rda"))
```

Wrangle the data.

We begin by sub-setting the Oregon Coast Coho ESU from the full salmon data set.

```{r}
## Packages
library(MARSS)
library(ggplot2)
library(dplyr)
library(forecast)
library(tidyverse)
library(zoo)
library(corrplot)
```

```{r}
### Data exploration
esu <- unique(esa.salmon$esu_dps)
esunum <- which(esu == esu[7])
esuname <- esu[esunum]
```

```{r}
## ---- new-plotesu -----------------------------------------------------------
plotesu <- function(esuname, keep_set = NULL) {
  df <- esa.salmon %>%
    filter(esu_dps %in% esuname) %>%
    { if (!is.null(keep_set)) filter(., esapopname %in% keep_set) else . }

  ggplot(df, aes(x = spawningyear,
                 y = log(value),
                 colour = majorpopgroup)) +
    geom_point(size = 0.2, na.rm = TRUE) +
    facet_wrap(~esapopname) +
    theme(strip.text.x = element_text(size = 3),
          axis.text.x  = element_text(size = 5, angle = 90)) +
    ggtitle(paste0(esuname, collapse = "\n"))
}
```




```{r}
## ---- pick-a-handful --------------------------------------------------------
keep <- head(unique(esa.salmon$esapopname[esa.salmon$esu_dps == esu[7]]), 6)

## ---- plot-the-handful ------------------------------------------------------
plotesu(esu[7], keep)


## Create data frames for our ESU
esa_filtered <- esa.salmon %>%
  filter(esu_dps == esuname, esapopname != "") %>%
  mutate(log.spawner = log(value)) %>%
  mutate(log.spawner = replace(log.spawner, is.infinite(log.spawner), NA))
```

```{r}
#Lookup data frame for major pop groups
pop.groups <- esa_filtered %>%
  select(esapopname, majorpopgroup) %>%
  distinct()%>%
  as.matrix()
```

```{r}
# Time series matrix for MARSS
coho<- esa_filtered %>%
  select(esapopname, spawningyear, log.spawner) %>%
  pivot_wider(
    names_from = "esapopname",
    values_from = "log.spawner",
    values_fn = mean
  ) %>%
  column_to_rownames(var = "spawningyear") %>%
  as.matrix() %>%
  t()
coho[is.na(coho)] <- NA
head(coho)
```

```{r}
#clean up row names- our 26 sub-pops
tmp <- rownames(coho)
tmp <- stringr::str_replace(tmp, "Steelhead [(]Upper Columbia River DPS[)]", "")
tmp <- stringr::str_replace(tmp, "River - summer", "")
tmp <- stringr::str_trim(tmp)
rownames(coho) <- tmp
rownames(pop.groups)<-tmp
#subset colnames in case we need them later
cohocolnames<-colnames(coho)
```


```{r}
#### Question 1####
#Part 1: Filling in NA values
#use marss, save to object and get y-hat
# `R="diagonal and equal"` and `A="scaling"
mod.list<-list(
  R="diagonal and equal",
  A="scaling"
)
obj<-MARSS(coho, mod.list)
cohofit<-obj$states.se
rownames(cohofit)<-tmp
colnames(cohofit)<-cohocolnames
head(cohofit)

```

```{r}
## Part 2: Estimates of decrease from historical abundance
#random walk model, Z is a diagonal matrix
cohoU<-coef(obj, type="matrix")$U
hist(cohoU, main="Coho Population Trends", xlab="U coefficient")
colnames(cohoU)<-c("U")
```

```{r}
#Which populations are declining?
decliningpop<-subset(cohoU, cohoU<0)
#Salmon, coho (Oregon Coast ESU) Floras Creek/New River - fall
#And Sixes River- fall are the only groups that are declining
```

```{r}
meantrend<-mean(cohoU)
print(meantrend)
#So on average, this ESU group is experiencing slight increases from historical
#abundances (average U=0.0359 when fitting biased random walk models to ts data
```

Question 2: Evaluate support for the major population groups. Are the populations in the groups more correlated than outside the groups?

The Oregon Coast Coho ESU is divided into five “major population groups” (MPGs), based on biological or geographical similarities. In this question, we test whether these groups reflect real ecological patterns—specifically, whether populations within the same MPG show more similar (i.e., correlated) abundance trends than populations in different groups.

We evaluate this by fitting three MARSS models with different assumptions about grouping and comparing them using AICc. The model with the lowest AICc provides the best support for its grouping hypothesis.

```{r}
# Helper function: converts a grouping vector into a Z matrix
make_Z <- function(group_vec) {
  model.matrix(~ 0 + factor(group_vec))
}

# Extract the major population group (MPG) label for each population
gvec <- pop.groups[rownames(coho), "majorpopgroup"]

# Z matrix for the 5-group MPG model
Z.grp <- make_Z(gvec)

# Z matrix for the one-population model (all sites share a state)
Z.one <- matrix(1, nrow(coho), 1)

# Z matrix for the site-level model (each site has its own state)
Z.site <- diag(1, nrow(coho))

```

```{r}
# Base model components (shared across all groupings)
base.list <- list(
  B = "identity",            # random walk
  U = "zero",                # no drift
  Q = "equalvarcov",         # hidden states allowed to correlate
  R = "diagonal and unequal",# each site has its own observation error
  A = "scaling"              # site-specific intercepts
)

```

```{r}
# Fit model with 5 major population groups
fit.grp <- MARSS(coho, c(base.list, list(Z = Z.grp)), silent = TRUE)

# Fit model with one shared hidden state
fit.one <- MARSS(coho, c(base.list, list(Z = Z.one)), silent = TRUE)

# Fit model with one hidden state per site
fit.site <- MARSS(coho, c(base.list, list(Z = Z.site)), silent = TRUE)

```

```{r}
library(dplyr)

AICc.tab <- tibble(
  model = c("5-group MPG", "1-population", "Site-level"),
  num_states = c(ncol(Z.grp), ncol(Z.one), ncol(Z.site)),
  AICc = c(fit.grp$AICc, fit.one$AICc, fit.site$AICc)
) %>%
  arrange(AICc)

AICc.tab

```

### Interpretation of Results (Question 2)

To evaluate whether the five major population groups (MPGs) are meaningful in explaining Coho population dynamics, we fit three MARSS models with different grouping assumptions: one with a hidden state for each MPG, one treating all populations as a single group, and one giving each site its own independent state.

The model with the lowest AICc was the **5-group MPG model (AICc = 1457)**, with a substantial improvement over both the **site-level model (AICc = 1688)** and the **1-population model (AICc = 1706)**. These results strongly support the biological relevance of the MPG structure: populations within a group show similar dynamics, while those in different groups behave differently.

This provides ecological justification for modeling these populations as grouped rather than treating all sites as independent or pooled together.

Visualization of the correlation results:
To better understand the relationships between major population groups, we visualized the correlation matrix (Q) from the MARSS model that assigned one hidden state to each group. The Q matrix reflects how similarly the groups behave over time.

In the heatmap below, stronger red indicates higher positive correlation (groups move together), while blue indicates negative correlation (groups move in opposite directions), and white indicates weak or no correlation.

This visual allows us to assess whether some MPGs are tightly linked or behaving independently, even within the group structure.

```{r}
# Extract the process error covariance matrix from the 5-group model
Q_matrix <- coef(fit.grp, type = "matrix")$Q

# Convert it to a correlation matrix
Q_cor <- cov2cor(Q_matrix)

# Print it to inspect the values
Q_cor

```

```{r}
# Visualize the correlation matrix
library(corrplot)

corrplot(Q_cor,
         method = "color",      # fill cells with colors
         type = "upper",        # show only upper triangle
         tl.col = "black",      # label color
         tl.cex = 0.8,          # label size
         col = colorRampPalette(c("blue", "white", "red"))(200),
         title = "Correlation Between Major Population Groups",
         mar = c(0,0,2,0))

```
The correlation heatmap from the MARSS Q matrix shows strong positive relationships among all five major population groups. The diagonal (r = 1.0) reflects each group correlated with itself, while the off-diagonal values are all moderately to strongly positive (no blue or white values).

This indicates that while each group has its own underlying trend, all groups are generally moving in the same direction over time. This may reflect regional-scale environmental or management drivers that influence the entire ESU, such as climate patterns or hatchery practices.

Overall, this visual reinforces the AICc results: the MPG grouping is supported, but also shows that no group is acting entirely independently.


3. Evaluate the evidence of cycling in the data. *We will talk about how to do this on the Tuesday after lab.*


# Methods

Address the following in your methods

* Describe your assumptions about the x and how the data time series are related to x.

   - How are the x and y (data) related? 1 x for 1 y or will you assume 1 x for all y or 1 x for each major population group? How will you choose? 
   - What will you assume about the U for the x's?
   - What will you assume about the Q matrix?
   
* Write out your assumptions as different models **in matrix form**, fit each and then compare these with AIC or AICc.

* Do your estimates differ depending on the assumptions you make about the structure of the data, i.e. you assumptions about the x's, Q, and U.

### Tips

**Simplify**

If your ESU has many populations, start with a smaller set of 4-7 populations.

**Assumptions**

You can assume that `R="diagonal and equal"` and `A="scaling"`. Assume that "historical" means the earliest years available for your group.

**States**

Your abundance estimate is the "x" or "state" estimates. You can get this from
```
fit$states
```
or 
```
tsSmooth(fit)
```
where `fit` is from `fit <- MARSS()`

**plotting**

Estimate of the mean of the spawner counts based on your x model.
```
autoplot(fit, plot.type="fitted.ytT")
```

**diagnostics**
```
autoplot(fit, plot.type="residuals")
```

# Results

# Discussion

# Description of each team member's contributions

Example: "All team members helped decide on the goal and ran the analyses for the individual regions. Team members 2 & 3 wrote most of the code for the analysis of the regions. Team member 4 researched approaches for measuring accuracy of forecasts in [Hyndman & Athanasopoulos[OTexts.com/fpp2] and team member 2 added code for that to the methods. Team member 4 also researched tests for stationarity and worked with team member 2 to code that up. Team member 1 worked on the plotting section of the report using and adapting code that team member 3 wrote. All team members helped edit the report and wrote the discussion together."

