---
title: Team 1 - Lab 4
subtitle: Hidden Markov Models
author: "Name 1, Name 2, Name 3"
date: May XX, 2025
output: 
  html_document:
    code-folding: true
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
options(dplyr.summarise.inform = FALSE)
```

###############################################################
############ Pt 1: PDO Data #############
###############################################################
###############################################################

```{r}
library(dplyr)
#install.packages("rsoi")
pdo <- rsoi::download_pdo()

### Adjust the data 
pdo$Year[which(pdo$Month%in%c("Oct","Nov","Dec"))] <- pdo$Year[which(pdo$Month%in%c("Oct","Nov","Dec"))] + 1
pdo <- dplyr::group_by(pdo, Year) %>%
  dplyr::summarize(winter_pdo = mean(PDO[which(Month %in% c("Oct","Nov","Dec","Jan","Feb"))])) %>% 
  dplyr::select(winter_pdo, Year)
# The first year will be missing Oct-Dec
pdo <- pdo[-1,]

```

# Plot winter PDO vs year
```{r}
library(ggplot2)

plot <- ggplot(pdo, aes(x = Year, y = winter_pdo)) +
  geom_line() +
  labs(title = "Winter PDO vs Year", x = "Year", y = "Winter PDO") +
  theme_minimal()

plot

```

### Fit a two state HMM to the annual indices of PDO 
```{r}

library(hmmTMB)

pdo <- data.frame(
  ID = 1,
  time = pdo$Year,  # Use actual years as time
  winter_pdo = pdo$winter_pdo
)

# Set up hidden states (2 states for different PDO regimes)
hidden <- MarkovChain$new(data = pdo, n_states = 2)

# Use normal distribution for continuous PDO values
dists <- list(winter_pdo = "norm")

# Initial parameters for normal distribution (mean and sd for each state)
par0 <- list(winter_pdo = list(mean = c(-0.5, 0.5),    # Different means for each state
                               sd = c(0.8, 0.8)))      # Standard deviations

# Create observation model
obs_model <- Observation$new(data = pdo,
                             dists = dists,
                             n_states = 2,
                             par = par0)

# Fit the HMM
hmm <- HMM$new(obs = obs_model, hid = hidden)
hmm$fit(silent = TRUE)
```


#Examine the output 
```{r}
hmm
hmm$par

pdo$est_state <- factor(paste0("State", hmm$viterbi()))

ggplot(pdo, aes(time, winter_pdo, col = est_state)) + 
  geom_line(aes(group = 1), color = "black") + 
  geom_point(size = 2) + # This line will connect all the points
  ylab("Rained?") + 
  theme_bw()


```
Try fitting the model 10-20 times. Does the likelihood seem reasonably stable? (Note logLik() gets you the log-likelihood from model fits in R).
```{r}
library(hmmTMB)
library(ggplot2)

# Placeholder for storing models and log-likelihoods
models <- vector("list", 20)
log_liks <- numeric(20)

# Run 20 iterations
for (i in 1:20) {
  
  # Optionally randomize initial parameters slightly
  par0 <- list(winter_pdo = list(
    mean = rnorm(2, mean = c(-0.5, 0.5), sd = 0.2),  # jittered means
    sd = runif(2, min = 0.5, max = 1.2)              # jittered sds
  ))
  
  # Set up models
  hidden <- MarkovChain$new(data = pdo, n_states = 2)
  obs_model <- Observation$new(data = pdo, dists = list(winter_pdo = "norm"),
                               n_states = 2, par = par0)
  
  # Fit the model
  hmm <- HMM$new(obs = obs_model, hid = hidden)
  hmm$fit(silent = TRUE)
  
  # Store the model and log-likelihood
  models[[i]] <- hmm
  log_liks[i] <- logLik(hmm)
}

# Create a data frame for plotting
loglik_df <- data.frame(Iteration = 1:20, LogLikelihood = log_liks)

# Plot log-likelihoods
ggplot(loglik_df, aes(x = Iteration, y = LogLikelihood)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(title = "Log-Likelihood Across HMM Fits", y = "Log-Likelihood", x = "Iteration")

# Identify best model
best_index <- which.max(log_liks)
best_model <- models[[best_index]]
cat("Best model is at iteration", best_index, "with log-likelihood", log_liks[best_index], "\n")


```



Plot the posterior probability of being in the various states from your best model (e.g. probability of being in state 1 over time)

```{r}

# Extract the state probabilities 
state_probs<-best_model$state_probs()

# plot each of these 
library(ggplot2)


ggplot(pdo, aes(x = time)) +
  geom_line(aes(y = state_probs[,1], col = "State 1")) +
  geom_line(aes(y = state_probs[,2], col = "State 2")) +
  labs(title = "Posterior Probability of States", x = "Year", y = "Probability") +
  theme_minimal() +
  scale_color_manual(values = c("State 1" = "blue", "State 2" = "red")) +
  theme(legend.title = element_blank())


```

What is the long-run probability that the PDO is in state 1 versus state 2? You can calculate this from the transition matrix.
```{r}




```


#################################
##################### Pt. 2 #####
#################################


```{r load_data}
stoplight <- read.csv(here::here("Lab-4", "stoplight.csv"))

# Select only the local biological 
biological<- stoplight %>%
  filter(Type == "Local Biological")


# Normalize each row to range from 0 to 1
biological_normalized <- biological
biological_normalized[, -(1:2)] <- t(apply(biological[, -(1:2)], 1, function(x) {
  (x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
}))

# log transform the results 
biological_normalized[, -(1:2)] <- log(biological_normalized[, -(1:2)])


# Check the results
print("Normalized data (first few rows):")
print(biological_normalized[1:3, 1:6])



# Create plot to visualize normalized data
library(ggplot2)
library(tidyr)

# Pivot the data to long format for plotting
biological_long <- biological_normalized %>%
  pivot_longer(cols = -(1:2), names_to = "Year", values_to = "Value")

# Create the plot faceted by Ecosystem.Indicators
ggplot(biological_long, aes(x = Year, y = Value, group = 1)) +
  geom_line(color = "blue", alpha = 0.7) +
  geom_point(color = "darkblue", size = 0.8) +
  facet_wrap(~ Ecosystem.Indicators, scales = "free_x") +
  labs(title = "Normalized Biological Indicators (0-1 Scale)",
       x = "Year", 
       y = "Normalized Value",
       subtitle = "Each indicator normalized to range from 0 to 1") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        strip.text = element_text(size = 10),
        plot.title = element_text(size = 14))


```

######
###### Correlate 
```{r}
# Reshape the biological data
library(dplyr)
library(tidyr)

# Method 1: Using pivot_longer and pivot_wider
biological_reshaped <- biological_normalized %>%
  # Remove the Type column
  select(-Type) %>%
  # Convert to long format first
  pivot_longer(cols = -Ecosystem.Indicators, names_to = "Year", values_to = "Value") %>%
  # Remove the 'X' prefix from year names and convert to numeric
  mutate(time = as.numeric(gsub("^X", "", Year))) %>%
  mutate(ID = 1) %>%
  # Convert to wide format with Ecosystem.Indicators as column names
  pivot_wider(names_from = Ecosystem.Indicators, values_from = Value)

# Print the first few rows to check
print("Reshaped data (first few rows):")
print(head(biological_reshaped))

######### THERE ARE SOME -INF here... why. 


```
### Create a multivariate HMM with the biological data

```{r}


# 2 state 
# Updated 2-state HMM model
set.seed(123)

# Create hidden Markov chain
hidden <- MarkovChain$new(data = biological_reshaped, n_states = 2)

# Define distributions for each indicator
dists <- list(
 Copepod_richness = "norm", 
 N_copepod = "norm", 
 S_copepod = "norm",
 Biological_transition = "norm", 
 Nearshore_Ichthyoplankton = "norm",
 Ichthy_community_index = "norm",
 Chinook_salmon_juv = "norm",
 Coho_salmon_juv = "norm"
)

# Define initial parameters for each indicator
# Need to match the names in dists, not par0
par0 <- list(
  Copepod_richness = list(mean = c(0.3, 0.7), sd = c(0.2, 0.2)),
  N_copepod = list(mean = c(0.3, 0.7), sd = c(0.2, 0.2)),
  S_copepod = list(mean = c(0.3, 0.7), sd = c(0.2, 0.2)),
  Biological_transition = list(mean = c(0.3, 0.7), sd = c(0.2, 0.2)),
  Nearshore_Ichthyoplankton = list(mean = c(0.3, 0.7), sd = c(0.2, 0.2)),
  Ichthy_community_index = list(mean = c(0.3, 0.7), sd = c(0.2, 0.2)),
  Chinook_salmon_juv = list(mean = c(0.3, 0.7), sd = c(0.2, 0.2)),
  Coho_salmon_juv = list(mean = c(0.3, 0.7), sd = c(0.2, 0.2))
)

# Create observation model using the correct data
# Should use biological_normalized, not calcofi
obs_model <- Observation$new(data = biological_reshaped,
                            dists = dists,
                            n_states = 2,
                            par = par0)

# Create HMM
hmm <- HMM$new(obs = obs_model, 
               hid = hidden)

# Fit the model
hmm$fit(silent = TRUE)

# Print results
print("Model fitting completed")
print(hmm)


# Collect the state estimates 

biological_reshaped$est_state <- factor(paste0("State", hmm$viterbi()))


ggplot(biological_reshaped, aes(Year, as.numeric(est_state), col = est_state)) + 
  geom_line(aes(group = 1), color = "black") + 
  geom_point(size = 3) +
  scale_y_continuous(breaks = c(1, 2), labels = c("State 1", "State 2")) +
  labs(title = "HMM State Changes Over Time",
       x = "Year", 
       y = "State",
       color = "Est. State") + 
  theme_bw()


```
# Same as above, but with three states 
```{r}

# 3 state HMM model
set.seed(123)

# First, ensure data is prepared correctly (exclude Year from model)
data_for_hmm <- biological_reshaped %>% select(-Year)

# Create hidden Markov chain
hidden <- MarkovChain$new(data = data_for_hmm, n_states = 3)

# Define distributions for each indicator
dists <- list(
  Copepod_richness = "norm", 
  N_copepod = "norm", 
  S_copepod = "norm",
  Biological_transition = "norm", 
  Nearshore_Ichthyoplankton = "norm",
  Ichthy_community_index = "norm",
  Chinook_salmon_juv = "norm",
  Coho_salmon_juv = "norm"
)

# Define initial parameters for 3 states
# State 1: Low values (mean ≈ 0.2)
# State 2: Medium values (mean ≈ 0.5)
# State 3: High values (mean ≈ 0.8)
par0 <- list(
  Copepod_richness = list(mean = c(0.2, 0.5, 0.8), sd = c(0.15, 0.15, 0.15)),
  N_copepod = list(mean = c(0.2, 0.5, 0.8), sd = c(0.15, 0.15, 0.15)),
  S_copepod = list(mean = c(0.2, 0.5, 0.8), sd = c(0.15, 0.15, 0.15)),
  Biological_transition = list(mean = c(0.2, 0.5, 0.8), sd = c(0.15, 0.15, 0.15)),
  Nearshore_Ichthyoplankton = list(mean = c(0.2, 0.5, 0.8), sd = c(0.15, 0.15, 0.15)),
  Ichthy_community_index = list(mean = c(0.2, 0.5, 0.8), sd = c(0.15, 0.15, 0.15)),
  Chinook_salmon_juv = list(mean = c(0.2, 0.5, 0.8), sd = c(0.15, 0.15, 0.15)),
  Coho_salmon_juv = list(mean = c(0.2, 0.5, 0.8), sd = c(0.15, 0.15, 0.15))
)

# Create observation model
obs_model <- Observation$new(data = data_for_hmm,
                            dists = dists,
                            n_states = 3,
                            par = par0)

# Create HMM
hmm <- HMM$new(obs = obs_model, 
               hid = hidden)

# Fit the model
print("Fitting 3-state HMM model...")
hmm$fit(silent = TRUE)

# Print results
print("Model fitting completed!")
print(hmm$par)

# Collect the state estimates 
biological_reshaped$est_state <- factor(paste0("State ", hmm$viterbi()))

# Plot the results for 3 states
ggplot(biological_reshaped, aes(Year, as.numeric(est_state), col = est_state)) + 
  geom_line(aes(group = 1), color = "black", alpha = 0.7) + 
  geom_point(size = 3) +
  scale_y_continuous(breaks = c(1, 2, 3), labels = c("State 1", "State 2", "State 3")) +
  labs(title = "3-State HMM: Biological System States Over Time",
       x = "Year", 
       y = "State",
       color = "Estimated State") + 
  theme_bw() +
  theme(legend.position = "bottom")





```





# Does it converge? 
# How many states seem to be supported 
@ transition probabilities 
# Anything else interesting? 


calcofi$ln = log(calcofi$m) # ln transform
calcofi <- dcast(melt(calcofi[,c("year","ln","species")], 
            id.vars = c("year", "species")), 
            year ~ species)
names(calcofi)[2:3] = c("L_stilbius","L_ochotensis")
head(calcofi)








### Choose a couple of covariates and make a multi state HMM 

#### Biological Data 





### Do we have to normalize the data in any way? 



##############################################################




Make sure to label your final write-up with "final" in the title so we know which one is the final one.

# Data

As part of the California Current Integrated Ecosystem Report, NOAA
scientists do annual updates of [stoplight charts for ecosystem
indicators](https://www.fisheries.noaa.gov/west-coast/science-data/ocean-conditions-indicators-trends).

We have included the `stoplight.csv` dataset for this week. One of the
columns divides indicators into groups (e.g. Local Physical, Local
Biological, etc). Please pick a type of indicators, and develop a 2- or
3-state multivariate HMM. A few tips:

Describe what plankton groups and covariates you used and indicate any temporal 
subsetting you chose. Also describe any standardization, centering or scaling 
that was used. 

## Load the data

One of the
columns divides indicators into groups (e.g. Local Physical, Local
Biological, etc). 


## Wrangle the data

```{r wrangle_data}
## add some code here
```


# General tasks

Each group has the same general tasks, but you will adapt them as you work on the data.

Please pick a type of indicators, and develop a 2- or 3-state multivariate HMM. 

# Methods

A few tips:

-   Assume all responses are Gaussian.

-   You're welcome to include covariates (year? Climate variables?) --
    but fitting a simple model without covariates is also totally fine

# Results

Summarize the model you've created. Specifically,

-   Does it converge?

-   How many states seem to be most supported?

-   Plot the time series of estimated states. What does this mean?

-   What are the transition probabilities?

-   If included, what are the covariate effects? What do these mean?

-   Anything else interesting that you've discovered?





# Discussion


# Team contributions

Example: "All team members helped decide on the goal and ran the analyses for the individual regions. Team members 2 & 3 wrote most of the code for the analysis. Team member 1 worked on the plotting section of the report using and adapting code that team member 3 wrote. All team members helped edit the report and wrote the discussion together."


#### We should normalize the data. 
### Scale in r, z normalize 
### When you scale the data, then log transform. 













