---
title: Team 3 - Lab 4
subtitle: Hidden Markov Models
author: "Ben Makhlouf, Brian McGreal, Anna Sulc"
date: May 20, 2025
output: 
  html_document:
    code-folding: true
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
options(dplyr.summarise.inform = FALSE)
```


# Data

<!-- As part of the California Current Integrated Ecosystem Report, NOAA -->
<!-- scientists do annual updates of [stoplight charts for ecosystem -->
<!-- indicators](https://www.fisheries.noaa.gov/west-coast/science-data/ocean-conditions-indicators-trends). -->

<!-- We have included the `stoplight.csv` dataset for this week. One of the -->
<!-- columns divides indicators into groups (e.g. Local Physical, Local -->
<!-- Biological, etc). Please pick a type of indicators, and develop a 2- or -->
<!-- 3-state multivariate HMM. A few tips: -->

<!-- Describe what plankton groups and covariates you used and indicate any temporal  -->
<!-- subsetting you chose. Also describe any standardization, centering or scaling  -->
<!-- that was used.  -->

We chose to work with the PDO variable for the univariate hidden markov model. We worked on a yearly scale and focused on the winter month period to ensure data were aligned on a temporal scale. 

For the Stoplight data set, we chose to focus on local biological variables as we thought it may be interesting to see whether all biological data followed the same trends or if there were opposite trends occuring for different parts of the system. We were interested in how this affected models with different states. 

<!-- Load the data -->

```{r load_data and packages, include=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(hmmTMB)

#install.packages("rsoi")
pdo <- rsoi::download_pdo()
stoplight <- read.csv(here::here("Lab-4", "stoplight.csv"))

# Select only the local biological 
biological <- stoplight %>%
  filter(Type == "Local Biological")
```

## Wrangle the data

For the biological variables, we chose to normalize data as scales varied largely (copepod abundance vs salmon). Additionally, we used log(abundance/counts) in our model to ensure we kept our variables in a positive space. As data was normalized, we replaced any zeros with very small numbers to avoid -INF or NaNs in the data rather than dropping that information entirely. 

```{r wrangle_data, include=FALSE}

### Adjust the data for PDO 
pdo$Year[which(pdo$Month%in%c("Oct","Nov","Dec"))] <- pdo$Year[which(pdo$Month%in%c("Oct","Nov","Dec"))] + 1
pdo <- dplyr::group_by(pdo, Year) %>%
  dplyr::summarize(winter_pdo = mean(PDO[which(Month %in% c("Oct","Nov","Dec","Jan","Feb"))])) %>% 
  dplyr::select(winter_pdo, Year)
# The first year will be missing Oct-Dec
pdo <- pdo[-1,]


plot <- ggplot(pdo, aes(x = Year, y = winter_pdo)) +
  geom_line() +
  labs(title = "Winter PDO vs Year", x = "Year", y = "Winter PDO") +
  theme_classic()

plot


# Normalize each row to range from 0 to 1
biological_normalized <- biological
biological_normalized[, -(1:2)] <- t(apply(biological[, -(1:2)], 1, function(x) {
  (x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
}))



# log transform the results 
biological_normalized[, -(1:2)] <- log(pmax(biological_normalized[, -(1:2)], 1e-6))
# Check the results
print("Normalized data (first few rows):")
print(biological_normalized[1:3, 1:6])


# Pivot the data to long format for plotting
biological_long <- biological_normalized %>%
  pivot_longer(cols = -(1:2), names_to = "Year", values_to = "Value")

# Create the plot faceted by Ecosystem.Indicators
ggplot(biological_long, aes(x = Year, y = Value, group = 1)) +
  geom_line(color = "blue", alpha = 0.7) +
  geom_point(color = "darkblue", size = 0.8) +
  facet_wrap(~ Ecosystem.Indicators, scales = "free_x") +
  labs(title = "Normalized Biological Indicators (0-1 Scale)",
       x = "Year", 
       y = "Normalized Value",
       subtitle = "Each indicator normalized to range from 0 to 1") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        strip.text = element_text(size = 10),
        plot.title = element_text(size = 14))


```

# Methods

## Univariate - PDO

We assumed all responses are Gaussian in our models. We chose to fit a simple model with no covariates in both instances and instead and instead focused on the impacts of assuming different numbers of states and exploring model diagnostics and performance. 


Here we fit a two state HMM to the annual indices of PDO 
```{r, echo=FALSE}

library(hmmTMB)

pdo <- data.frame(
  ID = 1,
  time = pdo$Year,  # Use actual years as time
  winter_pdo = pdo$winter_pdo
)

# Set up hidden states (2 states for different PDO regimes)
hidden <- MarkovChain$new(data = pdo, n_states = 2)

# Use normal distribution for continuous PDO values
dists <- list(winter_pdo = "norm")

# Initial parameters for normal distribution (mean and sd for each state)
par0 <- list(winter_pdo = list(mean = c(-0.5, 0.5),    # Different means for each state
                               sd = c(0.8, 0.8)))      # Standard deviations

# Create observation model
obs_model <- Observation$new(data = pdo,
                             dists = dists,
                             n_states = 2,
                             par = par0)

# Fit the HMM
hmm <- HMM$new(obs = obs_model, hid = hidden)
hmm$fit(silent = TRUE)

```
    
Examine the model output 
```{r, echo=FALSE}
hmm

hmm$par()
```

```{r plot, echo=FALSE}
pdo$est_state <- factor(paste0("State", hmm$viterbi()))

ggplot(pdo, aes(time, winter_pdo, col = est_state)) + 
  geom_point(size = 2) + # This line will connect all the points
  ylab("Warm or Cold?") + 
  xlab("") + 
  theme_classic()
```

We fit the model 20 times and explored the likelihood. Plotting the likelihood demonstrated that is was quite stable and remained in a similar range across all 20 iterations. This may be one way to check convergence - if we were located in a local minimum, running a couple hundred iterations might show a likelihod that jumps between two significantly different values. 

```{r, echo=FALSE}
# Placeholder for storing models and log-likelihoods
models <- vector("list", 20)
log_liks <- numeric(20)

# Run 20 iterations
for (i in 1:20) {
  
  # Optionally randomize initial parameters slightly
  par0 <- list(winter_pdo = list(
    mean = rnorm(2, mean = c(-0.5, 0.5), sd = 0.2),  # jittered means
    sd = runif(2, min = 0.5, max = 1.2)              # jittered sds
  ))
  
  # Set up models
  hidden <- MarkovChain$new(data = pdo, n_states = 2)
  obs_model <- Observation$new(data = pdo, dists = list(winter_pdo = "norm"),
                               n_states = 2, par = par0)
  
  # Fit the model
  hmm <- HMM$new(obs = obs_model, hid = hidden)
  hmm$fit(silent = TRUE)
  
  # Store the model and log-likelihood
  models[[i]] <- hmm
  log_liks[i] <- logLik(hmm)
}

# Create a data frame for plotting
loglik_df <- data.frame(Iteration = 1:20, LogLikelihood = log_liks)

# Plot log-likelihoods
ggplot(loglik_df, aes(x = Iteration, y = LogLikelihood)) +
  geom_point() +
  theme_classic() +
  labs(title = "Log-Likelihood Across HMM Fits", y = "Log-Likelihood", x = "Iteration")

# Identify best model
best_index <- which.max(log_liks)
best_model <- models[[best_index]]
cat("Best model is at iteration", best_index, "with log-likelihood", log_liks[best_index], "\n")
```

**Plot the posterior probability of being in the various states from your best model (e.g. probability of being in state 1 over time)**

```{r, echo=FALSE}

# Extract the state probabilities 
state_probs<-best_model$state_probs()

# plot each of these 
library(ggplot2)

ggplot(pdo, aes(x = time)) +
  geom_line(aes(y = state_probs[,1], col = "State 1")) +
  geom_point(aes(y = state_probs[,1], col = "State 1")) +
  geom_line(aes(y = state_probs[,2], col = "State 2")) +
  geom_point(aes(y = state_probs[,2], col = "State 2")) +
  labs(title = "Posterior Probability of States", x = "Year", y = "Probability") +
  theme_classic() +
  scale_color_manual(values = c("State 1" = "blue", "State 2" = "red")) +
  theme(legend.title = element_blank())


```

**What is the long-run probability that the PDO is in state 1 versus state 2?**

The long-run probabilities represent the proportion of time the PDO is expected to be in each state over a very long period. This can be calculated using the eigenvalues of the transition matrix. Our results show that we will be in State 1 approximately 2/3s of the time in a long term future.

```{r, echo=FALSE}

trans_mat <- as.data.frame(best_model$par()$tpm)

#Can also get it this way but can't get df
#best_model$print_tpm()

calculate_stationary_distribution <- function(P) { # must be a square matrix

  eigen_analysis <- eigen(t(P)) # Transpose the matrix
  eigenvalues <- eigen_analysis$values
  eigenvectors <- eigen_analysis$vectors

  # Find the eigenvalue closest to 1
  target_eigenvalue_index <- which.min(abs(eigenvalues - 1))
  target_eigenvector <- eigenvectors[, target_eigenvalue_index]

  # Normalize the eigenvector to sum to 1
  stationary_distribution <- target_eigenvector / sum(target_eigenvector)

  # Check that probabilities are real and positive
  if(any(Im(stationary_distribution) != 0) || any(Re(stationary_distribution) < 0)){
    n_states <- nrow(P)
    initial_distribution <- rep(1/n_states, n_states) # Start with equal probabilities
    for (i in 1:1000) {  # Iterate until convergence (adjust as needed)
      initial_distribution <- initial_distribution %*% P
    }
    stationary_distribution <- as.vector(initial_distribution)
  } else {
    stationary_distribution <- Re(stationary_distribution)
  }
  return(stationary_distribution)
}

# Calculate the stationary distribution
long_run_probs <- calculate_stationary_distribution(trans_mat)

# Print the probs
cat("Long-run probability of being in State 1:", long_run_probs[1], "\n")
cat("Long-run probability of being in State 2:", long_run_probs[2], "\n")

cat("The PDO is predicted to be in State 1 approximately", round(long_run_probs[1]*100,2), "% of the time, and in State 2 approximately", round(long_run_probs[2]*100,2), "% of the time.\n")


```

**Using slide 15 of the HMM lecture as a template and the matrix in Q3, write out your fitted HMM model as equations.**

Our PDO model has two states and assumes a gaussian distribution

1. Hidden State Process (Markov Chain)

Let \( z_t \in \{1, 2\} \) denote the hidden state at time \( t \). The hidden states follow a first-order Markov chain:

\[
\Pr(z_1 = i) = \pi_i, \quad i \in \{1, 2\}
\]

\[
\Pr(z_t = j \mid z_{t-1} = i) = a_{ij}, \quad i,j \in \{1, 2\}
\]

where \( \pi_i \) are the initial state probabilities and \( a_{ij} \) are the elements of the transition probability matrix \( A \).

2. Observation Model 

Let \( y_t \) be the observed PDO value at time \( t \). Conditional on the hidden state \( z_t = k \), the observation is normally distributed:

\[
y_t \mid z_t = k \sim \mathcal{N}(\mu_k, \sigma_k^2)
\]

for \( k \in \{1, 2\} \), where \( \mu_k \) and \( \sigma_k^2 \) are the state-specific mean and variance of the normal distribution.


The transition matrix \( A \) for the model:

\[
A = 
\begin{pmatrix}
0.8316750 & 0.1683250 \\
0.2928184 & 0.7071816
\end{pmatrix}
\]

This matrix represents the probabilities of transitioning between two hidden states, **State 1** and **State 2**. In other words: 

- From **State 1**, the probability of remaining in **State 1** is \( 0.8316750 \), and the probability of transitioning to **State 2** is \( 0.1683250 \).
- From **State 2**, the probability of transitioning to **State 1** is \( 0.2928184 \), and the probability of remaining in **State 2** is \( 0.7071816 \).




## Multivariate - "Stoplight" Indicator Data

```{r, echo=FALSE}


biological_reshaped <- biological_normalized %>%

  select(-Type) %>%
  # Convert to long format first
  pivot_longer(cols = -Ecosystem.Indicators, names_to = "Year", values_to = "Value") %>%
  mutate(time = as.numeric(gsub("^X", "", Year))) %>%
  mutate(ID = 1) %>%
  # Convert to wide format with Ecosystem.Indicators as column names
  pivot_wider(names_from = Ecosystem.Indicators, values_from = Value)

# Print the first few rows to check
print("Reshaped data (first few rows):")
print(head(biological_reshaped))

```

## 2 state vs. 3 state HMMs

#### 2 states
```{r 2state, echo=FALSE}
library(data.table)
stoplight <- read.csv(here::here("Lab-4", "stoplight.csv"))

# Select only the local biological 
data <- stoplight %>%
  filter(Type == "Local Biological")

data <- data[-c(2)]
years <- colnames(data[2:length(data)])
species <- c(data$Ecosystem.Indicators)
data <- data[-c(1)]
data <- transpose(data)
colnames(data) <- species
cor(data)
data$year <- years
data$year <- substring(data$year, 2)

# specify markov model

set.seed(1)

# 2 state model
hidden2 <- MarkovChain$new(data = data, n_states = 2)
dists2 <- list(Copepod_richness = "norm",
              N_copepod = "norm",
              S_copepod = "norm",
              Biological_transition = "norm",
              Nearshore_Ichthyoplankton = "norm",
              Ichthy_community_index = "norm",
              Chinook_salmon_juv = "norm",
              Coho_salmon_juv = "norm")
par02 <- list(Copepod_richness = list(mean=c(mean(data$Copepod_richness)+sd(data$Copepod_richness),
                                             mean(data$Copepod_richness)-sd(data$Copepod_richness)), 
                                      sd=c(sd(data$Copepod_richness),
                                           sd(data$Copepod_richness))),
             N_copepod = list(mean=c(mean(data$N_copepod)+sd(data$N_copepod),
                                     mean(data$N_copepod)-sd(data$N_copepod)), 
                              sd=c(sd(data$N_copepod),
                                   sd(data$N_copepod))),
             S_copepod = list(mean=c(mean(data$S_copepod)+sd(data$S_copepod),
                                     mean(data$S_copepod)-sd(data$S_copepod)), 
                              sd=c(sd(data$S_copepod),
                                   sd(data$S_copepod))),
             Biological_transition = list(mean=c(mean(data$Biological_transition)+sd(data$Biological_transition),
                                                 mean(data$Biological_transition)-sd(data$Biological_transition)), 
                                          sd=c(sd(data$Biological_transition),
                                               sd(data$Biological_transition))),
             Nearshore_Ichthyoplankton = list(mean=c(mean(data$Nearshore_Ichthyoplankton)+sd(data$Nearshore_Ichthyoplankton),
                                                     mean(data$Nearshore_Ichthyoplankton)-sd(data$Nearshore_Ichthyoplankton)), 
                                              sd=c(sd(data$Nearshore_Ichthyoplankton),
                                                   sd(data$Nearshore_Ichthyoplankton))),
             Ichthy_community_index = list(mean=c(mean(data$Ichthy_community_index)+sd(data$Ichthy_community_index),
                                                  mean(data$Ichthy_community_index)-sd(data$Ichthy_community_index)), 
                                           sd=c(sd(data$Ichthy_community_index),
                                                sd(data$Ichthy_community_index))),
             Chinook_salmon_juv = list(mean=c(mean(data$Chinook_salmon_juv)+sd(data$Chinook_salmon_juv),
                                              mean(data$Chinook_salmon_juv)-sd(data$Chinook_salmon_juv)), 
                                       sd=c(sd(data$Chinook_salmon_juv),
                                            sd(data$Chinook_salmon_juv))),
             Coho_salmon_juv = list(mean=c(mean(data$Coho_salmon_juv)+sd(data$Coho_salmon_juv),
                                           mean(data$Coho_salmon_juv)-sd(data$Coho_salmon_juv)), 
                                    sd=c(sd(data$Coho_salmon_juv),
                                         sd(data$Coho_salmon_juv)))             
             )
obs_model2 <- Observation$new(data = data,
                             dists = dists2,
                             n_states = 2,
                             par = par02)
hmm2 <- HMM$new(obs = obs_model2,
               hid = hidden2)
hmm2$fit(silent=FALSE)
hmm2

data$est_state2 <- factor(paste0("State", hmm2$viterbi()))

ggplot(data, aes(x=year, y=est_state2, col = est_state2)) + 
  geom_point() + 
  ylab("2 state") + 
  xlab("")+
  theme_classic()
```

#### 3 states
```{r 3state, echo=FALSE}

# 3 state model
hidden3 <- MarkovChain$new(data = data, n_states = 3)
dists3 <- list(Copepod_richness = "norm",
              N_copepod = "norm",
              S_copepod = "norm",
              Biological_transition = "norm",
              Nearshore_Ichthyoplankton = "norm",
              Ichthy_community_index = "norm",
              Chinook_salmon_juv = "norm",
              Coho_salmon_juv = "norm")
par03 <- list(Copepod_richness = list(mean=c(mean(data$Copepod_richness)+1.5*sd(data$Copepod_richness),
                                             mean(data$Copepod_richness),
                                             mean(data$Copepod_richness)-1.5*sd(data$Copepod_richness)),
                                      sd=c(sd(data$Copepod_richness),
                                           sd(data$Copepod_richness),
                                           sd(data$Copepod_richness))),
             N_copepod = list(mean=c(mean(data$N_copepod)+1.5*sd(data$N_copepod),
                                     mean(data$N_copepod),
                                     mean(data$N_copepod)-1.5*sd(data$N_copepod)), 
                              sd=c(sd(data$N_copepod),
                                   sd(data$N_copepod),
                                   sd(data$N_copepod))),
             S_copepod = list(mean=c(mean(data$S_copepod)+1.5*sd(data$S_copepod),
                                     mean(data$S_copepod),
                                     mean(data$S_copepod)-1.5*sd(data$S_copepod)), 
                              sd=c(sd(data$S_copepod),
                                   sd(data$S_copepod),
                                   sd(data$S_copepod))),
             Biological_transition = list(mean=c(mean(data$Biological_transition)+1.5*sd(data$Biological_transition),
                                                 mean(data$Biological_transition),
                                                 mean(data$Biological_transition)-1.5*sd(data$Biological_transition)), 
                                          sd=c(sd(data$Biological_transition),
                                               sd(data$Biological_transition),
                                               sd(data$Biological_transition))),
             Nearshore_Ichthyoplankton = list(mean=c(mean(data$Nearshore_Ichthyoplankton)+1.5*sd(data$Nearshore_Ichthyoplankton),
                                                     mean(data$Nearshore_Ichthyoplankton),
                                                     mean(data$Nearshore_Ichthyoplankton)-1.5*sd(data$Nearshore_Ichthyoplankton)), 
                                              sd=c(sd(data$Nearshore_Ichthyoplankton),
                                                   sd(data$Nearshore_Ichthyoplankton),
                                                   sd(data$Nearshore_Ichthyoplankton))),
             Ichthy_community_index = list(mean=c(mean(data$Ichthy_community_index)+1.5*sd(data$Ichthy_community_index),
                                                  mean(data$Ichthy_community_index),
                                                  mean(data$Ichthy_community_index)-1.5*sd(data$Ichthy_community_index)), 
                                           sd=c(sd(data$Ichthy_community_index),
                                                sd(data$Ichthy_community_index),
                                                sd(data$Ichthy_community_index))),
             Chinook_salmon_juv = list(mean=c(mean(data$Chinook_salmon_juv)+1.5*sd(data$Chinook_salmon_juv),
                                              mean(data$Chinook_salmon_juv),
                                              mean(data$Chinook_salmon_juv)-1.5*sd(data$Chinook_salmon_juv)), 
                                       sd=c(sd(data$Chinook_salmon_juv),
                                            sd(data$Chinook_salmon_juv),
                                            sd(data$Chinook_salmon_juv))),
             Coho_salmon_juv = list(mean=c(mean(data$Coho_salmon_juv)+1.5*sd(data$Coho_salmon_juv),
                                           mean(data$Coho_salmon_juv),
                                           mean(data$Coho_salmon_juv)-1.5*sd(data$Coho_salmon_juv)), 
                                    sd=c(sd(data$Coho_salmon_juv),
                                         sd(data$Coho_salmon_juv),
                                         sd(data$Coho_salmon_juv)))             
             )
obs_model3 <- Observation$new(data = data,
                             dists = dists3,
                             n_states = 3,
                             par = par03)
hmm3 <- HMM$new(obs = obs_model3,
               hid = hidden3)
hmm3$fit(silent=FALSE)
hmm3

data$est_state3 <- factor(paste0("State", hmm3$viterbi()))

ggplot(data, aes(x=year, est_state3, col = est_state3)) + 
  geom_point() + 
  ylab("3 state") + 
  xlab("")+
  theme_classic()

# ggplot(hmm3_series, aes(Year, as.numeric(est_state), col = est_state)) +
#    geom_line(aes(group = 1), color = "black") +
#    geom_point(size = 3) +
#    scale_y_continuous(breaks = c(1, 2), labels = c("State 1", "State 2")) +
#    labs(title = "HMM State Changes Over Time",
#         x = "Year",
#         y = "State",
#         color = "Est. State") +
#    theme_classic()
```
# Results

We tested two forms of models. In a first step we experimented with the PDO data and a two state model that assumed a normal response of the data. 

In the second part of the lab, we experimented with a 1, 2 and 3 state model for the Local Biological data sets of the stoplight dataset. We also assumed a normal distribution for all data.

The following section summarizes the outputs of our analysis.

### Convergence

```{r, echo=FALSE}


#For initial hmm
pseudo <- best_model$pseudores()

ggplot(as.data.frame(pseudo), aes(sample = pdo$winter_pdo)) +
  geom_qq() +
  geom_qq_line(col="red") +
  ggtitle("QQ Plot")

best_model$AIC_marginal()

best_model$AIC_conditional()


#For biological data

# hmm1$AIC_marginal()
# 
# hmm2$AIC_marginal()
# 
# hmm3$AIC_marginal()


```

### States


```{r, echo=FALSE}



```


### -   Plot the time series of estimated states. What does this mean?

In the case of the PDO model, we seem to be in the second state more than the first. This is consistent with the estimates of long term probabilities where we spend 2/3s of the time in one state and 1/3 in the other. 

```{r, echo=FALSE}

pdo$est_state <- factor(paste0("State", best_model$viterbi()))

ggplot(pdo, aes(time, winter_pdo, col = est_state)) + 
  geom_line(aes(group = 1), color = "black") + 
  geom_point(size = 2) + # This line will connect all the points
  ylab("Rained?") + 
  theme_classic()
```

### Transition Probabilities

```{r, echo=FALSE}

trans_mat

```




# Discussion

## PDO HMM

As a first experience with HMMs, this lab was a great way to appreciate their strengths and also inquire about potential limitations. We used the PDO example to explore the diagnostics and performance of a HMM. The results demonstrated a 60/30 split in states (for long term probabilities, individual model runs could vary between 50/50 and 80/20 for example) which for PDO seemed intuitive given it is an oscillating even but remains uneven in duration and in its phases. 

While the model performed well and our 20 iteration "check" using the loglikelihood seemed to show a stable global minimum, the remainder of diagnostics raised the question of convergence of HMMs. It seemed that while it is easy to compare the performance of different HMMs and their fit, checking convergence is less direct. The loglikelihood test is effective however a question may be raised for how many iterations are sufficient before one decides the model has converged. We explored the TMB object of the fitted model but it was unclear whether other diagnostics were available beyond those listed on the slides.  

## Multistate Biological Data HMM


# Team contributions

All team members worked through each lab sections and reconvened to discuss results and the general direction of the lab. Based on this discussion, we then collectively combined our code and each contributed to the write up. 

