---
title: Team 1 - Lab 4
subtitle: Hidden Markov Models
author: "Name 1, Name 2, Name 3"
date: May XX, 2025
output: 
  html_document:
    code-folding: true
    toc: true
    toc_float: true
---



```{r setup, include = FALSE}
options(dplyr.summarise.inform = FALSE)
```

###############################################################
############ Pt 1: PDO Data #############
###############################################################
###############################################################

```{r}
library(dplyr)
#install.packages("rsoi")
pdo <- rsoi::download_pdo()

### Adjust the data 
pdo$Year[which(pdo$Month%in%c("Oct","Nov","Dec"))] <- pdo$Year[which(pdo$Month%in%c("Oct","Nov","Dec"))] + 1
pdo <- dplyr::group_by(pdo, Year) %>%
  dplyr::summarize(winter_pdo = mean(PDO[which(Month %in% c("Oct","Nov","Dec","Jan","Feb"))])) %>% 
  dplyr::select(winter_pdo, Year)
# The first year will be missing Oct-Dec
pdo <- pdo[-1,]

```

# Plot winter PDO vs year
```{r}
library(ggplot2)

plot <- ggplot(pdo, aes(x = Year, y = winter_pdo)) +
  geom_line() +
  labs(title = "Winter PDO vs Year", x = "Year", y = "Winter PDO") +
  theme_minimal()

plot

```
### Add a time column 

```{r}

pdo <- data.frame(
  ID = 1,
  time = pdo$Year,  # Use actual years as time
  winter_pdo = pdo$winter_pdo
)



```


### Fit a two state HMM to the annual indices of PDO 
```{r}

library(hmmTMB)

# Assume your data is in a data frame called 'pdo_data'
# with columns 'winter_pdo' and 'Year'

# Set up hidden states (2 states for different PDO regimes)
hidden <- MarkovChain$new(data = pdo, n_states = 2)

# Use normal distribution for continuous PDO values
dists <- list(winter_pdo = "norm")

# Initial parameters for normal distribution (mean and sd for each state)
par0 <- list(winter_pdo = list(mean = c(-0.5, 0.5),    # Different means for each state
                               sd = c(0.8, 0.8)))      # Standard deviations

# Create observation model
obs_model <- Observation$new(data = pdo,
                             dists = dists,
                             n_states = 2,
                             par = par0)

# Fit the HMM
hmm <- HMM$new(obs = obs_model, hid = hidden)
hmm$fit(silent = TRUE)
```


#Examine the output 
```{r}
hmm
hmm$par

pdo$est_state <- factor(paste0("State", hmm$viterbi()))

ggplot(pdo, aes(time, winter_pdo, col = est_state)) + 
  geom_line(aes(group = 1), color = "black") + 
  geom_point(size = 2) + # This line will connect all the points
  ylab("Rained?") + 
  theme_bw()

logLik(hmm)

```
Try fitting the model 10-20 times. Does the likelihood seem reasonably stable? (Note logLik() gets you the log-likelihood from model fits in R).
```{r}
library(hmmTMB)
library(ggplot2)

# Placeholder for storing models and log-likelihoods
models <- vector("list", 20)
log_liks <- numeric(20)

# Run 20 iterations
for (i in 1:20) {
  
  # Optionally randomize initial parameters slightly
  par0 <- list(winter_pdo = list(
    mean = rnorm(2, mean = c(-0.5, 0.5), sd = 0.2),  # jittered means
    sd = runif(2, min = 0.5, max = 1.2)              # jittered sds
  ))
  
  # Set up models
  hidden <- MarkovChain$new(data = pdo, n_states = 2)
  obs_model <- Observation$new(data = pdo, dists = list(winter_pdo = "norm"),
                               n_states = 2, par = par0)
  
  # Fit the model
  hmm <- HMM$new(obs = obs_model, hid = hidden)
  hmm$fit(silent = TRUE)
  
  # Store the model and log-likelihood
  models[[i]] <- hmm
  log_liks[i] <- logLik(hmm)
}

# Create a data frame for plotting
loglik_df <- data.frame(Iteration = 1:20, LogLikelihood = log_liks)

# Plot log-likelihoods
ggplot(loglik_df, aes(x = Iteration, y = LogLikelihood)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(title = "Log-Likelihood Across HMM Fits", y = "Log-Likelihood", x = "Iteration")

# Identify best model
best_index <- which.max(log_liks)
best_model <- models[[best_index]]
cat("Best model is at iteration", best_index, "with log-likelihood", log_liks[best_index], "\n")


```

##################################
############ Pt2 #################
##################################

```{r load_data}
stoplight <- read.csv(here::here("Lab-4", "stoplight.csv"))
```

### Choose a couple of covariates and make a multi state HMM 

### Do we have to normalize the data in any way? 



##############################################################




Make sure to label your final write-up with "final" in the title so we know which one is the final one.

# Data

As part of the California Current Integrated Ecosystem Report, NOAA
scientists do annual updates of [stoplight charts for ecosystem
indicators](https://www.fisheries.noaa.gov/west-coast/science-data/ocean-conditions-indicators-trends).

We have included the `stoplight.csv` dataset for this week. One of the
columns divides indicators into groups (e.g. Local Physical, Local
Biological, etc). Please pick a type of indicators, and develop a 2- or
3-state multivariate HMM. A few tips:

Describe what plankton groups and covariates you used and indicate any temporal 
subsetting you chose. Also describe any standardization, centering or scaling 
that was used. 

## Load the data

One of the
columns divides indicators into groups (e.g. Local Physical, Local
Biological, etc). 


## Wrangle the data

```{r wrangle_data}
## add some code here
```


# General tasks

Each group has the same general tasks, but you will adapt them as you work on the data.

Please pick a type of indicators, and develop a 2- or 3-state multivariate HMM. 

# Methods

A few tips:

-   Assume all responses are Gaussian.

-   You're welcome to include covariates (year? Climate variables?) --
    but fitting a simple model without covariates is also totally fine

# Results

Summarize the model you've created. Specifically,

-   Does it converge?

-   How many states seem to be most supported?

-   Plot the time series of estimated states. What does this mean?

-   What are the transition probabilities?

-   If included, what are the covariate effects? What do these mean?

-   Anything else interesting that you've discovered?

# Discussion


# Team contributions

Example: "All team members helped decide on the goal and ran the analyses for the individual regions. Team members 2 & 3 wrote most of the code for the analysis. Team member 1 worked on the plotting section of the report using and adapting code that team member 3 wrote. All team members helped edit the report and wrote the discussion together."

