---
title: Team 2 - Lab 2
subtitle: MARSS models
author: "Brian McGreal, Lilac Hong, Emma Christman"
date: April 23, 2025
output: 
  html_document:
    code-folding: true
    toc: true
    toc_float: true
---

# Data
As team 2, we were assigned the Snake River Spring and Summer Run Chinook Data. This data has three seasons of runs: spring, spring/summer, and summer, divided among 5 major populations and when separated by run timing, major population and smaller tributary, 28 different populations.

Load the data.
```{r}
library(atsalibrary)
library(here)
library(MARSS)
library(panelr)
library(readr)
library(tidyverse)
library(vcvComp)
library(zoo)

load(here::here("Lab-2", "Data_Images", "esa-salmon.rda"))
```

Wrangle the data.

**Add your wrangling code here**
```{r data, echo=FALSE}
# check out the data
esu <- unique(esa.salmon$esu_dps)
head(esa.salmon)
esu

# subset esu data
esu2 <- subset(esa.salmon, esu_dps == "Salmon, Chinook (Snake River spring/summer-run ESU)")
esu2$popname <- substr(esu2$esapopname, 53, nchar(esu2$esapopname)) 

# log transform and set data wide
esu2$logvalue <- log(esu2$value + 1)
esu2 <- esu2 %>% mutate_all(~ifelse(is.nan(.), NA, .))

esu2_widen <- esu2[-c(1:6, 8)]
w_esu2 <- panel_data(esu2_widen, id = popname, wave = spawningyear)
w_esu2 <- widen_panel(w_esu2, separator = "_")

# grab years, population names, and n
years <- names(w_esu2)
years <- years[-1]
years <- substring(years, first=10, last=13)

pops <- w_esu2$popname

# grab years, population names, and n
years <- names(w_esu2)
years <- years[-1]
years <- substring(years, first=10, last=13)

pops <- w_esu2$popname

# convert counts to matrix
data <- data.matrix(w_esu2[2:ncol(w_esu2)])
data

## get pdo data
# read in data from table
if(!file.exists(here::here("Lab-2", "Team-2", "data", "pdo_raw.rda"))){
  pdo <- read.table("https://www.ncei.noaa.gov/pub/data/cmb/ersst/v5/index/ersst.v5.pdo.dat",
                    sep = "",
                    header = TRUE,
                    skip = 1)
  save(pdo, file=here::here("Lab-2", "Team-2", "data", "pdo_raw.rda"))
} 

if(file.exists(here::here("Lab-2", "Team-2", "data", "pdo_raw.rda"))) {
  load(here::here("Lab-2", "Team-2", "data", "pdo_raw.rda"))
}

#annual average
# annual average - is this smart?
annual <- rowMeans(pdo[ ,2:13])
annual
pdo$annual <- annual

# drop month vars and years before 1960
pdo <- pdo[-c(2:13)]
pdo <- pdo %>% filter(Year >= 1946)
pdo <- pdo %>% filter(Year < 2025)

# create lagged average (taking average of three prior years)
pdo$pdo_03ya <- zoo::rollmean(pdo$annual ,k=3, align="right", fill=NA)
pdo$spawningyear <- pdo$Year + 1
head(pdo)
  # convinced?
pdo <- pdo[-c(1,2)]
pdo <- pdo %>% filter(spawningyear >= 1949)
pdo <- pdo %>% filter(spawningyear < 2025)

# set wide
pdo_yrs <- pdo$spawningyear
pdo_idx <- pdo$pdo_03ya

Wpdo <- as.data.frame(matrix(pdo_idx, nrow=1, byrow=TRUE))
names(Wpdo) <- pdo_yrs
Wpdo <- data.matrix(Wpdo)
```

# General Questions

Each group has the same general tasks, but you will adapt them as you work on the data.

As a group, we've decided to dig into the estimates during the missing years, as outlined in the first question, as well as investigate the relationship between the different populations to see how they are/are not correlated. This comes from the second question, to which we are investigating not only the role of major population group, but run timing, and esa population name. This is allowing us to test several hypotheses:

H1: There is no difference between any of the groups (observing the same population different rivers and times)
H2: There is a difference between all of the separate esa populations, major populations, and run timing, meaning 28 different populations
H3: The same as H2, however there is temporal correlation 
H4: Run timing is the only distinguishing factor, meaning that there are only 3 correlated populations (and are temporally correlated)
H5: Major population groups is the only distinguishing factor, meaning that there are only 5 correlated populations 
H6: Major population group and run timing are the distinguishing factors, meaning that there are 9 correlated populations (and are temporally correlated).

1. Create estimates of spawner abundance for all missing years and provide estimates of the decline from the historical abundance.

2. Evaluate support for the major population groups. Are the populations in the groups more correlated than outside the groups?

3. Evaluate the evidence of cycling in the data. *We will talk about how to do this on the Tuesday after lab.*


# Methods and Code

Address the following in your methods

* Describe your assumptions about the x and how the data time series are related to x.

   - How are the x and y (data) related? 1 x for 1 y or will you assume 1 x for all y or 1 x for each major population group? How will you choose? 
   - What will you assume about the U for the x's?
   - What will you assume about the Q matrix?
   
* Write out your assumptions as different models **in matrix form**, fit each and then compare these with AIC or AICc.

* Do your estimates differ depending on the assumptions you make about the structure of the data, i.e. you assumptions about the x's, Q, and U.


#Code
## Hypothesis 1
```{r H1, echo=FALSE}
#H1: There is no correlation between the populations
mod.list.1 <- list(B = "identity", 
                   U = matrix("u"),
                   Q = matrix("q"),
                   Z = matrix(1,28,1),
                   A = "zero",
                   R = "diagonal and equal",
                   x0 = matrix("mu"),
                   tinitx = 0)

if(!file.exists(here::here("Lab-2", "Team-2", "data", "fit1.rds"))){
fit.1 <- MARSS(data, model = mod.list.1)
  saveRDS(fit.1, file=here::here("Lab-2", "Team-2", "data", "fit1.rds"))
}
# load in fit.1
fit.1 <- readRDS(file=here::here("Lab-2", "Team-2", "data", "fit1.rds"))

# grabbing data for figures
states <- fit.1$states
statesSE <- fit.1$states.se

states <- as.data.frame(states)
statesSE <- as.data.frame(statesSE)

states_long <- states %>% 
  pivot_longer("V1":"V76", names_to="Year", values_to="fitted")
states_long <- states_long %>% 
  mutate(Year = years)

statesSE_long <- statesSE %>% 
  pivot_longer("V1":"V76", names_to="Year", values_to="se")

states_long <- states_long %>% 
  mutate(se = statesSE_long$se)

states_long$lb <- states_long$fitted - states_long$se
states_long$ub <- states_long$fitted + states_long$se

# save long state estimates\
save(states_long, file=here::here("Lab-2", "Team-2", "data", "fit1_long.Rda"))

plot <- ggplot(data = states_long, aes(x = Year, y = fitted, group=1)) +
  geom_line() +
  labs(x = "", 
       y="State Estimate, ln(abundance)",
       title='Spring Summer Chinook Abundance, Upper Columbia/Snake River') +
  geom_ribbon(aes(ymin=lb, ymax=ub), alpha=0.35, linetype=0) +
  theme_classic() +
  theme(legend.position="bottom") + 
  scale_x_discrete(breaks = c(1950, 1964, 1978, 1992, 2006, 2020))
plot
```

## Hypothesis 2
```{r H2, echo=FALSE}
#H2: There is a correlation within the populations (28 different populations)
mod.list.2 <- list(B = "identity", 
                   U = matrix(paste0("u", seq(nrow(data)))),
                   Q = "diagonal and equal",
                   Z = "identity",
                   A = "zero",
                   R = "diagonal and equal",
                   x0 = "unequal",
                   tinitx = 0)

if(!file.exists(here::here("Lab-2", "Team-2", "data", "fit2.rds"))){
fit.2 <- MARSS::MARSS(data, model = mod.list.2)
  saveRDS(fit.2, file=here::here("Lab-2", "Team-2", "data", "fit2.rds"))
}
# load in fit.2
fit.2 <- readRDS(file=here::here("Lab-2", "Team-2", "data", "fit2.rds"))

# grabbing data for figures
states <- fit.2$states
statesSE <- fit.2$states.se

states <- as.data.frame(states)
statesSE <- as.data.frame(statesSE)

states_long <- states %>% 
  pivot_longer("V1":"V76", names_to="Year", values_to="fitted")
states_long <- states_long %>% 
  mutate(Year = rep(years, nrow(data)))

states_long <- states_long %>% 
  mutate(state = c(rep(pops[1], 76),
                   rep(pops[2], 76),
                   rep(pops[3], 76),
                   rep(pops[4], 76),
                   rep(pops[5], 76),
                   rep(pops[6], 76),
                   rep(pops[7], 76),
                   rep(pops[8], 76),
                   rep(pops[9], 76),
                   rep(pops[10], 76),
                   rep(pops[11], 76),
                   rep(pops[12], 76),
                   rep(pops[13], 76),
                   rep(pops[14], 76),
                   rep(pops[15], 76),
                   rep(pops[16], 76),
                   rep(pops[17], 76),
                   rep(pops[18], 76),
                   rep(pops[19], 76),
                   rep(pops[20], 76),
                   rep(pops[21], 76),
                   rep(pops[22], 76),
                   rep(pops[23], 76),
                   rep(pops[24], 76),
                   rep(pops[25], 76),
                   rep(pops[26], 76),
                   rep(pops[27], 76),
                   rep(pops[28], 76)
                   ))

statesSE_long <- statesSE %>% 
  pivot_longer("V1":"V76", names_to="Year", values_to="se")

states_long <- states_long %>% 
  mutate(se = statesSE_long$se)

states_long$lb <- states_long$fitted - states_long$se
states_long$ub <- states_long$fitted + states_long$se

# save long state estimates\
save(states_long, file=here::here("Lab-2", "Team-2", "data", "fit2_long.Rda"))

plot <- ggplot(data = states_long, aes(x = Year, y = fitted, group = state)) +
  geom_line() +
  labs(x = "", 
       y="State Estimate",
       title='Spring Summer Chinook Abundance, Upper Columbia/Snake River',
       subtitle="28 individual populations, equal process variance") +
  geom_ribbon(aes(ymin=lb, ymax=ub, fill=state), alpha=0.35, linetype=0) +
  theme_classic() +
  theme(legend.position="bottom") + 
  scale_x_discrete(breaks = c(1950, 1964, 1978, 1992, 2006, 2020))
plot

# cov <- matrix(c(states_long$fitted), 76, ncol = nrow(data))
# fit2.cor <- cor(cov)
# unique(states_long$state)
# fit2.cor
```

## Hypothesis 3
```{r H3, echo=FALSE}
#H3: Same as H2 but the errors are temporally correlated
mod.list.3 <- mod.list.2
mod.list.3$Q <- "equalvarcov"

if(!file.exists(here::here("Lab-2", "Team-2", "data", "fit3.rds"))){
fit.3 <- MARSS:: MARSS(data, model = mod.list.3)
  saveRDS(fit.3, file=here::here("Lab-2", "Team-2", "data", "fit3.rds"))
}
# load in fit.3
fit.3 <- readRDS(file=here::here("Lab-2", "Team-2", "data", "fit3.rds"))

# grabbing data for figures
states <- fit.3$states
statesSE <- fit.3$states.se

states <- as.data.frame(states)
statesSE <- as.data.frame(statesSE)

states_long <- states %>% 
  pivot_longer("V1":"V76", names_to="Year", values_to="fitted")
states_long <- states_long %>% 
  mutate(Year = rep(years, nrow(data)))

states_long <- states_long %>% 
  mutate(state = c(rep(pops[1], 76),
                   rep(pops[2], 76),
                   rep(pops[3], 76),
                   rep(pops[4], 76),
                   rep(pops[5], 76),
                   rep(pops[6], 76),
                   rep(pops[7], 76),
                   rep(pops[8], 76),
                   rep(pops[9], 76),
                   rep(pops[10], 76),
                   rep(pops[11], 76),
                   rep(pops[12], 76),
                   rep(pops[13], 76),
                   rep(pops[14], 76),
                   rep(pops[15], 76),
                   rep(pops[16], 76),
                   rep(pops[17], 76),
                   rep(pops[18], 76),
                   rep(pops[19], 76),
                   rep(pops[20], 76),
                   rep(pops[21], 76),
                   rep(pops[22], 76),
                   rep(pops[23], 76),
                   rep(pops[24], 76),
                   rep(pops[25], 76),
                   rep(pops[26], 76),
                   rep(pops[27], 76),
                   rep(pops[28], 76)
                   ))

statesSE_long <- statesSE %>% 
  pivot_longer("V1":"V76", names_to="Year", values_to="se")

states_long <- states_long %>% 
  mutate(se = statesSE_long$se)

states_long$lb <- states_long$fitted - states_long$se
states_long$ub <- states_long$fitted + states_long$se

# save long state estimates\
save(states_long, file=here::here("Lab-2", "Team-2", "data", "fit3_long.Rda"))

plot <- ggplot(data = states_long, aes(x = Year, y = fitted, group = state)) +
  geom_line() +
  labs(x = "", 
       y="State Estimate",
       title='Spring Summer Chinook Abundance, Upper Columbia/Snake River',
       subtitle="28 individual populations, equal process variance and covariance") +
  geom_ribbon(aes(ymin=lb, ymax=ub, fill=state), alpha=0.35, linetype=0) +
  theme_classic() +
  theme(legend.position="bottom") + 
  scale_x_discrete(breaks = c(1950, 1964, 1978, 1992, 2006, 2020))
plot

# cov <- matrix(c(states_long$fitted), 76, ncol = nrow(data))
# fit3.cor <- cor(cov)
# unique(states_long$state)
# fit3.cor
```

## Hypothesis 4a
```{r H4a, echo=FALSE}
#H4: Run timing is the only distinguishing factor, meaning that there are only 3 correlated populations (and are temporally correlated)

#3x28 Z Matrix for H4
Zmatrix4 <- matrix(c(
  # Row 1
  1,0,0,
  # Row 2
  1,0,0,
  # Row 3
  0,1,0,
  #Row 4
  0,1,0,
  #Row 5
  1,0,0,
  # Row 6
  1,0,0,
  #Row 7
  0,1,0,
  #Row 8
  0,0,1,
  #Row 9
  1,0,0,
  #Row 10
  0,1,0,
  #Row 11
  1,0,0,
  #Row 12
  0,1,0,
  #Row 13
  1,0,0,
  #Row 14
  0,1,0,
  #Row 15
  1,0,0,
  #Row 16
  1,0,0,
  #Row 17
  1,0,0,
  #Row 18
  0,0,1,
  #Row 19
  0,1,0,
  #Row 20
  1,0,0,
  #Row 21
  0,0,1,
  #Row 22
  0,0,1,
  #Row 23
  1,0,0,
  #Row 24
  1,0,0,
  #Row 25
  0,1,0,
  #Row 26
  1,0,0,
  #Row 27
  1,0,0,
  #Row 28
  1,0,0
), nrow = 28, byrow = TRUE)

# equal process variance and covariance
mod.list.4a <- list(B = "identity",  
                   U = matrix(paste0("u", seq(ncol(Zmatrix4)))),
                   Q = "equalvarcov",
                   Z = Zmatrix4,
                   A = "zero",
                   R = "diagonal and equal", #Observation variance
                   x0 = "unequal",
                   tinitx = 0)

if(!file.exists(here::here("Lab-2", "Team-2", "data", "fit4a.rds"))){
fit.4a <- MARSS::MARSS(data, model = mod.list.4a)
  saveRDS(fit.4a, file=here::here("Lab-2", "Team-2", "data", "fit4a.rds"))
}
# load in fit.4a
fit.4a <- readRDS(file=here::here("Lab-2", "Team-2", "data", "fit4a.rds"))

# grabbing data for figures
states <- fit.4a$states
statesSE <- fit.4a$states.se

states <- as.data.frame(states)
statesSE <- as.data.frame(statesSE)

states_long <- states %>% 
  pivot_longer("V1":"V76", names_to="Year", values_to="fitted")
states_long <- states_long %>% 
  mutate(Year = rep(years, ncol(Zmatrix4)))

states_long <- states_long %>% 
  mutate(state = c(rep("Spring", 76),
                   rep("Spring/summer", 76),
                   rep("Summer", 76)
  ))

statesSE_long <- statesSE %>% 
  pivot_longer("V1":"V76", names_to="Year", values_to="se")

states_long <- states_long %>% 
  mutate(se = statesSE_long$se)

states_long$lb <- states_long$fitted - states_long$se
states_long$ub <- states_long$fitted + states_long$se

# save long state estimates\
save(states_long, file=here::here("Lab-2", "Team-2", "data", "fit4a_long.Rda"))

plot <- ggplot(data = states_long, aes(x = Year, y = fitted, group = state)) +
  geom_line() +
  labs(x = "", 
       y="State Estimate",
       title='Spring Summer Chinook Abundance, Upper Columbia/Snake River',
       subtitle="3 runs, equal process variance and covariance") +
  geom_ribbon(aes(ymin=lb, ymax=ub, fill=state), alpha=0.35, linetype=0) +
  theme_classic() +
  theme(legend.position="bottom") + 
  scale_x_discrete(breaks = c(1950, 1964, 1978, 1992, 2006, 2020))
plot

cov <- matrix(c(states_long$fitted), 76, ncol = ncol(Zmatrix4))
fit4a.cor <- cor(cov)
unique(states_long$state)
fit4a.cor
```

## Hypothesis 4b
```{r H4b, echo=FALSE}
#H4b: Run timing is the only distinguishing factor, meaning that there are only 3 correlated populations (and are NOT temporally correlated)

# diagonal and unequal process variance 
mod.list.4b <- list(B = "identity",  
                   U = "equal",
                   Q = "diagonal and equal",
                   Z = Zmatrix4,
                   A = "zero",
                   R = "diagonal and equal", #Observation variance
                   x0 = "unequal",
                   tinitx = 0)

if(!file.exists(here::here("Lab-2", "Team-2", "data", "fit4b.rds"))){
fit.4b <- MARSS::MARSS(data, model = mod.list.4b)
  saveRDS(fit.4b, file=here::here("Lab-2", "Team-2", "data", "fit4b.rds"))
}
# load in fit.4b
fit.4b <- readRDS(file=here::here("Lab-2", "Team-2", "data", "fit4b.rds"))

# grabbing data for figures
states <- fit.4b$states
statesSE <- fit.4b$states.se

states <- as.data.frame(states)
statesSE <- as.data.frame(statesSE)

states_long <- states %>% 
  pivot_longer("V1":"V76", names_to="Year", values_to="fitted")
states_long <- states_long %>% 
  mutate(Year = rep(years, ncol(Zmatrix4)))

states_long <- states_long %>% 
  mutate(state = c(rep("Spring", 76),
                   rep("Spring/summer", 76),
                   rep("Summer", 76)
  ))

statesSE_long <- statesSE %>% 
  pivot_longer("V1":"V76", names_to="Year", values_to="se")

states_long <- states_long %>% 
  mutate(se = statesSE_long$se)

states_long$lb <- states_long$fitted - states_long$se
states_long$ub <- states_long$fitted + states_long$se

# save long state estimates\
save(states_long, file=here::here("Lab-2", "Team-2", "data", "fit4b_long.Rda"))

plot <- ggplot(data = states_long, aes(x = Year, y = fitted, group = state)) +
  geom_line() +
  labs(x = "", 
       y="State Estimate",
       title='Spring Summer Chinook Abundance, Upper Columbia/Snake River',
       subtitle="3 runs, equal process variance") +
  geom_ribbon(aes(ymin=lb, ymax=ub, fill=state), alpha=0.35, linetype=0) +
  theme_classic() +
  theme(legend.position="bottom") + 
  scale_x_discrete(breaks = c(1950, 1964, 1978, 1992, 2006, 2020))
plot

cov <- matrix(c(states_long$fitted), 76, ncol = ncol(Zmatrix4))
fit4b.cor <- cor(cov)
unique(states_long$state)
fit4b.cor
```

## Hypothesis 5a
```{r H5a, echo=FALSE}
#H5a: Major population group is distinguishing factors, meaning that there are 5 correlated populations (and are temporally correlated)

#5x28 Z matrix for H5
Zmatrix5 <- matrix(c(
  # Row 1
  0,1,0,0,0,
  # Row 2
  0,0,1,0,0,
  # Row 3
  0,0,1,0,0,
  #Row 4
  0,0,1,0,0,
  #Row 5
  1,0,0,0,0,
  # Row 6
  0,0,1,0,0,
  #Row 7
  0,0,0,0,1,
  #Row 8
  0,0,0,1,0,
  #Row 9
  1,0,0,0,0,
  #Row 10
  1,0,0,0,0,
  #Row 11
  0,0,0,0,1,
  #Row 12
  0,0,1,0,0,
  #Row 13
  0,0,1,0,0,
  #Row 14
  0,0,1,0,0,
  #Row 15
  0,0,1,0,0,
  #Row 16
  1,0,0,0,0,
  #Row 17
  0,0,0,0,1,
  #Row 18
  0,0,0,0,1,
  #Row 19
  0,0,0,0,1,
  #Row 20
  0,0,0,0,1,
  #Row 21
  0,0,0,1,0,
  #Row 22
  0,0,0,1,0,
  #Row 23
  0,0,1,0,0,
  #Row 24
  0,1,0,0,0,
  #Row 25
  0,0,0,0,1,
  #Row 26
  1,0,0,0,0,
  #Row 27
  1,0,0,0,0,
  #Row 28
  0,0,0,0,1
), nrow = 28, byrow = TRUE)

# equal process variance and covariance
mod.list.5a <- list(B = "identity",  
                   U = matrix(paste0("u", seq(ncol(Zmatrix5)))),
                   Q = "equalvarcov",
                   Z = Zmatrix5,
                   A = "zero",
                   R = "diagonal and equal", #Observation variance
                   x0 = "unequal",
                   tinitx = 0)

if(!file.exists(here::here("Lab-2", "Team-2", "data", "fit5a.rds"))){
fit.5a <- MARSS::MARSS(data, model = mod.list.5a)
  saveRDS(fit.5a, file=here::here("Lab-2", "Team-2", "data", "fit5a.rds"))
}
# load in fit.5a
fit.5a <- readRDS(file=here::here("Lab-2", "Team-2", "data", "fit5a.rds"))

# grabbing data for figures
states <- fit.5a$states
statesSE <- fit.5a$states.se

states <- as.data.frame(states)
statesSE <- as.data.frame(statesSE)

states_long <- states %>% 
  pivot_longer("V1":"V76", names_to="Year", values_to="fitted")
states_long <- states_long %>% 
  mutate(Year = rep(years, ncol(Zmatrix5)))

states_long <- states_long %>% 
  mutate(state = c(rep("lower snake", 76),
                   rep("middle fork snake", 76),
                   rep("gradne ronde", 76),
                   rep("upper salmon", 76),
                   rep("south fork salmon", 76)
  ))

statesSE_long <- statesSE %>% 
  pivot_longer("V1":"V76", names_to="Year", values_to="se")

states_long <- states_long %>% 
  mutate(se = statesSE_long$se)

states_long$lb <- states_long$fitted - states_long$se
states_long$ub <- states_long$fitted + states_long$se

# save long state estimates\
save(states_long, file=here::here("Lab-2", "Team-2", "data", "fit5a_long.Rda"))

plot <- ggplot(data = states_long, aes(x = Year, y = fitted, group = state)) +
  geom_line() +
  labs(x = "", 
       y="State Estimate",
       title='Spring Summer Chinook Abundance, Upper Columbia/Snake River',
       subtitle="5 major populations, equal process variance and covariance") +
  geom_ribbon(aes(ymin=lb, ymax=ub, fill=state), alpha=0.35, linetype=0) +
  theme_classic() +
  theme(legend.position="bottom") + 
  scale_x_discrete(breaks = c(1950, 1964, 1978, 1992, 2006, 2020))
plot

cov <- matrix(c(states_long$fitted), 76, ncol = ncol(Zmatrix5))
fit5a.cor <- cor(cov)
unique(states_long$state)
fit5a.cor
```

## Hypothesis 5b
```{r H5b, echo= FALSE}
#H5b: Major population group is distinguishing factors, meaning that there are 5 correlated populations (and are NOT temporally correlated)

# diagonal and unequal process variance 
mod.list.5b <- list(B = "identity",  
                   U = "equal",
                   Q = "diagonal and equal",
                   Z = Zmatrix5,
                   A = "zero",
                   R = "diagonal and equal", #Observation variance
                   x0 = "unequal",
                   tinitx = 0)

if(!file.exists(here::here("Lab-2", "Team-2", "data", "fit5b.rds"))){
fit.5b <- MARSS::MARSS(data, model = mod.list.5b)
  saveRDS(fit.5b, file=here::here("Lab-2", "Team-2", "data", "fit5b.rds"))
}
# load in fit.5b
fit.5b <- readRDS(file=here::here("Lab-2", "Team-2", "data", "fit5b.rds"))

# grabbing data for figures
states <- fit.5b$states
statesSE <- fit.5b$states.se

states <- as.data.frame(states)
statesSE <- as.data.frame(statesSE)

states_long <- states %>% 
  pivot_longer("V1":"V76", names_to="Year", values_to="fitted")
states_long <- states_long %>% 
  mutate(Year = rep(years, ncol(Zmatrix5)))

states_long <- states_long %>% 
  mutate(state = c(rep("lower snake", 76),
                   rep("middle fork snake", 76),
                   rep("gradne ronde", 76),
                   rep("upper salmon", 76),
                   rep("south fork salmon", 76)
  ))

statesSE_long <- statesSE %>% 
  pivot_longer("V1":"V76", names_to="Year", values_to="se")

states_long <- states_long %>% 
  mutate(se = statesSE_long$se)

states_long$lb <- states_long$fitted - states_long$se
states_long$ub <- states_long$fitted + states_long$se

# save long state estimates\
save(states_long, file=here::here("Lab-2", "Team-2", "data", "fit5b_long.Rda"))

plot <- ggplot(data = states_long, aes(x = Year, y = fitted, group = state)) +
  geom_line() +
  labs(x = "", 
       y="State Estimate",
       title='Spring Summer Chinook Abundance, Upper Columbia/Snake River',
       subtitle="5 major populations, equal process variance") +
  geom_ribbon(aes(ymin=lb, ymax=ub, fill=state), alpha=0.35, linetype=0) +
  theme_classic() +
  theme(legend.position="bottom") + 
  scale_x_discrete(breaks = c(1950, 1964, 1978, 1992, 2006, 2020))
plot

cov <- matrix(c(states_long$fitted), 76, ncol = ncol(Zmatrix5))
fit5b.cor <- cor(cov)
unique(states_long$state)
fit5b.cor
```

## Hypothesis 6a
```{r H6a, echo=FALSE}
#H6a: Major population group and run timing are the distinguishing factors, meaning that there are 9 correlated populations (and are temporally correlated)

#9x28 Z matrix
Zmatrix6 <- matrix(c(
  # Row 1
  0,0,1,0,0,0,0,0,0,
  # Row 2
  0,0,0,1,0,0,0,0,0,
  # Row 3
  0,0,0,0,1,0,0,0,0,
  #Row 4
  0,0,0,0,1,0,0,0,0,
  #Row 5
  1,0,0,0,0,0,0,0,0,
  # Row 6
  0,0,0,1,0,0,0,0,0,
  #Row 7
  0,0,0,0,0,0,0,0,1,
  #Row 8
  0,0,0,0,0,1,0,0,0,
  #Row 9
  1,0,0,0,0,0,0,0,0,
  #Row 10
  0,1,0,0,0,0,0,0,0,
  #Row 11
  0,0,0,0,0,0,1,0,0,
  #Row 12
  0,0,0,0,1,0,0,0,0,
  #Row 13
  0,0,0,1,0,0,0,0,0,
  #Row 14
  0,0,0,0,1,0,0,0,0,
  #Row 15
  0,0,0,1,0,0,0,0,0,
  #Row 16
  1,0,0,0,0,0,0,0,0,
  #Row 17
  0,0,0,0,0,0,1,0,0,
  #Row 18
  0,0,0,0,0,0,0,1,0,
  #Row 19
  0,0,0,0,0,0,0,0,1,
  #Row 20
  0,0,0,0,0,0,1,0,0,
  #Row 21
  0,0,0,0,0,0,0,1,0,
  #Row 22
  0,0,0,0,0,0,0,1,0,
  #Row 23
  0,0,0,1,0,0,0,0,0,
  #Row 24
  0,0,1,0,0,0,0,0,0,
  #Row 25
  0,0,0,0,0,0,0,0,1,
  #Row 26
  1,0,0,0,0,0,0,0,0,
  #Row 27
  1,0,0,0,0,0,0,0,0,
  #Row 28
  0,0,0,0,0,0,1,0,0
), nrow = 28, byrow = TRUE)

# equal process variance and covariance
mod.list.6a <- list(B = "identity",  
                   U = matrix(paste0("u", seq(ncol(Zmatrix6)))),
                   Q = "equalvarcov",
                   Z = Zmatrix6,
                   A = "zero",
                   R = "diagonal and equal", #Observation variance
                   x0 = "unequal",
                   tinitx = 0)

if(!file.exists(here::here("Lab-2", "Team-2", "data", "fit6a.rds"))){
fit.6a <- MARSS::MARSS(data, model = mod.list.6a)
  saveRDS(fit.6a, file=here::here("Lab-2", "Team-2", "data", "fit6a.rds"))
}
# load in fit.6a
fit.6a <- readRDS(file=here::here("Lab-2", "Team-2", "data", "fit6a.rds"))

# grabbing data for figures
states <- fit.6a$states
statesSE <- fit.6a$states.se

states <- as.data.frame(states)
statesSE <- as.data.frame(statesSE)

states_long <- states %>% 
  pivot_longer("V1":"V76", names_to="Year", values_to="fitted")
states_long <- states_long %>% 
  mutate(Year = rep(years, ncol(Zmatrix6)))

states_long <- states_long %>% 
  mutate(state = c(rep("Lower Snake, Spring", 76),
                   rep("Middle Fork Salmon, Spring", 76),
                   rep("Middle Fork Salmon, Spring/summer", 76),
                   rep("Grande Ronde/Imnaha, Spring", 76),
                   rep("Upper Salmon, Spring/summer", 76),
                   rep("South Fork Salmon, Summer", 76),
                   rep("Grande Ronde/Imnaha, Spring/summer", 76),
                   rep("Upper Salmon, Spring", 76),
                   rep("Upper Salmon, Summer", 76)
  ))

statesSE_long <- statesSE %>% 
  pivot_longer("V1":"V76", names_to="Year", values_to="se")

states_long <- states_long %>% 
  mutate(se = statesSE_long$se)

states_long$lb <- states_long$fitted - states_long$se
states_long$ub <- states_long$fitted + states_long$se

# save long state estimates\
save(states_long, file=here::here("Lab-2", "Team-2", "data", "fit6a_long.Rda"))

plot <- ggplot(data = states_long, aes(x = Year, y = fitted, group = state)) +
  geom_line() +
  labs(x = "", 
       y="State Estimate",
       title='Spring Summer Chinook Abundance, Upper Columbia/Snake River',
       subtitle="9 groups - 3 runs and 5 major populations, equal process variance and covariance") +
  geom_ribbon(aes(ymin=lb, ymax=ub, fill=state), alpha=0.35, linetype=0) +
  theme_classic() +
  theme(legend.position="bottom") + 
  scale_x_discrete(breaks = c(1950, 1964, 1978, 1992, 2006, 2020))
plot

cov <- matrix(c(states_long$fitted), 76, ncol = ncol(Zmatrix6))
fit6a.cor <- cor(cov)
unique(states_long$state)
fit6a.cor
```

## Hypothesis 6b
```{r H6b, echo = FALSE}
#H6b: Major population group and run timing are the distinguishing factors, meaning that there are 9 correlated populations (and are NOT temporally correlated)

# diagonal and unequal process variance 
mod.list.6b <- list(B = "identity",  
                   U = "equal",
                   Q = "diagonal and equal",
                   Z = Zmatrix6,
                   A = "zero",
                   R = "diagonal and equal", #Observation variance
                   x0 = "unequal",
                   tinitx = 0)

if(!file.exists(here::here("Lab-2", "Team-2", "data", "fit6b.rds"))){
fit.6b <- MARSS::MARSS(data, model = mod.list.6b)
  saveRDS(fit.6b, file=here::here("Lab-2", "Team-2", "data", "fit6b.rds"))
}
# load in fit.6b
fit.6b <- readRDS(file=here::here("Lab-2", "Team-2", "data", "fit6b.rds"))

# grabbing data for figures
states <- fit.6b$states
statesSE <- fit.6b$states.se

states <- as.data.frame(states)
statesSE <- as.data.frame(statesSE)

states_long <- states %>% 
  pivot_longer("V1":"V76", names_to="Year", values_to="fitted")
states_long <- states_long %>% 
  mutate(Year = rep(years, ncol(Zmatrix6)))

states_long <- states_long %>% 
  mutate(state = c(rep("Lower Snake, Spring", 76),
                   rep("Middle Fork Salmon, Spring", 76),
                   rep("Middle Fork Salmon, Spring/summer", 76),
                   rep("Grande Ronde/Imnaha, Spring", 76),
                   rep("Upper Salmon, Spring/summer", 76),
                   rep("South Fork Salmon, Summer", 76),
                   rep("Grande Ronde/Imnaha, Spring/summer", 76),
                   rep("Upper Salmon, Spring", 76),
                   rep("Upper Salmon, Summer", 76)
  ))

statesSE_long <- statesSE %>% 
  pivot_longer("V1":"V76", names_to="Year", values_to="se")

states_long <- states_long %>% 
  mutate(se = statesSE_long$se)

states_long$lb <- states_long$fitted - states_long$se
states_long$ub <- states_long$fitted + states_long$se

# save long state estimates\
save(states_long, file=here::here("Lab-2", "Team-2", "data", "fit6b_long.Rda"))

plot <- ggplot(data = states_long, aes(x = Year, y = fitted, group = state)) +
  geom_line() +
  labs(x = "", 
       y="State Estimate",
       title='Spring Summer Chinook Abundance, Upper Columbia/Snake River',
       subtitle="9 groups - 3 runs and 5 major populations, equal process variance") +
  geom_ribbon(aes(ymin=lb, ymax=ub, fill=state), alpha=0.35, linetype=0) +
  theme_classic() +
  theme(legend.position="bottom") + 
  scale_x_discrete(breaks = c(1950, 1964, 1978, 1992, 2006, 2020))
plot

cov <- matrix(c(states_long$fitted), 76, ncol = ncol(Zmatrix6))
fit6b.cor <- cor(cov)
unique(states_long$state)
fit6b.cor
```

## PDO
```{r pdo, echo = FALSE}
# # to add PDO to whichever models we want...
# c.model <- matrix("pdo", ncol(ZMATRIX), 1)
#   # add to model build  
# C = c.model, c = Wpdo,
#   # add to modlist
```
### Tips

**Simplify**

If your ESU has many populations, start with a smaller set of 4-7 populations.

**Assumptions**

You can assume that `R="diagonal and equal"` and `A="scaling"`. Assume that "historical" means the earliest years available for your group.

**States**

Your abundance estimate is the "x" or "state" estimates. You can get this from
```
fit$states
```
or 
```
tsSmooth(fit)
```
where `fit` is from `fit <- MARSS()`

**plotting**

Estimate of the mean of the spawner counts based on your x model.
```
autoplot(fit, plot.type="fitted.ytT")
```

**diagnostics**
```
autoplot(fit, plot.type="residuals")
```

# Results

# Discussion

# Description of each team member's contributions

Example: All team members met together twice to talk about goals and how to divide work. Brian designed the code and question for the first and third question and Emma designed and contributed code for the second question about population grouping. All members contributed to the discussion and writing of the report.

