---
title: Team 2 - Lab 2
subtitle: MARSS models
author: "Brian McGreal, Lilac ?, Emma Christman"
date: April 23, 2025
output: 
  html_document:
    code-folding: true
    toc: true
    toc_float: true
---

# Data
As team 2, we were assigned the Snake River Spring and Summer Run Chinook Data. This data has three seasons of runs: spring, spring/summer, and summer, divided among 5 major populations and when separated by run timing, major population and smaller tributary, 28 different populations.

Load the data.
```{r}
load(here::here("Lab-2", "Data_Images", "esa-salmon.rda"))
```

Wrangle the data.

**Add your wrangling code here**
```{r}
#Filter for just the "Salmon, Chinook (Snake River spring/summer-run ESU)"
esa.salmon <- esa.salmon %>% filter(esu_dps == "Salmon, Chinook (Snake River spring/summer-run ESU)")
unique(esa.salmon$majorpopgroup)
#Remove rows that have a "NaN" in the value column
esa.salmon <- esa.salmon %>% filter(value != "NaN")

#Setupa. matrix

# Create a matrix for each population, filling in missing years as NA
data_matrix <- esa.salmon %>%
  spread(key = spawningyear, value = value) %>%
  select(-esapopname) %>%
  as.matrix()

# Fill any missing values in the matrix (if any)
data_matrix[is.na(data_matrix)] <- NA
```

# General Questions

Each group has the same general tasks, but you will adapt them as you work on the data.

As a group, we've decided to dig into the estimates during the missing years, as outlined in the first question, as well as investigate the relationship between the different populations to see how they are/are not correlated. This comes from the second question, to which we are investigating not only the role of major population group, but run timing, and esa population name. This is allowing us to test several hypotheses:

H1: There is no difference between any of the groups (observing the same population different rivers and times)
H2: There is a difference between all of the separate esa populations, major populations, and run timing, meaning 28 different populations
H3: The same as H2, however there is temporal correlation 
H4: Run timing is the only distinguishing factor, meaning that there are only 3 correlated populations (and are temporally correlated)
H5: Major population groups is the only distinguishing factor, meaning that there are only 5 correlated populations 
H6: Major population group and run timing are the distinguishing factors, meaning that there are 9 correlated populations (and are temporally correlated).

1. Create estimates of spawner abundance for all missing years and provide estimates of the decline from the historical abundance.

2. Evaluate support for the major population groups. Are the populations in the groups more correlated than outside the groups?

3. Evaluate the evidence of cycling in the data. *We will talk about how to do this on the Tuesday after lab.*


# Methods and Code

Address the following in your methods

* Describe your assumptions about the x and how the data time series are related to x.

   - How are the x and y (data) related? 1 x for 1 y or will you assume 1 x for all y or 1 x for each major population group? How will you choose? 
   - What will you assume about the U for the x's?
   - What will you assume about the Q matrix?
   
* Write out your assumptions as different models **in matrix form**, fit each and then compare these with AIC or AICc.

* Do your estimates differ depending on the assumptions you make about the structure of the data, i.e. you assumptions about the x's, Q, and U.


#Code for Question One

#Code for Question Two
```{r}
#H1: There is no correlation between the populations
mod.list.1 <- list(B = "identity", 
                   U = matric ("u"),
                   Q = matrix("q"),
                   Z = matrix(1,28,1),
                   A = "scaling",
                   R = "diagonal and equal",
                   x0 = matrix("mu"),
                   tinitx = 0)
fit.1 <- MARSS(data, model = mod.list.1)

#H2: There is a correlation within the populations (28 different populations)
mod.list.2 <- list(B = "identity", 
                   U = "equal",
                   Q = "diagonal and equal",
                   Z = "identity",
                   A = "scaling",
                   R = "diagonal and equal",
                   x0 = matrix("mu"),
                   tinitx = 0)
fit.2 <- MARSS::MARSS(data, model = mod.list.2)

#H3: Same as H2 but the errors are temporally correlated
mod.list.3 <- mod.list.2
mod.list.3$Q <- "equalvarcov"
fit.3 <- MARSS:: MARSS(data, model = mod.list.3)

#H4: Run timing is the only distinguishing factor, meaning that there are only 3 correlated populations (and are temporally correlated)
#3x28 Z Matrix for H4
Zmatrix4 <- matrix(c(
  # Row 1
  1,0,0,
  # Row 2
  1,0,0,
  # Row 3
  0,1,0,
  #Row 4
  0,1,0,
  #Row 5
  1,0,0,
  # Row 6
  1,0,0,
  #Row 7
  0,1,0,
  #Row 8
  0,0,1,
  #Row 9
  1,0,0,
  #Row 10
  0,1,0,
  #Row 11
  1,0,0,
  #Row 12
  0,1,0,
  #Row 13
  1,0,0,
  #Row 14
  0,1,0,
  #Row 15
  1,0,0,
  #Row 16
  1,0,0,
  #Row 17
  1,0,0,
  #Row 18
  0,0,1,
  #Row 19
  0,1,0,
  #Row 20
  1,0,0,
  #Row 21
  0,0,1,
  #Row 22
  0,0,1,
  #Row 23
  1,0,0,
  #Row 24
  1,0,0,
  #Row 25
  0,1,0,
  #Row 26
  1,0,0,
  #Row 27
  1,0,0,
  #Row 28
  1,0,0
), nrow = 28, byrow = TRUE)

mod.list.4 <- list(B = "identity",  
                   U = "equal",
                   Q = "equalvarcov",
                   Z = ZMatrix4,
                   A = "scaling",
                   R = "diagonal and equal", #Observation variance
                   x0 = "unequal",
                   tinitx = 0)

fit.4 <- MARSS::MARSS(data, model = mod.list.4)

#H5 - Brian's code here -----
#5x28 Z matrix for H5
Zmatrix5 <- matrix(c(
  # Row 1
  0,1,0,0,0,
  # Row 2
  0,0,1,0,0,
  # Row 3
  0,0,1,0,0,
  #Row 4
  0,0,1,0,0,
  #Row 5
  1,0,0,0,0,
  # Row 6
  0,0,1,0,0,
  #Row 7
  0,0,0,0,1,
  #Row 8
  0,0,0,1,0,
  #Row 9
  1,0,0,0,0,
  #Row 10
  1,0,0,0,0,
  #Row 11
  0,0,0,0,1,
  #Row 12
  0,0,1,0,0,
  #Row 13
  0,0,1,0,0,
  #Row 14
  0,0,1,0,0,
  #Row 15
  0,0,1,0,0,
  #Row 16
  1,0,0,0,0,
  #Row 17
  0,0,0,0,1,
  #Row 18
  0,0,0,0,1,
  #Row 19
  0,0,0,0,1,
  #Row 20
  0,0,0,0,1,
  #Row 21
  0,0,0,1,0,
  #Row 22
  0,0,0,1,0,
  #Row 23
  0,0,1,0,0,
  #Row 24
  0,1,0,0,0,
  #Row 25
  0,0,0,0,1,
  #Row 26
  1,0,0,0,0,
  #Row 27
  1,0,0,0,0,
  #Row 28
  0,0,0,0,1
), nrow = 28, byrow = TRUE)
#H6: Major population group and run timing are the distinguishing factors, meaning that there are 9 correlated populations (and are temporally correlated)

#9x28 Z matrix
Zmatrix6 <- matrix(c(
  # Row 1
  0,0,1,0,0,0,0,0,0,
  # Row 2
  0,0,0,1,0,0,0,0,0,
  # Row 3
  0,0,0,0,1,0,0,0,0,
  #Row 4
  0,0,0,0,1,0,0,0,0,
  #Row 5
  1,0,0,0,0,0,0,0,0,
  # Row 6
  0,0,0,1,0,0,0,0,0,
  #Row 7
  0,0,0,0,0,0,0,0,1,
  #Row 8
  0,0,0,0,0,1,0,0,0,
  #Row 9
  1,0,0,0,0,0,0,0,0,
  #Row 10
  0,1,0,0,0,0,0,0,0,
  #Row 11
  0,0,0,0,0,0,1,0,0,
  #Row 12
  0,0,0,0,1,0,0,0,0,
  #Row 13
  0,0,0,1,0,0,0,0,0,
  #Row 14
  0,0,0,0,1,0,0,0,0,
  #Row 15
  0,0,0,1,0,0,0,0,0,
  #Row 16
  1,0,0,0,0,0,0,0,0,
  #Row 17
  0,0,0,0,0,0,1,0,0,
  #Row 18
  0,0,0,0,0,0,0,1,0,
  #Row 19
  0,0,0,0,0,0,0,0,1,
  #Row 20
  0,0,0,0,0,0,1,0,0,
  #Row 21
  0,0,0,0,0,0,0,1,0,
  #Row 22
  0,0,0,0,0,0,0,1,0,
  #Row 23
  0,0,0,1,0,0,0,0,0,
  #Row 24
  0,0,1,0,0,0,0,0,0,
  #Row 25
  0,0,0,0,0,0,0,0,1,
  #Row 26
  1,0,0,0,0,0,0,0,0,
  #Row 27
  1,0,0,0,0,0,0,0,0,
  #Row 28
  0,0,0,0,0,0,1,0,0
), nrow = 28, byrow = TRUE)

mod.list.6 <- list(B = "identity", 
                   U = "equal",
                   Q = "equalvarcov",
                   Z = Zmatrix6,
                   A = "scaling",
                   R = "diagonal and equal",
                   x0 = "unequal",
                   tinitx = 0)

fit.6 <- MARSS::MARSS(data, model = mod.list.6)

#Code for Question Three

```
### Tips

**Simplify**

If your ESU has many populations, start with a smaller set of 4-7 populations.

**Assumptions**

You can assume that `R="diagonal and equal"` and `A="scaling"`. Assume that "historical" means the earliest years available for your group.

**States**

Your abundance estimate is the "x" or "state" estimates. You can get this from
```
fit$states
```
or 
```
tsSmooth(fit)
```
where `fit` is from `fit <- MARSS()`

**plotting**

Estimate of the mean of the spawner counts based on your x model.
```
autoplot(fit, plot.type="fitted.ytT")
```

**diagnostics**
```
autoplot(fit, plot.type="residuals")
```

# Results

# Discussion

# Description of each team member's contributions

Example: All team members met together twice to talk about goals and how to divide work. Brian designed the code and question for the first and third question and Emma designed and contributed code for the second question about population grouping. All members contributed to the discussion and writing of the report.

