---
title: Team X - Lab 5
subtitle: Dynamic Linear Models (DLMs) - Time-varying productivity of salmon
author: "Name 1, Name 2, Name 3"
date: May XX, 2025
output: 
  html_document:
    code-folding: true
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
options(dplyr.summarise.inform = FALSE)
```

## Load the data

```{r dlm-load-atsa, eval=FALSE}
## library(devtools)
## Windows users will likely need to set this
## Sys.setenv("R_REMOTES_NO_ERRORS_FROM_WARNINGS" = "true")
#devtools::install_github("nwfsc-timeseries/atsalibrary")

## get data
data(KvichakSockeye, package="atsalibrary")
SR_data <- KvichakSockeye
```

## Wrangle the data

```{r wrangle_data}

library(dplyr)

# Remove NA's using dplyr
SR_clean <- SR_data %>%
  filter(!is.na(spawners) & !is.na(recruits)) %>%
  mutate(log_RS = log(recruits/spawners),
         brood_year = as.numeric(brood_year))


# Make matrices of the log recruits/spawners and log spawners 

logRS<- matrix(SR_clean$log_RS, nrow=1)
log_spawners<- matrix(log(SR_clean$spawners), nrow=1)


# Quick plot log recruits/spawners vs log spawners 
plot(logRS~log_spawners, xlab="log spawners", ylab="log recruits/spawners")


```

1. Begin by fitting a reduced form of Ricker's model that includes only a time-varying level ($\alpha_t$) and observation error ($v_t$). That is,

\begin{align*}
\text{log}(R_t) &= \alpha_t + \text{log}(S_t) + v_t \\
\text{log}(R_t/S_t) &= \alpha_t + v_t
\end{align*}

This model assumes no density-dependent survival in that the number of recruits is an ascending function of spawners. Plot the ts of $\alpha_t$ and note the AICc for this model. Also plot appropriate model diagnostics.

```{r}

# Model 1: Time-varying intercept only
# This is the simplest DLM - just one hidden state (α_t) that changes over time
# log(R_t/S_t) = α_t + v_t  (observation equation)
# α_t = α_{t-1} + w_t        (state equation - random walk)
# No covariates, no density dependence - just tracking how productivity drifts over time

library(MARSS)

mod.list <- list(
  B = matrix(1),      
  U = matrix(0),        
  Q = "unconstrained",  
  Z = matrix(1),        
  A = matrix(0),        
  R = "unconstrained",  
  x0 = "unconstrained"  
)

model1<-MARSS(logRS,mod.list)

# Extract results
print(model1)

# MARSS fit is
# Estimation method: kem 
# Convergence test: conv.test.slope.tol = 0.5, abstol = 0.001
# Estimation converged in 24 iterations. 
# Log-likelihood: -51.87179 
# AIC: 109.7436   AICc: 110.359   
#  
#       Estimate
# R.R      0.211
# Q.Q      0.303
# x0.x0    1.002
# Initial states (x0) defined at t=0

cat("AICc:", model1$AICc, "\n")

# Extract the time-varying states (α_t estimates)
alpha_t <- model1$states[1, ]
alpha_se <- model1$states.se[1, ]

# Create plots
par(mfrow = c(2, 2))

# 1. Time series of α_t 
plot(years, alpha_t, type = "l", lwd = 2,
     main = expression(paste("Time-varying Productivity (", alpha[t], ")")),
     xlab = "Brood Year", ylab = expression(alpha[t]),
     ylim = range(alpha_t + c(-2, 2) * alpha_se))

### ADD DIAGNOSTICS 


```
2. Fit the full Ricker model with a time-varying intercept and a static effect of spawners. For this model, obtain the time series of $\alpha_t$, which is an estimate of the stock productivity in the absence of density-dependent effects. How do these estimates of productivity compare to those from the previous question? Plot the ts of $\alpha_t$ and note the AICc for this model. Also plot appropriate model diagnostics. ($Hint$: If you don't want a parameter to vary with time, what does that say about its process variance?)

```{r}
# Model 2: Full Ricker with time-varying intercept + static density dependence
# This adds the effect of spawner density (negative effect on productivity)
# log(R_t/S_t) = α_t - β*S_t + v_t  (observation equation)
# α_t = α_{t-1} + w_t                (state equation - α varies over time)
# β = constant                       (density effect is static - no process variance)

# key difference is adding spawners as a covariate, meaning the relationship between year and log(R/S) can vary by density 

spawners_cov <- matrix(SR_clean$spawners, nrow = 1) # need spawners as a covariate
# should be logged 
log_spawners_cov <- matrix(log(SR_clean$spawners), nrow = 1) # need spawners as a covariate

mod.list.2 <- list(
  B = matrix(1),        # alpha_t = alpha_{t-1} + w_t
  U = matrix(0),
  Q = "unconstrained",  # alpha_t has process error
  Z = matrix(1),        # observation = alpha_t + beta*covariate + v_t
  A = matrix(0),
  R = "unconstrained",
  x0 = "unconstrained",
  D = matrix(1),        # coefficient of covariate beta (use D not C)
  d = log_spawners_cov  # covariate matrix of logged spawner counts (use d not c)
)

model2 <- MARSS(logRS, mod.list.2)  # Also fix the model list name here

```




years <- SR_clean$brood_year #years, x variable 
TT <- length(years)
dat <- matrix(SR_clean$log_RS, nrow = 1) # Response variable




```

```{r}
library(MARSS)
#### EXAMPLE FROM TEXT 
data(SalmonSurvCUI)
years <- SalmonSurvCUI[, 1]
TT <- length(years)
# response data: logit(survival)
dat <- matrix(SalmonSurvCUI[, 2], nrow = 1)



#### In our data 
# our response variable is log recruits/spawner 

# for each brood year, calculate log recruits/spawner 







## add some code here
```


# General tasks

Use the information and data in the previous section to answer the following questions. Note that if any model is not converging, then you will need to increase the `maxit` parameter in the `control` argument/list that gets passed to `MARSS()`. For example, you might try `control=list(maxit=2000)`.

1. Begin by fitting a reduced form of Ricker's model that includes only a time-varying level ($\alpha_t$) and observation error ($v_t$). That is,

\begin{align*}
\text{log}(R_t) &= \alpha_t + \text{log}(S_t) + v_t \\
\text{log}(R_t/S_t) &= \alpha_t + v_t
\end{align*}

This model assumes no density-dependent survival in that the number of recruits is an ascending function of spawners. Plot the ts of $\alpha_t$ and note the AICc for this model. Also plot appropriate model diagnostics.

2. Fit the full Ricker model with a time-varying intercept and a static effect of spawners. For this model, obtain the time series of $\alpha_t$, which is an estimate of the stock productivity in the absence of density-dependent effects. How do these estimates of productivity compare to those from the previous question? Plot the ts of $\alpha_t$ and note the AICc for this model. Also plot appropriate model diagnostics. ($Hint$: If you don't want a parameter to vary with time, what does that say about its process variance?)

3. Fit the expanded model that includes the summer PDO index as the covariate (`pdo_summer_t2`). What is the mean level of productivity? Plot the ts of $\delta_t$ and note the AICc for this model. Also plot appropriate model diagnostics.

4. Fit the expanded model that includes the winter PDO index as the covariate (`pdo_winter_t2`). What is the mean level of productivity? Plot the ts of $\delta_t$ and note the AICc for this model. Also plot appropriate model diagnostics.

5. Based on AICc, which of the models above is the most parsimonious? Is it well behaved (i.e., are the model assumptions met)? Plot the model forecasts for the best model. Is this a good forecast model?

6. Consider other environmental factors that could influence the time-varying intrinsic productivity of these salmon. For example, are there other large-scale indices of ocean conditions (e.g., sea-surface temperature, El Niño Southern Oscillation)? You could also consider the possible influence of other salmon stocks such as those we examined in Lab 1.



# Methods

Please address the following in your methods:

* What environmental or dummy variables did you include and why?

* What forms of models did you fit (ie, write them out in matrix form)?

* What sort of model diagnostics did you use to examine model assumptions?


# Results


# Discussion


# Team contributions

Example: "All team members helped decide on the goal and ran the analyses for the individual regions. Team members 2 & 3 wrote most of the code for the analysis. Team member 1 worked on the plotting section of the report using and adapting code that team member 3 wrote. All team members helped edit the report and wrote the discussion together."

