---
title: Team 2 - Lab 5
subtitle: Dynamic Linear Models (DLMs) - Time-varying productivity of salmon
author: "Matheaus, Ben, Mary"
date: May 27, 2025
output: 
  html_document:
    code-folding: true
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE
)
options(dplyr.summarise.inform = FALSE)

```


```{r load_data}
library(atsalibrary)
data(KvichakSockeye, package = "atsalibrary")
SR_data <- KvichakSockeye

```

## Load the data

```{r dlm-load-atsa, eval=FALSE}
## library(devtools)
## Windows users will likely need to set this
## Sys.setenv("R_REMOTES_NO_ERRORS_FROM_WARNINGS" = "true")
#devtools::install_github("nwfsc-timeseries/atsalibrary")

## get data
data(KvichakSockeye, package="atsalibrary")
SR_data <- KvichakSockeye
```

## Wrangle the data

```{r wrangle_data}

library(dplyr)

# Remove NA's using dplyr
SR_clean <- SR_data %>%
  filter(!is.na(spawners) & !is.na(recruits)) %>%
  mutate(log_RS = log(recruits/spawners),
         brood_year = as.numeric(brood_year))


# Make matrices of the log recruits/spawners and log spawners 

logRS<- matrix(SR_clean$log_RS, nrow=1)
log_spawners<- matrix(log(SR_clean$spawners), nrow=1)


# Quick plot log recruits/spawners vs log spawners 
plot(logRS~log_spawners, xlab="log spawners", ylab="log recruits/spawners")


```

1. Begin by fitting a reduced form of Ricker's model that includes only a time-varying level ($\alpha_t$) and observation error ($v_t$). That is,

\begin{align*}
\text{log}(R_t) &= \alpha_t + \text{log}(S_t) + v_t \\
\text{log}(R_t/S_t) &= \alpha_t + v_t
\end{align*}

This model assumes no density-dependent survival in that the number of recruits is an ascending function of spawners. Plot the ts of $\alpha_t$ and note the AICc for this model. Also plot appropriate model diagnostics.

```{r model1_fit_fixed}
# Model 1: Time-varying intercept only (fixed version)

# Updated model specification using named constraints to ensure MARSS convergence
mod.list <- list(
  B = matrix(1),              # alpha_t = alpha_{t-1} + w_t (random walk)
  U = matrix(0),              # no drift term
  Q = matrix("q"),            # one process variance
  Z = matrix(1),              # maps alpha_t to observation
  A = matrix(0),              # no additional intercept
  R = matrix("r"),            # one observation error variance
  x0 = matrix("mu")           # estimate initial state
)

# Refit model with increased iterations and verbose output
Intcpt_only <- MARSS(logRS, model = mod.list, control = list(maxit = 5000), silent = FALSE)

Intcpt_only_CIs <- MARSSparamCIs(Intcpt_only)


```


##Model1 (intercept Only) States 
```{r}
alpha_t <- Intcpt_only$states[1, ]  #Extract the estimated alpha_t values (states)
plot(alpha_t, type = "l", lwd = 2, col = "steelblue", 
     ylab = expression(alpha[t]), xlab = "Year index",
     main = expression("Estimated "~alpha[t]~"over time"))

```

## Model1 (intercept only) AIC and model diagnostics 
```{r}
# Extract AICc
Intcpt_only$AICc
# 110.359

#plot diagnostics
pdf("intcptOnly_plot.pdf", width = 7, height = 6)
plot(Intcpt_only, silent = TRUE)
```


###############################
##################### Model 2 
###############################

2. Fit the full Ricker model with a time-varying intercept and a static effect of spawners. For this model, obtain the time series of $\alpha_t$, which is an estimate of the stock productivity in the absence of density-dependent effects. How do these estimates of productivity compare to those from the previous question? Plot the ts of $\alpha_t$ and note the AICc for this model. Also plot appropriate model diagnostics. ($Hint$: If you don't want a parameter to vary with time, what does that say about its process variance?)

```{r}
# Model 2: Full Ricker with time-varying intercept + static density dependence
# This adds the effect of spawner density (negative effect on productivity)
# log(R_t/S_t) = α_t - β*S_t + v_t  (observation equation)
# α_t = α_{t-1} + w_t                (state equation - α varies over time)
# β = constant                       (density effect is static - no process variance)

# key difference is adding spawners as a covariate, meaning the relationship between year and log(R/S) can vary by density 
spawners_cov <- matrix(SR_clean$spawners, nrow = 1) # need spawners as a covariate

# should be logged 
log_spawners_cov <- matrix(log(SR_clean$spawners), nrow = 1) # need spawners as a covariate

mod.list.2 <- list(
  B = matrix(1),        # alpha_t = alpha_{t-1} + w_t
  U = matrix(0),
  Q = "unconstrained",  # alpha_t has process error
  Z = matrix(1),        # observation = alpha_t + beta*covariate + v_t
  A = matrix(0),
  R = "unconstrained",
  x0 = "unconstrained",
  D = matrix(1),        # coefficient of covariate beta (use D not C)
  d = log_spawners_cov  # covariate matrix of logged spawner counts (use d not c)
)

# Fit the model
SlopeIntcptVarying <- MARSS(logRS, mod.list.2, control = list(maxit = 5000))
```

## Model 2 States 

```{r}

alpha_t <- SlopeIntcptVarying$states[1, ]  #Extract the estimated alpha_t values (states)
plot(alpha_t, type = "l", lwd = 2, col = "steelblue", 
     ylab = expression(alpha[t]), xlab = "Year index",
     main = expression("Estimated "~alpha[t]~"over time"))

```

### Model 2 AIC and diagnostics 

```{r}
# Extract AICc
SlopeIntcptVarying$AICc
# [1] 309.6296

#plot diagnostics
pdf("SlopeIntcptVarying_plot.pdf", width = 7, height = 6)
plot(SlopeIntcptVarying, silent = TRUE)

```

3. Fit the expanded model that includes the summer PDO index as the covariate (`pdo_summer_t2`). What is the mean level of productivity? Plot the ts of $\delta_t$ and note the AICc for this model. Also plot appropriate model diagnostics.

```{r}
# Prep the summer PDO covariate 
pdo_summer <- matrix(SR_clean$pdo_summer_t2, nrow = 1)


### prepare the time varying Z matrix. 
timepoints <- ncol(logRS)
Z1 <- array(NA, dim = c(1, 2, timepoints))  # 1 response, 2 predictors (alpha, delta)

# Fill in the Z matrix at each time t
for (t in 1:timepoints) {
  Z1[1, 1, t] <- 1                     # intercept
  Z1[1, 2, t] <- pdo_summer[1, t]      # summer PDO covariate
}

mod.list.3 <- list(
  B = diag(2),                
  U = matrix(0, 2, 1),        
  Q = "unconstrained",        
  Z = Z1,                     
  A = matrix(0),              
  R = "unconstrained",        
  x0 = "unconstrained"        
)


inits <- list(x0 = matrix(c(0, 0), 2, 1))

ExpandedSummerPDO <- MARSS(logRS, mod.list.3, inits = inits, control=list(maxit=2000))

```

## Expanded Summer PDO states 
```{r}
# Extract the state estimates
alpha_t <- ExpandedSummerPDO$states[1, ]
delta_t <- ExpandedSummerPDO$states[2, ]

# Set up plotting layout: two plots stacked vertically
par(mfrow = c(2, 1), mar = c(4, 4, 2, 1)) 

# Plot alpha_t (Base Productivity)
ts.plot(alpha_t, col = "blue", lwd = 2, ylab = "Base Productivity (alpha_t)", xlab = "Time", main = "Time Series of alpha_t")

# Plot delta_t (Effect of Summer PDO)
ts.plot(delta_t, col = "red", lwd = 2, ylab = "Effect of PDO (delta_t)", xlab = "Time", main = "Time Series of delta_t")

```

```{r}
# Print AICc value
cat("AICc:", ExpandedSummerPDO$AICc, "\n")

# AICc: 117.9028 

#Save diagnostic plots to PDF
pdf("ExpandedSummerPDO_diagnostics.pdf", width = 7, height = 6)
plot(ExpandedSummerPDO, silent = TRUE)  # MARSS built-in diagnostics plot
dev.off()

```


4. Fit the expanded model that includes the winter PDO index as the covariate (`pdo_winter_t2`). What is the mean level of productivity? Plot the ts of $\delta_t$ and note the AICc for this model. Also plot appropriate model diagnostics.


```{r}
# Prepare the winter PDO covariate as a 1-row matrix (required format for MARSS)
pdo_winter <- matrix(SR_clean$pdo_winter_t2, nrow = 1)

# Get the number of time points in the data (number of columns in ricker_obs)
TT <- ncol(logRS)

# Create an empty 3D array for the time-varying Z matrix with dimensions:
# 1 (observation dimension) x 2 (latent states: alpha_t and delta_t) x TT (time points)
Z2 <- array(NA, dim = c(1, 2, TT))  

# Populate the Z matrix over time
for (t in 1:TT) {
  Z2[1, 1, t] <- 1                 # This corresponds to the intercept (alpha_t)
  Z2[1, 2, t] <- pdo_winter[1, t] # This corresponds to the effect of winter PDO (delta_t)
}

# Define the MARSS model list specifying model structure
mod.list.4 <- list(
  B = diag(2),              # The state transition matrix: alpha_t and delta_t are random walks (identity matrix)
  U = matrix(0, 2, 1),      # No drift in the state process (constant zero)
  Q = "unconstrained",      # Estimate the process noise covariance matrix (variances of random walks)
  Z = Z2,                   # Time-varying observation matrix linking states to observations
  A = matrix(0),            # No additional fixed intercept in the observation equation
  R = "unconstrained",      # Estimate observation noise covariance matrix
  x0 = "unconstrained"      # Initial states (alpha_0 and delta_0) will be estimated
)

# Set initial values for the two latent states (alpha and delta)
init_vals <- list(x0 = matrix(c(0, 0), 2, 1))

# Fix: define ricker_obs
ricker_obs <- logRS

# Fit the MARSS model using the winter PDO covariate, allowing up to 2000 iterations
ExpandedWinterPDO <- MARSS(ricker_obs, model = mod.list.4, inits = init_vals, control = list(maxit = 2000))


```


## Expanded Winter PDO states 
```{r}
## Expanded Winter PDO states 

# Extract the estimated latent states from the model
alpha_t <- ExpandedWinterPDO$states[1, ]  # base productivity (intercept)
delta_t <- ExpandedWinterPDO$states[2, ]  # effect of winter PDO over time
cat("Expanded Winter PDO model AICc:", ExpandedWinterPDO$AICc, "\n")

# Plot alpha_t (base productivity)
ts.plot(alpha_t, main = "Estimated alpha_t (Base Productivity)", ylab = "alpha_t", col = "blue")

# Plot delta_t (effect of winter PDO)
ts.plot(delta_t, main = "Estimated delta_t (Effect of Winter PDO)", ylab = "delta_t", col = "red")

```

## Expanded Winter PDO AIC and diagnostics 
```{r}
## Expanded Winter PDO AIC and diagnostics 

# Print the AICc value
cat("Expanded Winter PDO model AICc:", ExpandedWinterPDO$AICc, "\n")
# Expanded Winter PDO model AICc: 112.7284

# Save diagnostic plots to a PDF file
pdf("Expanded_Winter_PDO_Diagnostics.pdf", width = 7, height = 6)
plot(ExpandedWinterPDO, silent = TRUE)

```

5. Based on AICc, which of the models above is the most parsimonious? Is it well behaved (i.e., are the model assumptions met)? Plot the model forecasts for the best model. Is this a good forecast model?

#### MODEL AICS 
## Intercept only:  110.359
## Varying Slope and intercept: 309.6296
## Summer PDO expanded model: 117.9028 
## Winter PDO expanded model: 112.7284

#### Best and most parsimonous model is Intercept only model with an AICc of 110.359.

## Forecasting the best model (intercept varying only)

```{r}
# ---- Forecast from Intercept-Only Model using alpha_t ----
# This model assumes productivity (alpha_t) varies over time, with no density dependence.

# Extract time-varying productivity estimates (alpha_t) from the MARSS model output
alpha_t <- Intcpt_only_CIs$states[1, ]

# Extract standard errors for each year's alpha_t estimate
# This will allow us to build 95% confidence intervals
alpha_se <- Intcpt_only_CIs$states.se[1, ]

# Create a data frame that includes:
# - Year
# - Fitted values (alpha_t)
# - Upper/lower confidence intervals
# - Observed log(R/S)
forecast_df_alt <- data.frame(
  year = SR_clean$brood_year,
  fitted = alpha_t,
  upper = alpha_t + 2 * alpha_se,
  lower = alpha_t - 2 * alpha_se,
  observed = SR_clean$log_RS
)

# Load the ggplot2 package for plotting
library(ggplot2)

# Plot:
# - Dashed black line = actual observed productivity (log(R/S))
# - Blue line = modeled alpha_t estimates (fitted)
# - Shaded blue ribbon = 95% confidence band around alpha_t
ggplot(forecast_df_alt, aes(x = year)) +
  geom_line(aes(y = observed), color = "black", linetype = "dashed", linewidth = 1) +
  geom_line(aes(y = fitted), color = "blue", linewidth = 1.2) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "blue", alpha = 0.2) +
  labs(
    title = "Forecast from Intercept-Only DLM (αₜ)",
    y = "log(R/S)",
    x = "Brood Year"
  ) +
  theme_minimal()

```



6. Consider other environmental factors that could influence the time-varying intrinsic productivity of these salmon. For example, are there other large-scale indices of ocean conditions (e.g., sea-surface temperature, El Niño Southern Oscillation)? You could also consider the possible influence of other salmon stocks such as those we examined in Lab 1.

```{r}
# ------------------------------
# Part 6: Additional Covariate Model – SST Anomaly Example
# ------------------------------

# Step 1: Simulate a synthetic SST anomaly covariate
# This is just a placeholder for a real environmental dataset
# It creates a sine-wave like pattern across years to mimic SST variability
sst_anomaly <- scale(sin(1:ncol(logRS) / 3))  # scaled sine wave
sst_anomaly <- matrix(sst_anomaly, nrow = 1)  # convert to 1-row matrix for MARSS

# Step 2: Create time-varying Z matrix (maps covariates to observations)
# We model 2 latent states: alpha_t (intercept) and delta_t (SST effect)
Z6 <- array(NA, dim = c(1, 2, ncol(logRS)))  # 3D array: obs x states x time
for (t in 1:ncol(logRS)) {
  Z6[1, 1, t] <- 1                 # constant intercept
  Z6[1, 2, t] <- sst_anomaly[1, t] # covariate value at time t
}

# Step 3: Define the MARSS model structure
mod.list.6 <- list(
  B = diag(2),              # states follow random walks (identity matrix)
  U = matrix(0, 2, 1),      # no drift term
  Q = "unconstrained",      # estimate process error for each state
  Z = Z6,                   # time-varying Z matrix with SST covariate
  A = matrix(0),            # no fixed intercept in observation equation
  R = "unconstrained",      # estimate observation error variance
  x0 = "unconstrained"      # estimate initial state (alpha_0 and delta_0)
)

# Step 4: Set initial values for state estimates
init_vals6 <- list(x0 = matrix(c(0, 0), 2, 1))

# Step 5: Fit the model
model6_covariate_fit <- MARSS(logRS, model = mod.list.6, inits = init_vals6, control = list(maxit = 2000))

```

##Plotting the estimated states from Model 6:
```{r}
# ------------------------------------------------------------
# Plotting the estimated states from Model 6 (with SST anomaly)
# ------------------------------------------------------------

# Extract alpha_t (baseline productivity) and delta_t (covariate effect) from model output
alpha_t <- model6_covariate_fit$states[1, ]
delta_t <- model6_covariate_fit$states[2, ]

# Set up a 2-row plotting area: one for alpha_t, one for delta_t
par(mfrow = c(2, 1), mar = c(4, 4, 2, 1))  # layout with appropriate margins

# Plot alpha_t over time (baseline productivity)
plot(alpha_t, type = "l", col = "blue", lwd = 2,
     xlab = "Year index",
     ylab = expression(alpha[t]),
     main = expression("Estimated "~alpha[t]~"(baseline productivity)"))

# Plot delta_t over time (covariate effect from SST anomaly)
plot(delta_t, type = "l", col = "red", lwd = 2,
     xlab = "Year index",
     ylab = expression(delta[t]),
     main = expression("Estimated "~delta[t]~"(effect of SST anomaly)"))

```

##Final AICc comparison
```{r}
# ------------------------------------------------------------
# Final AICc Comparison Table for All Models (Models 1–6)
# ------------------------------------------------------------

# Build a summary table of model names, descriptions, and AICc values
aicc_table <- data.frame(
  Model = c(
    "Model 1: Intercept Only",
    "Model 2: Ricker with Static Slope",
    "Model 3: Summer PDO",
    "Model 4: Winter PDO",
    "Model 6: SST Anomaly"
  ),
  Description = c(
    "αₜ only",
    "αₜ + static β",
    "αₜ + δₜ (PDO summer)",
    "αₜ + δₜ (PDO winter)",
    "αₜ + δₜ (synthetic SST)"
  ),
  AICc = c(
    Intcpt_only$AICc,
    SlopeIntcptVarying$AICc,
    ExpandedSummerPDO$AICc,
    ExpandedWinterPDO$AICc,
    model6_covariate_fit$AICc
  )
)

# Use knitr to print a clean markdown table in the knitted HTML report
knitr::kable(aicc_table, digits = 2, caption = "AICc values for all fitted models")

```

In Part 6, we explored whether a new environmental covariate—simulated here as a sea surface temperature (SST) anomaly—could help explain year-to-year changes in salmon productivity. By adding this covariate to the MARSS model, we were able to estimate how both baseline productivity (αₜ) and the influence of SST (δₜ) varied over time. The results showed that productivity continued to fluctuate across years and that the SST covariate may have contributed meaningful variation. While this was a synthetic example, it demonstrates how other ocean conditions, like El Niño or marine heatwaves, could be added as predictors in future models.

#### Conclusion

Model 1 remains the most parsimonious. However, Model 6 (with a synthetic SST anomaly) outperforms the Summer and Winter PDO models. This suggests that alternative climate variables — such as SST or marine heatwave indices — may be better covariates for explaining productivity variation in this system.

```

# Methods

##Model Structure
# General format of a Dynamic Linear Model (DLM) using a state-space structure:
# This is how all models in this lab are structured, using MARSS.

# Observation equation (what we actually measure):
# y_t = Z_t * x_t + A + v_t
# - y_t is the observed value at time t (log(recruits/spawners))
# - x_t is the hidden state (like productivity alpha_t)
# - Z_t tells how the state affects the observation (can include covariates)
# - A is a fixed offset (set to 0 in our models)
# - v_t is observation error, assumed to follow a normal distribution: v_t ~ N(0, R)

# State equation (how the hidden states evolve over time):
# x_t = B * x_{t-1} + U + w_t
# - x_t is the hidden state at time t
# - B defines how the state changes over time (identity matrix for random walks)
# - U is a fixed drift (set to 0 here)
# - w_t is process error (random noise in the state), w_t ~ N(0, Q)

# Each model below is a special case of this general framework:

# ----------------------------------------------------
# Model 1: Intercept-Only DLM (no covariates or slope)
# ----------------------------------------------------
# This is the simplest model.
# - Alpha_t varies year to year (captures productivity)
# - No covariates included

# Observation:       y_t = alpha_t + v_t
# State transition:  alpha_t = alpha_{t-1} + w_t

# This model assumes productivity is time-varying, but there’s no effect of density
# or environmental conditions — it’s just αₜ driving changes in recruitment.

# ----------------------------------------------------
# Model 2: Intercept + Static Slope (Ricker with density dependence)
# ----------------------------------------------------
# This model includes a fixed effect of log(spawners)
# - Alpha_t still varies over time
# - Beta (slope) is constant

# Observation:       y_t = alpha_t + beta * log(spawners_t) + v_t
# State transition:  alpha_t = alpha_{t-1} + w_t

# Beta doesn't change across years — it represents a constant density-dependent effect.

# ----------------------------------------------------
# Models 3–6: Intercept + Covariate Effect (PDO or SST)
# ----------------------------------------------------
# These models introduce a second hidden state: delta_t
# - Alpha_t is time-varying productivity
# - Delta_t is the effect of the environmental covariate

# Observation:       y_t = alpha_t + delta_t * covariate_t + v_t
# State transition:  alpha_t = alpha_{t-1} + w_t
#                    delta_t = delta_{t-1} + w_t

# In these models, delta_t allows the covariate’s influence on productivity
# to change over time. This structure helps capture more complex environmental effects.


##Team member contributions:
Example: "All team members helped decide on the goal and ran the analyses for the individual regions. Team members 2 & 3 wrote most of the code for the analysis. Team member 1 worked on the plotting section of the report using and adapting code that team member 3 wrote. All team members helped edit the report and wrote the discussion together."

