---
title: "Team 2 - Lab 5"
subtitle: "Dynamic Linear Models (DLMs) - Time-varying productivity of salmon"
author: "Mary, Ben, Magnus"
date: "May 2025"
output: 
  html_document:
    code-folding: true
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
options(dplyr.summarise.inform = FALSE)
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# Data

We analyzed the Kvichak Sockeye salmon dataset from the `atsalibrary` package. This dataset includes yearly observations of brood-year salmon spawner abundance and resulting recruits. Additionally, it contains environmental covariates representing ocean conditions—the Pacific Decadal Oscillation (PDO) indices—from two years prior to each brood year. Specifically, `pdo_summer_t2` and `pdo_winter_t2` reflect summer and winter PDO values, respectively, lagged to account for early life-stage exposure of the salmon to ocean conditions.

## Load the data

```{r dlm-read-data}
library(atsalibrary)
data(KvichakSockeye, package="atsalibrary")
SR_data <- KvichakSockeye
```

```{r dlm-data-head}
head(SR_data)
```

## Wrangle the data

```{r wrangle_data}
library(dplyr)

SR_clean <- SR_data %>%
  filter(!is.na(spawners) & !is.na(recruits)) %>%
  mutate(
    log_RS = log(recruits / spawners),
    brood_year = as.numeric(brood_year),
    logR = log(recruits),
    logS = log(spawners)
  )

plot(SR_clean$log_RS ~ SR_clean$spawners,
     xlab = "Number of spawners",
     ylab = "log [Recruits/Spawners]")
```

# General Tasks

## Question 1 – Reduced Ricker Model (time-varying alpha)

```{r model1_reduced_ricker}
library(MARSS)
years <- SR_clean$brood_year
logRS <- matrix(SR_clean$log_RS, nrow = 1)

mod1 <- MARSS(
  logRS,
  model = list(
    B = matrix(1),
    U = matrix(0),
    Q = matrix("q"),
    Z = matrix(1),
    A = matrix(0),
    R = matrix("r"),
    x0 = matrix("mu")
  ),
  control = list(maxit = 5000)
)

mod1$AICc
plot(mod1)
```

## Question 2 – Full Ricker Model (time-varying alpha + static beta)

```{r model2_full_ricker}
logR <- matrix(SR_clean$logR, nrow = 1)
logS <- matrix(SR_clean$logS, nrow = 1)

mod2 <- MARSS(
  logR,
  model = list(
    B = matrix(1),
    U = matrix(0),
    Q = matrix("q"),
    Z = matrix(1),
    A = matrix(0),
    R = matrix("r"),
    x0 = matrix(0),
    D = matrix("b", 1, 1),
    d = logS
  ),
  control = list(maxit = 5000)
)

mod2$AICc
plot(mod2)
```

## Question 3 – Add Summer PDO as Covariate

```{r model3_summer_pdo}
X <- matrix(SR_clean$pdo_summer_t2, nrow = 1)
St <- matrix(SR_clean$spawners, nrow = 1)
covariates <- rbind(X, log(St))

mod3 <- MARSS(
  logRS,
  model = list(
    B = matrix(1),
    U = matrix(0),
    Q = matrix("q.alpha"),
    Z = matrix(1),
    A = matrix(0),
    R = matrix("r"),
    x0 = matrix(0),
    D = matrix(c("beta1", "-beta2"), nrow = 1),
    d = covariates
  )
)

mod3CIs <- MARSSparamCIs(mod3)
mod3$AICc

library(ggplot2)
data.frame(
  Parameter = c("beta[1]", "beta[2]"),
  Mean = mod3CIs$parMean[1:2],
  SE = c(mod3CIs$par.se$A[1,1], mod3CIs$par.se$A[2,1])
) %>%
  ggplot(aes(x = Parameter, y = Mean)) +
  geom_point() +
  geom_errorbar(aes(ymin = Mean - SE, ymax = Mean + SE), width = 0.2) +
  coord_flip() +
  theme_minimal() +
  ggtitle("Fixed parameter estimates")
```

## Question 4 – Add Winter PDO as Covariate

```{r model4_winter_pdo}
X <- matrix(SR_clean$pdo_winter_t2, nrow = 1)
covariates <- rbind(X, log(St))

mod4 <- MARSS(
  logRS,
  model = list(
    B = matrix(1),
    U = matrix(0),
    Q = matrix("q.alpha"),
    Z = matrix(1),
    A = matrix(0),
    R = matrix("r"),
    x0 = matrix(0),
    D = matrix(c("beta1", "-beta2"), nrow = 1),
    d = covariates
  )
)

mod4CIs <- MARSSparamCIs(mod4)
mod4$AICc

data.frame(
  Parameter = c("beta[1]", "beta[2]"),
  Mean = mod4CIs$parMean[1:2],
  SE = c(mod4CIs$par.se$A[1,1], mod4CIs$par.se$A[2,1])
) %>%
  ggplot(aes(x = Parameter, y = Mean)) +
  geom_point() +
  geom_errorbar(aes(ymin = Mean - SE, ymax = Mean + SE), width = 0.2) +
  coord_flip() +
  theme_minimal() +
  ggtitle("Fixed parameter estimates")
```

---

# ✅ Mary picks up from here

You can now begin with Question 5: comparing AICc values, checking model diagnostics, and plotting forecasts for the best-fitting model. Let me know if you'd like help scripting that. Also, the methods, results, and discussion sections can now be filled based on the models above.
