---
title: "Team 1 - Lab 1"
author: "Nicole, Brian, Anna"
date: "April 21, 2025"
output:
  pdf_document:
    toc: yes
  html_document:
    code-folding: yes
    toc: yes
    toc_float: yes
subtitle: Forecasting with ARIMA and ETS models
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

\newpage
# Data

This lab used annual data recording salmon landings in six different US states covering the period from 1950 to 2016: Alaska, California, Oregon, Michigan, Pennsylvania, and Washington. Due to sparse data, Pennsylvania and Michigan were excluded from the analysis. Monthly data recording salmon landings as well as the value of fish landed was used from Washington, Oregon, and California covering the period from 1990 to 2022. 

# Research Questions

1. What model structure best fits the different states and regions (Total and Pacific Northwest) represented in the yearly data set? How does forecasting accuracy compare across these regions?

2. How does model performance change when sections of the time series are dropped? 

3. How does forecasting monthly value of catch compare with separate forecasts of monthly landings and prices being interacted to find overall value?

# Methods

We initially planned to fit ARIMA and ETS models to the different states and compare which model structure best fit each state or region. Noting the similarities between Washington and Oregon’s time series data, we also wanted to compare forecast accuracy for models trained on different data sets (e.g. compare Washington’s forecast accuracy on Oregon’s test data). We also planned on recovering monthly prices by dividing the value of fish landed by the landed weight in order to assess the efficiency of forecasting approaches.

## Question 1
To address question 1, we subset the data for each west coast state, Pacific Northwest which includes WA, OR, and CA, and the total of all states combined, into a training data set and test data set. The training data sets spanned the years 1950-2005, and the test data set was 2006-2016. When we plotted the data there were no obvious problems with stationarity. There might be a slight downwards trend, and there are a few gaps in the California data, but it is not immediately obvious if the data is non-stationary.


```{r, include=FALSE}
## Packages
library(atsalibrary)
library(dplyr)
library(ggplot2)
library(patchwork)
library(tidyverse)
library(zoo)
par(mfrow=c(1,1))
#Filter and plot data
data(chinook)
chinook.year<-chinook.year %>%
  filter(!State %in% c("Michigan", "Pennsylvania"))

## Subset test and training data for each region
# Washington
dat <- chinook.year %>%
  filter(State == "Washington") %>%
  mutate(lnreturns = log.metric.tons,
         year = Year) %>%
  select(year, lnreturns)
wats <- ts(dat$lnreturns, start=dat$year[1])
watrain <- window(wats, dat$year[1], dat$year[1]+55)
watest <- window(wats, dat$year[1]+56, dat$year[1]+66)
testlength<-length(watest)
# Oregon
dat <- chinook.year %>%
  filter(State == "Oregon") %>%
  mutate(lnreturns = log.metric.tons,
         year = Year) %>%
  select(year, lnreturns)
orts <- ts(dat$lnreturns, start=dat$year[1])
ortrain <- window(orts, dat$year[1], dat$year[1]+56)
ortest <- window(orts, dat$year[1]+56, dat$year[1]+66)
# California
dat <- chinook.year %>%
  filter(State == "California") %>%
  mutate(lnreturns = log.metric.tons,
         year = Year) %>%
  select(year, lnreturns)
cats <- ts(dat$lnreturns, start=dat$year[1])
catrain <- window(cats, dat$year[1], dat$year[1]+55)
catest <- window(cats, dat$year[1]+56, dat$year[1]+66)
# PNW Region
chinook.pnw.total <- chinook.year %>%
  filter(!State == "Alaska")%>%
  mutate(metric.tons = exp(log.metric.tons)) %>%  # undo the log
  group_by(Year) %>%
  summarise(
    total.metric.tons = sum(metric.tons, na.rm = TRUE),
    log.total.metric.tons = log(total.metric.tons)
  )
dat <- chinook.pnw.total %>%
  mutate(lnreturns = log.total.metric.tons,
         year = Year) %>%
  select(year, lnreturns)
pnwts <- ts(dat$lnreturns, start=dat$year[1])
pnwtrain <- window(pnwts, dat$year[1], dat$year[1]+55)
pnwtest <- window(pnwts, dat$year[1]+56, dat$year[1]+66)
# Alaska
dat <- chinook.year %>%
  filter(State=="Alaska") %>%
  mutate(lnreturns = log.metric.tons,
         year = Year) %>%
  select(year, lnreturns)
akts <- ts(dat$lnreturns, start=dat$year[1])
aktrain <- window(akts, dat$year[1], dat$year[1]+55)
aktest <- window(akts, dat$year[1]+56, dat$year[1]+66)

chinook.total <- chinook.year %>%
  filter(State%in%c("California", "Alaska", "Oregon", "Washington")) %>%
  mutate(metric.tons = exp(log.metric.tons)) %>%  # undo the log
  group_by(Year) %>%
  summarise(
    total.metric.tons = sum(metric.tons, na.rm = TRUE),
    log.total.metric.tons = log(total.metric.tons)
  )
dat <- chinook.total %>%
  mutate(lnreturns = log.total.metric.tons,
         year = Year) %>%
  select(year, lnreturns)
totalts<-ts(dat$lnreturns, start=dat$year[1])
totaltrain <- window(totalts, dat$year[1], dat$year[1]+55)
totaltest <- window(totalts, dat$year[1]+56, dat$year[1]+66)

```


```{r, echo=FALSE}
ggplot(chinook.year, aes(x = Year, y = log.metric.tons, color = State)) +
  geom_line() +
  labs(
    title = "Chinook Landings by State",
    x = "Year",
    y = "Log(Metric Tons)"
  ) +
  theme_minimal()
```


#### Use ACF and PACF

```{r, echo=FALSE}
par(mfrow=c(2,2))
acf(catrain, main= "California ACF")
acf(aktrain, main="Alaska ACF")
acf(watrain, main="Washington ACF")
acf(ortrain,main="Oregon ACF")

pacf(catrain, main= "California pACF")
pacf(aktrain, main="Alaska pACF")
pacf(watrain, main="Washington pACF")
pacf(ortrain,main="Oregon pACF")

acf(pnwtrain, main= "PNW ACF")
acf(totaltrain, main="Total ACF")
pacf(pnwtrain, main= "PNW pACF")
pacf(totaltrain, main="Total pACF")
par(mfrow=c(1,1))
```

#### Test for stationarity

When we run ndiffs on the states’ data, Oregon, Washington, the Pacific Northwest region determined one first-difference is required to make the data stationary.

```{r, echo=FALSE}
akndiffs<-ndiffs(aktrain)
candiffs<-ndiffs(catrain)
orndiffs<-ndiffs(ortrain)
wandiffs<-ndiffs(watrain)
totalndiffs<-ndiffs(totaltrain)
pnwndiffs<-ndiffs(pnwtrain)

ndiffsresults<-matrix(data=c(akndiffs, candiffs, orndiffs, wandiffs, totalndiffs, pnwndiffs), ncol=6, nrow=1)
colnames(ndiffsresults)<-c("Alaksa", "California", "Oregon", "Washington", "Total", "PNW")
print(ndiffsresults)
```

## Question 2

#### Yearly Data

```{r, echo=FALSE}

chinook.year.filtered <- chinook.year %>%
  filter(!State %in% c("Michigan", "Pennsylvania"))

ggplot(chinook.year.filtered, aes(x = Year, y = metric.tons)) +
  geom_line() +
  facet_wrap(~ State, ncol = 2, scales = "free_y") +
  labs(
    title = "Chinook Landings by State",
    x = "Year",
    y = "Log(Metric Tons)"
  ) +
  theme_minimal()


# Create the time series 
dat_CA <- ts(chinook.year.filtered$log.metric.tons[chinook.year.filtered$State == "California"], start = chinook.year.filtered$Year[chinook.year.filtered$State == "California"][1], )
dat_WA <-ts(chinook.year.filtered$log.metric.tons[chinook.year.filtered$State == "Washington"], start = chinook.year.filtered$Year[chinook.year.filtered$State == "Washington"][1], )
dat_AK <-ts(chinook.year.filtered$log.metric.tons[chinook.year.filtered$State == "Alaska"], start = chinook.year.filtered$Year[chinook.year.filtered$State == "Alaska"][1], )
dat_OR <- ts(chinook.year.filtered$log.metric.tons[chinook.year.filtered$State == "Oregon"], start = chinook.year.filtered$Year[chinook.year.filtered$State == "Oregon"][1], )


```

Check the type of relationship we are dealing with to formulate research questions. We checked the ACF and PACF using a custom function.

```{r, echo=FALSE}
check_acf <- function(ts_list, area_names) {
  # ts_list: a list of 4 time series (e.g., list(ts_CA, ts_OR, ts_WA, ts_AK))
  # area_names: character vector of 4 names (e.g., c("California", "Oregon", "Washington", "Alaska"))
  
  if (length(ts_list) != 4 | length(area_names) != 4) {
    stop("Provide exactly 4 time series and 4 area names.")
  }
  
  par(mfrow = c(4, 2), mar = c(3, 3, 3, 1))  # 4 rows, 2 columns layout
  
  for (i in 1:4) {
    ts <- na.interp(ts_list[[i]])
    
    # ACF plot
    acf(ts, main = paste("ACF -", area_names[i]))
    
    # PACF plot
    pacf(ts, main = paste("PACF -", area_names[i]))
  }
  
  # Reset layout to default
  par(mfrow = c(1,1))
}


list_area <- list(dat_AK,dat_CA,dat_OR,dat_WA)
area_names <- c("California", "Oregon", "Washington", "Alaska")

check_acf(ts_list = list_area,area_names)
```

We also checked the differencing necessary for each series.

```{r, echo=FALSE}

ndiffs(dat_CA, test='adf')
ndiffs(dat_AK, test='adf') #diff 1
ndiffs(dat_WA, test='adf') #diff 1
ndiffs(dat_OR, test='adf')
```

#### Monthly Data

```{r, echo=FALSE}
ch_month <- chinook.month %>%
  filter(!State %in% c("Michigan", "Pennsylvania"))


```

## Question 3
Preliminary analysis of question 3 involved first recovering price data and then plotting monthly landings, prices, and total value for all three states represented. 

```{r, echo=FALSE}
# create date column (first of each month)
chinook.month <- chinook.month %>%
  mutate(Date = as.Date(paste(Year, Month, "01", sep = "-"), format = "%Y-%b-%d"))

# back out price (US$/ton)
chinook.month$price <- chinook.month$value.usd/chinook.month$metric.tons

# plot with price by state
landings.monthly <- ggplot(chinook.month, aes(x = Date, y = log.metric.tons)) +
  geom_line() +
  facet_wrap(~ State, nrow = 1) +  # 3 panels vertically
  labs(title = "Monthly Chinook Salmon Landings",
       x = "Date",
       y = "Log(m. tons)") +
  theme_minimal()

chinook.month[sapply(chinook.month, is.infinite)] <- NA
chinook.month <- chinook.month %>% mutate_all(~ifelse(is.nan(.), NA, .))
price.monthly <- ggplot(chinook.month, aes(x = Date, y = price)) +
  geom_line() +
  facet_wrap(~ State, nrow = 1) +  # 3 panels horizontal
  labs(title = "Monthly Chinook Salmon Price",
       x = "Date",
       y = "US$/m. ton") +
  theme_minimal()

value.monthly <- ggplot(chinook.month, aes(x = Date, y = value.usd)) +
  geom_line() +
  facet_wrap(~ State, nrow = 1) +  # 3 panels horizontal
  labs(title = "Monthly Chinook Salmon Value",
       x = "Date",
       y = "US$") +
  theme_minimal()

landings.monthly/price.monthly/value.monthly
```

Next, diagnostics were performed on these data, evaluating ACFs and pACFs checking ndiffs for stationarity.

```{r, echo=FALSE}
# set time series
CAchinook.month <- subset(chinook.month, State == "CA")
ORchinook.month <- subset(chinook.month, State == "OR")
WAchinook.month <- subset(chinook.month, State == "WA")

#CA
CAlands.ts <- ts(CAchinook.month$log.metric.tons, start=c(1990,1),
                 frequency = 12)
CAlands.train <- window(CAlands.ts, c(1990, 1), c(2020, 12))
CAlands.test <- window(CAlands.ts, c(2021, 1), c(2022, 12))

CAprice.ts <- ts(CAchinook.month$price, start=c(1990,1),
                 frequency = 12)
CAprice.train <- window(CAprice.ts, c(1990, 1), c(2020, 12))
CAprice.test <- window(CAprice.ts, c(2021, 1), c(2022, 12))

CAvalue.ts <- ts(CAchinook.month$value.usd, start=c(1990,1),
                 frequency = 12)
CAvalue.train <- window(CAvalue.ts, c(1990, 1), c(2020, 12))
CAvalue.test <- window(CAvalue.ts, c(2021, 1), c(2022, 12))

#OR
ORlands.ts <- ts(ORchinook.month$log.metric.tons, start=c(1990,1),
                 frequency = 12)
ORlands.train <- window(ORlands.ts, c(1990, 1), c(2020, 12))
ORlands.test <- window(ORlands.ts, c(2021, 1), c(2022, 12))

ORprice.ts <- ts(ORchinook.month$price, start=c(1990,1),
                 frequency = 12)
ORprice.train <- window(ORprice.ts, c(1990, 1), c(2020, 12))
ORprice.test <- window(ORprice.ts, c(2021, 1), c(2022, 12))

ORvalue.ts <- ts(ORchinook.month$value.usd, start=c(1990,1),
                 frequency = 12)
ORvalue.train <- window(ORvalue.ts, c(1990, 1), c(2020, 12))
ORvalue.test <- window(ORvalue.ts, c(2021, 1), c(2022, 12))

#WA
WAlands.ts <- ts(WAchinook.month$log.metric.tons, start=c(1990,1),
                 frequency = 12)
WAlands.train <- window(WAlands.ts, c(1990, 1), c(2020, 12))
WAlands.test <- window(WAlands.ts, c(2021, 1), c(2022, 12))

WAprice.ts <- ts(WAchinook.month$price, start=c(1990,1),
                 frequency = 12)
WAprice.train <- window(WAprice.ts, c(1990, 1), c(2020, 12))
WAprice.test <- window(WAprice.ts, c(2021, 1), c(2022, 12))

WAvalue.ts <- ts(WAchinook.month$value.usd, start=c(1990,1),
                 frequency = 12)
WAvalue.train <- window(WAvalue.ts, c(1990, 1), c(2020, 12))
WAvalue.test <- window(WAvalue.ts, c(2021, 1), c(2022, 12))
```

```{r, echo=FALSE}
## Differencing and stationarity
CAlandings <- ndiffs(CAlands.train)
CAprices <- ndiffs(CAprice.train)
CAvalue <- ndiffs(CAvalue.train)

ORlandings <- ndiffs(ORlands.train)
ORprices <- ndiffs(ORprice.train)
ORvalue <- ndiffs(ORvalue.train)

WAlandings <- ndiffs(WAlands.train)
WAprices <- ndiffs(WAprice.train)
WAvalue <- ndiffs(WAvalue.train)

ndiffsresults<-matrix(data=c(CAlandings, CAprices, CAvalue, ORlandings, ORprices, ORvalue, WAlandings, WAprices, WAvalue), ncol=3, nrow=3)
colnames(ndiffsresults) <- c("Landings", "Prices", "Value")
rownames(ndiffsresults) <- c("CA", "OR", "WA")
print(ndiffsresults)
```


```{r, echo=FALSE}
## ACF plots
par(mfrow=c(3,3))
CAl_acf <- acf(CAlands.train, main= "California Landings ACF", na.action = na.pass)
CAp_acf <- acf(CAprice.train, main= "California Prices ACF", na.action = na.pass)
CAv_acf <- acf(CAvalue.train, main= "California Values ACF", na.action = na.pass)

ORl_acf <- acf(ORlands.train, main= "Oregon Landings ACF", na.action = na.pass)
ORp_acf <- acf(ORprice.train, main= "Oregon Prices ACF", na.action = na.pass)
ORv_acf <- acf(ORvalue.train, main= "Oregon Values ACF", na.action = na.pass)

WAl_acf <- acf(WAlands.train, main= "Washington Landings ACF", na.action = na.pass)
WAp_acf <- acf(WAprice.train, main= "Washington Prices ACF", na.action = na.pass)
WAv_acf <- acf(WAvalue.train, main= "Washington Values ACF", na.action = na.pass)

#PACF
CAl_pacf <- pacf(CAlands.train, main= "California Landings pACF", na.action = na.pass)
CAp_pacf <- pacf(CAprice.train, main= "California Prices pACF", na.action = na.pass)
CAv_pacf <- pacf(CAvalue.train, main= "California Values pACF", na.action = na.pass)

ORl_pacf <- pacf(ORlands.train, main= "Oregon Landings pACF", na.action = na.pass)
ORp_pacf <- pacf(ORprice.train, main= "Oregon Prices pACF", na.action = na.pass)
ORv_pacf <- pacf(ORvalue.train, main= "Oregon Values pACF", na.action = na.pass)

WAl_pacf <- pacf(WAlands.train, main= "Washington Landings pACF", na.action = na.pass)
WAp_pacf <- pacf(WAprice.train, main= "Washington Prices pACF", na.action = na.pass)
WAv_pacf <- pacf(WAvalue.train, main= "Washington Values pACF", na.action = na.pass)
par(mfrow=c(1,1))
```

There is clear evidence of seasonal trends in the above plots, meaning any useful model will more than likely involve a seasonal component (sARIMA).

Finally, data from each of three state-level time series was subsetted into training and testing data sets. Training data spanned the years from 1990 through 2020 while test data covers the years 2021 & 2022. 

# Results

## Question 1

We fit multiple ARIMA and Holt models to the yearly time series data for each state, the PNW region (WA, OR, CA), and the total chinook landings for all states combined. ARIMA models where p, d, or q were greater than 1 had significantly higher AIC values than more parsimonious ARIMA models and are excluded from this write-up. We then forecasted 10 years of data using the six ARIMA models, tested the forecasts against the test data for each region. Using root mean squared error, we selected the best fit model for each state or region. The Holt model was the best fit for Alaska, California, and the Pacific Northwest region. An ARIMA(1,0,0) model was the best fit for Washington and Oregon. An ARIMA(0,1,0) model was the best fit for the entire Pacific Coast chinook landing data (log metric tons). See the figures below for the six regional time series and forecasts from the best fit model for each. We will also provide a brief interpretation of the best fit models, and check of the residuals using a Ljung-Box test. 


```{r, echo=FALSE}
auto.arima(aktrain) # returned (1,0 0)
akts.arima.0<-arima(aktrain, order=c(0,0,0))
akts.arima.1<-arima(aktrain, order=c(0,1,0))
akts.arima.2<-arima(aktrain, order=c(1,0,0)) #auto-arima model
akts.arima.3<-arima(aktrain, order=c(1,1,0))
akts.arima.4<-arima(aktrain, order=c(1,1,1))
akts.holt<-ets(aktrain, model="AAN")

# California
auto.arima(catrain) # returned (0,1,0)
cats.arima.0<-arima(catrain, order=c(0,0,0))
cats.arima.1<-arima(catrain, order=c(0,1,0))#auto-arima model
cats.arima.2<-arima(catrain, order=c(1,0,0))
cats.arima.3<-arima(catrain, order=c(1,1,0))
cats.arima.4<-arima(catrain, order=c(1,1,1))
cats.holt<-ets(catrain, model="AAN")

# Oregon 
auto.arima(ortrain) # returned (0,1,0)
orts.arima.0<-arima(ortrain, order=c(0,0,0))
orts.arima.1<-arima(ortrain, order=c(0,1,0))#auto-arima model
orts.arima.2<-arima(ortrain, order=c(1,0,0)) 
orts.arima.3<-arima(ortrain, order=c(1,1,0))
orts.arima.4<-arima(ortrain, order=c(1,1,1))
orts.holt<-ets(ortrain, model="AAN")

# Washington
auto.arima(watrain) # returned (0,1,0)
wats.arima.0<-arima(watrain, order=c(0,0,0))
wats.arima.1<-arima(watrain, order=c(0,1,0))#auto-arima model
wats.arima.2<-arima(watrain, order=c(1,0,0)) 
wats.arima.3<-arima(watrain, order=c(1,1,0))
wats.arima.4<-arima(watrain, order=c(1,1,1))
wats.holt<-ets(watrain, model="AAN")

# PNW
auto.arima(pnwtrain) # returned (0,1,0)
pnwts.arima.0<-arima(pnwtrain, order=c(0,0,0))
pnwts.arima.1<-arima(pnwtrain, order=c(0,1,0)) #auto-arima model
pnwts.arima.2<-arima(pnwtrain, order=c(1,0,0)) 
pnwts.arima.3<-arima(pnwtrain, order=c(1,1,0))
pnwts.arima.4<-arima(pnwtrain, order=c(1,1,1))
pnwts.holt<-ets(pnwtrain, model="AAN")

# Total
# Total
auto.arima(totaltrain)
totalts.arima.0<-arima(totaltrain, order=c(0,0,0))
totalts.arima.1<-arima(totaltrain, order=c(0,1,0)) #auto-arima model
totalts.arima.2<-arima(totaltrain, order=c(1,0,0)) 
totalts.arima.3<-arima(totaltrain, order=c(1,1,0))
totalts.arima.4<-arima(totaltrain, order=c(1,1,1))
totalts.holt<-ets(totaltrain, model="AAN")

## Forecasting for each model at each location
# Alaska
ak.forecast.0<-forecast(akts.arima.0, h=testlength)
ak.forecast.1<-forecast(akts.arima.1, h=testlength)
ak.forecast.2<-forecast(akts.arima.2, h=testlength)
ak.forecast.3<-forecast(akts.arima.3, h=testlength)
ak.forecast.4<-forecast(akts.arima.4, h=testlength)
ak.forecast.5<-forecast(akts.holt, h=testlength) #Holt model

# California
ca.forecast.0<-forecast(cats.arima.0, h=testlength)
ca.forecast.1<-forecast(cats.arima.1, h=testlength)
ca.forecast.2<-forecast(cats.arima.2, h=testlength)
ca.forecast.3<-forecast(cats.arima.3, h=testlength)
ca.forecast.4<-forecast(cats.arima.4, h=testlength)
ca.forecast.5<-forecast(cats.holt, h=testlength)

# Oregon
or.forecast.0<-forecast(orts.arima.0, h=testlength)
or.forecast.1<-forecast(orts.arima.1, h=testlength)
or.forecast.2<-forecast(orts.arima.2, h=testlength)
or.forecast.3<-forecast(orts.arima.3, h=testlength)
or.forecast.4<-forecast(orts.arima.4, h=testlength)
or.forecast.5<-forecast(orts.holt, h=testlength)

#Washington
wa.forecast.0<-forecast(wats.arima.0, h=testlength)
wa.forecast.1<-forecast(wats.arima.1, h=testlength)
wa.forecast.2<-forecast(wats.arima.2, h=testlength)
wa.forecast.3<-forecast(wats.arima.3, h=testlength)
wa.forecast.4<-forecast(wats.arima.4, h=testlength)
wa.forecast.5<-forecast(wats.holt, h=testlength)

# PNW
pnw.forecast.0<-forecast(pnwts.arima.0, h=testlength)
pnw.forecast.1<-forecast(pnwts.arima.1, h=testlength)
pnw.forecast.2<-forecast(pnwts.arima.2, h=testlength)
pnw.forecast.3<-forecast(pnwts.arima.3, h=testlength)
pnw.forecast.4<-forecast(pnwts.arima.4, h=testlength)
pnw.forecast.5<-forecast(pnwts.holt, h=testlength)

# Total
total.forecast.0<-forecast(totalts.arima.0, h=testlength)
total.forecast.1<-forecast(totalts.arima.1, h=testlength)
total.forecast.2<-forecast(totalts.arima.2, h=testlength)
total.forecast.3<-forecast(totalts.arima.3, h=testlength)
total.forecast.4<-forecast(totalts.arima.4, h=testlength)
total.forecast.5<-forecast(totalts.holt, h=testlength)
```

#### Alaska
The best fit model for the Alaska chinook landing data was a Holt model with the following equation:
$y_t=l_{t-1}+b_{t-1}+\epsilon_t$\
$l_t=l_{t-1}+b_{t-1}+0.9999*\epsilon_t$\
$b_t =b_{t-1} + 0.0001*\epsilon_t$ 
This model indicates a slight downward trend that remains fairly consistent over time (beta=0.0001), and gives almost all weight to the most recent observations (alpha=0.9999).

```{r, echo=FALSE}
aktest.0<-accuracy(ak.forecast.0, aktest)
aktest.1<-accuracy(ak.forecast.1, aktest)
aktest.2<-accuracy(ak.forecast.2, aktest)
aktest.3<-accuracy(ak.forecast.3, aktest)
aktest.4<-accuracy(ak.forecast.4, aktest)
aktest.5<-accuracy(ak.forecast.5, aktest)
ak.forecast.summary<-matrix(data=c(aktest.0[2, "RMSE"],aktest.1[2,"RMSE"], aktest.2[2,"RMSE"],
                                   aktest.3[2,"RMSE"], aktest.4[2, "RMSE"],
                                   aktest.5[2, "RMSE"]), nrow=1, ncol=6)
colnames(ak.forecast.summary)<-c("0,0,0", "0,1,0", "1,0,0", "1,1,0", "1,1,1", "Holt")
rownames(ak.forecast.summary)<-c("RMSE")
print(ak.forecast.summary) #The Holt model had lowest RMSE
checkresiduals(akts.holt)
```

#### California
The best fit model for California was also a Holt model with the following equation:
$y_t=l_{t-1}+b_{t-1}+\epsilon_t$
$l_t=l_{t-1}+b_{t-1}+0.0005*\epsilon_t$
$b_t=b_{t-1}+0.0005*\epsilon_t$ 

The model indicates a slightly declining trend, but small beta value (beta=0.0005) suggest the level and trend are relatively stable over time. The model also puts very little weight on recent observations (alpha=0.0005).

```{r, echo=FALSE}
# California
catest.0<-accuracy(ca.forecast.0, catest)
catest.1<-accuracy(ca.forecast.1, catest)
catest.2<-accuracy(ca.forecast.2, catest)
catest.3<-accuracy(ca.forecast.3, catest)
catest.4<-accuracy(ca.forecast.4, catest)
catest.5<-accuracy(ca.forecast.5, catest)
ca.forecast.summary<-matrix(data=c(catest.0[2,"RMSE"],catest.1[2,"RMSE"], catest.2[2,"RMSE"],
                                   catest.3[2,"RMSE"], catest.4[2, "RMSE"],
                                   catest.5[2, "RMSE"]), nrow=1, ncol=6)
colnames(ca.forecast.summary)<-c("0,0,0","0,1,0", "1,0,0", "1,1,0", "1,1,1", "Holt")
rownames(ca.forecast.summary)<-c("RMSE")
print(ca.forecast.summary) #The Holt had the most accurate forecast
checkresiduals(cats.holt)
```
#### Oregon
The best fit model for Oregon was a first-order ARIMA model with no differencing and no moving average. The time series is stationary and auto-correlated around a mean of 7.5190, with observations being moderately dependent on recent values (ar1<1).
$y_t=0.7303*y_{t-1}+2.026+\epsilon_t$
$\epsilon_t~N(0,\sigma^2=0.1876)$

```{r, echo=FALSE}
ortest.0<-accuracy(or.forecast.0, ortest)
ortest.1<-accuracy(or.forecast.1, ortest)
ortest.2<-accuracy(or.forecast.2, ortest)
ortest.3<-accuracy(or.forecast.3, ortest)
ortest.4<-accuracy(or.forecast.4, ortest)
ortest.5<-accuracy(or.forecast.5, ortest)

or.forecast.summary<-matrix(data=c(ortest.0[2,"RMSE"], ortest.1[2,"RMSE"], ortest.2[2,"RMSE"],
                                   ortest.3[2,"RMSE"], ortest.4[2, "RMSE"],
                                   ortest.5[2, "RMSE"]), nrow=1, ncol=6)
colnames(or.forecast.summary)<-c("0,0,0","0,1,0", "1,0,0", "1,1,0", "1,1,1", "Holt")
rownames(or.forecast.summary)<-c("RMSE")
print(or.forecast.summary) #The ARIMA(1,0,0) model had the best forecast
checkresiduals(orts.arima.2)
```

#### Washington
Like Oregon, the best fit model for Washington was a first-order ARIMA model with no differencing and not moving average. The values differ slightly, with Washington having a slightly higher average than Oregon (intercept=7.9066), and autoregression component (ar1) of 0.8878, meaning the series shows strong positive correlation.

$y_t=0.8878*y_{t-1}+0.8871+\epsilon_t$
$\epsilon_t~N(0,\sigma^2=0.0436)$
```{r, echo=FALSE}
watest.0<-accuracy(wa.forecast.0, watest)
watest.1<-accuracy(wa.forecast.1, watest)
watest.2<-accuracy(wa.forecast.2, watest)
watest.3<-accuracy(wa.forecast.3, watest)
watest.4<-accuracy(wa.forecast.4, watest)
watest.5<-accuracy(wa.forecast.5, watest)
wa.forecast.summary<-matrix(data=c(watest.0[2,"RMSE"], watest.1[2,"RMSE"], watest.2[2,"RMSE"],
                                   watest.3[2,"RMSE"], watest.4[2, "RMSE"],
                                   watest.5[2, "RMSE"]), nrow=1, ncol=6)
colnames(wa.forecast.summary)<-c("0,0,0","0,1,0", "1,0,0", "1,1,0", "1,1,1", "Holt")
rownames(wa.forecast.summary)<-c("RMSE")
print(wa.forecast.summary) #The ARIMA(1,0,0) model had the best forecast
checkresiduals(wats.arima.2)
```
#### PNW Region
The model that best fit the Pacific Northwest region as a whole was a Holt model with the following equation:
$y_t=l_{t-1}+b_{t-1}+\epsilon_t$\
$l_t=l_{t-1}+b_{t-1}+0.9619*\epsilon_t$\
$b_t =b_{t-1} + 0.0001*\epsilon_t$ 

The small beta value suggests that the trend is stable over time, and the alpha value being close to 1 indicates that the model assigns greater weight to recent observations. The model has identified a slightly negative trend in chinook landings over time (b=-0.0108).

```{r, echo=FALSE}

# PNW
pnwtest.0<-accuracy(pnw.forecast.0, pnwtest)
pnwtest.1<-accuracy(pnw.forecast.1, pnwtest)
pnwtest.2<-accuracy(pnw.forecast.2, pnwtest)
pnwtest.3<-accuracy(pnw.forecast.3, pnwtest)
pnwtest.4<-accuracy(pnw.forecast.4, pnwtest)
pnwtest.5<-accuracy(pnw.forecast.5, pnwtest)
pnw.forecast.summary<-matrix(data=c(pnwtest.0[2,"RMSE"], pnwtest.1[2,"RMSE"], pnwtest.2[2,"RMSE"],
                                   pnwtest.3[2,"RMSE"], pnwtest.4[2, "RMSE"],
                                   pnwtest.5[2, "RMSE"]), nrow=1, ncol=6)
colnames(pnw.forecast.summary)<-c("0,0,0","0,1,0", "1,0,0", "1,1,0", "1,1,1", "Holt")
rownames(pnw.forecast.summary)<-c("RMSE")
print(pnw.forecast.summary) #The Holt model had lowest RMSE
checkresiduals(pnwts.holt)
```

#### Total chinook landings
The best fit model for total chinook landings was a first-order ARIMA with first difference since the data was non-stationary, and no seasonality:
$y_t = 0.8878*y_{t-1}+0.8871+\epsilon_t$\
$\epsilon_t~N(0,\sigma^2=0.0436)$\

The alpha value suggests weak, negative autocorrelation, but it is not statistically significant.
```{r, echo=FALSE}
totaltest.0<-accuracy(total.forecast.0, totaltest)
totaltest.1<-accuracy(total.forecast.1, totaltest)
totaltest.2<-accuracy(total.forecast.2, totaltest)
totaltest.3<-accuracy(total.forecast.3, totaltest)
totaltest.4<-accuracy(total.forecast.4, totaltest)
totaltest.5<-accuracy(total.forecast.5, totaltest)
total.forecast.summary<-matrix(data=c(totaltest.0[2,"RMSE"], totaltest.1[2,"RMSE"], totaltest.2[2,"RMSE"],
                                   totaltest.3[2,"RMSE"], totaltest.4[2, "RMSE"],
                                   totaltest.5[2, "RMSE"]), nrow=1, ncol=6)
colnames(total.forecast.summary)<-c("0,0,0","0,1,0", "1,0,0", "1,1,0", "1,1,1", "Holt")
rownames(total.forecast.summary)<-c("RMSE")
print(total.forecast.summary)
checkresiduals(totalts.arima.3)

# Plotting best forecast from each region
# Alaska
par(mfrow=c(3,2))
plot(ak.forecast.5, main="Alaksa Chinook Landings", 
     ylab="Log Volume Biometric Tons", xlab="Year")
points(aktest, pch=20)
# California
plot(ca.forecast.5, main="California Chinook Landings", 
     ylab="Log Volume Biometric Tons", xlab="Year")
points(catest, pch=20)
# Oregon
plot(or.forecast.2, main="Oregon Chinook Landings", 
     ylab="Log Volume Biometric Tons", xlab="Year")
points(ortest, pch=20)
# Washington
plot(wa.forecast.2, main="Washington Chinook Landings", 
     ylab="Log Volume Biometric Tons", xlab="Year")
points(watest, pch=20)
# PNW
plot(pnw.forecast.5, main="PNW Chinook Landings", 
     ylab="Log Volume Biometric Tons", xlab="Year")
points(pnwtest, pch=20)
#Total
plot(total.forecast.3, main="Total Chinook Landings", ylab="Log Volume Biometric Tons", xlab="Year")
points(totaltest, pch=20)
```

In summary, none of these models contained a seasonality component. The model structure differed slightly between regions, but none of the models found significant changes in trend over time. Many of them behaved functionally like a random walk. 

*Forecasting Accuracy Comparison*

When we compared forecasting accuracy between models fit to one region against test data from another region, we found some interesting results. In some cases, another state’s forecasted data was a better fit than the best fit model for that state. This was the case for Alaska, whose test data was more accurately forecasted by the ARIMA model fit to the Washington data. Only the Washington and Oregon model forecasts were the best fit for their state’s test data. This is the opposite of what we would have expected, given our original hypothesis that Washington and Oregon’s time series look very similar. We would have expected forecasts from these two states to not be accurate when tested against other states’ data, since California and Alaska’s data looked different, but the Washington and Oregon models performed fairly well across the board. The large regional models (PNW and total) did not perform well, which does support the idea that forecasts for salmon fisheries should be made on smaller regional scales that are ecologically relevant to salmon populations. 

```{r, echo=FALSE}
forecastrmse<-matrix(nrow=6, ncol=6)
rownames(forecastrmse)<-c("AK Forecast", "CA Forecast", 
                         "OR Forecast", "WA Forecast", "PNW Forecast", "Total Forecast")
colnames(forecastrmse)<-c("AK Model", "CA Model", 
                         "OR Model", "WA Model", "PNW Model", "Total Model")
forecastrmse[1,1]<-accuracy(ak.forecast.5, aktest)[2, "RMSE"]
forecastrmse[1,2]<-accuracy(ca.forecast.5, aktest)[2, "RMSE"]
forecastrmse[1,3]<-accuracy(or.forecast.2, aktest)[2, "RMSE"]
forecastrmse[1,4]<-accuracy(wa.forecast.2, aktest)[2, "RMSE"]
forecastrmse[1,5]<-accuracy(pnw.forecast.5, aktest)[2, "RMSE"]
forecastrmse[1,6]<-accuracy(total.forecast.3, aktest)[2, "RMSE"]
forecastrmse[2,1]<-accuracy(ak.forecast.5, catest)[2, "RMSE"]
forecastrmse[2,2]<-accuracy(ca.forecast.5, catest)[2, "RMSE"]
forecastrmse[2,3]<-accuracy(or.forecast.2, catest)[2, "RMSE"]
forecastrmse[2,4]<-accuracy(wa.forecast.2, catest)[2, "RMSE"]
forecastrmse[2,5]<-accuracy(pnw.forecast.5, catest)[2, "RMSE"]
forecastrmse[2,6]<-accuracy(total.forecast.3, catest)[2, "RMSE"]
forecastrmse[3,1]<-accuracy(ak.forecast.5, ortest)[2, "RMSE"]
forecastrmse[3,2]<-accuracy(ca.forecast.5, ortest)[2, "RMSE"]
forecastrmse[3,3]<-accuracy(or.forecast.2, ortest)[2, "RMSE"]
forecastrmse[3,4]<-accuracy(wa.forecast.2, ortest)[2, "RMSE"]
forecastrmse[3,5]<-accuracy(pnw.forecast.5,ortest)[2, "RMSE"]
forecastrmse[3,6]<-accuracy(total.forecast.3, ortest)[2, "RMSE"]
forecastrmse[4,1]<-accuracy(ak.forecast.5, watest)[2, "RMSE"]
forecastrmse[4,2]<-accuracy(ca.forecast.5, watest)[2, "RMSE"]
forecastrmse[4,3]<-accuracy(or.forecast.2, watest)[2, "RMSE"]
forecastrmse[4,4]<-accuracy(wa.forecast.2, watest)[2, "RMSE"]
forecastrmse[4,5]<-accuracy(pnw.forecast.5,watest)[2, "RMSE"]
forecastrmse[4,6]<-accuracy(total.forecast.3, watest)[2, "RMSE"]
forecastrmse[5,1]<-accuracy(ak.forecast.5, pnwtest)[2, "RMSE"]
forecastrmse[5,2]<-accuracy(ca.forecast.5, pnwtest)[2, "RMSE"]
forecastrmse[5,3]<-accuracy(or.forecast.2, pnwtest)[2, "RMSE"]
forecastrmse[5,4]<-accuracy(wa.forecast.2, pnwtest)[2, "RMSE"]
forecastrmse[5,5]<-accuracy(pnw.forecast.5,pnwtest)[2, "RMSE"]
forecastrmse[5,6]<-accuracy(total.forecast.3, pnwtest)[2, "RMSE"]
forecastrmse[6,1]<-accuracy(ak.forecast.5, totaltest)[2, "RMSE"]
forecastrmse[6,2]<-accuracy(ca.forecast.5, totaltest)[2, "RMSE"]
forecastrmse[6,3]<-accuracy(or.forecast.2, totaltest)[2, "RMSE"]
forecastrmse[6,4]<-accuracy(wa.forecast.2, totaltest)[2, "RMSE"]
forecastrmse[6,5]<-accuracy(pnw.forecast.5,totaltest)[2, "RMSE"]
forecastrmse[6,6]<-accuracy(total.forecast.3, totaltest)[2, "RMSE"]
print(forecastrmse)
```

## Question 2

#### Initial Step - Forecasting with Maximum Knowledge (ARIMA, ETS and HW) 

As part of our analysis process, We wanted to determine which model structure and respective function was best suited for our analysis. We started with a simple ARIMA model to see if the auto regressive process with full knowledge could be used to produce accurate forecasts. We used auto.arima to select the best model for each data set. We began with the yearly time series and found poor forecasting performance along with simple models so chose to pivot away to monthly data.

```{r, echo=FALSE}

#Yearly time series 
arima_ff <- function(ts_list, labels) {
  

  results <- list()
  par(mfrow = c(2, 2), mar = c(4, 4, 3, 2))  
  
  for (i in 1:length(ts_list)) {
    ts_data <- ts_list[[i]]
    
    # Set training and testing data
    train <- window(ts_data, start = start(ts_data), end = c(start(ts_data)[1] + 20))
    test  <- window(ts_data, start = c(start(ts_data)[1] + 20), end = c(start(ts_data)[1] + 66))
    
    # Fit and forecast
    model <- auto.arima(train)
    pred <- forecast(model, h = 3)
    acc <- accuracy(pred, test)
    
    model_type <- paste0("ARIMA(", paste(arimaorder(model), collapse = ","), ")")
    
      results[[i]] <- list(
      Label = labels[i],
      AIC = AIC(model),
      BIC = BIC(model),
      MAPE = acc[2, "MAPE"],
      RMSE = acc[2, "RMSE"],
      Model = model_type)
    
    # Plot forecast and compare
    plot(pred, main = paste("Forecast -", labels[i]), ylab = "Value", xlab = "Time")
    points(test, col = "red", pch = 16)
  }
  
  par(mfrow = c(1, 1))  # Reset layout
    final_results <- do.call(rbind, lapply(results, as.data.frame))
  return(final_results)
  
}

arima_ff(list_area,area_names)
```

Our results showed that with yearly data only CA had any autocorrelation in the data while WA and AK were simple random walks. This was demonstrated by the ballooning of the confidence intervals in the forecasts like we had seen in class.

Given the poor results for the yearly time series (although true values still remained within the confidence intervals) we considered using a seasonal component to inform the forecast so shifted to exploring ETS and Holt-Winters models. We also re-ran the function for the monthly data

```{r, echo=FALSE}

# Create the time series 
datmon_CA <- ts(ch_month$metric.tons[ch_month$State == "CA"], start = c(1, 1), frequency = 12 )
datmon_WA <-ts(ch_month$metric.tons[ch_month$State == "WA"], start = c(1, 1), frequency = 12 )
datmon_OR <- ts(ch_month$metric.tons[ch_month$State == "OR"], start = c(1, 1), frequency = 12 )

list_area_month <- list(datmon_CA,datmon_OR,datmon_WA )
area_names_m <- c("California", "Oregon", "Washington")


ets_ff <- function(ts_list, labels = NULL) {

  
  par(mfrow = c(2, 2), mar = c(4, 4, 3, 2))  # 2x2 panel layout
  results <- list()
  for (i in 1:length(ts_list)) {
    ts_data <- na.interp(ts_list[[i]])
    
    # Define train/test split (adjust if needed based on length of your data)
    train <- window(ts_data, start = start(ts_data), end = c(start(ts_data)[1] + 20))
    test  <- window(ts_data, start = c(start(ts_data)[1] + 20), end = c(start(ts_data)[1] + 66))
    
    # Fit ARIMA and forecast
    model <- ets(train)
    pred <- forecast(model, h = 2*12)
    
    acc <- accuracy(pred, test)
    
    # Plot forecast with test data
    plot(pred, main = paste("Forecast -", labels[i]), ylab = "Value", xlab = "Time")
    points(test, col = "red", pch = 16)
    
    results[[i]] <- list(
      Label = labels[i],
      AIC = AIC(model),
      BIC = BIC(model),
      MAPE = acc[2, "MAPE"],
      RMSE = acc[2, "RMSE"],
      Model = model$method)
    
  }
  
  
  par(mfrow = c(1, 1))  # Reset plotting for later
  final_results <- do.call(rbind, lapply(results, as.data.frame))
  return(final_results)
}


ets_ff(list_area_month,area_names_m)
```

The ETS analysis with perfect knowledge produced three models with Additive Error, No Trend, Additive Seasonality (ANA). This indicates that data oscillates around a constant mean with seasonality but no other trends are detected. As a result, the model cannot forecast

We also tried Holt-Winters to explore if model performance improved.

```{r, echo=FALSE}

hw_ff <- function(ts_list,labels) {

  
  par(mfrow = c(2, 2), mar = c(4, 4, 3, 2)) 
  results <- list()
  for (i in 1:length(ts_list)) {
    ts_data <- na.interp(ts_list[[i]])
    

    train <- window(ts_data, start = start(ts_data), end = c(start(ts_data)[1] + 20))
    test  <- window(ts_data, start = c(start(ts_data)[1] + 20), end = c(start(ts_data)[1] + 66))
    
    # Fit and forecast
    model <- hw(train)

    pred <- forecast(model, h = 2*12)
    acc <- accuracy(pred, test)

    
    # Plot forecast with test data
    plot(pred, main = paste("Forecast HW -", labels[i]), ylab = "Value", xlab = "Time")
    points(test, col = "red", pch = 16)
    
    results[[labels[i]]] <- list(
      model_structure = model$method,
      accuracy_metrics = acc
    )
  }
  
  final_results <- do.call(rbind, lapply(results, as.data.frame))
  par(mfrow = c(1, 1))  # Reset plotting for later

  return(final_results)
}



hw_ff(list_area_month,area_names_m)


```

While Holt-Winters showed slight improvement the model also reverted to this seasonal trend around the mean. Given the simplicity of observing changes in ETS structure and performance, we chose to use that structure for the exercise is data loss to evaluate whether data loss may change model structure.

#### Impacts of missing data

In this section, we explored the impacts of dropping a window of data. In this case, we choose to drop 33 years of data. Originally, our plan was to place this function in a larger loop with windows that changed sizes and moved along the time series but this goal proved to be too ambitious for the length of this lab and with the issues we discovered along the way. We settled on using this window to demonstrate impacts of missing data

```{r, echo=FALSE}

#block size is how much is dropped, drop_start is where it happens
ets_ff_sensitivity <- function(ts_list, labels = NULL, block_size, drop_start) {
  

  
  results <- list()
  par(mfrow = c(2, 2), mar = c(4, 4, 3, 2))  # Plot setup
  
  for (i in seq_along(ts_list)) {
    ts_orig <- ts_list[[i]]
    ts_mod <- ts_orig
    
    # Drop the block and interpolate the data
    drop_end <- min(drop_start + block_size - 1, length(ts_mod))
    ts_mod[drop_start:drop_end] <- NA
    ts_clean <- na.interp(ts_mod)
    
    # Save original missing indexes for plotting
    missing_idx <- which(is.na(ts_mod))
    interpolated_times <- time(ts_clean)[missing_idx]
    interpolated_values <- ts_clean[missing_idx]
    
    # Define time range
    ts_times <- time(ts_clean)
    
    # Split into train/test
    train <- window(ts_clean, start = start(ts_clean), end = c(start(ts_clean)[1] + 20))
    test  <- window(ts_clean, start = c(start(ts_clean)[1] + 20), end = c(start(ts_clean)[1] + 66))
    
    # Fit ETS model
    model <- ets(train, model = "ZZZ")
    fc <- forecast(model, h = length(test))
    
    # Accuracy
    acc <- accuracy(fc, test)
    
    # Store results so we can compare them later
    results[[i]] <- list(
      Label = labels[i],
      AIC = AIC(model),
      BIC = BIC(model),
      MAPE = acc[2, "MAPE"],
      RMSE = acc[2, "RMSE"],
      Model = model$method
    )
    
    # Plot forecast
    plot(fc, main = paste(labels[i], "- Drop:", drop_start, "to", drop_end), ylab = "Value", xlab = "Time")
    
    # Learning to shade the dropped regions
    drop_time_start <- ts_times[drop_start]
    drop_time_end   <- ts_times[drop_end]
    usr <- par("usr")  # plotting range
    rect(xleft = drop_time_start, xright = drop_time_end,
         ybottom = usr[3], ytop = usr[4],
         col = rgb(0.9, 0.9, 0.9, 0.5), border = NA)
    

    
    # Add test points and interpolated points
    lines(test, col = "lightgreen")
    points(interpolated_times, interpolated_values, col = "blue")
    
    # Re-plot forecast line on top
    lines(fc$mean, col = "red")
    

    legend("topright",
           legend = c("Forecast", "Test", "Interpolated", "Dropped"),
           col = c("red", "lightgreen", "blue", rgb(0.9, 0.9, 0.9, 0.5)),
           pch = c(NA, NA, 1, 15), lty = c(1, 1, NA, NA),
           pt.cex = 0.7,  # Size of points/symbols
           cex = 0.7,     # Size of text
           bty = "n",     # No box
           y.intersp = 0.7,  # Vertical spacing between lines
           x.intersp = 0.5) 
  }
  
  par(mfrow = c(1, 1))  # Reset plot layout
  
  # Compile results into a data frame
  df_results <- do.call(rbind, lapply(results, as.data.frame))
  return(df_results)
}

ets_ff_sensitivity(list_area_month, labels=area_names_m, drop_start = 30,block_size=96)


```

Dropping data was most consequential for the CA data set as data was already sparse. For the CA time series, the model structure changed compared to the control case of full knowledge where the model was no longer able to detect the same auto correlative trend (A,N,A) as it did originally. This was expected and partially motivated this exercise as Brian observed some of his models doing this so we hoped we could replicate a "switch" in model types once data became sparse.

For the OR and WA data set, we observed that as data decreases, the models started to settle on a mean seasonal behavior. From a practical perspective, these models with such a large block of missing data could only be used to forecast a "general" behavior rather than accounting for any sort of variation that may have been present. In the case of Washington, as the model was trained on a period of exceptionally high data, it forecasts a seasonal trend that is more dramatic than the past fluctuations.

#### Data Sparsity - Exclude All High Values

Masking a block data can simulate a loss of funding, pause in project activities, etc. We hoped to explore a more persistent loss and chose to simulate how losing certain values across a data set impacts model results. If we decrease the magnitude of the seasonal trend, how will the model respond. For this case, we modified the function above to replace all values over 200 with an NA value that is then interpolated. The result is a dampened seasonal trend. Again, the objective was to see if this kind of data modification could cause for model specification changes and if these changes varied among models.

```{r, echo=FALSE}
ets_target_sensitivity <- function(ts_list, labels = NULL, block_size, drop_start, target) {
  
  if (is.null(labels)) {
    labels <- paste("Series", 1:length(ts_list))
  }
  
  results <- list()
  par(mfrow = c(2, 2), mar = c(4, 4, 3, 2))  # Plot prep
  
  for (i in seq_along(ts_list)) {
    ts_orig <- ts_list[[i]]
    ts_mod <- ts_orig
    
    # Replace values above 200 with NA in the block 
    drop_end <- min(length(ts_mod), drop_start + block_size - 1)  #-1 to make sure we're in bounds
    ts_mod[drop_start:drop_end][ts_mod[drop_start:drop_end] > 200] <- NA
    
    # Interpolate the missing values in the modified data
    ts_clean <- na.interp(ts_mod)
    
    # Save original missing indexes for plotting
    miss_i <- which(is.na(ts_mod))
    interpolated_times <- time(ts_clean)[miss_i]
    interpolated_values <- ts_clean[miss_i]
    
    # Define time range for train/test
    ts_times <- time(ts_clean)
    
    # Split into train/test
    train <- window(ts_clean, start = start(ts_clean), end = c(start(ts_clean)[1] + 20))
    test  <- window(ts_orig, start = c(start(ts_orig)[1] + 20), end = c(start(ts_orig)[1] + 66))
    
    # Fit model
    model <- ets(train, model = "ZZZ")
    pred <- forecast(model, h = length(test))
    
    # Accuracy stats
    acc <- accuracy(pred, test)
    
    # Store results 
    results[[i]] <- list(
      Label = labels[i],
      AIC = AIC(model),
      BIC = BIC(model),
      MAPE = acc[2, "MAPE"],
      RMSE = acc[2, "RMSE"],
      Model = model$method
    )
    
    #True training set
    plot(ts_orig, main = paste(labels[i], "- Drop:", target), ylab = "Value", xlab = "Time", col = "black")
    #Interpolated values
    lines(ts_clean, col = "blue", lwd = 2)  
    
    
    # True values
    lines(test, col = "lightgreen", pch = 16)
    
    # Forecasts
    lines(pred$mean, col = "red")
    
    legend("topright", legend = c("Original", "Interpolated", "Forecast", "Test"),
           col = c("black", "blue", "red", "lightgreen"),
          lty = c(1, 1, 1, 1), pt.cex = 1.5, bty = "n")
  }
  
  par(mfrow = c(1, 1))  
  
  # Data
  df_results <- do.call(rbind, lapply(results, as.data.frame))
  return(df_results)
}

# Example call to the function
ets_target_sensitivity(list_area_month, labels = area_names_m, drop_start = 1, block_size = 33 * 12, target = 200)
```

The results of this modification proved to be impact model specification more than block elimination of data. We were surprised by this outcome as we would have expected these changes to conserve more of the original structure of the data compared to dropping a block as the seasonal component remains intact and only magnitude changes.

However, this modification resulted in a (AAA), (ANA) and (MNM) model types. It seemed that only OR preserved the seasonal structure and was able to produce forecasts that may prove relevant for a practical application. The CA data produced an (AAA) model structure as the sparsity of the data prior to modification paired with the loss of values over 200 created a weak seasonal effect. This means the model had Additive Error, Additive Trend, Additive Seasonality or that seasonal trends over time are considered to be constant and predictable.The results for WA were most surprising as the large landings of salmon made it such that the greatest proportion of data was lost making the seasonal effect very weak. However, this model also changed to Multiplicative error with No trend, Multiplicative seasonality suggesting that even with the data loss the model would expect large changes in the data. Simultaneously the model also predicted the weakest seasonal trend. Perhaps the multiplicative relationship produced an very dampened prediction?

#### Summarizing Results

Finally, we consolidated all results into a data frame to examine how changes to the data affected model type and model performance.

```{r, echo=FALSE}

ets_full <- ets_ff(list_area_month,area_names_m)

ets_sens <- ets_ff_sensitivity(list_area_month, labels=area_names_m, drop_start = 30,block_size=96)

ets_targ_sens <- ets_target_sensitivity(list_area_month, labels = area_names_m, drop_start = 1, block_size = 33 * 12, target = 200)


#Label each result for comparison
ets_full$type <- "test"
ets_sens$type <- "Block"
ets_targ_sens$type <- "Target200"


summary_sens <- rbind(ets_sens,ets_targ_sens,ets_full)

summary_sens

```

The comparison shows that the best performing models by AIC are those that are subject to the targeted data drops. This is unsurprising as we dampened the variability in these data sets by eliminating high values. However, it is also this kind of data loss that made them the least realistic when forecasting and changed their structure. For the dropping of blocks of data, model performance was still higher than the test series by AIC, but RMSE proved worse. Model structure was also conserved compared to the initial series .

## Question 3

The Forecast package's auto.arima function was used to determine which ARIMA models best fit the training data related to landings, prices, and value of fish landed. 

```{r, echo=FALSE}
#CA
CAlands.fit <- forecast::auto.arima(CAlands.train)
CAprice.fit <- forecast::auto.arima(CAprice.train)
CAvalue.fit <- forecast::auto.arima(CAvalue.train)

#OR
ORlands.fit <- forecast::auto.arima(ORlands.train)
ORprice.fit <- forecast::auto.arima(ORprice.train)
ORvalue.fit <- forecast::auto.arima(ORvalue.train)

#WA
WAlands.fit <- forecast::auto.arima(WAlands.train)
WAprice.fit <- forecast::auto.arima(WAprice.train)
WAvalue.fit <- forecast::auto.arima(WAvalue.train)
```

```{r, echo=FALSE}
CAlands.fit
CAprice.fit
CAvalue.fit

ORlands.fit
ORprice.fit
ORvalue.fit

WAlands.fit
WAprice.fit
WAvalue.fit
```

The models describing Chinook market data in California fail to find any seasonal trends, most likely due to data from California being much more scarce than data from Oregon and Washington. Moving forward California will be omitted from the process of forecasting landings, prices, and value.

```{r}
checkresiduals(ORvalue.fit)
```
\newpage

```{r}
checkresiduals(WAvalue.fit)
```

In general, models fitting landed weight, prices, and value show well behaved residuals distributed symmetrically around a mean of zero with little auto-correlation. The value models for both Oregon and Washington show the most autocorrelation and also a very tight distribution of error terms. More 'well behaved' residual charts for landings and prices are not shown for brevity.

```{r, echo=FALSE}
# project forecasts

#OR
par(mfrow=c(2,1))
ORlands.fc <- forecast::forecast(ORlands.fit, h=24)
ORprice.fc <- forecast::forecast(ORprice.fit, h=24)
ORvalue.fc <- forecast::forecast(ORvalue.fit, h=24)
plot(ORvalue.fc, main="Forecast Value of Chinook Landed, Oregon")
points(ORvalue.test)

#WA
WAlands.fc <- forecast::forecast(WAlands.fit, h=24)
WAprice.fc <- forecast::forecast(WAprice.fit, h=24)
WAvalue.fc <- forecast::forecast(WAvalue.fit, h=24)
plot(WAvalue.fc, main="Forecast Value of Chinook Landed, Washington")
points(WAvalue.test)
par(mfrow=c(1,1))
```

The above figure shows forecasted value of Chinook landed in Washington and Oregon based on the values reported in the data, with test data represented . This will be contrasted with separate forecasts of landings and prices being interacted. 

```{r, echo=FALSE}
#comparing forecasts
#OR
ORlands.compoundfc <- exp(ORlands.fc$mean)
ORvalue.compoundfc <- ORlands.compoundfc*ORprice.fc$mean
 
#WA
WAlands.compoundfc <- exp(WAlands.fc$mean)
WAvalue.compoundfc <- WAlands.compoundfc*WAprice.fc$mean

# plotting
#OR
par(mfrow=c(2,1))
plot(ORvalue.fc$mean, main = "Forecast Value of Chinook Landed, Oregon", ylab = "", xlab = "")
points(ORvalue.test)

plot(ORvalue.compoundfc, main = "Interacted Forecasts, Chinook Landed and Prices, Oregon", ylab = "")
points(ORvalue.test)

#WA
plot(WAvalue.fc$mean, main = "Forecast Value of Chinook Landed, Washington", ylab = "", xlab = "")
points(WAvalue.test)

plot(WAvalue.compoundfc, main = "Interacted Forecasts, Chinook Landed and Prices, Washington", ylab = "")
points(WAvalue.test)
par(mfrow=c(1,1))
```

Visually comparing the value forecast to the interacted forecasts of landings and prices it is clear that the interacted forecasts track the test data much better than the value forecast in Oregon. Visual inspection of data from Washington is not as clear, so we next compare mean square error. 

```{r, echo=FALSE}
OR <- data.frame(date=as.Date(as.yearmon(time(ORvalue.fc$mean))), 
                 value=as.matrix(ORvalue.fc$mean),
                 compound=as.matrix(ORvalue.compoundfc),
                 test=as.matrix(ORvalue.test))

OR$value_error <- (OR$test - OR$value)^2
OR$compound_error <- (OR$test - OR$compound)^2

MSE_ORvalue <- mean(OR$value_error, na.rm=TRUE)
MSE_ORcompound <- mean(OR$compound_error, na.rm=TRUE)

WA <- data.frame(date=as.Date(as.yearmon(time(WAvalue.fc$mean))), 
                 value=as.matrix(WAvalue.fc$mean),
                 compound=as.matrix(WAvalue.compoundfc),
                 test=as.matrix(WAvalue.test))

WA$value_error <- (WA$test - WA$value)^2
WA$compound_error <- (WA$test - WA$compound)^2

MSE_WAvalue <- mean(WA$value_error, na.rm=TRUE)
MSE_WAcompound <- mean(WA$compound_error, na.rm=TRUE)

forecastmse<-matrix(nrow=2, ncol=2)
rownames(forecastmse)<-c("Oregon", "Washington")
colnames(forecastmse)<-c("Value forecast", "Interacted forecast")
forecastmse[1,1] <- MSE_ORvalue
forecastmse[1,2] <- MSE_ORcompound
forecastmse[2,1] <- MSE_WAvalue
forecastmse[2,2] <- MSE_WAcompound
print(forecastmse)
```
Comparing MSE we can see that the forecast of value of Chinook landed performs very slightly better than the interacted forecast in Washington, while the interacted forecast radically outperforms the value forecast for Oregon. 

# Discussion

In summary, none of these models contained a seasonality component. The model structure differed slightly between regions, but none of the models found significant changes in trend over time. Many of them behaved functionally like a random walk. 

When we compared forecasting accuracy between models fit to one region against test data from another region, we found some interesting results. In some cases, another state’s forecasted data was a better fit than the best fit model for that state. This was the case for Alaska, whose test data was more accurately forecasted by the ARIMA model fit to the Washington data. Only the Washington and Oregon model forecasts were the best fit for their state’s test data. This is the opposite of what we would have expected, given our original hypothesis that Washington and Oregon’s time series look very similar. I would have expected forecasts from these two states to not be accurate when tested against other states’ data, since California and Alaska’s data looked different, but the Washington and Oregon models performed fairly well across the board. The large regional models (PNW and total) did not perform well, which does support the idea that forecasts for salmon fisheries should be made on smaller regional scales that are ecologically relevant to salmon populations.

Although we were unable to complete our initial objectives of completing a sensitivity analysis that allows us to explore what influences the way ETS and auto.arima select models, we were able to use the basic building blocks for these functions to gain insights on behavior if this was possible. We were able to understand how these models operate and see their limitations compared to other types of models we have encountered.

If we had more time, we would have liked to explore different methods of data loss, perhaps focused more on proportions rather than a fixed value or different ways of replacing the data that was lost (drawing from a distribution, etc). Finally, our ideal of looping over these functions and changing the modification window would be a great next step. This exercise helped us reflect on the limitations of these models, especially given that we are more familiar with state space models so we perhaps overestimated the abilities of these model types.

The comparison of forecasted value and the interaction of forecasted landings and prices yielded fruitful results. While the Washington value forecast very slightly outperformed the interacted forecast, in Orgeon interacting forecasts of landings and prices proved much more accurate than the forecast of value. This is likely because, while related, prices and landings each have their own tenancies that are subsumed into the value data. It seems reasonable for any agencies or researchers to walk through this two-step interaction approach when trying to project the value of landings into the future.

# Team member contributions

Each member of the team took on one issue and completed the code and the analysis. Halfway through, we consulted other team members for feedback and adjust analysis based on the joint discussion. All team members participated in the lab write up and review the results.

