---
title: "Team 1 - Lab 1"
author: "Nicole, Brian, Anna"
date: "April 21, 2025"
output:
  pdf_document:
    toc: yes
  html_document:
    code-folding: yes
    toc: yes
    toc_float: yes
subtitle: Forecasting with ARIMA and ETS models
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

\newpage
# Data

This lab used annual data recording salmon landings in six different US states covering the period from 1950 to 2016: Alaska, California, Oregon, Michigan, Pennsylvania, and Washington. Due to sparse data, Pennsylvania and Michigan were excluded from the analysis. Monthly data recording salmon landings as well as the value of fish landed was used from Washington, Oregon, and California covering the period from 1990 to 2022. 

# Research Questions

1. What model structure best fits the different states and regions (Total and Pacific Northwest) represented in the yearly data set? How does forecasting accuracy compare across these regions?

2. How does model performance change when sections of the time series are dropped? 

3. How does forecasting monthly value of catch compare with separate forecasts of monthly landings and prices being interacted to find overall value?

# Methods

## Initial plan
We initially planned to fit ARIMA and ETS models to the different states and compare which model structure best fit each state or region. Noting the similarities between Washington and Oregon’s time series data, we also wanted to compare forecast accuracy for models trained on different data sets (e.g. compare Washington’s forecast accuracy on Oregon’s test data). We also planned on recovering monthly prices by dividing the value of fish landed by the landed weight in order to assess the efficiency of forecasting approaches.

# Diagnostics and preliminary exploration - Question 1
To address question 1, we subset the data for each west coast state, Pacific Northwest which includes WA, OR, and CA, and the total of all states combined, into a training data set and test data set. The training data sets spanned the years 1950-2005, and the test data set was 2006-2016. When we plotted the data there were no obvious problems with stationarity. There might be a slight downwards trend, and there are a few gaps in the California data, but it is not immediately obvious if the data is non-stationary.


```{r, include=FALSE}
## Packages
library(atsalibrary)
library(dplyr)
library(ggplot2)
library(patchwork)
library(tidyverse)
library(zoo)
par(mfrow=c(1,1))
#Filter and plot data
data(chinook)
chinook.year<-chinook.year %>%
  filter(!State %in% c("Michigan", "Pennsylvania"))

## Subset test and training data for each region
# Washington
dat <- chinook.year %>%
  filter(State == "Washington") %>%
  mutate(lnreturns = log.metric.tons,
         year = Year) %>%
  select(year, lnreturns)
wats <- ts(dat$lnreturns, start=dat$year[1])
watrain <- window(wats, dat$year[1], dat$year[1]+55)
watest <- window(wats, dat$year[1]+56, dat$year[1]+66)
testlength<-length(watest)
# Oregon
dat <- chinook.year %>%
  filter(State == "Oregon") %>%
  mutate(lnreturns = log.metric.tons,
         year = Year) %>%
  select(year, lnreturns)
orts <- ts(dat$lnreturns, start=dat$year[1])
ortrain <- window(orts, dat$year[1], dat$year[1]+56)
ortest <- window(orts, dat$year[1]+56, dat$year[1]+66)
# California
dat <- chinook.year %>%
  filter(State == "California") %>%
  mutate(lnreturns = log.metric.tons,
         year = Year) %>%
  select(year, lnreturns)
cats <- ts(dat$lnreturns, start=dat$year[1])
catrain <- window(cats, dat$year[1], dat$year[1]+55)
catest <- window(cats, dat$year[1]+56, dat$year[1]+66)
# PNW Region
chinook.pnw.total <- chinook.year %>%
  filter(!State == "Alaska")%>%
  mutate(metric.tons = exp(log.metric.tons)) %>%  # undo the log
  group_by(Year) %>%
  summarise(
    total.metric.tons = sum(metric.tons, na.rm = TRUE),
    log.total.metric.tons = log(total.metric.tons)
  )
dat <- chinook.pnw.total %>%
  mutate(lnreturns = log.total.metric.tons,
         year = Year) %>%
  select(year, lnreturns)
pnwts <- ts(dat$lnreturns, start=dat$year[1])
pnwtrain <- window(pnwts, dat$year[1], dat$year[1]+55)
pnwtest <- window(pnwts, dat$year[1]+56, dat$year[1]+66)
# Alaska
dat <- chinook.year %>%
  filter(State=="Alaska") %>%
  mutate(lnreturns = log.metric.tons,
         year = Year) %>%
  select(year, lnreturns)
akts <- ts(dat$lnreturns, start=dat$year[1])
aktrain <- window(akts, dat$year[1], dat$year[1]+55)
aktest <- window(akts, dat$year[1]+56, dat$year[1]+66)

chinook.total <- chinook.year %>%
  filter(State%in%c("California", "Alaska", "Oregon", "Washington")) %>%
  mutate(metric.tons = exp(log.metric.tons)) %>%  # undo the log
  group_by(Year) %>%
  summarise(
    total.metric.tons = sum(metric.tons, na.rm = TRUE),
    log.total.metric.tons = log(total.metric.tons)
  )
dat <- chinook.total %>%
  mutate(lnreturns = log.total.metric.tons,
         year = Year) %>%
  select(year, lnreturns)
totalts<-ts(dat$lnreturns, start=dat$year[1])
totaltrain <- window(totalts, dat$year[1], dat$year[1]+55)
totaltest <- window(totalts, dat$year[1]+56, dat$year[1]+66)

```


```{r, echo=FALSE}
ggplot(chinook.year, aes(x = Year, y = log.metric.tons, color = State)) +
  geom_line() +
  labs(
    title = "Chinook Landings by State",
    x = "Year",
    y = "Log(Metric Tons)"
  ) +
  theme_minimal()
```


## Use ACF and PACF

```{r, echo=FALSE}
par(mfrow=c(2,2))
acf(catrain, main= "California ACF")
acf(aktrain, main="Alaska ACF")
acf(watrain, main="Washington ACF")
acf(ortrain,main="Oregon ACF")


pacf(catrain, main= "California pACF")
pacf(aktrain, main="Alaska pACF")
pacf(watrain, main="Washington pACF")
pacf(ortrain,main="Oregon pACF")

acf(pnwtrain, main= "PNW ACF")
acf(totaltrain, main="Total ACF")
pacf(pnwtrain, main= "PNW pACF")
pacf(totaltrain, main="Total pACF")
par(mfrow=c(1,1))
```

What did you learn? Also try decomposition.
*I might need help interpreting these plots*

## Test for stationarity

When we run ndiffs on the states’ data, Oregon, Washington, the Pacific Northwest region determined one first-difference is required to make the data stationary.

```{r, echo=FALSE}
akndiffs<-ndiffs(aktrain)
candiffs<-ndiffs(catrain)
orndiffs<-ndiffs(ortrain)
wandiffs<-ndiffs(watrain)
totalndiffs<-ndiffs(totaltrain)
pnwndiffs<-ndiffs(pnwtrain)

ndiffsresults<-matrix(data=c(akndiffs, candiffs, orndiffs, wandiffs, totalndiffs, pnwndiffs), ncol=6, nrow=1)
colnames(ndiffsresults)<-c("Alaksa", "California", "Oregon", "Washington", "Total", "PNW")
print(ndiffsresults)
```

# Diagnostics and preliminary exploration - Question 3
Preliminary analysis of question 2 involved first recovering price data and then plotting monthly landings, prices, and total value for all three states represented. 

```{r, echo=FALSE}
# create date column (first of each month)
chinook.month <- chinook.month %>%
  mutate(Date = as.Date(paste(Year, Month, "01", sep = "-"), format = "%Y-%b-%d"))

# back out price (US$/ton)
chinook.month$price <- chinook.month$value.usd/chinook.month$metric.tons

# plot with price by state
landings.monthly <- ggplot(chinook.month, aes(x = Date, y = log.metric.tons)) +
  geom_line() +
  facet_wrap(~ State, nrow = 1) +  # 3 panels vertically
  labs(title = "Monthly Chinook Salmon Landings",
       x = "Date",
       y = "Log(m. tons)") +
  theme_minimal()

chinook.month[sapply(chinook.month, is.infinite)] <- NA
chinook.month <- chinook.month %>% mutate_all(~ifelse(is.nan(.), NA, .))
price.monthly <- ggplot(chinook.month, aes(x = Date, y = price)) +
  geom_line() +
  facet_wrap(~ State, nrow = 1) +  # 3 panels horizontal
  labs(title = "Monthly Chinook Salmon Price",
       x = "Date",
       y = "US$/m. ton") +
  theme_minimal()

value.monthly <- ggplot(chinook.month, aes(x = Date, y = value.usd)) +
  geom_line() +
  facet_wrap(~ State, nrow = 1) +  # 3 panels horizontal
  labs(title = "Monthly Chinook Salmon Value",
       x = "Date",
       y = "US$") +
  theme_minimal()

landings.monthly/price.monthly/value.monthly
```

Next, diagnostics were performed on these data, evaluating ACFs and pACFs checking ndiffs for stationarity.

```{r, echo=FALSE}
# set time series
CAchinook.month <- subset(chinook.month, State == "CA")
ORchinook.month <- subset(chinook.month, State == "OR")
WAchinook.month <- subset(chinook.month, State == "WA")

#CA
CAlands.ts <- ts(CAchinook.month$log.metric.tons, start=c(1990,1),
                 frequency = 12)
CAlands.train <- window(CAlands.ts, c(1990, 1), c(2020, 12))
CAlands.test <- window(CAlands.ts, c(2021, 1), c(2022, 12))

CAprice.ts <- ts(CAchinook.month$price, start=c(1990,1),
                 frequency = 12)
CAprice.train <- window(CAprice.ts, c(1990, 1), c(2020, 12))
CAprice.test <- window(CAprice.ts, c(2021, 1), c(2022, 12))

CAvalue.ts <- ts(CAchinook.month$value.usd, start=c(1990,1),
                 frequency = 12)
CAvalue.train <- window(CAvalue.ts, c(1990, 1), c(2020, 12))
CAvalue.test <- window(CAvalue.ts, c(2021, 1), c(2022, 12))

#OR
ORlands.ts <- ts(ORchinook.month$log.metric.tons, start=c(1990,1),
                 frequency = 12)
ORlands.train <- window(ORlands.ts, c(1990, 1), c(2020, 12))
ORlands.test <- window(ORlands.ts, c(2021, 1), c(2022, 12))

ORprice.ts <- ts(ORchinook.month$price, start=c(1990,1),
                 frequency = 12)
ORprice.train <- window(ORprice.ts, c(1990, 1), c(2020, 12))
ORprice.test <- window(ORprice.ts, c(2021, 1), c(2022, 12))

ORvalue.ts <- ts(ORchinook.month$value.usd, start=c(1990,1),
                 frequency = 12)
ORvalue.train <- window(ORvalue.ts, c(1990, 1), c(2020, 12))
ORvalue.test <- window(ORvalue.ts, c(2021, 1), c(2022, 12))

#WA
WAlands.ts <- ts(WAchinook.month$log.metric.tons, start=c(1990,1),
                 frequency = 12)
WAlands.train <- window(WAlands.ts, c(1990, 1), c(2020, 12))
WAlands.test <- window(WAlands.ts, c(2021, 1), c(2022, 12))

WAprice.ts <- ts(WAchinook.month$price, start=c(1990,1),
                 frequency = 12)
WAprice.train <- window(WAprice.ts, c(1990, 1), c(2020, 12))
WAprice.test <- window(WAprice.ts, c(2021, 1), c(2022, 12))

WAvalue.ts <- ts(WAchinook.month$value.usd, start=c(1990,1),
                 frequency = 12)
WAvalue.train <- window(WAvalue.ts, c(1990, 1), c(2020, 12))
WAvalue.test <- window(WAvalue.ts, c(2021, 1), c(2022, 12))
```

```{r, echo=FALSE}
## Differencing and stationarity
CAlandings <- ndiffs(CAlands.train)
CAprices <- ndiffs(CAprice.train)
CAvalue <- ndiffs(CAvalue.train)

ORlandings <- ndiffs(ORlands.train)
ORprices <- ndiffs(ORprice.train)
ORvalue <- ndiffs(ORvalue.train)

WAlandings <- ndiffs(WAlands.train)
WAprices <- ndiffs(WAprice.train)
WAvalue <- ndiffs(WAvalue.train)

ndiffsresults<-matrix(data=c(CAlandings, CAprices, CAvalue, ORlandings, ORprices, ORvalue, WAlandings, WAprices, WAvalue), ncol=3, nrow=3)
colnames(ndiffsresults) <- c("Landings", "Prices", "Value")
rownames(ndiffsresults) <- c("CA", "OR", "WA")
print(ndiffsresults)
```


```{r, echo=FALSE}
## ACF plots
par(mfrow=c(3,3))
CAl_acf <- acf(CAlands.train, main= "California Landings ACF", na.action = na.pass)
CAp_acf <- acf(CAprice.train, main= "California Prices ACF", na.action = na.pass)
CAv_acf <- acf(CAvalue.train, main= "California Values ACF", na.action = na.pass)

ORl_acf <- acf(ORlands.train, main= "Oregon Landings ACF", na.action = na.pass)
ORp_acf <- acf(ORprice.train, main= "Oregon Prices ACF", na.action = na.pass)
ORv_acf <- acf(ORvalue.train, main= "Oregon Values ACF", na.action = na.pass)

WAl_acf <- acf(WAlands.train, main= "Washington Landings ACF", na.action = na.pass)
WAp_acf <- acf(WAprice.train, main= "Washington Prices ACF", na.action = na.pass)
WAv_acf <- acf(WAvalue.train, main= "Washington Values ACF", na.action = na.pass)

#PACF
CAl_pacf <- pacf(CAlands.train, main= "California Landings pACF", na.action = na.pass)
CAp_pacf <- pacf(CAprice.train, main= "California Prices pACF", na.action = na.pass)
CAv_pacf <- pacf(CAvalue.train, main= "California Values pACF", na.action = na.pass)

ORl_pacf <- pacf(ORlands.train, main= "Oregon Landings pACF", na.action = na.pass)
ORp_pacf <- pacf(ORprice.train, main= "Oregon Prices pACF", na.action = na.pass)
ORv_pacf <- pacf(ORvalue.train, main= "Oregon Values pACF", na.action = na.pass)

WAl_pacf <- pacf(WAlands.train, main= "Washington Landings pACF", na.action = na.pass)
WAp_pacf <- pacf(WAprice.train, main= "Washington Prices pACF", na.action = na.pass)
WAv_pacf <- pacf(WAvalue.train, main= "Washington Values pACF", na.action = na.pass)
par(mfrow=c(1,1))
```

There is clear evidence of seasonal trends in the above plots, meaning any useful model will more than likely involve a seasonal component (sARIMA).

Finally, data from each of three state-level time series was subsetted into training and testing data sets. Training data spanned the years from 1990 through 2020 while test data covers the years 2021 & 2022. 

# Results

## Question 1

We fit multiple ARIMA and Holt models to the yearly time series data for each state, the PNW region (WA, OR, CA), and the total chinook landings for all states combined. ARIMA models where p, d, or q were greater than 1 had significantly higher AIC values than more parsimonious ARIMA models and are excluded from this write-up. We then forecasted 10 years of data using the six ARIMA models, tested the forecasts against the test data for each region. Using root mean squared error, we selected the best fit model for each state or region. The Holt model was the best fit for Alaska, California, and the Pacific Northwest region. An ARIMA(1,0,0) model was the best fit for Washington and Oregon. An ARIMA(0,1,0) model was the best fit for the entire Pacific Coast chinook landing data (log metric tons). See the figures below for the six regional time series and forecasts from the best fit model for each. We will also provide a brief interpretation of the best fit models, and check of the residuals using a Ljung-Box test. 


```{r, echo=FALSE}
auto.arima(aktrain) # returned (1,0 0)
akts.arima.0<-arima(aktrain, order=c(0,0,0))
akts.arima.1<-arima(aktrain, order=c(0,1,0))
akts.arima.2<-arima(aktrain, order=c(1,0,0)) #auto-arima model
akts.arima.3<-arima(aktrain, order=c(1,1,0))
akts.arima.4<-arima(aktrain, order=c(1,1,1))
akts.holt<-ets(aktrain, model="AAN")

# California
auto.arima(catrain) # returned (0,1,0)
cats.arima.0<-arima(catrain, order=c(0,0,0))
cats.arima.1<-arima(catrain, order=c(0,1,0))#auto-arima model
cats.arima.2<-arima(catrain, order=c(1,0,0))
cats.arima.3<-arima(catrain, order=c(1,1,0))
cats.arima.4<-arima(catrain, order=c(1,1,1))
cats.holt<-ets(catrain, model="AAN")

# Oregon 
auto.arima(ortrain) # returned (0,1,0)
orts.arima.0<-arima(ortrain, order=c(0,0,0))
orts.arima.1<-arima(ortrain, order=c(0,1,0))#auto-arima model
orts.arima.2<-arima(ortrain, order=c(1,0,0)) 
orts.arima.3<-arima(ortrain, order=c(1,1,0))
orts.arima.4<-arima(ortrain, order=c(1,1,1))
orts.holt<-ets(ortrain, model="AAN")

# Washington
auto.arima(watrain) # returned (0,1,0)
wats.arima.0<-arima(watrain, order=c(0,0,0))
wats.arima.1<-arima(watrain, order=c(0,1,0))#auto-arima model
wats.arima.2<-arima(watrain, order=c(1,0,0)) 
wats.arima.3<-arima(watrain, order=c(1,1,0))
wats.arima.4<-arima(watrain, order=c(1,1,1))
wats.holt<-ets(watrain, model="AAN")

# PNW
auto.arima(pnwtrain) # returned (0,1,0)
pnwts.arima.0<-arima(pnwtrain, order=c(0,0,0))
pnwts.arima.1<-arima(pnwtrain, order=c(0,1,0)) #auto-arima model
pnwts.arima.2<-arima(pnwtrain, order=c(1,0,0)) 
pnwts.arima.3<-arima(pnwtrain, order=c(1,1,0))
pnwts.arima.4<-arima(pnwtrain, order=c(1,1,1))
pnwts.holt<-ets(pnwtrain, model="AAN")

# Total
# Total
auto.arima(totaltrain)
totalts.arima.0<-arima(totaltrain, order=c(0,0,0))
totalts.arima.1<-arima(totaltrain, order=c(0,1,0)) #auto-arima model
totalts.arima.2<-arima(totaltrain, order=c(1,0,0)) 
totalts.arima.3<-arima(totaltrain, order=c(1,1,0))
totalts.arima.4<-arima(totaltrain, order=c(1,1,1))
totalts.holt<-ets(totaltrain, model="AAN")

## Forecasting for each model at each location
# Alaska
ak.forecast.0<-forecast(akts.arima.0, h=testlength)
ak.forecast.1<-forecast(akts.arima.1, h=testlength)
ak.forecast.2<-forecast(akts.arima.2, h=testlength)
ak.forecast.3<-forecast(akts.arima.3, h=testlength)
ak.forecast.4<-forecast(akts.arima.4, h=testlength)
ak.forecast.5<-forecast(akts.holt, h=testlength) #Holt model

# California
ca.forecast.0<-forecast(cats.arima.0, h=testlength)
ca.forecast.1<-forecast(cats.arima.1, h=testlength)
ca.forecast.2<-forecast(cats.arima.2, h=testlength)
ca.forecast.3<-forecast(cats.arima.3, h=testlength)
ca.forecast.4<-forecast(cats.arima.4, h=testlength)
ca.forecast.5<-forecast(cats.holt, h=testlength)

# Oregon
or.forecast.0<-forecast(orts.arima.0, h=testlength)
or.forecast.1<-forecast(orts.arima.1, h=testlength)
or.forecast.2<-forecast(orts.arima.2, h=testlength)
or.forecast.3<-forecast(orts.arima.3, h=testlength)
or.forecast.4<-forecast(orts.arima.4, h=testlength)
or.forecast.5<-forecast(orts.holt, h=testlength)

#Washington
wa.forecast.0<-forecast(wats.arima.0, h=testlength)
wa.forecast.1<-forecast(wats.arima.1, h=testlength)
wa.forecast.2<-forecast(wats.arima.2, h=testlength)
wa.forecast.3<-forecast(wats.arima.3, h=testlength)
wa.forecast.4<-forecast(wats.arima.4, h=testlength)
wa.forecast.5<-forecast(wats.holt, h=testlength)

# PNW
pnw.forecast.0<-forecast(pnwts.arima.0, h=testlength)
pnw.forecast.1<-forecast(pnwts.arima.1, h=testlength)
pnw.forecast.2<-forecast(pnwts.arima.2, h=testlength)
pnw.forecast.3<-forecast(pnwts.arima.3, h=testlength)
pnw.forecast.4<-forecast(pnwts.arima.4, h=testlength)
pnw.forecast.5<-forecast(pnwts.holt, h=testlength)

# Total
total.forecast.0<-forecast(totalts.arima.0, h=testlength)
total.forecast.1<-forecast(totalts.arima.1, h=testlength)
total.forecast.2<-forecast(totalts.arima.2, h=testlength)
total.forecast.3<-forecast(totalts.arima.3, h=testlength)
total.forecast.4<-forecast(totalts.arima.4, h=testlength)
total.forecast.5<-forecast(totalts.holt, h=testlength)
```

#### Alaska
The best fit model for the Alaska chinook landing data was a Holt model with the following equation:
$y_t=l_{t-1}+b_{t-1}+\epsilon_t$\
$l_t=l_{t-1}+b_{t-1}+0.9999*\epsilon_t$\
$b_t =b_{t-1} + 0.0001*\epsilon_t$ 
This model indicates a slight downward trend that remains fairly consistent over time (beta=0.0001), and gives almost all weight to the most recent observations (alpha=0.9999).

```{r, echo=FALSE}
aktest.0<-accuracy(ak.forecast.0, aktest)
aktest.1<-accuracy(ak.forecast.1, aktest)
aktest.2<-accuracy(ak.forecast.2, aktest)
aktest.3<-accuracy(ak.forecast.3, aktest)
aktest.4<-accuracy(ak.forecast.4, aktest)
aktest.5<-accuracy(ak.forecast.5, aktest)
ak.forecast.summary<-matrix(data=c(aktest.0[2, "RMSE"],aktest.1[2,"RMSE"], aktest.2[2,"RMSE"],
                                   aktest.3[2,"RMSE"], aktest.4[2, "RMSE"],
                                   aktest.5[2, "RMSE"]), nrow=1, ncol=6)
colnames(ak.forecast.summary)<-c("0,0,0", "0,1,0", "1,0,0", "1,1,0", "1,1,1", "Holt")
rownames(ak.forecast.summary)<-c("RMSE")
print(ak.forecast.summary) #The Holt model had lowest RMSE
checkresiduals(akts.holt)
```

#### California
The best fit model for California was also a Holt model with the following equation:
$y_t=l_{t-1}+b_{t-1}+\epsilon_t$
$l_t=l_{t-1}+b_{t-1}+0.0005*\epsilon_t$
$b_t=b_{t-1}+0.0005*\epsilon_t$ 

The model indicates a slightly declining trend, but small beta value (beta=0.0005) suggest the level and trend are relatively stable over time. The model also puts very little weight on recent observations (alpha=0.0005).

```{r, echo=FALSE}
# California
catest.0<-accuracy(ca.forecast.0, catest)
catest.1<-accuracy(ca.forecast.1, catest)
catest.2<-accuracy(ca.forecast.2, catest)
catest.3<-accuracy(ca.forecast.3, catest)
catest.4<-accuracy(ca.forecast.4, catest)
catest.5<-accuracy(ca.forecast.5, catest)
ca.forecast.summary<-matrix(data=c(catest.0[2,"RMSE"],catest.1[2,"RMSE"], catest.2[2,"RMSE"],
                                   catest.3[2,"RMSE"], catest.4[2, "RMSE"],
                                   catest.5[2, "RMSE"]), nrow=1, ncol=6)
colnames(ca.forecast.summary)<-c("0,0,0","0,1,0", "1,0,0", "1,1,0", "1,1,1", "Holt")
rownames(ca.forecast.summary)<-c("RMSE")
print(ca.forecast.summary) #The Holt had the most accurate forecast
checkresiduals(cats.holt)
```
#### Oregon
The best fit model for Oregon was a first-order ARIMA model with no differencing and no moving average. The time series is stationary and auto-correlated around a mean of 7.5190, with observations being moderately dependent on recent values (ar1<1).
$y_t=0.7303*y_{t-1}+2.026+\epsilon_t$
$\epsilon_t~N(0,\sigma^2=0.1876)$

```{r, echo=FALSE}
ortest.0<-accuracy(or.forecast.0, ortest)
ortest.1<-accuracy(or.forecast.1, ortest)
ortest.2<-accuracy(or.forecast.2, ortest)
ortest.3<-accuracy(or.forecast.3, ortest)
ortest.4<-accuracy(or.forecast.4, ortest)
ortest.5<-accuracy(or.forecast.5, ortest)

or.forecast.summary<-matrix(data=c(ortest.0[2,"RMSE"], ortest.1[2,"RMSE"], ortest.2[2,"RMSE"],
                                   ortest.3[2,"RMSE"], ortest.4[2, "RMSE"],
                                   ortest.5[2, "RMSE"]), nrow=1, ncol=6)
colnames(or.forecast.summary)<-c("0,0,0","0,1,0", "1,0,0", "1,1,0", "1,1,1", "Holt")
rownames(or.forecast.summary)<-c("RMSE")
print(or.forecast.summary) #The ARIMA(1,0,0) model had the best forecast
checkresiduals(orts.arima.2)
```

#### Washington
Like Oregon, the best fit model for Washington was a first-order ARIMA model with no differencing and not moving average. The values differ slightly, with Washington having a slightly higher average than Oregon (intercept=7.9066), and autoregression component (ar1) of 0.8878, meaning the series shows strong positive correlation.

$y_t=0.8878*y_{t-1}+0.8871+\epsilon_t$
$\epsilon_t~N(0,\sigma^2=0.0436)$
```{r, echo=FALSE}
watest.0<-accuracy(wa.forecast.0, watest)
watest.1<-accuracy(wa.forecast.1, watest)
watest.2<-accuracy(wa.forecast.2, watest)
watest.3<-accuracy(wa.forecast.3, watest)
watest.4<-accuracy(wa.forecast.4, watest)
watest.5<-accuracy(wa.forecast.5, watest)
wa.forecast.summary<-matrix(data=c(watest.0[2,"RMSE"], watest.1[2,"RMSE"], watest.2[2,"RMSE"],
                                   watest.3[2,"RMSE"], watest.4[2, "RMSE"],
                                   watest.5[2, "RMSE"]), nrow=1, ncol=6)
colnames(wa.forecast.summary)<-c("0,0,0","0,1,0", "1,0,0", "1,1,0", "1,1,1", "Holt")
rownames(wa.forecast.summary)<-c("RMSE")
print(wa.forecast.summary) #The ARIMA(1,0,0) model had the best forecast
checkresiduals(wats.arima.2)
```
#### PNW Region
The model that best fit the Pacific Northwest region as a whole was a Holt model with the following equation:
$y_t=l_{t-1}+b_{t-1}+\epsilon_t$\
$l_t=l_{t-1}+b_{t-1}+0.9619*\epsilon_t$\
$b_t =b_{t-1} + 0.0001*\epsilon_t$ 

The small beta value suggests that the trend is stable over time, and the alpha value being close to 1 indicates that the model assigns greater weight to recent observations. The model has identified a slightly negative trend in chinook landings over time (b=-0.0108).

```{r, echo=FALSE}

# PNW
pnwtest.0<-accuracy(pnw.forecast.0, pnwtest)
pnwtest.1<-accuracy(pnw.forecast.1, pnwtest)
pnwtest.2<-accuracy(pnw.forecast.2, pnwtest)
pnwtest.3<-accuracy(pnw.forecast.3, pnwtest)
pnwtest.4<-accuracy(pnw.forecast.4, pnwtest)
pnwtest.5<-accuracy(pnw.forecast.5, pnwtest)
pnw.forecast.summary<-matrix(data=c(pnwtest.0[2,"RMSE"], pnwtest.1[2,"RMSE"], pnwtest.2[2,"RMSE"],
                                   pnwtest.3[2,"RMSE"], pnwtest.4[2, "RMSE"],
                                   pnwtest.5[2, "RMSE"]), nrow=1, ncol=6)
colnames(pnw.forecast.summary)<-c("0,0,0","0,1,0", "1,0,0", "1,1,0", "1,1,1", "Holt")
rownames(pnw.forecast.summary)<-c("RMSE")
print(pnw.forecast.summary) #The Holt model had lowest RMSE
checkresiduals(pnwts.holt)
```

#### Total chinook landings
The best fit model for total chinook landings was a first-order ARIMA with first difference since the data was non-stationary, and no seasonality:
$y_t = 0.8878*y_{t-1}+0.8871+\epsilon_t$\
$\epsilon_t~N(0,\sigma^2=0.0436)$\

The alpha value suggests weak, negative autocorrelation, but it is not statistically significant.
```{r, echo=FALSE}
totaltest.0<-accuracy(total.forecast.0, totaltest)
totaltest.1<-accuracy(total.forecast.1, totaltest)
totaltest.2<-accuracy(total.forecast.2, totaltest)
totaltest.3<-accuracy(total.forecast.3, totaltest)
totaltest.4<-accuracy(total.forecast.4, totaltest)
totaltest.5<-accuracy(total.forecast.5, totaltest)
total.forecast.summary<-matrix(data=c(totaltest.0[2,"RMSE"], totaltest.1[2,"RMSE"], totaltest.2[2,"RMSE"],
                                   totaltest.3[2,"RMSE"], totaltest.4[2, "RMSE"],
                                   totaltest.5[2, "RMSE"]), nrow=1, ncol=6)
colnames(total.forecast.summary)<-c("0,0,0","0,1,0", "1,0,0", "1,1,0", "1,1,1", "Holt")
rownames(total.forecast.summary)<-c("RMSE")
print(total.forecast.summary)
checkresiduals(totalts.arima.3)

# Plotting best forecast from each region
# Alaska
par(mfrow=c(3,2))
plot(ak.forecast.5, main="Alaksa Chinook Landings", 
     ylab="Log Volume Biometric Tons", xlab="Year")
points(aktest, pch=20)
# California
plot(ca.forecast.5, main="California Chinook Landings", 
     ylab="Log Volume Biometric Tons", xlab="Year")
points(catest, pch=20)
# Oregon
plot(or.forecast.2, main="Oregon Chinook Landings", 
     ylab="Log Volume Biometric Tons", xlab="Year")
points(ortest, pch=20)
# Washington
plot(wa.forecast.2, main="Washington Chinook Landings", 
     ylab="Log Volume Biometric Tons", xlab="Year")
points(watest, pch=20)
# PNW
plot(pnw.forecast.5, main="PNW Chinook Landings", 
     ylab="Log Volume Biometric Tons", xlab="Year")
points(pnwtest, pch=20)
#Total
plot(total.forecast.3, main="Total Chinook Landings", ylab="Log Volume Biometric Tons", xlab="Year")
points(totaltest, pch=20)
```

In summary, none of these models contained a seasonality component. The model structure differed slightly between regions, but none of the models found significant changes in trend over time. Many of them behaved functionally like a random walk. 

*Forecasting Accuracy Comparison*

When we compared forecasting accuracy between models fit to one region against test data from another region, we found some interesting results. In some cases, another state’s forecasted data was a better fit than the best fit model for that state. This was the case for Alaska, whose test data was more accurately forecasted by the ARIMA model fit to the Washington data. Only the Washington and Oregon model forecasts were the best fit for their state’s test data. This is the opposite of what I would have expected, given our original hypothesis that Washington and Oregon’s time series look very similar. I would have expected forecasts from these two states to not be accurate when tested against other states’ data, since California and Alaska’s data looked different, but the Washington and Oregon models performed fairly well across the board. The large regional models (PNW and total) did not perform well, which does support the idea that forecasts for salmon fisheries should be made on smaller regional scales that are ecologically relevant to salmon populations. 

```{r, echo=FALSE}
forecastrmse<-matrix(nrow=6, ncol=6)
rownames(forecastrmse)<-c("AK Forecast", "CA Forecast", 
                         "OR Forecast", "WA Forecast", "PNW Forecast", "Total Forecast")
colnames(forecastrmse)<-c("AK Model", "CA Model", 
                         "OR Model", "WA Model", "PNW Model", "Total Model")
forecastrmse[1,1]<-accuracy(ak.forecast.5, aktest)[2, "RMSE"]
forecastrmse[1,2]<-accuracy(ca.forecast.5, aktest)[2, "RMSE"]
forecastrmse[1,3]<-accuracy(or.forecast.2, aktest)[2, "RMSE"]
forecastrmse[1,4]<-accuracy(wa.forecast.2, aktest)[2, "RMSE"]
forecastrmse[1,5]<-accuracy(pnw.forecast.5, aktest)[2, "RMSE"]
forecastrmse[1,6]<-accuracy(total.forecast.3, aktest)[2, "RMSE"]
forecastrmse[2,1]<-accuracy(ak.forecast.5, catest)[2, "RMSE"]
forecastrmse[2,2]<-accuracy(ca.forecast.5, catest)[2, "RMSE"]
forecastrmse[2,3]<-accuracy(or.forecast.2, catest)[2, "RMSE"]
forecastrmse[2,4]<-accuracy(wa.forecast.2, catest)[2, "RMSE"]
forecastrmse[2,5]<-accuracy(pnw.forecast.5, catest)[2, "RMSE"]
forecastrmse[2,6]<-accuracy(total.forecast.3, catest)[2, "RMSE"]
forecastrmse[3,1]<-accuracy(ak.forecast.5, ortest)[2, "RMSE"]
forecastrmse[3,2]<-accuracy(ca.forecast.5, ortest)[2, "RMSE"]
forecastrmse[3,3]<-accuracy(or.forecast.2, ortest)[2, "RMSE"]
forecastrmse[3,4]<-accuracy(wa.forecast.2, ortest)[2, "RMSE"]
forecastrmse[3,5]<-accuracy(pnw.forecast.5,ortest)[2, "RMSE"]
forecastrmse[3,6]<-accuracy(total.forecast.3, ortest)[2, "RMSE"]
forecastrmse[4,1]<-accuracy(ak.forecast.5, watest)[2, "RMSE"]
forecastrmse[4,2]<-accuracy(ca.forecast.5, watest)[2, "RMSE"]
forecastrmse[4,3]<-accuracy(or.forecast.2, watest)[2, "RMSE"]
forecastrmse[4,4]<-accuracy(wa.forecast.2, watest)[2, "RMSE"]
forecastrmse[4,5]<-accuracy(pnw.forecast.5,watest)[2, "RMSE"]
forecastrmse[4,6]<-accuracy(total.forecast.3, watest)[2, "RMSE"]
forecastrmse[5,1]<-accuracy(ak.forecast.5, pnwtest)[2, "RMSE"]
forecastrmse[5,2]<-accuracy(ca.forecast.5, pnwtest)[2, "RMSE"]
forecastrmse[5,3]<-accuracy(or.forecast.2, pnwtest)[2, "RMSE"]
forecastrmse[5,4]<-accuracy(wa.forecast.2, pnwtest)[2, "RMSE"]
forecastrmse[5,5]<-accuracy(pnw.forecast.5,pnwtest)[2, "RMSE"]
forecastrmse[5,6]<-accuracy(total.forecast.3, pnwtest)[2, "RMSE"]
forecastrmse[6,1]<-accuracy(ak.forecast.5, totaltest)[2, "RMSE"]
forecastrmse[6,2]<-accuracy(ca.forecast.5, totaltest)[2, "RMSE"]
forecastrmse[6,3]<-accuracy(or.forecast.2, totaltest)[2, "RMSE"]
forecastrmse[6,4]<-accuracy(wa.forecast.2, totaltest)[2, "RMSE"]
forecastrmse[6,5]<-accuracy(pnw.forecast.5,totaltest)[2, "RMSE"]
forecastrmse[6,6]<-accuracy(total.forecast.3, totaltest)[2, "RMSE"]
print(forecastrmse)
```

## Question 2

The Forecast package's auto.arima function was used to determine which ARIMA models best fit the training data related to landings, prices, and value of fish landed. 

```{r, echo=FALSE}
#CA
CAlands.fit <- forecast::auto.arima(CAlands.train)
CAprice.fit <- forecast::auto.arima(CAprice.train)
CAvalue.fit <- forecast::auto.arima(CAvalue.train)

#OR
ORlands.fit <- forecast::auto.arima(ORlands.train)
ORprice.fit <- forecast::auto.arima(ORprice.train)
ORvalue.fit <- forecast::auto.arima(ORvalue.train)

#WA
WAlands.fit <- forecast::auto.arima(WAlands.train)
WAprice.fit <- forecast::auto.arima(WAprice.train)
WAvalue.fit <- forecast::auto.arima(WAvalue.train)
```

```{r, echo=FALSE}
CAlands.fit
CAprice.fit
CAvalue.fit

ORlands.fit
ORprice.fit
ORvalue.fit

WAlands.fit
WAprice.fit
WAvalue.fit
```

The models describing Chinook market data in California fail to find any seasonal trends, most likely due to data from California being much more scarce than data from Oregon and Washington. Moving forward California will be omitted from the process of forecasting landings, prices, and value.

```{r}
checkresiduals(ORlands.fit)
```
\newpage

```{r}
checkresiduals(ORprice.fit)
```
\newpage

```{r}
checkresiduals(ORvalue.fit)
```
\newpage

```{r}
checkresiduals(WAlands.fit)
```
\newpage

```{r}
checkresiduals(WAprice.fit)
```
\newpage

```{r}
checkresiduals(WAvalue.fit)
```

In general, models fitting landed weight, prices, and value show well behaved residuals distributed symmetrically around a mean of zero with little auto-correlation. The value models for both Oregon and Washington show the most autocorrelation and also a very tight distribution of error terms. 

```{r, echo=FALSE}
# project forecasts
#CA
CAlands.fc <- forecast::forecast(CAlands.fit, h=24)
plot(CAlands.fc)
points(CAlands.test)

CAprice.fc <- forecast::forecast(CAprice.fit, h=24)
plot(CAprice.fc)
points(CAprice.test)

CAvalue.fc <- forecast::forecast(CAvalue.fit, h=24)
plot(CAvalue.fc)
points(CAvalue.test)

#OR
ORlands.fc <- forecast::forecast(ORlands.fit, h=24)
plot(ORlands.fc)
points(ORlands.test)

ORprice.fc <- forecast::forecast(ORprice.fit, h=24)
plot(ORprice.fc)
points(ORprice.test)

ORvalue.fc <- forecast::forecast(ORvalue.fit, h=24)
plot(ORvalue.fc)
points(ORvalue.test)

#WA
WAlands.fc <- forecast::forecast(WAlands.fit, h=24)
plot(WAlands.fc)
points(WAlands.test)

WAprice.fc <- forecast::forecast(WAprice.fit, h=24)
plot(WAprice.fc)
points(WAprice.test)

WAvalue.fc <- forecast::forecast(WAvalue.fit, h=24)
plot(WAvalue.fc)
points(WAvalue.test)

#comparing forecasts
#CA
CAlands.compoundfc <- exp(CAlands.fc$mean)
CAvalue.compoundfc <- CAlands.compoundfc*CAprice.fc$mean
CAvalue.compoundfc
CAvalue.fc$mean
CAvalue.test

#OR
ORlands.compoundfc <- exp(ORlands.fc$mean)
ORvalue.compoundfc <- ORlands.compoundfc*ORprice.fc$mean
ORvalue.compoundfc
ORvalue.fc$mean
ORvalue.test

#WA
WAlands.compoundfc <- exp(WAlands.fc$mean)
WAvalue.compoundfc <- WAlands.compoundfc*WAprice.fc$mean
WAvalue.compoundfc
WAvalue.fc$mean
WAvalue.test

# plotting
#CA
plot(CAvalue.compoundfc)
points(CAvalue.test)

plot(CAvalue.fc$mean)
points(CAvalue.test)

#OR
plot(ORvalue.compoundfc)
points(ORvalue.test)

plot(ORvalue.fc$mean)
points(ORvalue.test)

#WA
plot(WAvalue.compoundfc)
points(WAvalue.test)

plot(WAvalue.fc$mean)
points(WAvalue.test)
```

# Discussion

In summary, none of these models contained a seasonality component. The model structure differed slightly between regions, but none of the models found significant changes in trend over time. Many of them behaved functionally like a random walk. 

When we compared forecasting accuracy between models fit to one region against test data from another region, we found some interesting results. In some cases, another state’s forecasted data was a better fit than the best fit model for that state. This was the case for Alaska, whose test data was more accurately forecasted by the ARIMA model fit to the Washington data. Only the Washington and Oregon model forecasts were the best fit for their state’s test data. This is the opposite of what I would have expected, given our original hypothesis that Washington and Oregon’s time series look very similar. I would have expected forecasts from these two states to not be accurate when tested against other states’ data, since California and Alaska’s data looked different, but the Washington and Oregon models performed fairly well across the board. The large regional models (PNW and total) did not perform well, which does support the idea that forecasts for salmon fisheries should be made on smaller regional scales that are ecologically relevant to salmon populations.

# Description of each team member's contributions

Each member of the team took on one issue and completed the code and the analysis. Halfway through, we consulted other team members for feedback and adjust analysis based on the joint discussion. All team members participated in the lab write up and review the results.

