---
title: Team 2 - Lab 1
subtitle: Forecasting with ARIMA and ETS models
author: "Lilac Hong"
date: April 20, 2025
output: 
  html_document:
    code-folding: true
    toc: true
    toc_float: true
---

```{r echo=FALSE}
library(ggplot2)
library(dplyr)
library(lubridate)
library(forecast)
```

# Data

Discuss what parts of the data sets you will work with.

# Question your team will address

What will your team do with the data? You can do many things with the data. The only constraints are that you

* fit ARIMA and ETS models. You can fit other models in addition to these if you want (lm, glm, gam, etc)
* do diagnostics for ARIMA and ETS models
* make forecasts
* test how good your forecasts or compare forecasts (many options here)

Example, "Our team decided to compare the accuracy of forecasts using best fit ARIMA models for WA, OR and CA commercial yearly landings. Our question is whether forecast accuracy is different for simple (1st order) versus 2nd to 4th order ARIMA models."

# Method you will use

## Initial plan

Describe what you plan to do to address your question.


## What you actually did

# Diagnostics and preliminary exploration

## Plot the data

Plot the data and discuss any obvious problems with stationarity from your visual test.

### Monthly

The monthly data is `chinook.month` and has 3 states `r unique(chinook.month$State)`.

```{r echo=FALSE}
# Create a date column (first of each month)
chinook.month <- chinook.month %>%
  mutate(Date = as.Date(paste(Year, Month, "01", sep = "-"), format = "%Y-%b-%d"))

# Plot with facets by State
ggplot(chinook.month, aes(x = Date, y = log.metric.tons)) +
  geom_line() +
  facet_wrap(~ State, ncol = 1) +  # 3 panels vertically
  labs(title = "Monthly Chinook Salmon Landings",
       x = "Date",
       y = "Log(Metric Tons)") +
  theme_minimal()
```

California (CA) exhibits strong signs of non-stationarity. There are large gaps in the data, including an extended period of missing values around 2008–2010. The means are higher in the early years and lower in more recent years, with noticeable inconsistencies in variance.

Oregon (OR) also appears to be non-stationary, though some seasonality is evident. The overall mean shows a slight downward trend over time.

Washington (WA) appears to be the most stable of the three. A clear seasonal pattern is present, and both the mean and variance remain relatively consistent over time, especially in comparison to CA and OR.

#### Monthly data summed across states
```{r}
chinook.month.wc <- chinook.month %>%
  group_by(Year, Month) %>%
  summarise(
    metric.tons = sum(metric.tons, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Date = as.Date(paste(Year, Month, "01", sep = "-"), format = "%Y-%b-%d"),
    log.metric.tons = log(metric.tons)
  ) %>%
  arrange(Date)

# Plot with facets by State
ggplot(chinook.month.wc, aes(x = Date, y = log.metric.tons)) +
  geom_line() +
  labs(title = "Monthly Chinook Salmon Landings, Sum of CA, OR, and WA",
       x = "Date",
       y = "Log(Metric Tons)") +
  theme_minimal()
```

After aggregating data across states, the time series appears not strictly stationary, though it shows greater stationarity from around 2000 onward. A strong and consistent seasonal pattern is evident. The early 1990s exhibit higher variability and a potential decline in the mean. From the 2000s forward, there is no clear long-term upward or downward trend.

### Yearly

The yearly data is `chinook.year` and has 6 states `r unique(chinook.year$State)`.

```{r yearly}
chinook.year.filtered <- chinook.year %>%
  filter(!State %in% c("Michigan", "Pennsylvania", "Alaska"))

ggplot(chinook.year.filtered, aes(x = Year, y = log.metric.tons)) +
  geom_line() +
  facet_wrap(~ State, ncol = 2, scales = "free_y") +
  labs(
    title = "Chinook Landings by State",
    x = "Year",
    y = "Log(Metric Tons)"
  ) +
  theme_minimal()
```

The California plot appears fairly stable until a dramatic drop in the most recent years, with generally consistent variance prior to the decline. Therefore, the California time series exhibits non-stationarity due to this abrupt structural change.

In both the Oregon and Washington plots, there are visible changes in both the mean and variance over time, indicating that these time series are also non-stationary.


## Use ACF and PACF

Use the ACF and PACF plots to explore the time series. What did you learn? Also try decomposition.

```{r}
# Filter WA data
chinook.month.WA <- chinook.month %>%
  filter(State == "WA")

# Create time-series objects
ts_WA <- ts(chinook.month.WA, 
            start = c(1990,1), end = c(2022, 12), frequency = 12)

# ACF and PACF plots
par(mfrow=c(1,2))
acf(ts_WA[,"metric.tons"], na.action = na.pass, las = 1)
pacf(ts_WA[,"metric.tons"], na.action = na.pass, las = 1)
```

The spikes at lag 12 and 24 (1.0 and 2.0 here for monthly data) in both ACF and PACF highlight seasonal components. A sharp cut-off after lag 1, followed by minor or mostly insignificant values, suggests that an AR(1) model might capture the short-term structure.

```{r}
# Filter CA data
chinook.month.CA <- chinook.month %>%
  filter(State == "CA")

# Create time-series objects
ts_CA <- ts(chinook.month.CA, 
            start = c(1990,1), end = c(2022, 12), frequency = 12)

# ACF and PACF plots
par(mfrow=c(1,2))
acf(ts_CA[,"metric.tons"], na.action = na.pass, las = 1)
pacf(ts_CA[,"metric.tons"], na.action = na.pass, las = 1)
```



```{r}


ts_clean_WA <- tsclean(ts_WA[,"log.metric.tons"])
decomp_WA <- decompose(ts_clean_WA[,"log.metric.tons"], type = "additive")
```


## Test for stationarity

Run tests using `forecast::ndiffs` and discuss any stationarity issues that are revealed.

```{r}
ndiffs(ts_WA[, "log.metric.tons"])
ndiffs(ts_CA[, "log.metric.tons"]) 
```

If `ndiffs()` returns:
* `0`: Series is already stationary (no differencing needed).
* `1`: Series is non-stationary, needs one difference.
* `2`: Strong non-stationarity, needs two differences.



```{r}
fit_arima_WA <- auto.arima(ts_WA[, "metric.tons"])
summary(fit_arima_WA)

fit_ets_WA <- ets(ts_WA[, "metric.tons"])
summary(fit_ets_WA)

accuracy(fit_arima_WA)
accuracy(fit_ets_WA)

forecast_arima_WA <- forecast(fit_arima_WA, h = 10)
forecast_ets_WA <- forecast(fit_ets_WA, h = 10)

plot(forecast_arima_WA, main = "ARIMA Forecast")
plot(forecast_ets_WA, main = "ETS Forecast")
```

```{r}
library(mgcv)
fit_gam_WA <- gam(ts_WA[, "metric.tons"] ~ s(Date))
summary(fit_gam_WA)
plot(fit_gam_WA)
```

# Results

# Discussion

# Description of each team member's contributions

Example: "All team members helped decide on the goal and ran the analyses for the individual regions. Team members 2 & 3 wrote most of the code for the analysis of the regions. Team member 4 researched approaches for measuring accuracy of forecasts in [Hyndman & Athanasopoulos[OTexts.com/fpp2] and team member 2 added code for that to the methods. Team member 4 also researched tests for stationarity and worked with team member 2 to code that up. Team member 1 worked on the plotting section of the report using and adapting code that team member 3 wrote. All team members helped edit the report and wrote the discussion together."

