---
title: Lab 1 Forecasting with ARIMA and ETS models
author: Eli Holmes
output: 
  html_document:
    code-folding: true
    toc: true
    toc_float: true
---

```{r include=FALSE}
library(tidyverse)
options(dplyr.summarise.inform = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

For this lab you will use the material you have learned on ARMA and ETS models to explore features of time series of commercial landings of Chinook salmon in WA, CA, and OR. Then you will use these models to create forecasts and ask a research or forecasting question with those forecasts.

You will find this Rmd file in the [Fish550-2025 repository](https://github.com/atsa-es/fish550-2025/tree/main/Lab-1). Clone the repository and then make sure you can Knit the `Lab1-ARIMA-2025.Rmd` file on your computer. Install any packages that it complains about.

## Teams

1. Team 1: Anna, Brian, Nicole
2. Team 2: Matheus, Lilac, Mary
3. Team 3: Ben, Emma, Tina

## References

Holmes, E. E., M. D. Scheuerell, and E. J. Ward. Applied Time Series Analysis for Fisheries and Environmental data. [Box-Jenkins Method chapter](https://atsa-es.github.io/atsa-labs/chap-boxjenkins-.html) and [time series chapter](https://atsa-es.github.io/atsa-labs/chap-tslab.html
).

Holmes, E. E. (2020) Fisheries Catch Forecasting <https://fish-forecast.github.io/Fish-Forecast-Bookdown>

Hyndman, R.J., & Athanasopoulos, G. (2018) Forecasting: principles and practice, 2nd edition, OTexts: Melbourne, Australia. <https://otexts.com/fpp2/>.

Last year's team write-ups [Fish550-2023](https://atsa-es.github.io/fish550-2023/Lab-1/Lab1-ARIMA.html)

Plus the lecture material on the ATSA website.

## Type of questions you might ask

"Compare the accuracy of forecasts using ARIMA models versus best fit ETS models. Are some types better than others for forecast accuracy?"

"Compare the accuracy of forecasts trained on log landings versus total landings."

"Do the accuracy of forecasts change over time?"

"Look at the autoregressive structure of for the 3 states. Are there differences by region (WA vs OR vs CA)?"

"Compare the forecasts using 5, 10, 15, and 20 years of training data. Does forecast accuracy increase with more training data?"

"Compare the forecasts using monthly versus yearly data. Are forecasts better using one or the other?"

"There is something off with the California yearly data. How would you deal with that?"

"Using monthly data, are forecasts for some months better than others?"

"How does forecast accuracy change the farther out you try to forecast?"

"Create 1-year forecasts using all the data (for your state or all if you want). Is forecast error correlated with the PDO or other ocean indices?"

## Chinook commercial landings data

The dataset has commercial landings data for WA, OR and CA. There is a monthly and a yearly dataset.

Load the data. The data is in the `atsalibrary` R package.

```{r message=FALSE, warning=FALSE}
# if you need to install
# install.packages('atsalibrary', repos = c('https://atsa-es.r-universe.dev', 'https://cloud.r-project.org'))
library(atsalibrary)
```

### Monthly

The monthly data is `chinook.month` and has 3 states `r unique(chinook.month$State)`.

```{r echo=FALSE}
library(ggplot2)
library(dplyr)
library(lubridate)

# Create a date column (first of each month)
chinook.month <- chinook.month %>%
  mutate(Date = as.Date(paste(Year, Month, "01", sep = "-"), format = "%Y-%b-%d"))

# Plot with facets by State
ggplot(chinook.month, aes(x = Date, y = log.metric.tons)) +
  geom_line() +
  facet_wrap(~ State, ncol = 1) +  # 3 panels vertically
  labs(title = "Monthly Chinook Salmon Landings",
       x = "Date",
       y = "Log(Metric Tons)") +
  theme_minimal()
```

Monthly data summed across states.

```{r}
library(dplyr)
library(lubridate)

chinook.month.wc <- chinook.month %>%
  group_by(Year, Month) %>%
  summarise(
    metric.tons = sum(metric.tons, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Date = as.Date(paste(Year, Month, "01", sep = "-"), format = "%Y-%b-%d"),
    log.metric.tons = log(metric.tons)
  ) %>%
  arrange(Date)
```

### Yearly

The yearly data is `chinook.year` and has 6 states `r unique(chinook.year$State)`.

```{r}
library(dplyr)
library(ggplot2)

chinook.year.filtered <- chinook.year %>%
  filter(!State %in% c("Michigan", "Pennsylvania"))

ggplot(chinook.year.filtered, aes(x = Year, y = log.metric.tons)) +
  geom_line() +
  facet_wrap(~ State, ncol = 2, scales = "free_y") +
  labs(
    title = "Chinook Landings by State",
    x = "Year",
    y = "Log(Metric Tons)"
  ) +
  theme_minimal()
```

## Example analysis

Get one time series and split into train and test. Each with 10 years.

```{r}
library(dplyr)

# Create a month number from the abbreviated month name
dat <- chinook.month %>%
  filter(State == "WA") %>%
  mutate(
    lnreturns = log.metric.tons,
    year = Year,
    month = match(Month, month.abb)  # Convert "Jan", "Feb", ... to 1, 2, ...
  ) %>%
 dplyr::select(year, month, lnreturns)


# Create the time series object with monthly frequency
datts <- ts(dat$lnreturns, start = c(dat$year[1], dat$month[1]), frequency = 12)

# Create 10-year train and test windows
train <- window(datts, start = c(dat$year[1], dat$month[1]), end = c(dat$year[1]+9, dat$month[1]))
test <- window(datts, start = c(dat$year[1]+10, dat$month[1]), end = c(dat$year[1]+19, dat$month[1]))
```

```{r}
train
```


Fit a model with `auto.arima()` in the forecast package.
```{r message=FALSE}
library(forecast)
mod <- auto.arima(train)
mod
```

Plot a 10-year forecast against the test data.
```{r}
library(zoo)
fr <- forecast(mod, h=10*12)
autoplot(fr) + geom_point(aes(x=x, y=y), data=fortify(test))
```
## Train / test for yearly data

```{r}
dat <- chinook.year %>%
  filter(State == "Washington") %>%
  mutate(lnreturns = log.metric.tons,
         year = Year) %>%
  dplyr::select(year, lnreturns)
datts <- ts(dat$lnreturns, start=dat$year[1])
train <- window(datts, dat$year[1], dat$year[1]+9)
test <- window(datts, dat$year[1]+10, dat$year[1]+10+9)
```

```{r}
train
```
```{r}

```

```{r}
```{r define-clean-chinookts}
chinook.month.wc <- chinook.month %>%
  group_by(Year, Month) %>%
  summarise(metric.tons = sum(metric.tons, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    log.metric.tons = ifelse(metric.tons <= 0, NA, log(metric.tons)),
    Date = as.Date(paste(Year, Month, "01", sep = "-"), format = "%Y-%m-%d")
  ) %>%
  arrange(Date)

chinookts <- ts(
  chinook.month.wc$log.metric.tons,
  start = c(1990, 1),
  frequency = 12
)

```

```{r}
```{r fit-auto-arima-class}
traindat <- window(chinookts, start = c(1990, 10), end = c(1998, 12))
mod_class <- auto.arima(traindat)
mod_class
```

```{r}
## Manual ARIMA(1,1,1) model

```{r fit-manual-arima-111}
mod_manual_111 <- Arima(traindat, order = c(1,1,1))
mod_manual_111
```

```{r}
## Manual ARIMA(2,0,0) model

```{r fit-manual-arima-200}
mod_manual_200 <- Arima(traindat, order = c(2,0,0))
mod_manual_200
```

```{r}
## Forecast all models 1 year ahead

```{r forecast-all-models}
library(forecast)

fc_class <- forecast(mod_class, h = 12)
fc_111   <- forecast(mod_manual_111, h = 12)
fc_200   <- forecast(mod_manual_200, h = 12)
```

```{r}
## Plot all forecasts together

```{r plot-forecast-comparison}
autoplot(fc_class) +
  autolayer(fc_111$mean, series = "ARIMA(1,1,1)", color = "red") +
  autolayer(fc_200$mean, series = "ARIMA(2,0,0)", color = "blue") +
  ggtitle("Forecast Comparison: Auto ARIMA vs Manual Models") +
  ylab("Log(Metric Tons)") +
  xlab("Time") +
  theme_minimal()
```

```{r}
```{r plot-auto-arima}
autoplot(fc_class) +
  ggtitle("Auto ARIMA Forecast") +
  ylab("Log(Metric Tons)") +
  theme_minimal()
```

```

```
```{r}
```{r plot-manual-111}
autoplot(fc_111) +
  ggtitle("Manual ARIMA(1,1,1) Forecast") +
  ylab("Log(Metric Tons)") +
  theme_minimal()
```

```

```
```{r}
```{r plot-manual-200}
autoplot(fc_200) +
  ggtitle("Manual ARIMA(2,0,0) Forecast") +
  ylab("Log(Metric Tons)") +
  theme_minimal()
```





###(Mary)How does the structure of an ARIMA model affect its ability to forecast salmon landings when trained on 25 years of data?

##(Mary) Which ARIMA model structure best captures long-term patterns in Chinook salmon landings, and why?

(Mary) After discussing with my teamates and reviewing best results, I decided to follow Matheus' approach of training on a 25-year dataset, since our findings showed this time frame produced the most reliable forecasts. For my section of the lab, I wil use the same framework for consistency across our analyses. My goal was to define a 25-year training set (1980–2004) and use it to fit multiple ARIMA models — starting with an automatically selected model using auto.arima(), followed by two manually specified ARIMA models. This allowed me to compare how model structure impacts forecast performance when applied to monthly Chinook salmon landings.

```{r}
```{r clean-monthly-data-1980}
library(atsalibrary)
library(forecast)
library(zoo)
library(tidyverse)
library(ggpubr)

# Create Date column and safely log-transform metric.tons
chinook.month <- chinook.month %>%
  mutate(
    Date = as.Date(paste(Year, Month, "01", sep = "-"), format = "%Y-%b-%d"),
    log.metric.tons = ifelse(metric.tons <= 0, NA, log(metric.tons))
  )
```

```{r}
```{r ts-and-train-test-25yrs}
# Create time series from monthly log data starting in 1980
month_ts <- ts(chinook.month$log.metric.tons, start = c(1980, 1), frequency = 12)

# 25-year training: Jan 1980 – Dec 2004
train_25 <- window(month_ts, start = c(1980, 1), end = c(2004, 12))

# Test: Jan 2005 – Dec 2022 (or last available year)
test_25 <- window(month_ts, start = c(2005, 1), end = c(2022, 12))
```

(Mary) In the following section, I fit three different ARIMA models to the 25-year training data from above:
- The first is chosen automatically using the auto.arima() function, which finds the best model based on AIC.
- The second is a manually defined ARIMA(1,1,1) model, which includes both AR and MA components with one difference.
- The third is ARIMA(2,0,0), a manually specified model with more autoregressive terms but no differencing.
These manual models are included to test whether expert-chosen structures can outperform the automatically selected model.

```{r}
```{r fit-three-arima-25yrs}
# Fit auto ARIMA to 25-year training data
mod_auto_25 <- auto.arima(train_25)

# Manual ARIMA(1,1,1)
mod_manual_111 <- Arima(train_25, order = c(1,1,1))

# Manual ARIMA(2,0,0)
mod_manual_200 <- Arima(train_25, order = c(2,0,0))
```

This section generates forecasts from all three fitted models for the same length as the test set.
The output will be used to visualize and evaluate each model's ability to accurately forecast future events.
```{r}
```{r forecast-arima-25yrs}
# Forecast all models for the length of the test set
fc_auto_25   <- forecast(mod_auto_25, h = length(test_25))
fc_manual_111 <- forecast(mod_manual_111, h = length(test_25))
fc_manual_200 <- forecast(mod_manual_200, h = length(test_25))
```

This plot shows the forecast from the auto ARIMA model overlaid with the actual observed data from the test period.The shaded area represents the model’s forecast uncertainty.This visual comparison helps us assess how closely the model captures the true pattern of Chinook landings.
```{r}
```{r plot-forecast-auto-25}
autoplot(fc_auto_25) +
  autolayer(test_25, series = "Observed", color = "black") +
  ggtitle("Auto ARIMA Forecast (25-Year Training)") +
  ylab("Log(Metric Tons)") +
  theme_minimal()
```

This plot shows the forecast from the manual ARIMA(1,1,1) model, with the test data overlaid in black. This model assumes moderate memory and differencing, and is commonly used in many forecasting applications.Comparing this plot to the auto model helps us evaluate whether a standard manual model performs better or worse.
```{r}
```{r plot-forecast-manual-111-25}
autoplot(fc_manual_111) +
  autolayer(test_25, series = "Observed", color = "black") +
  ggtitle("Manual ARIMA(1,1,1) Forecast (25-Year Training)") +
  ylab("Log(Metric Tons)") +
  theme_minimal()
```

This plot shows the forecast from the manual ARIMA(2,0,0) model, which relies more heavily on autoregressive terms. There is no differencing in this model, so we are testing how well the model can handle trend-like behavior directly. Again, the black line shows actual test data, providing a basis for visual evaluation.
```{r}
```{r plot-forecast-manual-200-25}
autoplot(fc_manual_200) +
  autolayer(test_25, series = "Observed", color = "black") +
  ggtitle("Manual ARIMA(2,0,0) Forecast (25-Year Training)") +
  ylab("Log(Metric Tons)") +
  theme_minimal()
```

This chunk calculates a set of forecast accuracy metrics for each model using the test set as the reference. These include Root Mean Square Error (RMSE), Mean Absolute Error (MAE), and MASE, among others.Lower values indicate better predictive performance.
The results are combined into a single table so we can directly compare which model forecasted the test data most accurately.
```{r}
```{r compare-accuracy-25yr}
# Compare forecast accuracy using test data
acc_auto   <- accuracy(fc_auto_25, test_25)
acc_111    <- accuracy(fc_manual_111, test_25)
acc_200    <- accuracy(fc_manual_200, test_25)

# Combine into a data frame for easy comparison
accuracy_results <- rbind(
  `Auto ARIMA` = acc_auto,
  `Manual ARIMA(1,1,1)` = acc_111,
  `Manual ARIMA(2,0,0)` = acc_200
)

accuracy_results
```

```
```{r}
```{r clean-accuracy-table}
# Convert accuracy outputs to data frames
acc_df <- function(model_name, acc_matrix) {
  data.frame(
    Model = model_name,
    Dataset = c("Training", "Test"),
    ME = acc_matrix[, "ME"],
    RMSE = acc_matrix[, "RMSE"],
    MAE = acc_matrix[, "MAE"],
    MASE = acc_matrix[, "MASE"],
    ACF1 = acc_matrix[, "ACF1"],
    TheilsU = acc_matrix[, "Theil's U"]
  )
}

# Combine all models into one clean data frame
accuracy_table <- rbind(
  acc_df("Auto ARIMA", acc_auto),
  acc_df("Manual ARIMA(1,1,1)", acc_111),
  acc_df("Manual ARIMA(2,0,0)", acc_200)
)

# Display the table
kable(accuracy_table, digits = 3, caption = "Forecast Accuracy Metrics by Model and Dataset")
```

Based on the test data, the auto ARIMA model made the most accurate forecasts. It had the smallest errors, meaning its predictions were closest to the actual salmon landings. The manual ARIMA(1,1,1) model performed the worst. This model includes both autoregressive (AR) and moving average (MA) components, along with differencing (I), which means it's constantly adjusting based on past values and past errors. In this case, the model may have tried too hard to “chase” short-term fluctuations, which led to poor performance on future data — a problem known as overfitting.

The ARIMA(2,0,0) model was slightly better, but it had two AR terms and no differencing. That means it relied mostly on momentum from the last two months to predict the next one, without adjusting for longer-term trends in the data. Since the salmon landings likely have gradual seasonal or declining trends, not using differencing made this model less flexible. In contrast, the auto ARIMA model included just the right amount of seasonal adjustment and trend (via drift), keeping it simple and better suited to the actual structure of the data.



