---
title: Team 3 - Lab 5
subtitle: Dynamic Linear Models (DLMs) - Time-varying productivity of salmon
author: "Emma Christman, Brian McGreal, Anna Sulc"
date: May 27, 2025
output: 
  html_document:
    code-folding: true
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
options(dplyr.summarise.inform = FALSE)
```


```{r silent=TRUE, include=FALSE}
library(atsalibrary)
library(dplyr)
library(ggplot2)
library(here)
library(knitr)
library(MARSS)
library(tidyverse)
```

# Data

## Load the data

This exercise involves spawner-recruit data for sockeye salmon (Oncorhynchus nerka) from the Kvichak River in SW Alaska that span the years 1952-2005. We have a dataframe with columns for brood year (`brood_year`), number of spawners (`spawners`), number of recruits (`recruits`) and PDO at year $t-2$ in summer (`pdo_summer_t2`) and in winter (`pdo_winter_t2`).



```{r dlm-load-atsa, echo=FALSE}
## get data
data(KvichakSockeye, package="atsalibrary")
data <- KvichakSockeye

head(data)

```

We'll take a quick look at spawners and recruits over time in the data.

```{r plot, echo = FALSE, message=FALSE, warnings=FALSE}
#Plot Data 
suppressWarnings({
colors <- c("Recruits" = "blue", "Spawners" = "black")
gorp <- ggplot(data=data, aes(x=brood_year)) +
  geom_line(aes(y = spawners), color = "black", size = 1) +
  geom_line(aes(y = recruits), color = "blue", size = 1) +    
  labs(title = "Kvichak River Sockeye, 1950-1989",
       subtitle = "Spawners (black) and Recruits (blue)",
       x="",
       y = "Fish",
       color = "Legend") +
    scale_color_manual(values = colors)
gorp
})
```

## Wrangle the data

Before constructing DLMs to be solved using a MARSS model framework, some data cleaning is in order. A variable $y$ is constructed by taking the natural log of recruits divided by spawners ($\ln(R_t/S_t)$). The spawner variable $S_t$ is then scaled to be used as an inidcator variable in the model specification. Finally, because $S_t$ will act as an indicator variable, any observations with missing values in $S_t$ must be dropped, meaning in this case losing the first four observations in the time series. Because $y$ is constructed using the term $(R_t/S_t)$, any observations missing recruit values will also be dropped, as the resulting $y$ terms are NA, meaning we lose the last seven observations in the time series. This leaves us with complete data for 43 time steps. 

```{r wrangle_data}
# transform data
data$y <- log(data$recruits/data$spawners)
data <- data[!is.na(data$spawners),]
data <- data[!is.na(data$recruits),]

years <- data$brood_year
TT <- length(years)
dat <- matrix(data$y, nrow=1)

s <- data$spawners
s_z <- matrix((s - mean(s))/sqrt(var(s)), nrow = 1)
m <- dim(s_z)[1] + 1
```

# Methods

### Base Model structure

We began by fitting a reduced form of Ricker's model that including only a time-varying intercept ($\alpha_t$) as well as a time-varying effect of spawner abundance ($/beta_t$). That is,

\[
\ln\left(\frac{R_t}{S_t}\right) = \ln(\alpha_t) + \beta_tS_t+ v_t
\]

By setting $\ln\left(\frac{R_t}{S_t}\right)$ = $y_t$ and $\ln(\alpha_t)$ = $a_t$ we are left with the following observation equation:

\begin{align*}
\text{Linear Model:} \quad & y_t = a_t + \beta_tS_t+ v_t \\
                          & \quad v_t \sim \mathcal{N}(0, R) \\[1em]
\end{align*}

By then assuming that $a_t$ and $\beta_t$ are results of two unoberserved underlying processes, we may specify the following state equations:

\begin{align*}
\text{Parameters:} \quad       & a_t = a_{t-1} + w_{a,t}, \quad w_t \sim \mathcal{N}(0, Q_a) \\
                          & \beta_t = \beta_{t-1} + w_{\beta,t}, \quad w_t \sim \mathcal{N}(0, Q_\beta) \\ 
\end{align*}

This model assumes no density-dependent survival in that the number of recruits is an ascending function of spawners. We did not include any environmental variables for this step. 


```{r dlm-build}
# build MARSS model
# process
B <- diag(m)
U <- matrix(0, nrow=m, ncol=1)
Q <- matrix(list(0), m, m)
diag(Q) <- c("q.alpha", "q.beta")

# obs
Z <- array(NA, c(1, m, TT))
Z[1,1,] <- rep(1, TT)
Z[1,2,] <- s_z
A <- matrix(0)
R <- matrix("r")

# initital values & controls
inits <- list(x0 = matrix(c(0,0), nrow = m))
con_list <- list(maxit = 2146)
  # 2146 seems to be the minimum number of iterations at which all beta parameter varainces converge to zero

# model list
dlm.list <- list(B=B, U=U, Q=Q, Z=Z, A=A, R=R)

# fit model
if(!file.exists(here::here("Lab-5", "Team-3", "data", "dlm.rds"))){
  dlm <- MARSS(dat, model= dlm.list, inits = inits, method = "kem", control = con_list)
  saveRDS(dlm, file=here::here("Lab-5", "Team-3", "data", "dlm.rds"))
}
# load in ssm.phyto_month
dlm <- readRDS(file=here::here("Lab-5", "Team-3", "data", "dlm.rds"))
```

```{r dlm, echo = FALSE}
dlm
```

Below is a plot of estimated $a_t$ values across the study period. The model's 95% confidence interval is fairly tight around the mean estimate at all points.


```{r plotA, echo=FALSE}
## plot
alpha <- as.numeric(dlm$states[1,])
alpha.se <- as.numeric(dlm$states.se[1,])

# alpha
plotA <- plot(alpha~data$brood_year,
              main = "Time varying alpha estimates, 2146 iterations",
              xlab = "",
              ylab = "Alpha estimate",
              type='l', 
              ylim=c(-3,3), 
              xlim= c(1956, 2005))
lines(alpha+1.96*alpha.se~data$brood_year, lty="dashed")
lines(alpha-1.96*alpha.se~data$brood_year, lty="dashed")
```

When estimating a time varying $\beta$ term we found an interesting pattern. As we increased the maximum number of iterations the model could run through, the term $Q_\beta$ converges to zero. Below we have plots for models run with 500, 1000, 1750, and 2146 iterations (2146 being the minimum number of iterations necessary for all parameter estimates to converge.)

```{r plotB500, echo=FALSE}
# new dls - 500 iter
con_listALT <- list(maxit = 500)

# fit model
if(!file.exists(here::here("Lab-5", "Team-3", "data", "dlm500.rds"))){
  dlm500 <- MARSS(dat, model= dlm.list, inits = inits, method = "kem", control = con_listALT)
  saveRDS(dlm500, file=here::here("Lab-5", "Team-3", "data", "dlm500.rds"))
}
# load in ssm.phyto_month
dlm500 <- readRDS(file=here::here("Lab-5", "Team-3", "data", "dlm500.rds"))

#grab estimates
beta <- as.numeric(dlm500$states[2,])
beta.se <- as.numeric(dlm500$states.se[2,])

# beta
plot(beta~data$brood_year,
     main = "Time varying beta estimates, 500 iterations",
     xlab = "",
     ylab = "Beta estimate",
     type='l', 
     ylim=c(-0.5, 0.5),
     xlim= c(1956, 2005))
lines(beta+1.96*beta.se~data$brood_year, lty="dashed")
lines(beta-1.96*beta.se~data$brood_year, lty="dashed")
```

```{r plotB1000, echo=FALSE}
# new dls - 500 iter
con_listALT <- list(maxit = 1000)

# fit model
if(!file.exists(here::here("Lab-5", "Team-3", "data", "dlm1000.rds"))){
  dlm1000 <- MARSS(dat, model= dlm.list, inits = inits, method = "kem", control = con_listALT)
  saveRDS(dlm1000, file=here::here("Lab-5", "Team-3", "data", "dlm1000.rds"))
}
# load in ssm.phyto_month
dlm1000 <- readRDS(file=here::here("Lab-5", "Team-3", "data", "dlm1000.rds"))

#grab estimates
beta <- as.numeric(dlm1000$states[2,])
beta.se <- as.numeric(dlm1000$states.se[2,])

# beta
plot(beta~data$brood_year,
     main = "Time varying beta estimates, 1000 iterations",
     xlab = "",
     ylab = "Beta estimate",
     type='l', 
     ylim=c(-0.5, 0.5),
     xlim= c(1956, 2005))
lines(beta+1.96*beta.se~data$brood_year, lty="dashed")
lines(beta-1.96*beta.se~data$brood_year, lty="dashed")
```

```{r plotB1750, echo=FALSE}
# new dls - 500 iter
con_listALT <- list(maxit = 1750)

# fit model
if(!file.exists(here::here("Lab-5", "Team-3", "data", "dlm1750.rds"))){
  dlm1750 <- MARSS(dat, model= dlm.list, inits = inits, method = "kem", control = con_listALT)
  saveRDS(dlm1750, file=here::here("Lab-5", "Team-3", "data", "dlm1750.rds"))
}
# load in ssm.phyto_month
dlm1750 <- readRDS(file=here::here("Lab-5", "Team-3", "data", "dlm1750.rds"))

#grab estimates
beta <- as.numeric(dlm1750$states[2,])
beta.se <- as.numeric(dlm1750$states.se[2,])

# beta
plot(beta~data$brood_year,
     type='l',
     main = "Time varying beta estimates, 1750 iterations",
     xlab = "",
     ylab = "Beta estimate",
     ylim=c(-0.5, 0.5),
     xlim= c(1956, 2005))
lines(beta+1.96*beta.se~data$brood_year, lty="dashed")
lines(beta-1.96*beta.se~data$brood_year, lty="dashed")
```


```{r plotB2150, echo=FALSE}
# beta
beta <- as.numeric(dlm$states[2,])
beta.se <- as.numeric(dlm$states.se[2,])

plot(beta~data$brood_year,
     main = "Time varying beta estimates, 2146 iterations",
     xlab = "",
     ylab = "Beta estimate",
     type='l', 
     ylim=c(-0.5, 0.5),
     xlim= c(1956, 2005))
lines(beta+1.96*beta.se~data$brood_year, lty="dashed")
lines(beta-1.96*beta.se~data$brood_year, lty="dashed")
```

This may be due to the number of spawners having some "true" linear fixed effect on $\ln(R_t/S_t)$ which does not vary through time even when that flexibility is allowed by a modeling approach. 


### Covariates

Next,we fitted the same Ricker model but included covariate terms meant to control for the effect of ocean climate on the outcome spawner-recruit ratio. The model includes $d_t$ as input covariate data and $D$ as the effect of said covariates on the coutcome.

The equation for this model is now: 

\begin{align*}
\text{Linear Model:} \quad & y_t = a_t + \beta_tS_t+ Dd_t + v_t \\
                          & \quad v_t \sim \mathcal{N}(0, R) \\[1em]
\text{Parameters:} \quad       & a_t = a_{t-1} + w_{a,t}, \quad w_t \sim \mathcal{N}(0, Q_a) \\
                          & \beta_t = \beta_{t-1} + w_{\beta,t}, \quad w_t \sim \mathcal{N}(0, Q_\beta) \\ 
\end{align*}

We run models with time-varying $a_t$ and $\beta_t$ terms, allowing for 2146 iterations (the minimum number required for $Q_\beta$ to converge to zero). Three models are fit, one including two-year lagged winter PDO, one including two-year lagged summer PDO, and one including both. These will be compared with the null model fit in the last section. 

WINTER PDO

```{r dlmPw, echo = FALSE}
# winter pdo
D <- matrix("pdo", 1, 1)
pdo_w <- as.data.frame(matrix(data$pdo_winter_t2, nrow=1, byrow=TRUE))
pdo_w <- data.matrix(pdo_w)

dlmPw.list <- list(B=B, U=U, Q=Q, Z=Z, A=A, R=R, D=D, d=pdo_w)

if(!file.exists(here::here("Lab-5", "Team-3", "data", "dlmPw.rds"))){
  dlmPw <- MARSS(dat, model= dlmPw.list, inits = inits, method = "kem", control = con_list)
  saveRDS(dlmPw, file=here::here("Lab-5", "Team-3", "data", "dlmPw.rds"))
}

# load in ssm.phyto_month
dlmPw <- readRDS(file=here::here("Lab-5", "Team-3", "data", "dlmPw.rds"))
dlmPw
```

SUMMER PDO

Of interest here, this model converges quickly but unlike in the other three specifications, $Q_\beta$ does not converge at zero. 

```{r dlmPs, echo = FALSE}
# summer pdo
D <- matrix("pdo", 1, 1)
pdo_s <- as.data.frame(matrix(data$pdo_summer_t2, nrow=1, byrow=TRUE))
pdo_s <- data.matrix(pdo_s)

dlmPs.list <- list(B=B, U=U, Q=Q, Z=Z, A=A, R=R, D=D, d=pdo_s)

if(!file.exists(here::here("Lab-5", "Team-3", "data", "dlmPs.rds"))){
  dlmPs <- MARSS(dat, model= dlmPs.list, inits = inits, method = "kem", control = con_list)
  saveRDS(dlmPs, file=here::here("Lab-5", "Team-3", "data", "dlmPs.rds"))
}

# load in ssm.phyto_month
dlmPs <- readRDS(file=here::here("Lab-5", "Team-3", "data", "dlmPs.rds"))
dlmPs
```

WINTER AND SUMMER PDO

```{r dlmPb, echo=FALSE}
# winter and summer pdo
D <- matrix(c("pdoW", "pdoS"), 2, 1)
pdo_b <- rbind(pdo_w, pdo_s)
pdo_b <- data.matrix(pdo_b)

dlmPb.list <- list(B=B, U=U, Q=Q, Z=Z, A=A, R=R, D=D, d=pdo_b)

if(!file.exists(here::here("Lab-5", "Team-3", "data", "dlmPb.rds"))){
  dlmPb <- MARSS(dat, model= dlmPb.list, inits = inits, method = "kem", control = con_list)
  saveRDS(dlmPb, file=here::here("Lab-5", "Team-3", "data", "dlmPb.rds"))
}
# load in ssm.phyto_month
dlmPb <- readRDS(file=here::here("Lab-5", "Team-3", "data", "dlmPb.rds"))
dlmPb
```

Model fit is compared using AICc. These models are all very close in AICc score but the null model with no PDO covariates exhibits the best fit. This would seem to imply that the addition of PDO does not improve model fit substantially enough to make up for the additional parameters the model must estimate.

```{r aic, echo = FALSE}
## AICc
# an empty list to collect outputs
out_list <- list()
# Add each model's name and AICc 
out_list[[1]] <- data.frame(Model = "no covariates", AICc = dlm$AICc)
out_list[[2]] <- data.frame(Model = "winter PDO", AICc = dlmPw$AICc)
out_list[[3]] <- data.frame(Model = "summer PDO", AICc = dlmPs$AICc)
out_list[[4]] <- data.frame(Model = "winter and summer PDO", AICc = dlmPb$AICc)
out_list
```

Diagnostics on the on the models show above are compared here.
```{r diagnostics, echo = FALSE}

# diagnostics on base mode (no covariates)
resids_dlm <- residuals(dlm, type = "tT")

ggplot(subset(resids_dlm, name == "model"), aes(x = .std.resids)) +
  geom_histogram(bins = 20, fill = "skyblue", color = "black") +
  ggtitle("Histogram of Standardized Observation Residuals-dlm") +
  xlab("Standardized Residual")

ggplot(subset(resids_dlm, name == "model"), aes(sample = .std.resids)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  ggtitle("Q-Q Plot of Standardized Observation Residuals-dlm") +
  xlab("Theoretical Quantiles") +
  ylab("Sample Quantiles")

obs_resids_dlm <- subset(resids_dlm, name == "model")
dlm_resids <- na.omit(obs_resids_dlm$.std.resids)
acf(dlm_resids, main = "Autocorrelation of Standardized Observation Residuals_dlm")

# diagnostics on winter PDO model
resids_dlmPw <- residuals(dlmPw, type = "tT")

ggplot(subset(resids_dlmPw, name == "model"), aes(x = .std.resids)) +
  geom_histogram(bins = 20, fill = "skyblue", color = "black") +
  ggtitle("Histogram of Standardized Observation Residuals-dlmPw") +
  xlab("Standardized Residual")

ggplot(subset(resids_dlmPw, name == "model"), aes(sample = .std.resids)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  ggtitle("Q-Q Plot of Standardized Observation Residuals-dlmPw") +
  xlab("Theoretical Quantiles") +
  ylab("Sample Quantiles")

obs_resids_dlmPw <- subset(resids_dlmPw, name == "model")
dlmPw_resids <- na.omit(obs_resids_dlmPw$.std.resids)
acf(dlmPw_resids, main = "Autocorrelation of Standardized Observation Residuals_dlmPw")

# diagnostics on summer PDO model
resids_dlmPs <- residuals(dlmPs, type = "tT")

ggplot(subset(resids_dlmPs, name == "model"), aes(x = .std.resids)) +
  geom_histogram(bins = 20, fill = "skyblue", color = "black") +
  ggtitle("Histogram of Standardized Observation Residuals-dlmPs") +
  xlab("Standardized Residual")

ggplot(subset(resids_dlmPs, name == "model"), aes(sample = .std.resids)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  ggtitle("Q-Q Plot of Standardized Observation Residuals-dlmPs") +
  xlab("Theoretical Quantiles") +
  ylab("Sample Quantiles")

obs_resids_dlmPs <- subset(resids_dlmPs, name == "model")
dlmPs_resids <- na.omit(obs_resids_dlmPs$.std.resids)
acf(dlmPs_resids, main = "Autocorrelation of Standardized Observation Residuals_dlmPs")

# diagnostics on winter and summer PDO model
resids_dlmPb <- residuals(dlmPb, type = "tT")

ggplot(subset(resids_dlmPw, name == "model"), aes(x = .std.resids)) +
  geom_histogram(bins = 20, fill = "skyblue", color = "black") +
  ggtitle("Histogram of Standardized Observation Residuals-dlmPb") +
  xlab("Standardized Residual")

ggplot(subset(resids_dlmPb, name == "model"), aes(sample = .std.resids)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  ggtitle("Q-Q Plot of Standardized Observation Residuals-dlmPb") +
  xlab("Theoretical Quantiles") +
  ylab("Sample Quantiles")

obs_resids_dlmPb <- subset(resids_dlmPb, name == "model")
dlmPb_resids <- na.omit(obs_resids_dlmPb$.std.resids)
acf(dlmPb_resids, main = "Autocorrelation of Standardized Observation Residuals_dlmPb")

```


Similar to the null model, estiamtes of alpha in all models with covariates show fairly tight confidence intervals around a time-varying mean.

```{r covariancePLOT_alpha, echo = FALSE}
alphaPw <- as.numeric(dlmPw$states[1,])
betaPw <- as.numeric(dlmPw$states[2,])
alphaPw.se <- as.numeric(dlmPw$states.se[1,])
betaPw.se <- as.numeric(dlmPw$states.se[2,])

alphaPs <- as.numeric(dlmPs$states[1,])
betaPs <- as.numeric(dlmPs$states[2,])
alphaPs.se <- as.numeric(dlmPs$states.se[1,])
betaPs.se <- as.numeric(dlmPs$states.se[2,])

alphaPb <- as.numeric(dlmPb$states[1,])
betaPb <- as.numeric(dlmPb$states[2,])
alphaPb.se <- as.numeric(dlmPb$states.se[1,])
betaPb.se <- as.numeric(dlmPb$states.se[2,])

plot(alpha~data$brood_year,
     main = "Time varying alpha estimates",
     sub = "Null (black), Winter PDO (red), Summer PDO (blue), Full (green)",
     xlab = "",
     ylab = "Alpha estimate",
     type='l', 
     ylim=c(-3,3), 
     xlim= c(1956, 2005))
lines(alpha+1.96*alpha.se~data$brood_year, lty="dashed")
lines(alpha-1.96*alpha.se~data$brood_year, lty="dashed")

lines(alphaPw~data$brood_year, col = "red")
lines(alphaPw+1.96*alphaPw.se~data$brood_year, lty="dashed", col = "red")
lines(alphaPw-1.96*alphaPw.se~data$brood_year, lty="dashed", col = "red")

lines(alphaPs~data$brood_year, col = "blue")
lines(alphaPs+1.96*alphaPs.se~data$brood_year, lty="dashed", col = "blue")
lines(alphaPs-1.96*alphaPs.se~data$brood_year, lty="dashed", col = "blue")

lines(alphaPb~data$brood_year, col = "green")
lines(alphaPb+1.96*alphaPb.se~data$brood_year, lty="dashed", col = "green")
lines(alphaPb-1.96*alphaPb.se~data$brood_year, lty="dashed", col = "green")
```

Here we see that, in two covariate models along with the null model, $Q_\beta$ has converged to zero and the model has reutrned a time-invarying estimate of the effect of spawners on $\ln(R_t/S_t)$. The summer PDO model sees $Q_\beta$ converge at 0.0155, leaving the parameter some latitude to vary across the study period. This suggests that including summer PDO (and only summer PDO) controls for some of the time-varying structure of $\ln(R_t/S_t)$ and allows $\beta_t$ to vary throughout time. Nonetheless, the null model retains the best fit.

```{r covariancePLOT_beta, echo = FALSE}
# beta
plot(beta~data$brood_year,
     type='l', 
     main = "Time varying beta estimates",
     sub = "Null (black), Winter PDO (red), Summer PDO (blue), Full (green)",
     xlab = "",
     ylab = "Beta estimate",
     ylim=c(-0.5,0.5), 
     xlim= c(1956, 2005))
lines(beta+1.96*beta.se~data$brood_year, lty="dashed")
lines(beta-1.96*beta.se~data$brood_year, lty="dashed")

lines(betaPw~data$brood_year, col = "red")
lines(betaPw+1.96*betaPw.se~data$brood_year, lty="dashed", col = "red")
lines(betaPw-1.96*betaPw.se~data$brood_year, lty="dashed", col = "red")

lines(betaPs~data$brood_year, col = "blue")
lines(betaPs+1.96*betaPs.se~data$brood_year, lty="dashed", col = "blue")
lines(betaPs-1.96*betaPs.se~data$brood_year, lty="dashed", col = "blue")

lines(betaPb~data$brood_year, col = "green")
lines(betaPb+1.96*betaPb.se~data$brood_year, lty="dashed", col = "green")
lines(betaPb-1.96*betaPb.se~data$brood_year, lty="dashed", col = "green")
```


# Results
The results of the four DLMs show a pretty tight grouping in the AICc values, however the model with no covariates is the best fit for the data. The least-best fit according to the AICc is the model with both summer and winter values. This could be performing poorly because of the increased variability involved in trying to combine two different seasonal signatures into one covariate for the model. There was significant autocorrelation in the spawners covariate at the 5 and 10 year mark.

# Discussion
The results of these four models suggest that there is not a strong effect of PDO on the productivity of Kvichak sockeye, based only on the R/S rates over time. Although neither the winter or summer PDO models were considered the "best performing", the winter PDO model showed slightly better performance than the summer. This could be the case because winters at sea may act as a slight bottleneck (as assumed) for survival of Kvichak fish. Summers are typically reliably productive for salmon, as with winter is already a time of slowed growth for fish, and thus the effect of the winter PDO signal may have more of an effect on fish. Besides the effect of PDO, there are other strong environmental signals that may be affecting the productivity of Kvichak sockeye, such as other climate variables including SST/Marine Heatwaves. Additionally, at sea there there is both inter- and intra-specific competition for resources, especially in recent years where the number of sockeye salmon in the ocean has been increasing, as well as the presence of Pink salmon, a major competitor of sockeye. One factor not discussed in this lab as well is the effect of catch on these returning fish. Although escapement goals are set by ADF&G, the number of spawning fish that are able to return to spawning grounds may be affected by the amount of fish caught in the fishery in Bristol Bay.




# Team contributions

All team members contributed to the data wrangling and analysis. We each attempted all sections of the lab and then reconvened to compared differences and merge scripts. Each team member contributed text to results and discussion sections. 

