---
title: Team X - Lab 5
subtitle: Dynamic Linear Models (DLMs) - Time-varying productivity of salmon
author: "Emma Christensen, Brian McGreal, Anna Sulc"
date: May XX, 2025
output: 
  html_document:
    code-folding: true
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
options(dplyr.summarise.inform = FALSE)
```


```{r silent=TRUE, include=FALSE}
library(atsalibrary)
library(dplyr)
library(ggplot2)
library(here)
library(tidyverse)
library(MARSS)

```

# Data

## Load the data

This exercise involves spawner-recruit data for sockeye salmon (Oncorhynchus nerka) from the Kvichak River in SW Alaska that span the years 1952-2005. We have a dataframe with columns for brood year (`brood_year`), number of spawners (`spawners`), number of recruits (`recruits`) and PDO at year $t-2$ in summer (`pdo_summer_t2`) and in winter (`pdo_winter_t2`).



```{r dlm-load-atsa, echo=FALSE}
## library(devtools)
## Windows users will likely need to set this
## Sys.setenv("R_REMOTES_NO_ERRORS_FROM_WARNINGS" = "true")
# devtools::install_github("nwfsc-timeseries/atsalibrary")

## get data
data(KvichakSockeye, package="atsalibrary")
data <- KvichakSockeye

head(data)

```

We'll take a quick look at spawners and recruits over time in the data.

```{r plot, echo = FALSE}
#Plot Data 
colors <- c("Recruits" = "blue", "Spawners" = "black")
ggplot(data=data, aes(x=brood_year)) +
  geom_line(aes(y = spawners), color = "black", size = 1) +
  geom_line(aes(y = recruits), color = "blue", size = 1) +    
  labs(title = "Kvichak River Sockeye, 1950-1989",
       subtitle = "Spawners (black) and Recruits (blue)",
       x="",
       y = "Fish",
       color = "Legend") +
    scale_color_manual(values = colors)
```

## Wrangle the data

Before constructing DLMs to be solved using a MARSS model framework, some data cleaning is in order. A variable $y$ is constructed by taking the natural log of recruits divided by spawners ($\ln(R_t/S_t)$). The spawner variable $S_t$ is then scaled to be used as an inidcator variable in the model specification. Finally, because $S_t$ will act as an indicator variable, any observations with missing values in $S_t$ must be dropped, meaning in this case losing the first four observations in the time series.

```{r wrangle_data}
# transform data
data$y <- log(data$recruits/data$spawners)
data <- data[!is.na(data$spawners),]

years <- data$brood_year
TT <- length(years)
dat <- matrix(data$y, nrow=1)

s <- data$spawners
s_z <- matrix((s - mean(s))/sqrt(var(s)), nrow = 1)
m <- dim(s_z)[1] + 1
```

# Methods

### Base Model structure

We began by fitting a reduced form of Ricker's model that including only a time-varying intercept ($\alpha_t$) as well as a time-varying effect of spawner abundance ($/beta_t$). That is,

\[
\ln\left(\frac{R_t}{S_t}\right) = \ln(\alpha_t) + \beta_tS_t+ v_t
\]

By setting $\ln\left(\frac{R_t}{S_t}\right)$ = $y_t$ and $\ln(\alpha_t)$ = $a_t$ we are left with the following observation equation:

\begin{align*}
\text{Observation:} \quad & y_t = a_t + \beta_tS_t+ v_t \\
                          & \quad v_t \sim \mathcal{N}(0, R) \\[1em]
\end{align*}

By then assuming that $a_t$ and $\beta_t$ are results of two unoberserved underlying processes, we may specify the following state equations:

\begin{align*}
\text{State:} \quad       & a_t = a_{t-1} + w_{a,t}, \quad w_t \sim \mathcal{N}(0, Q_a) \\
                          & \beta_t = \beta_{t-1} + w_{\beta,t}, \quad w_t \sim \mathcal{N}(0, Q_\beta) \\ 
\end{align*}

This model assumes no density-dependent survival in that the number of recruits is an ascending function of spawners. We did not include any environmental variables for this step. 


```{r dlm-build}
# build MARSS model
# process
B <- diag(m)
U <- matrix(0, nrow=m, ncol=1)
Q <- matrix(list(0), m, m)
diag(Q) <- c("q.alpha", "q.beta")

# obs
Z <- array(NA, c(1, m, TT))
Z[1,1,] <- rep(1, TT)
Z[1,2,] <- s_z
A <- matrix(0)
R <- matrix("r")

# initital values & controls
inits <- list(x0 = matrix(c(0,0), nrow = m))
con_list <- list(maxit = 2146)
  # 2146 seems to be the minimum number of iterations at which all beta parameter varainces converge to zero

# model list
dlm.list <- list(B=B, U=U, Q=Q, Z=Z, A=A, R=R)

# fit model
if(!file.exists(here::here("Lab-5", "Team-3", "data", "dlm.rds"))){
  dlm <- MARSS(dat, model= dlm.list, inits = inits, method = "kem", control = con_list)
  saveRDS(dlm, file=here::here("Lab-5", "Team-3", "data", "dlm.rds"))
}
# load in ssm.phyto_month
dlm <- readRDS(file=here::here("Lab-5", "Team-3", "data", "dlm.rds"))
```

```{r dlm, echo = FALSE}
dlm
```

Plot the ts of $\alpha_t$ and note the AICc for this model. Also plot appropriate model diagnostics.


```{r plotA, echo=FALSE}
## plot
alpha <- as.numeric(dlm$states[1,])
alpha.se <- as.numeric(dlm$states.se[1,])

# alpha
plotA <- plot(alpha~data$brood_year,
              main = "Time varying beta estimates, 500 iterations",
              xlab = "",
              ylab = "Alpha estimate",
              type='l', 
              ylim=c(-3,3), 
              xlim= c(1956, 2005))
lines(alpha+1.96*alpha.se~data$brood_year, lty="dashed")
lines(alpha-1.96*alpha.se~data$brood_year, lty="dashed")
```

```{r plotB500, echo=FALSE}
# new dls - 500 iter
con_listALT <- list(maxit = 500)

# fit model
if(!file.exists(here::here("Lab-5", "Team-3", "data", "dlm500.rds"))){
  dlm500 <- MARSS(dat, model= dlm.list, inits = inits, method = "kem", control = con_listALT)
  saveRDS(dlm500, file=here::here("Lab-5", "Team-3", "data", "dlm500.rds"))
}
# load in ssm.phyto_month
dlm500 <- readRDS(file=here::here("Lab-5", "Team-3", "data", "dlm500.rds"))

#grab estimates
beta <- as.numeric(dlm500$states[2,])
beta.se <- as.numeric(dlm500$states.se[2,])

# beta
plot(beta~data$brood_year,
     main = "Time varying beta estimates, 500 iterations",
     xlab = "",
     ylab = "Beta estimate",
     type='l', 
     ylim=c(-0.5, 0.5),
     xlim= c(1956, 2005))
lines(beta+1.96*beta.se~data$brood_year, lty="dashed")
lines(beta-1.96*beta.se~data$brood_year, lty="dashed")
```

```{r plotB1000, echo=FALSE}
# new dls - 500 iter
con_listALT <- list(maxit = 1000)

# fit model
if(!file.exists(here::here("Lab-5", "Team-3", "data", "dlm1000.rds"))){
  dlm1000 <- MARSS(dat, model= dlm.list, inits = inits, method = "kem", control = con_listALT)
  saveRDS(dlm1000, file=here::here("Lab-5", "Team-3", "data", "dlm1000.rds"))
}
# load in ssm.phyto_month
dlm1000 <- readRDS(file=here::here("Lab-5", "Team-3", "data", "dlm1000.rds"))

#grab estimates
beta <- as.numeric(dlm1000$states[2,])
beta.se <- as.numeric(dlm1000$states.se[2,])

# beta
plot(beta~data$brood_year,
     main = "Time varying beta estimates, 1000 iterations",
     xlab = "",
     ylab = "Beta estimate",
     type='l', 
     ylim=c(-0.5, 0.5),
     xlim= c(1956, 2005))
lines(beta+1.96*beta.se~data$brood_year, lty="dashed")
lines(beta-1.96*beta.se~data$brood_year, lty="dashed")
```

```{r plotB1750, echo=FALSE}
# new dls - 500 iter
con_listALT <- list(maxit = 1750)

# fit model
if(!file.exists(here::here("Lab-5", "Team-3", "data", "dlm1750.rds"))){
  dlm1750 <- MARSS(dat, model= dlm.list, inits = inits, method = "kem", control = con_listALT)
  saveRDS(dlm1750, file=here::here("Lab-5", "Team-3", "data", "dlm1750.rds"))
}
# load in ssm.phyto_month
dlm1750 <- readRDS(file=here::here("Lab-5", "Team-3", "data", "dlm1750.rds"))

#grab estimates
beta <- as.numeric(dlm1750$states[2,])
beta.se <- as.numeric(dlm1750$states.se[2,])

# beta
plot(beta~data$brood_year,
     type='l',
     main = "Time varying beta estimates, 1750 iterations",
     xlab = "",
     ylab = "Beta estimate",
     ylim=c(-0.5, 0.5),
     xlim= c(1956, 2005))
lines(beta+1.96*beta.se~data$brood_year, lty="dashed")
lines(beta-1.96*beta.se~data$brood_year, lty="dashed")
```


```{r plotB2150, echo=FALSE}
# beta
beta <- as.numeric(dlm$states[2,])
beta.se <- as.numeric(dlm$states.se[2,])

plot(beta~data$brood_year,
     main = "Time varying beta estimates, 2146 iterations",
     xlab = "",
     ylab = "Beta estimate",
     type='l', 
     ylim=c(-0.5, 0.5),
     xlim= c(1956, 2005))
lines(beta+1.96*beta.se~data$brood_year, lty="dashed")
lines(beta-1.96*beta.se~data$brood_year, lty="dashed")
```


What environmental or dummy variables did you include and why?


### Covariates

Next,we fitted the Ricker model with a time-varying intercept and a static effect of spawners. 

For this model, we need the time series of $\alpha_t$, which is an estimate of the stock productivity in the absence of density-dependent effects. 

The equation for this model is now: 

\begin{align*}
\text{Observation:} \quad & y_t = a_t + \beta_tS_t+ Dd_t + v_t \\
                          & \quad v_t \sim \mathcal{N}(0, R) \\[1em]
\text{State:} \quad       & a_t = a_{t-1} + w_{a,t}, \quad w_t \sim \mathcal{N}(0, Q_a) \\
                          & \beta_t = \beta_{t-1} + w_{\beta,t}, \quad w_t \sim \mathcal{N}(0, Q_\beta) \\ 
\end{align*}


This means that $\alpha_t$ is time-varying and $\beta_t$ is static. 

```{r dlmPw, echo = FALSE}
# winter pdo
D <- matrix("pdo", 1, 1)
pdo_w <- as.data.frame(matrix(data$pdo_winter_t2, nrow=1, byrow=TRUE))
pdo_w <- data.matrix(pdo_w)

dlmPw.list <- list(B=B, U=U, Q=Q, Z=Z, A=A, R=R, D=D, d=pdo_w)

if(!file.exists(here::here("Lab-5", "Team-3", "data", "dlmPw.rds"))){
  dlmPw <- MARSS(dat, model= dlmPw.list, inits = inits, method = "kem", control = con_list)
  saveRDS(dlmPw, file=here::here("Lab-5", "Team-3", "data", "dlmPw.rds"))
}

# load in ssm.phyto_month
dlmPw <- readRDS(file=here::here("Lab-5", "Team-3", "data", "dlmPw.rds"))
dlmPw
```

```{r dlmPs, echo = FALSE}
# summer pdo
D <- matrix("pdo", 1, 1)
pdo_s <- as.data.frame(matrix(data$pdo_summer_t2, nrow=1, byrow=TRUE))
pdo_s <- data.matrix(pdo_s)

dlmPs.list <- list(B=B, U=U, Q=Q, Z=Z, A=A, R=R, D=D, d=pdo_s)

if(!file.exists(here::here("Lab-5", "Team-3", "data", "dlmPs.rds"))){
  dlmPs <- MARSS(dat, model= dlmPs.list, inits = inits, method = "kem", control = con_list)
  saveRDS(dlmPs, file=here::here("Lab-5", "Team-3", "data", "dlmPs.rds"))
}

# load in ssm.phyto_month
dlmPs <- readRDS(file=here::here("Lab-5", "Team-3", "data", "dlmPs.rds"))
dlmPs
```

```{r dlmPb, echo=FALSE}
# winter and summer pdo
D <- matrix(c("pdoW", "pdoS"), 2, 1)
pdo_b <- rbind(pdo_w, pdo_s)
pdo_b <- data.matrix(pdo_b)

dlmPb.list <- list(B=B, U=U, Q=Q, Z=Z, A=A, R=R, D=D, d=pdo_b)

if(!file.exists(here::here("Lab-5", "Team-3", "data", "dlmPb.rds"))){
  dlmPb <- MARSS(dat, model= dlmPb.list, inits = inits, method = "kem", control = con_list)
  saveRDS(dlmPb, file=here::here("Lab-5", "Team-3", "data", "dlmPb.rds"))
}
# load in ssm.phyto_month
dlmPb <- readRDS(file=here::here("Lab-5", "Team-3", "data", "dlmPb.rds"))
dlmPb
```

```{r aic, echo = FALSE}
## AICc
# an empty list to collect outputs
out_list <- list()
# Add each model's name and AICc 
out_list[[1]] <- data.frame(Model = "no covariates", AICc = dlm$AICc)
out_list[[2]] <- data.frame(Model = "winter PDO", AICc = dlmPw$AICc)
out_list[[3]] <- data.frame(Model = "summer PDO", AICc = dlmPs$AICc)
out_list[[4]] <- data.frame(Model = "winter and summer PDO", AICc = dlmPb$AICc)
out_list
```

```{r covariancePLOT, echo = FALSE}
alphaPw <- as.numeric(dlmPw$states[1,])
betaPw <- as.numeric(dlmPw$states[2,])
alphaPw.se <- as.numeric(dlmPw$states.se[1,])
betaPw.se <- as.numeric(dlmPw$states.se[2,])

alphaPs <- as.numeric(dlmPs$states[1,])
betaPs <- as.numeric(dlmPs$states[2,])
alphaPs.se <- as.numeric(dlmPs$states.se[1,])
betaPs.se <- as.numeric(dlmPs$states.se[2,])

alphaPb <- as.numeric(dlmPb$states[1,])
betaPb <- as.numeric(dlmPb$states[2,])
alphaPb.se <- as.numeric(dlmPb$states.se[1,])
betaPb.se <- as.numeric(dlmPb$states.se[2,])

plot(alpha~data$brood_year,
     main = "Time varying alpha estimates",
     sub = "Null (black), Winter PDO (red), Summer PDO (blue), Full (green)",
     xlab = "",
     ylab = "Alpha estimate",
     type='l', 
     ylim=c(-3,3), 
     xlim= c(1956, 2005))
lines(alpha+1.96*alpha.se~data$brood_year, lty="dashed")
lines(alpha-1.96*alpha.se~data$brood_year, lty="dashed")

lines(alphaPw~data$brood_year, col = "red")
lines(alphaPw+1.96*alphaPw.se~data$brood_year, lty="dashed", col = "red")
lines(alphaPw-1.96*alphaPw.se~data$brood_year, lty="dashed", col = "red")

lines(alphaPs~data$brood_year, col = "blue")
lines(alphaPs+1.96*alphaPs.se~data$brood_year, lty="dashed", col = "blue")
lines(alphaPs-1.96*alphaPs.se~data$brood_year, lty="dashed", col = "blue")

lines(alphaPb~data$brood_year, col = "green")
lines(alphaPb+1.96*alphaPb.se~data$brood_year, lty="dashed", col = "green")
lines(alphaPb-1.96*alphaPb.se~data$brood_year, lty="dashed", col = "green")

# beta
plot(beta~data$brood_year,
     type='l', 
     main = "Time varying beta estimates",
     sub = "Null (black), Winter PDO (red), Summer PDO (blue), Full (green)",
     xlab = "",
     ylab = "Beta estimate",
     ylim=c(-0.1,0.1), 
     xlim= c(1956, 2005))
lines(beta+1.96*beta.se~data$brood_year, lty="dashed")
lines(beta-1.96*beta.se~data$brood_year, lty="dashed")

lines(betaPw~data$brood_year, col = "red")
lines(betaPw+1.96*betaPw.se~data$brood_year, lty="dashed", col = "red")
lines(betaPw-1.96*betaPw.se~data$brood_year, lty="dashed", col = "red")

lines(betaPs~data$brood_year, col = "blue")
lines(betaPs+1.96*betaPs.se~data$brood_year, lty="dashed", col = "blue")
lines(betaPs-1.96*betaPs.se~data$brood_year, lty="dashed", col = "blue")

lines(betaPb~data$brood_year, col = "green")
lines(betaPb+1.96*betaPb.se~data$brood_year, lty="dashed", col = "green")
lines(betaPb-1.96*betaPb.se~data$brood_year, lty="dashed", col = "green")
```


How do these estimates of productivity compare to those from the previous question? Plot the ts of $\alpha_t$ and note the AICc for this model. Also plot appropriate model diagnostics. ($Hint$: If you don't want a parameter to vary with time, what does that say about its process variance?)





# Results


# Discussion



Consider other environmental factors that could influence the time-varying intrinsic productivity of these salmon. For example, are there other large-scale indices of ocean conditions (e.g., sea-surface temperature, El Niño Southern Oscillation)? You could also consider the possible influence of other salmon stocks such as those we examined in Lab 1.




# Team contributions

All team members contributed to the data wrangling and analysis. We each attempted all sections of the lab and then reconvened to compared differences and merge scripts. Each team member contributed text to results and discussion sections. 

