---
title: "Untitled"
output: html_document
---

### Addition of covariates

We fitted the expanded model that includes the summer PDO index as the covariate (`pdo_summer_t2`). 

The equation for this model is now: 

\[
\log\left(\frac{R_t}{S_t}\right) = \alpha_t - \beta S_t + \delta \cdot \text{PDO}_t + v_t
\]

\begin{align*}
\text{Observation:} \quad & y_t = Z \alpha_t + C_1 c_{1t} + C_2 c_{2t} + v_t \\
                          & y_t = 1 \cdot \alpha_t + (-1) \cdot S_t + \delta \cdot \text{PDO}_t + v_t, \quad v_t \sim \text{N}(0, R) \\[1em]
\text{State:} \quad       & \alpha_t = B \alpha_{t-1} + w_t \\
                          & \alpha_t = 1 \cdot \alpha_{t-1} + w_t, \quad w_t \sim \text{N}(0, Q)
\end{align*}

What is the mean level of productivity? Plot the ts of $\delta_t$ and note the AICc for this model. Also plot appropriate model diagnostics.

### Model Forecast for best model

Based on AICc, which of the models above is the most parsimonious? Is it well behaved (i.e., are the model assumptions met)? 

Based on the AICc, the best model is the PDO_summer covariate model.

```{r}



```



Plot the model forecasts for the best model. Is this a good forecast model?

```{r forecast best}



```

## Resulting Model forms (matrix form)


### Model 1: Reduced Ricker Model (time-varying productivity only)

$$
\boxed{
\begin{cases}
\mathbf{y}_t = \mathbf{Z} \mathbf{x}_t + \mathbf{v}_t \\
\mathbf{x}_t = \mathbf{B} \mathbf{x}_{t-1} + \mathbf{w}_t
\end{cases}
}
$$

where

$$
\mathbf{y}_t = \log\left(\frac{R_t}{S_t}\right), \quad
\mathbf{x}_t = \alpha_t
$$

and matrices/vectors:

$$
\mathbf{Z} = [1], \quad
\mathbf{B} = [1], \quad
\mathbf{v}_t \sim \mathbf{N}(0, R), \quad
\mathbf{w}_t \sim \mathbf{N}(0, Q)
$$

---

### Model 2: Full Ricker with static spawner effect

$$
\boxed{
\begin{cases}
\mathbf{y}_t = \mathbf{Z} \mathbf{x}_t + \mathbf{C} \mathbf{c}_t + \mathbf{v}_t \\
\mathbf{x}_t = \mathbf{B} \mathbf{x}_{t-1} + \mathbf{w}_t
\end{cases}
}
$$

where

$$
\mathbf{y}_t = \log\left(\frac{R_t}{S_t}\right), \quad
\mathbf{x}_t = \alpha_t, \quad
\mathbf{c}_t = S_t
$$

and matrices/vectors:

$$
\mathbf{Z} = [1], \quad
\mathbf{C} = [-1], \quad
\mathbf{B} = [1], \quad
\mathbf{v}_t \sim \mathbf{N}(0, R), \quad
\mathbf{w}_t \sim \mathbf{N}(0, Q)
$$

---

### Model 3: Expanded Ricker with PDO covariate

$$
\boxed{
\begin{cases}
\mathbf{y}_t = \mathbf{Z} \mathbf{x}_t + \mathbf{C} \mathbf{c}_t + \mathbf{v}_t \\
\mathbf{x}_t = \mathbf{B} \mathbf{x}_{t-1} + \mathbf{w}_t
\end{cases}
}
$$

where

$$
\mathbf{y}_t = \log\left(\frac{R_t}{S_t}\right), \quad
\mathbf{x}_t = \alpha_t, \quad
\mathbf{c}_t = \begin{bmatrix} S_t \\ \text{PDO}_t \end{bmatrix}
$$

and matrices/vectors:

$$
\mathbf{Z} = [1], \quad
\mathbf{C} = \begin{bmatrix} -1 & \delta \end{bmatrix}, \quad
\mathbf{B} = [1], \quad
\mathbf{v}_t \sim \mathbf{N}(0, R), \quad
\mathbf{w}_t \sim \mathbf{N}(0, Q)
$$


## Model diagnostics and fit


### Comparison of covariate performance

* What sort of model diagnostics did you use to examine model assumptions?
