---
title: "Lab_3"
editor: visual
---

```{r}
#Load all necessary data and packages
library(atsalibrary)
library(MARSS)
```

# Data

This lab used a monthly data set of plankton concentrations in Lake Washington from 1962-1994. We had data on 15 different plankton groups, both grazers and primary producers. The data set also contained three covariates:

### **Environmental data**

-   `Temp`: water temperature in degrees C

-   `TP`: total phosphorous concentration in mg m^-3^

-   `pH`: pH.

### **Phytoplankton**

-   `Cryptomonas`: small brown or green algae (edible)

-   `Diatoms`: small algae rich in silica (edible)

-   `Greens`: general class of algae (edible)

-   `Bluegreens`: cyanobacteria that can fix nitrogen (inedible)

-   `Unicells`: very small algae (edible)

-   `Other.algae`: generic catch-all for atypical species (edible)

### **Zooplankton**

-   `Conochilus`: colonial form of rotifer (grazer)

-   `Cyclops`: cyclopoid copepod (grazer)

-   `Daphnia`: cladoceran (grazer)

-   `Diaptomus`: calanoid copepod (grazer)

-   `Epischura`: very large calanoid copepod (predator)

-   `Leptodora`: very large cladoceran (predator)

-   `Neomysis`: opossum shrimp (predator)

-   `Non.daphnid.cladocerans`: catch-all for other cladocerans (grazers)

-   `Non.colonial.rotifers`: free-floating rotifers (grazers)

# Methods

### Approach to the Lab & Research Questions

Based on the general lab instructions and some of our own questions, we chose to structure our lab in the following way. Lab instructions were:

1)  Find the most parsimonious model among a set that examines the effects of environmental covariates and/or an indicator of seasonality, varying numbers of trends, and different forms of variance-covariance matrices for the observation errors.

2)  Plot trends and individual loadings for the model identified in task (1) above.

3)  Plot model fits and uncertainty intervals for the model identified in task (1) above.

4)  Describe the effects of environmental or dummy variables on (possibly seasonal) patterns in the data.

Using this, we formulated a few main questions:

1.  How do grazers and phytoplankton trends vary? Can the same/opposite factors be detected for both predators and prey? (we would expect opposite trends)

2.  How do different model error structures alter our results?

3.  How does peeling back time periods alter model results?

4.  Is using 1, 2 , or 3 trends in the DFA a better fit?

### Data Wrangling & Data Selection

For our analysis, we needed our data to be cleaned up and transformed for the MARSS model - this meant replacing zeros with NAs and z-scoring the data. It also involved transposing the data such that years were a single column.

-   Which plankton taxa did you choose and how did you choose them?

    In order to answer the question to our first and second question, we subset the data to include only the phytoplankton: Cryptomonas, Diatoms, Greens, Unicells, and Other Algae. This was done to be able to open the possibility of also exploring the relationships between the edible phytoplankton and the grazers.

-   What time period(s) did you examine and why?

We subset our data to 10 year block from 1980-1989.

-   What environmental or dummy variables did you include and why?

    For environmental covariates, we explored model alternatives with Temperature, Phosphorus, and a combination of Temperature and Phosphorus. To model scenarios without the presence of covariates, we also used cos and sin seasonal trends to compare.

```{r}

all_dat <- lakeWAplanktonTrans

## use only the 10 years from 1980-1989
yr_frst <- 1980
yr_last <- 1989
plank_dat <- all_dat[all_dat[, "Year"] >= yr_frst & 
                       all_dat[, "Year"] <= yr_last,]
## create vector of phytoplankton group names
phytoplankton <- c("Cryptomonas", "Diatoms", "Greens",
                   "Unicells", "Other.algae")
## get only the phytoplankton
dat_1980 <- plank_dat[, phytoplankton]

## ----dfa-trans-data-------------------------------------------------------------------------------------------
## transpose data so time goes across columns
dat_1980 <- t(dat_1980)
## get number of time series
N_ts <- dim(dat_1980)[1]
## get length of time series
TT <- dim(dat_1980)[2] 


## ----dfa-demean-data------------------------------------------------------------------------------------------
y_bar <- apply(dat_1980, 1, mean, na.rm = TRUE)
dat <- dat_1980 - y_bar
rownames(dat) <- rownames(dat_1980)


## ----dfa-plot-phytos, fig.height=9, fig.width=8, fig.cap='Demeaned time series of Lake Washington phytoplankton.'----
spp <- rownames(dat_1980)
clr <- c("brown","blue","darkgreen","darkred","purple")
cnt <- 1
par(mfrow = c(N_ts,1), mai = c(0.5,0.7,0.1,0.1), omi = c(0,0,0,0))
for(i in spp){
  plot(dat[i,],xlab = "",ylab="Abundance index", bty = "L", xaxt = "n", pch=16, col=clr[cnt], type="b")
  axis(1,12*(0:dim(dat_1980)[2])+1,yr_frst+0:dim(dat_1980)[2])
  title(i)
  cnt <- cnt + 1
}
```

### Model Forms

#### Initial Model (Test Case)

Our base state-space model, where we explored 3 possible factors for one type of phytoplankton, took on the following structure:

**Observation Equation:**

$\mathbf{y}_t = \mathbf{Z} \mathbf{x}_t + \mathbf{d} + \mathbf{v}_t$

where: - $\mathbf{y}_t$ is the $(N_t \times 1)$ vector of observations at time $t$. - $\mathbf{Z}$ is the $(N_t \times mm)$ loadings matrix. - $\mathbf{x}_t$ is the $(mm \times 1)$ vector of hidden states at time $t$. - $\mathbf{d}$ is the $(N_t \times 1)$ vector of observation offsets (set to zero in the code). - $\mathbf{v}_t$ is the $(N_t \times 1)$ vector of observation errors, with a variance-covariance matrix $\mathbf{R}$.

**State/Process Equation:**

$\mathbf{x}_t = \mathbf{B} \mathbf{x}_{t-1} + \mathbf{c} + \mathbf{w}_t$

where: - $\mathbf{x}_t$ is the $(mm \times 1)$ vector of hidden states at time $t$. - $\mathbf{B}$ is the $(mm \times mm)$ state transition matrix (here, an identity matrix). - $\mathbf{c}$ is the $(mm \times 1)$ vector of state offsets (set to zero). - $\mathbf{w}_t$ is the $(mm \times 1)$ vector of state process errors, with a variance-covariance matrix $\mathbf{Q}$.

**Matrix Definitions based on the R code:**

-   $\mathbf{Z} = \begin{bmatrix} z11 & 0 & 0 \\ z21 & z22 & 0 \\ z31 & z32 & z33 \\ z41 & z42 & z43 \\ z51 & z52 & z53 \end{bmatrix}$ (where $N_t = 5$ and $mm = 3$)

-   $\mathbf{d} = \begin{bmatrix} 0 \\ 0 \\ 0 \\ 0 \\ 0 \end{bmatrix}$ (a $(5 \times 1)$ zero vector)

-   $\mathbf{R}$ is a $(5 \times 5)$ diagonal matrix with unequal variances for the observation errors.

    $${
    \mathbf{R} = 
    \begin{bmatrix}
    \sigma_1^2 & 0 & 0 & 0 & 0 \\
    0 & \sigma_2^2 & 0 & 0 & 0 \\
    0 & 0 & \sigma_3^2 & 0 & 0 \\
    0 & 0 & 0 & \sigma_4^2 & 0 \\
    0 & 0 & 0 & 0 & \sigma_5^2
    \end{bmatrix}
    }$$

    $(\sigma_1^2)$, $(\sigma_2^2)$, $(\sigma_3^2)$, $(\sigma_4^2)$, and $(\sigma_5^2)$ are the unequal variances of the observation errors.

-   $mm = 3$ (number of processes/hidden states)

-   $\mathbf{B} = \begin{bmatrix} 1 & 0 & 0 \\ 0 & 1 & 0 \\ 0 & 0 & 1 \end{bmatrix}$ (a $(3 \times 3)$ identity matrix)

-   $\mathbf{c} = \begin{bmatrix} 0 \\ 0 \\ 0 \end{bmatrix}$ (a $(3 \times 1)$ zero vector)

-   $\mathbf{Q}$ is a $(3 \times 3)$ identity matrix.

In practice, this model looks like:

```{r}
## ----dfa-dfa-obs-eqn------------------------------------------------------------------------------------------
## 'ZZ' is loadings matrix
Z_vals <- list("z11",  0  ,  0  ,
               "z21","z22",  0  ,
               "z31","z32","z33",
               "z41","z42","z43",
               "z51","z52","z53")
ZZ <- matrix(Z_vals, nrow = N_ts, ncol = 3, byrow = TRUE)
ZZ
## 'aa' is the offset/scaling
aa <- "zero"
## 'DD' and 'd' are for covariates
DD <- "zero"  # matrix(0,mm,1)
dd <- "zero"  # matrix(0,1,wk_last)
## 'RR' is var-cov matrix for obs errors
RR <- "diagonal and unequal"


## ----dfa-dfa-proc-eqn-----------------------------------------------------------------------------------------
## number of processes
mm <- 3
## 'BB' is identity: 1's along the diagonal & 0's elsewhere
BB <- "identity"  # diag(mm)
## 'uu' is a column vector of 0's
uu <- "zero"  # matrix(0, mm, 1)
## 'CC' and 'cc' are for covariates
CC <- "zero"  # matrix(0, mm, 1)
cc <- "zero"  # matrix(0, 1, wk_last)
## 'QQ' is identity
QQ <- "identity"  # diag(mm)


## ----dfa-create-model-lists-----------------------------------------------------------------------------------
## list with specifications for model vectors/matrices
mod_list <- list(Z = ZZ, A = aa, D = DD, d = dd, R = RR,
                 B = BB, U = uu, C = CC, c = cc, Q = QQ)

## list with model inits
init_list <- list(x0 = matrix(rep(0, mm), mm, 1))

## list with model control parameters
con_list <- list(maxit = 3000, allow.degen = TRUE)


## ----dfa-fit-dfa-1, cache=TRUE--------------------------------------------------------------------------------
## fit MARSS
dfa_1 <- MARSS(y = dat, model = mod_list, inits = init_list, control = con_list)

```

#### Model Diagnostics

```{r}
resids <- MARSSresiduals(dfa_1, type = "tt1")
std_resids <- resids$standardized.residuals
str(std_resids)
dfa_1$convergence 
resids <- MARSSresiduals(dfa_1, type = "tt1")
str(resids)

std_resids_mat <- resids$std.residuals

# Set plotting layout: 4 rows, 2 columns (adjust as needed)
par(mfrow = c(ceiling(nrow(std_resids_mat) / 2), 2))

# Plot for each time series
for (i in 1:nrow(std_resids_mat)) {
  plot(std_resids_mat[i, ], type = "h",
       main = rownames(std_resids_mat)[i],
       xlab = "Time", ylab = "Standardized Residual",
       col = "steelblue", lwd = 2)
  abline(h = 0, lty = 2)
  abline(h = c(-2, 2), col = "red", lty = 3)
}
```

#### Alternate Error Structures

For this base model, we explored the impact of altering our assumptions about error structure and comparing model performance.

```{r}
# Step 1: Use same Z matrix and model list as before, but change R structure
Z_equal <- matrix(list(), nrow = 5, ncol = 2)  # assuming you're comparing to your 2-trend model
rownames(Z_equal) <- rownames(dat)
for(i in 1:5){
  Z_equal[i,1] <- paste0("z", i)
  Z_equal[i,2] <- paste0("z", i+5)
}

# Step 2: Define new model with equal error variance
model.equalvar <- list(
  Z = Z_equal,
  R = "diagonal and equal",  
  B = "identity",
  U = "zero",
  Q = "identity",
  x0 = "zero",
  V0 = "zero",
  A = "zero"
)

# Step 3: Fit the model
fit_equalvar <- MARSS(dat, model = model.equalvar, control = list(maxit = 5000))

# Step 4: View results
summary(fit_equalvar)
```

#### Alternate Model Forms

Scenario 1: Edible Phytoplankton Only (Base Scenario Above)

This served as our null or reference scenario.

Observation and Process equations remain the same. Only required changes are to the Z matrix.

$\mathbf{Z} = \begin{bmatrix} z_{11} & 0 & 0 \\ z_{21} & z_{22} & 0 \\ z_{31} & z_{32} & z_{33} \\ z_{41} & z_{42} & z_{43} \\ z_{51} & z_{52} & z_{53} \end{bmatrix}$

$\mathbf{d} = \begin{bmatrix} 0 \\ 0 \end{bmatrix}$

Scenario 3: Edible Phytoplankton and Phosphorus covariate

Adding covariates required altering model structure:

$$
\begin{aligned}
\mathbf{y}_t &= \mathbf{Z}\mathbf{x}_t + \mathbf{a} + \mathbf{D}\mathbf{d}_t + \mathbf{v}_t, \quad \text{where} \quad \mathbf{v}_t \sim \text{MVN}(\mathbf{0}, \mathbf{R}) \\
\mathbf{x}_t &= \mathbf{x}_{t-1} + \mathbf{w}_t, \quad \text{where} \quad \mathbf{w}_t \sim \text{MVN}(\mathbf{0}, \mathbf{Q})
\end{aligned}
$$

Where Z and d take on the following structure

$\mathbf{Z} = \begin{bmatrix} z_{11} & z_{12} & z_{13} \\ z_{21} & z_{22} & z_{23} \end{bmatrix}$

$\mathbf{d} = \begin{bmatrix} d_{1,Phosphorus} \\ d_{2,Phosphorus} \end{bmatrix} \text{Phosphorus}_t$

```{r}
# Adding the temperature and phosphorus covariates
temp <- t(plank_dat[,"Temp", drop = FALSE])
TP <- t(plank_dat[,"TP", drop = FALSE])


```

Scenario 4: Edible Phytoplankton with Phosphorus and Temperature covariates $$
\begin{aligned}
\mathbf{y}_t &= \mathbf{Z}\mathbf{x}_t + \mathbf{a} + \mathbf{D}_1\mathbf{d}_{1t} + \mathbf{D}_2\mathbf{d}_{2t} + \mathbf{v}_t, \quad \text{where} \quad \mathbf{v}_t \sim \text{MVN}(\mathbf{0}, \mathbf{R}) \\
\mathbf{x}_t &= \mathbf{x}_{t-1} + \mathbf{w}_t, \quad \text{where} \quad \mathbf{w}_t \sim \text{MVN}(\mathbf{0}, \mathbf{Q})
\end{aligned}
$$

Where Z and d take on the following structure

$\mathbf{Z} = \begin{bmatrix} z_{11} & z_{12} & z_{13} \\ z_{21} & z_{22} & z_{23} \end{bmatrix}$

$\mathbf{d} = \begin{bmatrix} d_{1,Phosphorus} \\ d_{2,Phosphorus} \end{bmatrix} \text{Phosphorus}_t$

$\mathbf{d} = \begin{bmatrix} d_{1,Temperature} \\ d_{2,Temperature} \end{bmatrix} \text{Temperature}_t$

Scenario 5: Dummy variables

Sometimes, if we do not have covariates available, we may choose to imitate seasonal cycles to explore the explanatory power of the trend. This yields the following code:

```{r}
cos_t <- cos(2 * pi * seq(TT) / 12)
sin_t <- sin(2 * pi * seq(TT) / 12)
dd <- rbind(cos_t,sin_t)
## list with specifications for model vectors/matrices

mod_list_seas <- list(m = 3)  # Only specify the number of trends

dfa_seas <- MARSS(dat, model = mod_list_seas, form = "dfa", 
                  z.score = FALSE, control = con_list, covariates = dd)
dfa_seas$AICc
```

Scenario 6: Comparing 1, 2, and 3 trends in the DFA

```{r}
#1 Trend
# Step 1: Create Z matrix for 1-trend model
Z <- matrix(list(), nrow = 5, ncol = 1)
rownames(Z) <- rownames(dat)
for(i in 1:5) {
  Z[i, 1] <- paste0("z", i)
}

# Step 2: Define DFA model structure
model.list <- list(
  Z = Z,
  R = "diagonal and unequal",
  B = "identity",
  U = "zero",
  Q = "identity",
  x0 = "zero",
  V0 = "zero",
  A = "zero"
)

# Step 3: Fit the model
fit_1trend <- MARSS(dat, model = model.list, control = list(maxit = 5000))


# Step 4: View results
summary(fit_1trend)

# Step 5: Extract the estimated trend (state variable X) from the model
dfa_trend <- fit_1trend$states[1, ]  # Since we only have 1 trend

# Step 6: Create a time variable that matches the 199-month window
# You originally subset rows 173:371, so the corresponding months are:
time_index <- 1:ncol(dat) 

# Step 7: Plot the trend

if (all(is.finite(dfa_trend))) {
  plot(time_index, dfa_trend, type = "l",
       lwd = 2, col = "darkblue",
       xlab = "Time (monthly index)",
       ylab = "Estimated latent trend",
       main = "Latent Temporal Trend Shared by 5 Plankton Taxa")
  abline(h = 0, lty = 2, col = "gray")
} else {
  warning("Trend contains non-finite values. Cannot plot.")
}

#2 Trends
# Step 1: Create Z matrix for 2-trend model
Z2 <- matrix(list(), nrow = 5, ncol = 2)
rownames(Z2) <- rownames(dat)
for(i in 1:5) {
  Z2[i, 1] <- paste0("z", (2*i - 1))
  Z2[i, 2] <- paste0("z", (2*i))
}

# Step 2: Define DFA model structure with 2 trends
model.list2 <- list(
  Z = Z2,
  R = "diagonal and unequal",
  B = "identity",
  U = "zero",
  Q = "identity",   # or "equalvarcov" if you want trends to covary
  x0 = "zero",
  V0 = "zero",
  A = "zero"
)

# Step 3: Fit the model
fit_2trend <- MARSS(dat, model = model.list2, control = list(maxit = 5000))

# Step 4: View results
summary(fit_2trend)

# Step 5: Extract the latent trends from the fitted model
trend_ts <- fit_2trend$states

# Step 6: Plot each trend separately
par(mfrow = c(2, 1))  # 2 rows, 1 column layout
plot(trend_ts[1, ], type = "l", col = "blue", lwd = 2,
     main = "Latent Trend 1 (X1)", ylab = "Value", xlab = "Time (months)")
plot(trend_ts[2, ], type = "l", col = "darkgreen", lwd = 2,
     main = "Latent Trend 2 (X2)", ylab = "Value", xlab = "Time (months)")

# Reset layout
par(mfrow = c(1, 1))

# Step 7: Plot the Z matrix loadings
Z_est <- coef(fit_2trend, type = "matrix")$Z
barplot(Z_est, beside = TRUE, col = c("blue", "darkgreen"),
        legend = c("Trend 1", "Trend 2"),
        main = "Taxon Loadings on Each Trend",
        ylab = "Loading Strength")

#3 Trend
# Step 1: Create Z matrix for 3-trend model
Z3 <- matrix(list(), nrow = 5, ncol = 3)
rownames(Z3) <- rownames(dat)
for(i in 1:5) {
  for(j in 1:3) {
    Z3[i, j] <- paste0("z", i, j)
  }
}

# Step 2: Define the model list
model.list3 <- list(
  Z = Z3,
  R = "diagonal and unequal",
  B = "identity",
  U = "zero",
  Q = "identity",
  x0 = "zero",
  V0 = "zero",
  A = "zero"
)

# Step 3: Fit the model (set higher maxit just in case)
fit_3trend <- MARSS(dat, model = model.list3, control = list(maxit = 5000))

# Step 4: View results
summary(fit_3trend)

# Step 5: Extract the estimated trends (states)
states_3trend <- fit_3trend$states

# Step 6: Plot the trends over time
ts.plot(t(states_3trend), col = 1:3, lty = 1, xlab = "Time", ylab = "Trend value", main = "Estimated DFA Trends (3-Trend Model)")


# Step 6: Plot trends using base R
matplot(t(states_3trend), type = "l", lty = 1, col = 1:3,
        xlab = "Time", ylab = "Trend value", main = "Estimated DFA Trends (3-Trend Model)")

# Add legend
#legend("topright", legend = paste("Trend", 1:3), col = 1:3, lty = 1)

# Step 7: Plot the loadings (Z matrix values)
Z_estimates <- coef(fit_3trend, type = "matrix")$Z
barplot(Z_estimates, beside = TRUE, col = 1:3,
        legend.text = paste("Trend", 1:3),
        args.legend = list(x = "topright", bty = "n"),
        main = "Factor Loadings (Z matrix)", ylab = "Loading strength", las = 2)
```

### Model Comparison

```{r}
# For models with similar structures, we used AIC to compare model performance
#Here we are comparing the inclusion of covariates:

mod_list = list(m = 3, R = "diagonal and unequal")

dfa_temp <- MARSS(dat, model = mod_list, form = "dfa", z.score = FALSE,
                  control = con_list, covariates = temp)

dfa_TP <- MARSS(dat, model = mod_list, form = "dfa", z.score = FALSE,
                control = con_list, covariates = TP)

dfa_both <- MARSS(dat, model = mod_list, form = "dfa", z.score = FALSE,
                  control = con_list, covariates = rbind(temp, TP))

print(cbind(model = c("no covars", "Temp", "TP", "Temp & TP", "Dummy"),
      AICc = round(c(dfa_1$AICc,
                     dfa_temp$AICc,
                     dfa_TP$AICc,
                     dfa_both$AICc,
                     dfa_seas$AICc))),
      quote = FALSE)

#Here we are comparing the models with 1,2, or 3 trends, and then comparing the alterations to the R Matix Error Structure
# Compare AICc values of equal vs unequal variance
cbind(
  model = c("2-trend unequal R", "2-trend equal R"),
  AICc = c(fit_2trend$AICc, fit_equalvar$AICc)
)
# Compare AICc values across all fitted models
aic_table <- data.frame(
  Model = c("1-Trend", "2-Trend", "3-Trend", "2-Trend Equal Error"),
  AICc = c(
    fit_1trend$AICc,
    fit_2trend$AICc,
    fit_3trend$AICc,
    fit_equalvar$AICc
  )
)

print(aic_table)
```

### Peeling Away Years

An interesting idea mentioned in lecture was exploring the impact of peeling away years on a DFA model specifically it's performance and the factors identified. Given our large data set, we chose to explore this question as a final part of this lab.

We made a function which fits our most simple model:

```{r}
#PEELING AWAY YEARS

# inputs are data in the correct structure (here dat), a MARSS model structure as a list, and a vector specifying which years to drop
run_dfa <- function(data, model, remove_years = NULL) {
  
  
  # Remove years if NULL is replaced with a vector to remove
  if (!is.null(remove_years)) {
    #Expanding to months
    remove_cols <- unlist(lapply(remove_years, function(y) ((y - 1) * 12 + 1):(y * 12)))
    
    use_years <- !(1:ncol(data) %in% remove_cols)
    data <- data[, use_years, drop = FALSE] #drop = FALSE to maintain matrix structure
  }
  
  
  # Model specification (basic DFA)
  
  model <- model
  #Model for MARSS with following general structure (example below)
  
  # model <- list(
  #   m = "states", 
  #   B = "identity",
  #   U = "zero",
  #   Q = "diagonal",
  #   A = "zero",
  #   R = "diagonal"
  # )
  
  # Fit the model
  dfa_mod <- MARSS(data, model = model,  silent = TRUE) # set silent = TRUE to suppress output
  
  # Extract AIC, number of states, and number of years
  num_years = ncol(data)
  AIC = dfa_mod$AIC
  
  #Extracting the loadings from the Z matrix
  Z_matrix = dfa_mod$par$Z
  print(Z_matrix)
  
  # Return relevant information
  return(list(AIC = AIC, num_years = num_years/12, loadings = Z_matrix))
}



```

Now we can experiment and see what the effects of different windows and window sizes are on the results:

```{r}
#FUNCTION CHUNK

results <- list()

years_to_remove <- list(NULL, 1:10, 11:20, 21:30,1:20,1:30) # Different year removal options

#Make the dataset for this
window_dat <- all_dat[, phytoplankton]

#Set up the model

## transpose data so time goes across columns
window_dat  <- t(window_dat )
## get number of time series
window_ts <- dim(window_dat )[1]
## get length of time series
window_TT <- dim(window_dat)[2]

#De-mean data
y_bar <- apply(window_dat , 1, mean, na.rm = TRUE)
window_dat_done <- window_dat  - y_bar
rownames(window_dat_done) <- rownames(window_dat )


#Make the model list

#OBSERVE

## 'aa' is the offset/scaling
aa <- "zero"

## 'DD' and 'd' are for covariates
DD <- "zero"  # matrix(0,mm,1)
dd <- "zero"  # matrix(0,1,wk_last)

## 'RR' is var-cov matrix for obs errors
RR <- "diagonal and unequal"

#PROCESS

## number of processes
mm <- 3

## 'BB' is identity: 1's along the diagonal & 0's elsewhere

BB <- "identity"  # diag(mm)

## 'uu' is a column vector of 0's
uu <- "zero"  # matrix(0, mm, 1)

## 'CC' and 'cc' are for covariates
CC <- "zero"  # matrix(0, mm, 1)
cc <- "zero"  # matrix(0, 1, wk_last)

## 'QQ' is identity
QQ <- "identity"  # diag(mm)

#Model

## list with specifications for model vectors/matrices
window_list <- list(Z = ZZ, A = aa, D = DD, d = dd, R = RR,
                    B = BB, U = uu, C = CC, c = cc, Q = QQ)
## list with model inits
init_list <- list(x0 = matrix(rep(0, mm), mm, 1))
## list with model control parameters
con_list <- list(maxit = 3000, allow.degen = TRUE)

# Loop through and run DFA for window
count <- 1 #make a counter


for (i in 1:length(years_to_remove)) {
  results[[count]] <- run_dfa(window_dat_done, window_list, years_to_remove[[i]])
  count <- count + 1
}


results

#Make a summary dataframe to make things easier
summary_df <- do.call(rbind, lapply(seq_along(results), function(i) {
  res <- results[[i]]
  yrs <- years_to_remove[[i]]
  data.frame(
    run_id = i,
    removed_years = if (is.null(yrs)) "none" else paste0(min(yrs), ":", max(yrs)),
    AIC = res$AIC,
    num_years = res$num_years,
    num_loadings = nrow(res$loadings),
    stringsAsFactors = FALSE
  )
}))

summary_df
```

# Results and Discussion

### Model Performance for Different Scenarios

Using the AIC values to compare among our models here (number of trends and error structure, and covariate inclusion), the models that best fit the phytoplankton taxa here are 2-Trend with Unequal Variance and the temperature-alone covariate model. Interestingly, the Dummy variable model performed better than the TP-alone model, and equally as well as the Temp and TP model. This suggests that while the combination of temperature (best performing model) explains a lot of the variance, there are other seasonal factors that further impact the abundance of plankton as a sinusoidal trend is capable of explaining more. This may be a factor such as amount of sunlight or day length. As a result, this may be a variable that is strongly collinear with temperature as you would expect sunlight to warm the water. This is further reinforced by the joint Temperature and Phosphorus model performing slightly worse than just temperature, suggesting that changes in phosphorus may be linked to plankton abundance but are not a driving factor of trends.

After fitting DFA models with 1, 2, and 3 latent trends, we compared their performance using AIC and AICc values. The 2-trend model had the lowest AICc, suggesting it provides the best fit to the observed plankton time series. Each trend represents an underlying pattern in community dynamics over time, and the factor loadings reveal how each taxon aligns with these trends. While all models reached only abstol convergence, the consistent incline in AICc across models supports the lower complexity of the 2-trend solution.

### Error Structure Impacts

We compared several Dynamic Factor Analysis models using AICc to evaluate their performance. The 2-trend model had the lowest AICc value, indicating the best fit to the data despite only reaching abstol convergence. The incline in AICc with increasing trends suggests that plankton dynamics in Lake Washington are best captured by a small few latent processes rather than a single common pattern. Additionally, when testing an alternative error structure using equal observation error variance, model fit worsened, reinforcing that allowing unequal error variance across taxa better reflects their unique variability. While the models did not fully converge, the AICc differences support using the 2-trend, unequal error variance model as the most parsimonious choice for interpreting system-wide trends.

### Peeling Away Years

Peeling away years and exploring DFA performance was a final step inspired by our curiosity to better understand the method. The results were confusing and suggest that perhaps the experiment was not properly set up. Given the quick and simple set up, there are factors such as which years were selected and what species of plankton that could have played a role in the performance of the model. There are also different ways to drop windows - for example, given the strong seasonal trend, we could have done DFA only on winter or summer abundances to try to find more subtle trends. 

The model with the least years performed best while a model with all years performed the most poorly. This seems improbable and incorrect and may be due to an overfitting issue. Another idea was the impacts of NAs on the ability of dfa to fir properly. However, all models have the same structure and should comparable as a result.

### Member Contributions

All team members participated in all steps of the lab. Following initial data exploration, we each took a different approach to each task and then checked in with our respective results . Following a discussion, we then took on combining our results in to a coherent lab report. For smaller tasks we split our work (plotting, error structure, or peeling away years) and then shared results with the rest of the team. 
