---
title: "Lab_3"
editor: visual
---

```{r}
#Load all necessary data and packages
library(atsalibrary)
library(MARSS)
```

# Data

This lab used a monthly data set of plankton concentrations in Lake Washington from 1962-1994. We had data on 15 different plankton groups, both grazers and primary producers. The data set also contained three covariates:

### **Environmental dat**

-   `Temp`: water temperature in degrees C

-   `TP`: total phosphorous concentration in mg m^-3^

-   `pH`: pH.

### **Phytoplankton**

-   `Cryptomonas`: small brown or green algae (edible)

-   `Diatoms`: small algae rich in silica (edible)

-   `Greens`: general class of algae (edible)

-   `Bluegreens`: cyanobacteria that can fix nitrogen (inedible)

-   `Unicells`: very small algae (edible)

-   `Other.algae`: generic catch-all for atypical species (edible)

### **Zooplankton**

-   `Conochilus`: colonial form of rotifer (grazer)

-   `Cyclops`: cyclopoid copepod (grazer)

-   `Daphnia`: cladoceran (grazer)

-   `Diaptomus`: calanoid copepod (grazer)

-   `Epischura`: very large calanoid copepod (predator)

-   `Leptodora`: very large cladoceran (predator)

-   `Neomysis`: opossum shrimp (predator)

-   `Non.daphnid.cladocerans`: catch-all for other cladocerans (grazers)

-   `Non.colonial.rotifers`: free-floating rotifers (grazers)

# Methods

### Approach to the Lab & Research Questions

Based on the general lab instructions and some of our own questions, we chose to structure our lab in the following way. Lab instructions were:

1)  Find the most parsimonious model among a set that examines the effects of environmental covariates and/or an indicator of seasonality, varying numbers of trends, and different forms of variance-covariance matrices for the observation errors.

2)  Plot trends and individual loadings for the model identified in task (1) above.

3)  Plot model fits and uncertainty intervals for the model identified in task (1) above.

4)  Describe the effects of environmental or dummy variables on (possibly seasonal) patterns in the data.

Using this, we formulated three main questions:

1.  How do grazers and phytoplankton trends vary? Can the same/opposite factors be detected for both predators and prey? (we would expect opposite trends)

2.  How do different model error structures alter our results?

3.  How does peeling back time periods alter model results?

### Data Wrangling & Data Selection

For our analysis, we needed our data to be cleaned up and transformed for the MARSS model - this meant replacing zeros with NAs and z-scoring the data. It also involved transposing the data such that years were a single column.

-   Which plankton taxa did you choose and how did you choose them?

    In order to answer the question to our first and second question, we subset the data to include only the phytoplankton: Cryptomonas, Diatoms, Greens, Unicells, and Other Algae. This was done to be able to open the possibility of also exploring the relationships between the edible phytoplankton and the grazers.

-   What time period(s) did you examine and why?

We subset our data to 10 year block from 1980-1989.

-   What environmental or dummy variables did you include and why?

    For environmental covariates, we explored model alternatives with Temperature, Phosphorus, and a combination of Temperature and Phosphorus. To model scenarios without the presence of covariates, we also used cos and sin seasonal trends to compare.

```{r}

all_dat <- lakeWAplanktonTrans

## use only the 10 years from 1980-1989
yr_frst <- 1980
yr_last <- 1989
plank_dat <- all_dat[all_dat[, "Year"] >= yr_frst & 
                       all_dat[, "Year"] <= yr_last,]
## create vector of phytoplankton group names
phytoplankton <- c("Cryptomonas", "Diatoms", "Greens",
                   "Unicells", "Other.algae")
## get only the phytoplankton
dat_1980 <- plank_dat[, phytoplankton]

## ----dfa-trans-data-------------------------------------------------------------------------------------------
## transpose data so time goes across columns
dat_1980 <- t(dat_1980)
## get number of time series
N_ts <- dim(dat_1980)[1]
## get length of time series
TT <- dim(dat_1980)[2] 


## ----dfa-demean-data------------------------------------------------------------------------------------------
y_bar <- apply(dat_1980, 1, mean, na.rm = TRUE)
dat <- dat_1980 - y_bar
rownames(dat) <- rownames(dat_1980)


## ----dfa-plot-phytos, fig.height=9, fig.width=8, fig.cap='Demeaned time series of Lake Washington phytoplankton.'----
spp <- rownames(dat_1980)
clr <- c("brown","blue","darkgreen","darkred","purple")
cnt <- 1
par(mfrow = c(N_ts,1), mai = c(0.5,0.7,0.1,0.1), omi = c(0,0,0,0))
for(i in spp){
  plot(dat[i,],xlab = "",ylab="Abundance index", bty = "L", xaxt = "n", pch=16, col=clr[cnt], type="b")
  axis(1,12*(0:dim(dat_1980)[2])+1,yr_frst+0:dim(dat_1980)[2])
  title(i)
  cnt <- cnt + 1
}
```

### Model Forms

#### Initial Model (Test Case)

Our base state-space model, where we explored 3 possible factors for one type of phytoplankton, took on the following structure:

**Observation Equation:**

$\mathbf{y}_t = \mathbf{Z} \mathbf{x}_t + \mathbf{d} + \mathbf{v}_t$

where: - $\mathbf{y}_t$ is the $(N_t \times 1)$ vector of observations at time $t$. - $\mathbf{Z}$ is the $(N_t \times mm)$ loadings matrix. - $\mathbf{x}_t$ is the $(mm \times 1)$ vector of hidden states at time $t$. - $\mathbf{d}$ is the $(N_t \times 1)$ vector of observation offsets (set to zero in the code). - $\mathbf{v}_t$ is the $(N_t \times 1)$ vector of observation errors, with a variance-covariance matrix $\mathbf{R}$.

**State/Process Equation:**

$\mathbf{x}_t = \mathbf{B} \mathbf{x}_{t-1} + \mathbf{c} + \mathbf{w}_t$

where: - $\mathbf{x}_t$ is the $(mm \times 1)$ vector of hidden states at time $t$. - $\mathbf{B}$ is the $(mm \times mm)$ state transition matrix (here, an identity matrix). - $\mathbf{c}$ is the $(mm \times 1)$ vector of state offsets (set to zero). - $\mathbf{w}_t$ is the $(mm \times 1)$ vector of state process errors, with a variance-covariance matrix $\mathbf{Q}$.

**Matrix Definitions based on the R code:**

-   $\mathbf{Z} = \begin{bmatrix} z11 & 0 & 0 \\ z21 & z22 & 0 \\ z31 & z32 & z33 \\ z41 & z42 & z43 \\ z51 & z52 & z53 \end{bmatrix}$ (where $N_t = 5$ and $mm = 3$)

-   $\mathbf{d} = \begin{bmatrix} 0 \\ 0 \\ 0 \\ 0 \\ 0 \end{bmatrix}$ (a $(5 \times 1)$ zero vector)

-   $\mathbf{R}$ is a $(5 \times 5)$ diagonal matrix with unequal variances for the observation errors.

    $${
    \mathbf{R} = 
    \begin{bmatrix}
    \sigma_1^2 & 0 & 0 & 0 & 0 \\
    0 & \sigma_2^2 & 0 & 0 & 0 \\
    0 & 0 & \sigma_3^2 & 0 & 0 \\
    0 & 0 & 0 & \sigma_4^2 & 0 \\
    0 & 0 & 0 & 0 & \sigma_5^2
    \end{bmatrix}
    }$$

    $(\sigma_1^2)$, $(\sigma_2^2)$, $(\sigma_3^2)$, $(\sigma_4^2)$, and $(\sigma_5^2)$ are the unequal variances of the observation errors.

-   $mm = 3$ (number of processes/hidden states)

-   $\mathbf{B} = \begin{bmatrix} 1 & 0 & 0 \\ 0 & 1 & 0 \\ 0 & 0 & 1 \end{bmatrix}$ (a $(3 \times 3)$ identity matrix)

-   $\mathbf{c} = \begin{bmatrix} 0 \\ 0 \\ 0 \end{bmatrix}$ (a $(3 \times 1)$ zero vector)

-   $\mathbf{Q}$ is a $(3 \times 3)$ identity matrix.

In practice, this model looks like:

```{r}
## ----dfa-dfa-obs-eqn------------------------------------------------------------------------------------------
## 'ZZ' is loadings matrix
Z_vals <- list("z11",  0  ,  0  ,
               "z21","z22",  0  ,
               "z31","z32","z33",
               "z41","z42","z43",
               "z51","z52","z53")
ZZ <- matrix(Z_vals, nrow = N_ts, ncol = 3, byrow = TRUE)
ZZ
## 'aa' is the offset/scaling
aa <- "zero"
## 'DD' and 'd' are for covariates
DD <- "zero"  # matrix(0,mm,1)
dd <- "zero"  # matrix(0,1,wk_last)
## 'RR' is var-cov matrix for obs errors
RR <- "diagonal and unequal"


## ----dfa-dfa-proc-eqn-----------------------------------------------------------------------------------------
## number of processes
mm <- 3
## 'BB' is identity: 1's along the diagonal & 0's elsewhere
BB <- "identity"  # diag(mm)
## 'uu' is a column vector of 0's
uu <- "zero"  # matrix(0, mm, 1)
## 'CC' and 'cc' are for covariates
CC <- "zero"  # matrix(0, mm, 1)
cc <- "zero"  # matrix(0, 1, wk_last)
## 'QQ' is identity
QQ <- "identity"  # diag(mm)


## ----dfa-create-model-lists-----------------------------------------------------------------------------------
## list with specifications for model vectors/matrices
mod_list <- list(Z = ZZ, A = aa, D = DD, d = dd, R = RR,
                 B = BB, U = uu, C = CC, c = cc, Q = QQ)

## list with model inits
init_list <- list(x0 = matrix(rep(0, mm), mm, 1))

## list with model control parameters
con_list <- list(maxit = 3000, allow.degen = TRUE)


## ----dfa-fit-dfa-1, cache=TRUE--------------------------------------------------------------------------------
## fit MARSS
dfa_1 <- MARSS(y = dat, model = mod_list, inits = init_list, control = con_list)

```

#### Model Diagnostics

#### Alternate Error Structures

For this base model, we explored the impact of altering our assumptions about error structure and comparing model performance. We explored three types of error matrices

#### Alternate Model Forms

Scenario 1: Edible Phytoplankton Only (Base Scenario Above)

This served as our null or reference scenario.

Scenario 2: Grazers and Edible Phytoplankton

Observation and Process equations remain the same. Only required changes are to the Z matrix?

$\mathbf{Z} = \begin{bmatrix} z_{11} & 0 & 0 \\ z_{21} & z_{22} & 0 \\ z_{31} & z_{32} & z_{33} \\ z_{41} & z_{42} & z_{43} \\ z_{51} & z_{52} & z_{53} \end{bmatrix}$

$\mathbf{d} = \begin{bmatrix} 0 \\ 0 \end{bmatrix}$

Scenario 3: Grazers and Edible Phytoplankton and Phosphorus covariate

Adding covariates required altering model structure:

$$
\begin{aligned}
\mathbf{y}_t &= \mathbf{Z}\mathbf{x}_t + \mathbf{a} + \mathbf{D}\mathbf{d}_t + \mathbf{v}_t, \quad \text{where} \quad \mathbf{v}_t \sim \text{MVN}(\mathbf{0}, \mathbf{R}) \\
\mathbf{x}_t &= \mathbf{x}_{t-1} + \mathbf{w}_t, \quad \text{where} \quad \mathbf{w}_t \sim \text{MVN}(\mathbf{0}, \mathbf{Q})
\end{aligned}
$$

Where Z and d take on the following structure

$\mathbf{Z} = \begin{bmatrix} z_{11} & z_{12} & z_{13} \\ z_{21} & z_{22} & z_{23} \end{bmatrix}$

$\mathbf{d} = \begin{bmatrix} d_{1,Phosphorus} \\ d_{2,Phosphorus} \end{bmatrix} \text{Phosphorus}_t$

```{r}
#Model Fitting


## get temperature and phosphorus covariates
temp <- t(plank_dat[,"Temp", drop = FALSE])
TP <- t(plank_dat[,"TP", drop = FALSE])


```

Scenario 4: Grazers and Edible Phytoplankton with Phosphorus and Temperature covariates $$
\begin{aligned}
\mathbf{y}_t &= \mathbf{Z}\mathbf{x}_t + \mathbf{a} + \mathbf{D}_1\mathbf{d}_{1t} + \mathbf{D}_2\mathbf{d}_{2t} + \mathbf{v}_t, \quad \text{where} \quad \mathbf{v}_t \sim \text{MVN}(\mathbf{0}, \mathbf{R}) \\
\mathbf{x}_t &= \mathbf{x}_{t-1} + \mathbf{w}_t, \quad \text{where} \quad \mathbf{w}_t \sim \text{MVN}(\mathbf{0}, \mathbf{Q})
\end{aligned}
$$

Where Z and d take on the following structure

$\mathbf{Z} = \begin{bmatrix} z_{11} & z_{12} & z_{13} \\ z_{21} & z_{22} & z_{23} \end{bmatrix}$

$\mathbf{d} = \begin{bmatrix} d_{1,Phosphorus} \\ d_{2,Phosphorus} \end{bmatrix} \text{Phosphorus}_t$

$\mathbf{d} = \begin{bmatrix} d_{1,Temperature} \\ d_{2,Temperature} \end{bmatrix} \text{Temperature}_t$

Scenario 5: Dummy variables

Sometimes, if we do not have covariates available, we may choose to imitate seasonal cycles to explore the explanatory power of the trend. This yields the following model structure:

And resulting code:

```{r}
cos_t <- cos(2 * pi * seq(TT) / 12)
sin_t <- sin(2 * pi * seq(TT) / 12)
dd <- rbind(cos_t,sin_t)
dfa_seas <- MARSS(dat, model = mod_list, form = "dfa", z.score = FALSE,
                  control = con_list, covariates = dd)
dfa_seas$AICc
```

### Model Comparison

```{r}
# For models with similar structures, we used AIC to compare model performance

mod_list = list(m = 3, R = "diagonal and unequal")

dfa_temp <- MARSS(dat, model = mod_list, form = "dfa", z.score = FALSE,
                  control = con_list, covariates = temp)

dfa_TP <- MARSS(dat, model = mod_list, form = "dfa", z.score = FALSE,
                control = con_list, covariates = TP)

dfa_both <- MARSS(dat, model = mod_list, form = "dfa", z.score = FALSE,
                  control = con_list, covariates = rbind(temp, TP))

print(cbind(model = c("no covars", "Temp", "TP", "Temp & TP", "Dummy"),
      AICc = round(c(dfa_1$AICc,
                     dfa_temp$AICc,
                     dfa_TP$AICc,
                     dfa_both$AICc,
                     dfa_seas$AICc))),
      quote = FALSE)
```

### Peeling Away Years

An interesting idea mentioned in lecture was exploring the impact of peeling away years on a DFA model specifically it's performance and the factors identified. Given our large data set, we chose to explore this question as a final part of this lab.

We made a function which fits our most simple model:

```{r}
# inputs are data in the correct structure (here dat), a MARSS model structure as a list, and a vector specifying which years to drop
run_dfa <- function(data, model, remove_years = NULL) {

  # Remove years if NULL is replaced with a vector to remove
  if (!is.null(remove_years)) {
    use_years <- which(!((1:ncol(data$Year)) %in% remove_years))
    data$Y <- data$Year[, use_years, drop = FALSE] #drop = FALSE to maintain matrix structure
  }
    
  # Model specification (basic DFA)
  
  model <- model
  #Model for MARSS with following general structure (example below)
  
  # model <- list(
  #   m = "states", 
  #   B = "identity",
  #   U = "zero",
  #   Q = "diagonal",
  #   A = "zero",
  #   R = "diagonal"
  # )

  # Fit the model
  dfa_mod <- MARSS(data$Y, model = model,  silent = TRUE) # set silent = TRUE to suppress output

  # Extract AIC, number of states, and number of years
  num_years = ncol(data$Year)
  AIC = dfa_mod$AIC
  
  #Extracting the loadings from the Z matrix
  Z_matrix = dfa_mod$states$Z
  
  # Return relevant information
  return(list(AIC = AIC, states = states, num_years = num_years, loadings = Z_matrix))
}

```

Now we can experiment and see what the effects of different windows and window sizes are on the results:

```{r}
# 
# results <- list()
# 
# years_to_remove <- list(NULL, 1:10, 11:20, 21:30,1:20,1:30) # Different year removal options
# 
# #Make the dataset for this
# window_dat <- all_dat[, phytoplankton]
# 
# #Set up the model
# 
# ## transpose data so time goes across columns
# window_dat  <- t(window_dat )
# ## get number of time series
# window_ts <- dim(window_dat )[1]
# ## get length of time series
# window_TT <- dim(window_dat)[2] 
# 
# #De-mean data
# y_bar <- apply(window_dat , 1, mean, na.rm = TRUE)
# window_dat_done <- window_dat  - y_bar
# rownames(window_dat_done) <- rownames(window_dat )
# 
# 
# #Make the model list
# 
# #OBSERVE
# 
# ## 'aa' is the offset/scaling
# aa <- "zero"
# 
# ## 'DD' and 'd' are for covariates
# DD <- "zero"  # matrix(0,mm,1)
# dd <- "zero"  # matrix(0,1,wk_last)
# 
# ## 'RR' is var-cov matrix for obs errors
# RR <- "diagonal and unequal"
# 
# #PROCESS
# 
# ## number of processes
# mm <- 3
# 
# ## 'BB' is identity: 1's along the diagonal & 0's elsewhere
# 
# BB <- "identity"  # diag(mm)
# 
# ## 'uu' is a column vector of 0's
# uu <- "zero"  # matrix(0, mm, 1)
# 
# ## 'CC' and 'cc' are for covariates
# CC <- "zero"  # matrix(0, mm, 1)
# cc <- "zero"  # matrix(0, 1, wk_last)
# 
# ## 'QQ' is identity
# QQ <- "identity"  # diag(mm)
# 
# #Model
# 
# ## list with specifications for model vectors/matrices
# window_list <- list(Z = ZZ, A = aa, D = DD, d = dd, R = RR,
#                  B = BB, U = uu, C = CC, c = cc, Q = QQ)
# ## list with model inits
# init_list <- list(x0 = matrix(rep(0, mm), mm, 1))
# ## list with model control parameters
# con_list <- list(maxit = 3000, allow.degen = TRUE)
# 
# # Loop through and run DFA for window
# count <- 1 #make a counter
# 
# 
# for (i in 1:length(years_to_remove)) {
#     results[[count]] <- run_dfa(dat_window, window_list, years_to_remove[[i]])
#     count <- count + 1
#   }




```

# Results

### Model Performance for Different Scenarios

-   Dummy variable explained the most

### Error Structure Impacts

### Peeling Away Years

# Discussion

### Member Contributions
