---
title: Team 1 - Lab 1
subtitle: Dynamic Factor Analysis (DFA) - Common trends in plankton data
author: "Name 1, Name 2, Name 3"
date: April XX, 2025
output: 
  html_document:
    code-folding: true
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
options(dplyr.summarise.inform = FALSE)
```

# Data

Describe what plankton groups and covariates you used and indicate any temporal subsetting you chose. 


## Load the data

```{r load_data}
#Packages
library(MARSS)

## Wrangle the data
```{r}
# Subset zooplankton data we will use
## load the raw data (there are 3 datasets contained here)
data(lakeWAplankton, package = "MARSS")
all_dat <- lakeWAplanktonTrans
unique(all_dat[,"Year"])
yr_frst <- 1985
yr_last <- 1994
plank_dat <- all_dat[all_dat[, "Year"] >= yr_frst & 
                       all_dat[, "Year"] <= yr_last,]

## create vector of phytoplankton group names
phytoplankton <- c("Cryptomonas", "Diatoms", "Greens", "Bluegreens",
                   "Unicells", "Other.algae")
zooplankton<- c("Conochilus", "Cyclops","Daphnia", "Diaptomus", "Epischura","Non.daphnid.cladocerans", "Non.colonial.rotifers")
## get only the phytoplankton
dat_phyto <- plank_dat[, phytoplankton]
dat_zoo<-plank_dat[,zooplankton]

par(mfrow=c(2,3))
plot(dat_zoo[,1], type="l", main=zooplankton[1])
plot(dat_zoo[,2], type="l", main=zooplankton[2])
plot(dat_zoo[,3], type="l", main=zooplankton[3])
plot(dat_zoo[,4], type="l", main=zooplankton[4])
plot(dat_zoo[,5], type="l", main=zooplankton[5])
plot(dat_zoo[,6], type="l", main=zooplankton[6])
lines(dat_zoo[,7], type="l", col="red")

```

* You can choose any taxa to include in your analyses, but make sure there are at least 5 groups. You can mix phytoplankton and zooplankton, if you'd like, but justify any choices you make. 

* You will want to subset the data to a time period with adequate observations for the taxa of interest. Your time window should include at lest 5 years (60 months).

```{r wrangle_data}
## add some code here
```


# General tasks

Each group has the same general tasks, but you will adapt them as you work on the data.

1) Find the most parsimonious model among a set that examines the effects of environmental covariates and/or an indicator of seasonality, varying numbers of trends, and different forms of variance-covariance matrices for the observation errors.

2) Plot trends and individual loadings for the model identified in task (1) above.

3) Plot model fits and uncertainty intervals for the model identified in task (1) above.

4) Describe the effects of environmental or dummy variables on (possibly seasonal) patterns in the data.


# Methods

Please address the following in your methods:

* Which plankton taxa did you choose and how did you choose them?
   
* What time period(s) did you examine and why?

* What environmental or dummy variables did you include and why?

* What forms of models did you fit (ie, write them out in matrix form)?

* What sort of model diagnostics did you use to examine model assumptions?


# Results


# Discussion


# Team contributions

Example: "All team members helped decide on the goal and ran the analyses for the individual regions. Team members 2 & 3 wrote most of the code for the analysis. Team member 1 worked on the plotting section of the report using and adapting code that team member 3 wrote. All team members helped edit the report and wrote the discussion together."

