---
title: Team 1 - Lab 1
subtitle: Dynamic Factor Analysis (DFA) - Common trends in plankton data
author: "Name 1, Name 2, Name 3"
date: April XX, 2025
output: 
  html_document:
    code-folding: true
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
options(dplyr.summarise.inform = FALSE)
```

# Data

We chose to use the seven zooplankton taxa that had complete observations from 1985 to 1994. Thus, Leptodora and Neomysis were excluded. We chose to examine the last ten years of the data set (from 1985 to 1994) because we were interested in the zooplankton community following all of the changes to the lake throughout the 1970's and 1980's (Ben could explain this better probably). The covariates we looked at were temperature, total phosphorus, and pH.


## Load the data

```{r load_data, echo=FALSE}
#Packages
library(MARSS)
library(forecast)
library(zoo)
library(RColorBrewer)

## Wrangle the data
# Subset zooplankton data we will use
## load the raw data (there are 3 datasets contained here)
data(lakeWAplankton, package = "MARSS")
all_dat <- lakeWAplanktonTrans
yr_frst <- 1985
yr_last <- 1994
plank_dat <- all_dat[all_dat[, "Year"] >= yr_frst & 
                       all_dat[, "Year"] <= yr_last,]
#subset covariates
temp<-plank_dat[,c("Temp")]
ph<-plank_dat[,c("pH")]
tp<-plank_dat[,c("TP")]
greens<-plank_dat[,c("Greens")]
greens<-na.approx(greens, na.rm = FALSE)
cos_t <- cos(2 * pi * seq(TT) / 12)
sin_t <- sin(2 * pi * seq(TT) / 12)
szn<- rbind(cos_t, sin_t)

## create vector of phytoplankton group names
zooplankton<- c("Conochilus", "Cyclops","Daphnia", "Diaptomus", "Epischura","Non.daphnid.cladocerans", "Non.colonial.rotifers")
## get only the phytoplankton
dat_zoo<-plank_dat[,zooplankton]
dat_zoo<-t(dat_zoo)

## get number of time series
N_ts <- dim(dat_zoo)[1]

## get length of time series
TT <- dim(dat_zoo)[2] 
y_bar <- apply(dat_zoo, 1, mean, na.rm = TRUE)

## subtract the means
dat <- dat_zoo - y_bar

spp <- rownames(dat)
dat<-t(apply(dat, 1, function(x) na.approx(x, na.rm = FALSE)))

```

```{r, echo=FALSE}
palette<-brewer.pal(6, "Blues")
palette2<-brewer.pal(6, "Greens")
color<-c(palette[2:6], palette2[3:6])
## initialize a counter
cnt <- 1
## set up plotting space & make plots
par(mfrow = c(N_ts, 1), mai = c(0.5, 0.7, 0.1, 0.1), omi = c(0, 0, 0, 0))

for(i in 1:length(spp)){
  plot(dat[i,],  bty = "L", xaxt = "n", pch = 16,
       xlab = "",
       ylab = "Abundance index", col=color[i], type="b")
  axis(1, 12 * (0:dim(dat)[2]) + 1, yr_frst + 0:dim(dat)[2])
  title(spp[i])
  cnt <- cnt + 1
  }
```

# General tasks

Each group has the same general tasks, but you will adapt them as you work on the data.

1) Find the most parsimonious model among a set that examines the effects of environmental covariates and/or an indicator of seasonality, varying numbers of trends, and different forms of variance-covariance matrices for the observation errors.
```{r}
### No covariates, m=3
## Observation model
# Loadings matrix
Z_vals <- list("z11", 0  ,  0  ,
               "z21","z22",  0  ,
               "z31","z32","z33",
               "z41","z42","z43",
               "z51","z52","z53",
               "z61","z62","z63",
               "z71","z72","z73")
ZZ <- matrix(Z_vals, nrow = N_ts, ncol = 3, byrow = TRUE)
## 'aa' is the offset/scaling
aa <- "zero"

## 'DD' and 'd' are for covariates
DD <- "zero"  # matrix(0,mm,1)
dd <- "zero"  # matrix(0,1,wk_last)
## 'RR' is var-cov matrix for obs errors
RR <- "diagonal and unequal"
## Process eq
## number of processes
mm <- 3
## 'BB' is identity: 1's along the diagonal & 0's elsewhere
BB <- "identity"  # diag(mm)
## 'uu' is a column vector of 0's
uu <- "zero"  # matrix(0, mm, 1)
## 'CC' and 'cc' are for covariates
CC <- "zero"  # matrix(0, mm, 1)
cc <- "zero"  # matrix(0, 1, wk_last)
## 'QQ' is identity
QQ <- "identity"  # diag(mm)

mod_list <- list(Z = ZZ, A = aa, D = DD, d = dd, R = RR,
                 B = BB, U = uu, C = CC, c = cc, Q = QQ)
## list with model inits
init_list <- list(x0 = matrix(rep(0, mm), mm, 1))
## list with model control parameters
con_list <- list(maxit = 3000, allow.degen = TRUE)
dfa_1 <- MARSS(y = dat, model = mod_list, inits = init_list, control = con_list)
mod_list$R<-"diagonal and equal"
dfa_2<-MARSS(y = dat, model = mod_list, inits = init_list, control = con_list)
mod_list$R<-"equalvarcov"
dfa_3<-MARSS(y = dat, model = mod_list, inits = init_list, control = con_list)
aicoutput<-print(cbind(model = c("diag and unequal", "diag and equal", "equalvarcov"),
      AICc = round(c(dfa_1$AICc,
                     dfa_2$AICc,
                     dfa_3$AICc)),
      quote = FALSE))
mod_list$R<-"diagonal and equal"
# diag and unequal R structure is best
#fit most complex covariate form
dfa_all <- MARSS(dat, model = list(m=3, R="diagonal and equal"), form = "dfa",z.score = FALSE,
                  control = con_list, covariates = rbind(temp, ph, tp, greens))
dfa_szn3<-MARSS(dat, model = list(m=3, R="diagonal and equal"), form = "dfa",z.score = FALSE,
                  control = con_list, covariates = rbind(szn))

#The dummy seasonal model performed best
aicoutput<-print(cbind(model = c("All four covariates", "Dummy seasonal"),
      AICc = round(c(dfa_all$AICc,
                     dfa_szn$AICc))))

# Now going to try changing level with the szn model
dfa_szn2<-MARSS(dat, model = list(m=2, R="diagonal and equal"), form = "dfa",z.score = FALSE,
                  control = con_list, covariates = rbind(szn))
dfa_szn1<-MARSS(dat, model = list(m=1, R="diagonal and equal"), form = "dfa",z.score = FALSE,
                  control = con_list, covariates = rbind(szn))
dfa_szn4<-MARSS(dat, model = list(m=4, R="diagonal and equal"), form = "dfa",z.score = FALSE,
                  control = con_list, covariates = rbind(szn))
dfa_szn5<-MARSS(dat, model = list(m=5, R="diagonal and equal"), form = "dfa",z.score = FALSE,
                  control = con_list, covariates = rbind(szn))
dfa_szn6<-MARSS(dat, model = list(m=6, R="diagonal and equal"), form = "dfa",z.score = FALSE,
                  control = con_list, covariates = rbind(szn))

aicoutput<-print(cbind(model = c("m=1", "m=2", "m=3", "m=4", "m=5", "m=6"),
      AICc = round(c(dfa_szn1$AICc,
                     dfa_szn2$AICc,
                     dfa_szn3$AICc, dfa_szn4$AICc,
                     dfa_szn5$AICc, dfa_szn6$AICc))))
# m=6 is the best model, now going to test cutting covariates
dfa_tptempph <- MARSS(dat, model = list(m=6, R="diagonal and equal"), form = "dfa",z.score = FALSE,
                  control = con_list, covariates = rbind(temp, ph, tp))
dfa_tptemp <- MARSS(dat, model = list(m=6, R="diagonal and equal"), form = "dfa",z.score = FALSE,
                  control = con_list, covariates = rbind(temp, tp))
dfa_tp <- MARSS(dat, model = list(m=6, R="diagonal and equal"), form = "dfa",z.score = FALSE,
                  control = con_list, covariates = rbind(tp))
dfa_tptempgreens<-MARSS(dat, model = list(m=6, R="diagonal and equal"), form = "dfa",z.score = FALSE,
                  control = con_list, covariates = rbind(temp, tp, greens))
dfa_tpgreens<-MARSS(dat, model = list(m=6, R="diagonal and equal"), form = "dfa",z.score = FALSE,
                  control = con_list, covariates = rbind(tp, greens))
dfa_tempgreens<-MARSS(dat, model = list(m=6, R="diagonal and equal"), form = "dfa",z.score = FALSE,
                  control = con_list, covariates = rbind(temp, greens))
dfa_tempph<-MARSS(dat, model = list(m=6, R="diagonal and equal"), form = "dfa",z.score = FALSE,
                  control = con_list, covariates = rbind(temp, ph))
dfa_temp<-MARSS(dat, model = list(m=6, R="diagonal and equal"), form = "dfa",z.score = FALSE,
                  control = con_list, covariates = rbind(temp))
dfa_ph<-MARSS(dat, model = list(m=6, R="diagonal and equal"), form = "dfa",z.score = FALSE,
                  control = con_list, covariates = rbind(ph))
dfa_greens<-MARSS(dat, model = list(m=6, R="diagonal and equal"), form = "dfa",z.score = FALSE,
                  control = con_list, covariates = rbind(greens))

aicoutput<-print(cbind(model = c("All four", "-greens", "-ph", "TP + greens", "TP+temp", "Temp+ph",
                                 "TP", "Temp", "Greens", "pH", "Dummy seasonal"),
      AICc = round(c(dfa_all$AICc,
                     dfa_tptempph$AICc,
                     dfa_tptempgreens$AICc, dfa_tpgreens$AICc,
                     dfa_tptemp$AICc, dfa_tempph$AICc,
                     dfa_tp$AICc, dfa_temp$AICc, dfa_greens$AICc, dfa_ph$AICc, dfa_szn6$AICc))))

# The temp+ph model performed best

# Re-checking R structures
dfa_tempph.Runequal<-MARSS(dat, model = list(m=6, R="diagonal and unequal"), form = "dfa",
                           z.score = FALSE,control = con_list, covariates = rbind(temp, ph))

dfa_tempph.Requalvarcov<-MARSS(dat, model = list(m=6, R="equalvarcov"), form = "dfa",
                           z.score = FALSE,control = con_list, covariates = rbind(temp, ph))
aicoutput<-print(cbind(model = c("diag and equal", "diag and unequal", "equalvarcov"),
      AICc = round(c(dfa_tempph$AICc, dfa_tempph.Runequal$AICc, dfa_tempph.Requalvarcov$AICc))))
#Checking m levels
dfa_tempph6<-MARSS(dat, model = list(m=6, R="diagonal and unequal"), form = "dfa",z.score = FALSE,
                  control = con_list, covariates = rbind(temp, ph))
dfa_tempph5<-MARSS(dat, model = list(m=5, R="diagonal and unequal"), form = "dfa",z.score = FALSE,
                  control = con_list, covariates = rbind(temp, ph))
dfa_tempph4<-MARSS(dat, model = list(m=4, R="diagonal and unequal"), form = "dfa",z.score = FALSE,
                  control = con_list, covariates = rbind(temp, ph))
dfa_tempph3<-MARSS(dat, model = list(m=3, R="diagonal and unequal"), form = "dfa",z.score = FALSE,
                  control = con_list, covariates = rbind(temp, ph))
dfa_tempph2<-MARSS(dat, model = list(m=2, R="diagonal and unequal"), form = "dfa",z.score = FALSE,
                  control = con_list, covariates = rbind(temp, ph))
dfa_tempph1<-MARSS(dat, model = list(m=1, R="diagonal and unequal"), form = "dfa",z.score = FALSE,
                  control = con_list, covariates = rbind(temp, ph))
aicoutput<-print(cbind(model = c("m=1", "m=2", "m=3", "m=4", "m=5", "m=6"),
      AICc = round(c(dfa_tempph1$AICc,
                     dfa_tempph2$AICc,
                     dfa_tempph3$AICc, dfa_tempph4$AICc,
                     dfa_tempph5$AICc,
                     dfa_tempph6$AICc))))

# the model with 6 levels is still "best", but performs about the same as m=5 so m=5 is most parsimonious
dfa_tempph5
```


2) Plot trends and individual loadings for the model identified in task (1) above.
```{r, echo=FALSE}
Z_est <- coef(dfa_tempph5, type = "matrix")$Z

## get the inverse of the rotation matrix
H_inv <- varimax(Z_est)$rotmat

Z_rot = Z_est %*% H_inv   

## rotate processes
proc_rot = solve(H_inv) %*% dfa_tempph5$states

ylbl <- zooplankton
w_ts <- seq(dim(dat)[2])
layout(matrix(c(1,2,3,4,5,6), mm, 2), widths = c(2,1))
par(mai = c(0.5, 0.5, 0.5, 0.1), omi = c(0, 0, 0, 0))

## plot the processes
for(i in 1:5) {
  ylm <- c(-1, 1) * max(abs(proc_rot[i,]))
  ## set up plot area
  plot(w_ts,proc_rot[i,], type = "n", bty = "L",
       ylim = ylm, xlab = "", ylab = "", xaxt = "n")
  ## draw zero-line
  abline(h = 0, col = "gray")
  ## plot trend line
  lines(w_ts, proc_rot[i,], lwd = 2)
  lines(w_ts, proc_rot[i,], lwd = 2)
  ## add panel labels
  mtext(paste("State",i), side = 3, line = 0.5)
  axis(1, 12 * (0:dim(dat_1980)[2]) + 1, yr_frst + 0:dim(dat_1980)[2])
}
## plot the loadings
minZ <- 0
ylm <- c(-1, 1) * max(abs(Z_rot))
for(i in 1:5) {
  plot(x = c(1:N_ts)[abs(Z_rot[,i])>minZ],
       y = as.vector(Z_rot[abs(Z_rot[,i])>minZ,i]),
       type = "h",
       lwd = 2, xlab = "", ylab = "", xaxt = "n", ylim = ylm,
       xlim = c(0.5, N_ts + 0.5), col = clr)
  for(j in 1:N_ts) {
    if(Z_rot[j,i] > minZ) {text(j, -0.03, ylbl[j], srt = 90, adj = 1, cex = 1.2, col = clr[j])}
    if(Z_rot[j,i] < -minZ) {text(j, 0.03, ylbl[j], srt = 90, adj = 0, cex = 1.2, col = clr[j])}
    abline(h = 0, lwd = 1.5, col = "gray")
  } 
  mtext(paste("Factor loadings on state", i), side = 3, line = 0.5)
}

## set up plotting area
par(mai = c(0.9,0.9,0.1,0.1))

## plot CCF's
ccf(proc_rot[1,],proc_rot[2,], lag.max = 12, main="")


```
3) Plot model fits and uncertainty intervals for the model identified in task (1) above.
```{r}
get_DFA_fits <- function(MLEobj, dd = NULL, alpha = 0.05) {
  ## empty list for results
  fits <- list()
  ## extra stuff for var() calcs
  Ey <- MARSS:::MARSShatyt(MLEobj)
  ## model params
  ZZ <- coef(MLEobj, type="matrix")$Z
  ## number of obs ts
  nn <- dim(Ey$ytT)[1]
  ## number of time steps
  TT <- dim(Ey$ytT)[2]
  ## get the inverse of the rotation matrix
  H_inv <- varimax(ZZ)$rotmat
  ## check for covars
  if(!is.null(dd)) {
    DD <- coef(MLEobj, type = "matrix")$D
    ## model expectation
    fits$ex <- ZZ %*% H_inv %*% MLEobj$states + DD %*% dd
  } else {
    ## model expectation
    fits$ex <- ZZ %*% H_inv %*% MLEobj$states
  }
  ## Var in model fits
  VtT <- MARSSkfss(MLEobj)$VtT
  VV <- NULL
  for(tt in 1:TT) {
    RZVZ <- coef(MLEobj, type = "matrix")$R - ZZ %*% VtT[,,tt] %*% t(ZZ)
    SS <- Ey$yxtT[,,tt] - Ey$ytT[,tt,drop = FALSE] %*% t(MLEobj$states[,tt,drop = FALSE])
    VV <- cbind(VV, diag(RZVZ + SS %*% t(ZZ) + ZZ %*% t(SS)))
  }
  SE <- sqrt(VV)
  ## upper & lower (1-alpha)% CI
  fits$up <- qnorm(1-alpha/2)*SE + fits$ex
  fits$lo <- qnorm(alpha/2)*SE + fits$ex
  return(fits)
}

mod_fit <- get_DFA_fits(dfa_tempph5)

## set up plotting area
par(mfrow = c(N_ts, 1), mai = c(0.5, 0.7, 0.1, 0.1), omi = c(0, 0, 0, 0))

## plot the fits
for(i in 1:N_ts) {
  up <- mod_fit$up[i,]
  mn <- mod_fit$ex[i,]
  lo <- mod_fit$lo[i,]
  plot(w_ts, mn, type = "n", ylim = c(min(lo), max(up)),
       cex.lab = 1.2,
       xlab = "", ylab = ylbl[i], xaxt = "n")
  axis(1, 12 * (0:dim(dat)[2]) + 1, yr_frst + 0:dim(dat)[2])
  points(w_ts,dat[i,], pch = 16, col = clr[i])
  lines(w_ts, up, col = "darkgray")
  lines(w_ts, mn, col = "black", lwd = 2)
  lines(w_ts, lo, col = "darkgray")
}
```
4) Describe the effects of environmental or dummy variables on (possibly seasonal) patterns in the data.


# Methods

Please address the following in your methods:

We chose to use the seven zooplankton taxa that had complete observations from 1985 to 1994. Thus, Leptodora and Neomysis were excluded. We chose to examine the last ten years of the data set (from 1985 to 1994) because we were interested in the zooplankton community following all of the changes to the lake throughout the 1970's and 1980's (Ben could explain this better probably). The covariates we looked at were temperature, total phosphorus, and pH.



* What environmental or dummy variables did you include and why?

* What forms of models did you fit (ie, write them out in matrix form)?

* What sort of model diagnostics did you use to examine model assumptions?


# Results


# Discussion


# Team contributions

Example: "All team members helped decide on the goal and ran the analyses for the individual regions. Team members 2 & 3 wrote most of the code for the analysis. Team member 1 worked on the plotting section of the report using and adapting code that team member 3 wrote. All team members helped edit the report and wrote the discussion together."

