---
title: Team 1 - Lab 1
subtitle: Dynamic Factor Analysis (DFA) - Common trends in plankton data
author: "Name 1, Name 2, Name 3"
date: April XX, 2025
output: 
  html_document:
    code-folding: true
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
options(dplyr.summarise.inform = FALSE)
```

# Data

We are interesting in looking at how all of the zooplankton species respond in the most recent years (1985 - 1994) other than Neomysis and __________, which there were very few measurements of during this time period. 

## Load the data

```{r load_data}
## load MARSS for data and analyses
library(MARSS)

## load the raw data (there are 3 datasets contained here)
data(lakeWAplankton, package = "MARSS")

## we want `lakeWAplanktonTrans`, which has been transformed
## so the 0's are replaced with NA's and the data z-scored
all_dat <- lakeWAplanktonTrans

```

## Wrangle the data

```{r wrangle_data}
library(dplyr)

data(lakeWAplankton)
# we want lakeWAplanktonTrans, which has been log-transformed
# and the 0s replaced with NAs
plankdat <- lakeWAplanktonTrans
years <- plankdat[, "Year"] >= 1985 & plankdat[, "Year"] < 1994
plankton <- c(
"Conochilus", "Cyclops", "Daphnia",
"Diaptomus", "Epischura","Non.daphnid.cladocerans","Non.colonial.rotifers"
)
plankton <- plankdat[years, plankton]
Temp <- t(plankdat[years, "Temp", drop = FALSE])
TP <- t(plankdat[years, "TP", drop = FALSE])
pH <- t(plankdat[years, "pH", drop = FALSE])

plankton <- t(plankton)
N.ts <- nrow(plankton)
TT <- ncol(plankton)


```



#According to lecture, the steps are first to... 
- fit the most complex model you can envision based on all of your possible
  covariates and random factors (states)
  
-keep the covariates fixed and choose the number of trends (states) using AICc

- keep the covariates & states fixed and choose the form for R

- sort out the covariates while keeping the states & R fixed




1. fit the most complex model you can envision based on all the possible covariates. 

```{r, echo = FALSE, results = FALSE}
# List of covariates
covariates <- rbind(Temp, TP, pH)

# Initialize control list
cntl.list <- list(maxit = 50000)

# Initialize AICc storage
aicc_values <- numeric(7)
state_numbers <- 1:7

# Loop over states from 1 to 7
for (m in state_numbers) {
  model.list <- list(m = m, R = "diagonal and unequal")
  
  # Fit the model for each number of states
  model <- suppressMessages(suppressWarnings(
    MARSS(plankton,
          model = model.list,
          z.score = TRUE, 
          form = "dfa", 
          control = cntl.list, 
          covariates = covariates, 
          method = "BFGS")
  ))
  
  # Store the AICc value
  aicc_values[m] <- model$AICc
}

# Print the results
print(cbind(
  model = paste(state_numbers, "States", sep = ""),
  AICc = round(aicc_values)
), quote = FALSE)

ordered_results <- cbind(
  model = paste(state_numbers, "States", sep = ""),
  AICc = round(aicc_values)
)

# Print the ordered results
print(ordered_results, quote = FALSE)

# model   AICc
# [1,] 6States 1546
# [2,] 7States 1548
# [3,] 5States 1552
# [4,] 4States 1566
# [5,] 3States 1578
# [6,] 2States 1595
# [7,] 1States 1644


```

## ok now we've decided on a six state model, and we can test different forms for R 

```{r}

m<-6

# "diagonal and equal" - Same observation error for all species
R_equal <- MARSS(plankton,
  model = list(m =m, R = "diagonal and equal"),
  z.score = TRUE, 
  form = "dfa", 
  control = cntl.list, 
  covariates = rbind(Temp, TP, pH), 
  method = "BFGS"
)

# "unconstrained" - Allow correlated observation errors
R_unconstrained <- MARSS(plankton,
  model = list(m = m  , R = "unconstrained"),
  z.score = TRUE, 
  form = "dfa", 
  control = cntl.list, 
  covariates = rbind(Temp, TP, pH), 
  method = "BFGS"
)

# Compare AICc values
print(cbind(
  model = c("diag_unequal", "diag_equal", "unconstrained"),
  AICc = round(c(
      aicc_values[1], 
      R_equal$AICc, 
      R_unconstrained$AICc)
  )
), quote = FALSE)

#    model         AICc
# [1,] diag_unequal  1644
# [2,] diag_equal    1566
# [3,] unconstrained 1555


```


# Now test covariate combinations 

```{r}
# Set up your control list for all models
cntl.list <- list(maxit = 500)
best_R <- "unconstrained"  # Your best R structure
best_m <- 6  # Your best number of states

# No covariates (baseline model)
No_Covars <- MARSS(plankton,
  model = list(m = best_m, R = best_R),
  z.score = TRUE, 
  form = "dfa", 
  control = cntl.list, 
  method = "BFGS"
)

# Single covariates
Temp_only <- MARSS(plankton,
  model = list(m = best_m, R = best_R),
  z.score = TRUE, 
  form = "dfa", 
  control = cntl.list, 
  covariates = rbind(Temp), 
  method = "BFGS"
)

TP_only <- MARSS(plankton,
  model = list(m = best_m, R = best_R),
  z.score = TRUE, 
  form = "dfa", 
  control = cntl.list, 
  covariates = rbind(TP), 
  method = "BFGS"
)

pH_only <- MARSS(plankton,
  model = list(m = best_m, R = best_R),
  z.score = TRUE, 
  form = "dfa", 
  control = cntl.list, 
  covariates = rbind(pH), 
  method = "BFGS"
)

# Pairs of covariates
Temp_TP <- MARSS(plankton,
  model = list(m = best_m, R = best_R),
  z.score = TRUE, 
  form = "dfa", 
  control = cntl.list, 
  covariates = rbind(Temp, TP), 
  method = "BFGS"
)

Temp_pH <- MARSS(plankton,
  model = list(m = best_m, R = best_R),
  z.score = TRUE, 
  form = "dfa", 
  control = cntl.list, 
  covariates = rbind(Temp, pH), 
  method = "BFGS"
)

TP_pH <- MARSS(plankton,
  model = list(m = best_m, R = best_R),
  z.score = TRUE, 
  form = "dfa", 
  control = cntl.list, 
  covariates = rbind(TP, pH), 
  method = "BFGS"
)

# All covariates (your current best model)
All_Covars <- MARSS(plankton,
  model = list(m = best_m, R = best_R),
  z.score = TRUE, 
  form = "dfa", 
  control = cntl.list, 
  covariates = rbind(Temp, TP, pH), 
  method = "BFGS"
)

# Compare AICc values
models_df <- data.frame(
  model = c("No_Covars", "Temp_only", "TP_only", "pH_only", 
            "Temp_TP", "Temp_pH", "TP_pH", "All_Covars"),
  AICc = c(No_Covars$AICc, Temp_only$AICc, TP_only$AICc, pH_only$AICc,
           Temp_TP$AICc, Temp_pH$AICc, TP_pH$AICc, All_Covars$AICc)
)

# Sort by AICc
models_df <- models_df[order(models_df$AICc),]
print(models_df)

# 8	All_Covars	1554.900		
# 6	Temp_pH	1555.025		
# 7	TP_pH	1591.661		
# 2	Temp_only	1598.316		
# 5	Temp_TP	1599.731		
# 4	pH_only	1604.539		
# 3	TP_only	1633.710		
# 1	No_Covars	1652.562


```

The most parsimonious model includes 6 states and all covariates 

# Plot trends and individual loadings 
```{r}

# Extract the key components from your best model
Best_Model <- All_Covars

# Get the Z matrix (factor loadings)
Z_est <- coef(Best_Model, type = "matrix")$Z

# Since you have only one state in your model, make sure Z is properly formatted
if(!is.matrix(Z_est)) {
  Z_est <- matrix(Z_est, ncol = 1)
}

# Number of time series (species)
N.ts <- nrow(plankton)
spp <- rownames(plankton)

# Set threshold for minimum loading to display
minZ <- 0.05

# Number of states/trends
m <- ncol(Z_est)

# Set y-axis limits
ylims <- c(-1.1 * max(abs(Z_est)), 1.1 * max(abs(Z_est)))

# Create plot
par(mfrow = c(ceiling(m/2), min(2, m)), mar = c(3, 4, 1.5, 0.5), oma = c(0.4, 1, 1, 1))

for(i in 1:m) {
  # Plot factor loadings as vertical lines
  plot(c(1:N.ts)[abs(Z_est[, i]) > minZ], 
       as.vector(Z_est[abs(Z_est[, i]) > minZ, i]), 
       type = "h", 
       lwd = 2, 
       xlab = "", 
       ylab = "", 
       xaxt = "n", 
       ylim = ylims, 
       xlim = c(0, N.ts+1))
  
  # Add species labels
  for(j in 1:N.ts) {
    if(Z_est[j, i] > minZ) {
      text(j, -0.05, spp[j], srt = 90, adj = 1, cex = 0.9)
    }
    if(Z_est[j, i] < -minZ) {
      text(j, 0.05, spp[j], srt = 90, adj = 0, cex = 0.9)
    }
  }
  
  # Add horizontal line at zero
  abline(h = 0, lwd = 1, col = "gray")
  
  # Add title
  mtext(paste("Factor loadings on trend", i, sep = " "), side = 3, line = .5)
}

```

# Plot Model fit and uncertainty 
```{r}

```

# Describe the effects of environmental or dummy variables on patterns in the data 

```{r}

```



# General tasks

Each group has the same general tasks, but you will adapt them as you work on the data.

1) Find the most parsimonious model among a set that examines the effects of environmental covariates and/or an indicator of seasonality, varying numbers of trends, and different forms of variance-covariance matrices for the observation errors.

2) Plot trends and individual loadings for the model identified in task (1) above.

3) Plot model fits and uncertainty intervals for the model identified in task (1) above.

4) Describe the effects of environmental or dummy variables on (possibly seasonal) patterns in the data.


# Methods

Please address the following in your methods:

* Which plankton taxa did you choose and how did you choose them?
   
* What time period(s) did you examine and why?

* What environmental or dummy variables did you include and why?

* What forms of models did you fit (ie, write them out in matrix form)?

* What sort of model diagnostics did you use to examine model assumptions?


# Results


# Discussion


# Team contributions

Example: "All team members helped decide on the goal and ran the analyses for the individual regions. Team members 2 & 3 wrote most of the code for the analysis. Team member 1 worked on the plotting section of the report using and adapting code that team member 3 wrote. All team members helped edit the report and wrote the discussion together."



````````````````````````````````






# the complicated way of doing things (all the matrix math and such)
```{r}


# Record the number of rows 
N.ts <- nrow(plankton)  # Number of observation time series (3 plankton species)
N.states <- 3  # Let's assume 3 hidden states (one per plankton group, but could be different)

# Set up the Z matrix (observation matrix)
# For 3 plankton species, 3 states
Z.vals <- list(
  "z11", 0, 0,        # Conochilus only affected by state 1
  "z21", "z22", 0,    # Cyclops affected by states 1 and 2
  "z31", "z32", "z33" # Daphnia affected by all three states
)
Z <- matrix(Z.vals, nrow = nrow(plankton), ncol = 3, byrow = TRUE)

Q <- B <- diag(1, 3)

# R.vals <- list(
# "r11", 0, 0, 
# 0, "r22", 0, 
# 0, 0, "r33"
# )

R <-"diagonal and unequal"
x0 <- U <- A <- "zero"
V0 <- diag(3, 3)

dfa.model <- list(
Z = Z, A = "zero", R = R, B = B, U = U,
Q = Q, x0 = x0, V0 = V0
)
cntl.list <- list(maxit = 500000)

ThreeStates<- MARSS(plankton , model = dfa.model, control = cntl.list)

```

# The much simpler way

```{r}

model.list <- list(m = 3, R = "diagonal and unequal")
cntl.list <- list(maxit = 50000)

ThreeStates <- MARSS(plankton,
  model = model.list,
  z.score = TRUE, 
  form = "dfa", 
  control = cntl.list, 
  method="BFGS"
)




```
