---
title: Team 3 - Lab 3
subtitle: Dynamic Factor Analysis (DFA) - Common trends in plankton data
<<<<<<< HEAD
author: "Ben Makhlouf, Nicole Doran, Lilac Hong"
date: May 6, 2025
=======
author: "Ben Makhlouf, Lilac Hong, Nicole Doran"
date: May6, 2025
>>>>>>> 761469d ( test)
output: 
  html_document:
    code-folding: true
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
options(dplyr.summarise.inform = FALSE)
```

# Data

We are interesting in looking at how all of the zooplankton species respond in the most recent years (1985 - 1994) other than Leptodora and Neomysis and, which there were very few measurements of during this time period. Despite the well documented changes to zooplankton community dynamics following the treatment of sewage early in the data, our group was curious about what was happening to phytoplankton abundances in the more recent time period. 

## Load the data

```{r load_data}
## load MARSS for data and analyses
library(MARSS)
library(RColorBrewer)
data(lakeWAplankton, package = "MARSS")
all_dat <- lakeWAplanktonTrans
```

## Wrangle the data
```{r wrangle_data}
library(dplyr)
plankdat <- lakeWAplanktonTrans
years <- plankdat[, "Year"] >= 1985 & plankdat[, "Year"] < 1994
plankton <- c(
"Conochilus", "Cyclops", "Daphnia",
"Diaptomus", "Epischura","Non.daphnid.cladocerans","Non.colonial.rotifers"
)
plankton <- plankdat[years, plankton]
Temp <- t(plankdat[years, "Temp", drop = FALSE])
TP <- t(plankdat[years, "TP", drop = FALSE])
pH <- t(plankdat[years, "pH", drop = FALSE])

plankton <- t(plankton)
N.ts <- nrow(plankton)
TT <- ncol(plankton)
```
## Visualizing the Data
```{r, echo=FALSE}
palette<-brewer.pal(6, "Blues")
palette2<-brewer.pal(6, "Greens")
color<-c(palette[2:6], palette2[3:6])
## initialize a counter
cnt <- 1
## set up plotting space & make plots
par(mfrow = c(N.ts, 1), mai = c(0.5, 0.7, 0.1, 0.1), omi = c(0, 0, 0, 0))

spp <- rownames(plankton)
## get length of time series
TT <- dim(plankton)[2] 
yr_frst<-1985
yr_last<-1994
for(i in 1:length(spp)){
  plot(plankton[i,],  bty = "L", xaxt = "n", pch = 16,
       xlab = "",
       ylab = "Abundance index", col=color[i], type="b")
  axis(1, 12 * (0:dim(plankton)[2]) + 1, yr_frst + 0:dim(plankton)[2])
  title(spp[i])
  cnt <- cnt + 1
  }
```
#According to lecture, the steps are to... 

- fit the most complex model you can envision based on all of your possible covariates and random factors (states)
  
-keep the covariates fixed and choose the number of trends (states) using AICc

- keep the covariates & states fixed and choose the form for R

- sort out the covariates while keeping the states & R fixed

#1. Fit the Dummy variable (Seasonal component) and the most complex model 

```{r, echo = FALSE, results = FALSE, include = FALSE}
# List of covariates
covariates <- rbind(Temp, TP, pH)

# Initialize control list
cntl.list <- list(maxit = 50000)

# First, all the covariates 
model.list <- list(m = 7, R = "diagonal and unequal") #All unique but
allcov <- suppressMessages(suppressWarnings(
    MARSS(plankton,
          model = model.list,
          z.score = TRUE, 
          form = "dfa", 
          control = cntl.list, 
          covariates = covariates, 
          method = "BFGS")))

# Then, Dummy variable (seasonal)

# Create the dummy variable 

cos_t <- cos(2 * pi * seq(TT) / 12)
sin_t <- sin(2 * pi * seq(TT) / 12)
szn<- rbind(cos_t, sin_t)

# Add this to the covariates

model.list <- list(m = 7, R = "diagonal and unequal") #All unique but
sznmodel <- suppressMessages(suppressWarnings(
    MARSS(plankton,
          model = model.list,
          z.score = TRUE, 
          form = "dfa", 
          control = cntl.list, 
          covariates = szn, 
          method = "BFGS")))

# Compare AICc
aicc_allcov <- allcov$AICc
aicc_szn <- sznmodel$AICc

# Print AICc values
print(cbind(
  model = c("All Covariates", "Seasonal Dummy"),
  AICc = round(c(aicc_allcov, aicc_szn))
), quote = FALSE)

# model          AICc
# [1,] All Covariates 1548
# [2,] Seasonal Dummy 1558
# The AICc for the model with all covariates is lower than the seasonal dummy model,
```


# All covariates works best. As such, we will now test the optimal number of states using all covariates 

```{r}
# Initialize AICc storage
aicc_values <- numeric(7)
state_numbers <- 1:7

# Loop over states from 1 to 7
for (m in state_numbers) {
  model.list <- list(m = m, R = "diagonal and unequal") 
  #All unique but uncorrelated 
  
  # Fit the model for each number of states
  model <- suppressMessages(suppressWarnings(
    MARSS(plankton,
          model = model.list,
          z.score = TRUE, 
          form = "dfa", 
          control = cntl.list, 
          covariates = covariates, 
          method = "BFGS")
  ))
  
  # Store the AICc value
  aicc_values[m] <- model$AICc
}
```
```{r, echo=FALSE}
# Print the results
print(cbind(
  model = paste(state_numbers, "States", sep = ""),
  AICc = round(aicc_values)
), quote = FALSE)

ordered_results <- cbind(
  model = paste(state_numbers, "States", sep = ""),
  AICc = round(aicc_values)
)

# Print the ordered results
print(ordered_results, quote = FALSE)
# model   AICc
# [1,] 1States 1644
# [2,] 2States 1595
# [3,] 3States 1578
# [4,] 4States 1566
# [5,] 5States 1552
# [6,] 6States 1546 # Best!
# [7,] 7States 1548 # Second best, close .... 

## ok now we've decided on a 6 state model, and we can test different forms for R
```


```{r, echo = FALSE, results = FALSE, include = FALSE}
# Set this equal to the number of best states 
m<-6

# "diagonal and equal" - Same observation error for all species
R_equal <- MARSS(plankton,
  model = list(m =m, R = "diagonal and equal"),
  z.score = TRUE, 
  form = "dfa", 
  control = cntl.list, 
  covariates = rbind(Temp, TP, pH), 
  method = "BFGS"
)


# 'diagonal and unequal' - Different observation error for each species
R_unequal <- MARSS(plankton,
  model = list(m = m, R = "diagonal and unequal"),
  z.score = TRUE, 
  form = "dfa", 
  control = cntl.list, 
  covariates = rbind(Temp, TP, pH), 
  method = "BFGS"
)

# "unconstrained" - Allow correlated observation errors
R_unconstrained <- MARSS(plankton,
  model = list(m = m  , R = "unconstrained"),
  z.score = TRUE, 
  form = "dfa", 
  control = cntl.list, 
  covariates = rbind(Temp, TP, pH), 
  method = "BFGS"
)

### I couldnt get this one to work... might have something to do with the unconstrained problem. 
# Diagonal and unequal varcov 
# Indpendent observation error and ..*figure out what this means* 

# R_unconstrained <- MARSS(plankton,
#   model = list(m = m  , R = "equalvarcov"),
#   z.score = TRUE, 
#   form = "dfa", 
#   control = cntl.list, 
#   covariates = rbind(Temp, TP, pH), 
#   method = "BFGS"
# )
```
```{r, echo=FALSE}
# Compare AICc values
print(cbind(
  model = c("diag_equal", "diag_unequal", "unconstrained"), #"equalvarcov"),
  AICc = round(c(
    R_equal$AICc,
    R_unequal$AICc,
    R_unconstrained$AICc
    #R_unconstrained$AICc
  ))
), quote = FALSE)

# model         AICc
# [1,] diag_equal    1566
# [2,] diag_unequal  1546 #Best 
# [3,] unconstrained 1555
# The best R Structure is diagonal and unequal, meaning each species has a different observation error and they are not correlated 
# Now test covariate combinations 
```



```{r, echo = FALSE, results = FALSE, include = FALSE}
# Set up your control list for all models
cntl.list <- list(maxit = 500)
best_R <- "diagonal and unequal"  # Your best R structure
best_m <- 6  # Your best number of states

# No covariates (baseline model)
No_Covars <- MARSS(plankton,
  model = list(m = best_m, R = best_R),
  z.score = TRUE, 
  form = "dfa", 
  control = cntl.list, 
  method = "BFGS"
)

# Single covariates
Temp_only <- MARSS(plankton,
  model = list(m = best_m, R = best_R),
  z.score = TRUE, 
  form = "dfa", 
  control = cntl.list, 
  covariates = rbind(Temp), 
  method = "BFGS"
)

TP_only <- MARSS(plankton,
  model = list(m = best_m, R = best_R),
  z.score = TRUE, 
  form = "dfa", 
  control = cntl.list, 
  covariates = rbind(TP), 
  method = "BFGS"
)

pH_only <- MARSS(plankton,
  model = list(m = best_m, R = best_R),
  z.score = TRUE, 
  form = "dfa", 
  control = cntl.list, 
  covariates = rbind(pH), 
  method = "BFGS"
)

# Pairs of covariates
Temp_TP <- MARSS(plankton,
  model = list(m = best_m, R = best_R),
  z.score = TRUE, 
  form = "dfa", 
  control = cntl.list, 
  covariates = rbind(Temp, TP), 
  method = "BFGS"
)

Temp_pH <- MARSS(plankton,
  model = list(m = best_m, R = best_R),
  z.score = TRUE, 
  form = "dfa", 
  control = cntl.list, 
  covariates = rbind(Temp, pH), 
  method = "BFGS"
)

TP_pH <- MARSS(plankton,
  model = list(m = best_m, R = best_R),
  z.score = TRUE, 
  form = "dfa", 
  control = cntl.list, 
  covariates = rbind(TP, pH), 
  method = "BFGS"
)

# All covariates (your current best model)
All_Covars <- MARSS(plankton,
  model = list(m = best_m, R = best_R),
  z.score = TRUE, 
  form = "dfa", 
  control = cntl.list, 
  covariates = rbind(Temp, TP, pH), 
  method = "BFGS"
)

# Compare AICc values
models_df <- data.frame(
  model = c("No_Covars", "Temp_only", "TP_only", "pH_only", 
            "Temp_TP", "Temp_pH", "TP_pH", "All_Covars"),
  AICc = c(No_Covars$AICc, Temp_only$AICc, TP_only$AICc, pH_only$AICc,
           Temp_TP$AICc, Temp_pH$AICc, TP_pH$AICc, All_Covars$AICc)
)
```
```{r, echo=FALSE}
# Sort by AICc
models_df <- models_df[order(models_df$AICc),]
print(models_df)

# 6	Temp_pH	1543.220 # BEST 		
# 8	All_Covars	1545.882		
# 2	Temp_only	1589.997		
# 7	TP_pH	1590.000		
# 5	Temp_TP	1590.515		
# 4	pH_only	1609.318		
# 3	TP_only	1632.277		
# 1	No_Covars	1656.677	

# ##### 
```
```{r, include=FALSE}
#Check if fewer states has lower AICc for temp-ph model
Temp_pH_5 <- MARSS(plankton,
  model = list(m = 5, R = best_R),
  z.score = TRUE, 
  form = "dfa", 
  control = cntl.list, 
  covariates = rbind(Temp, pH), 
  method = "BFGS"
)

Temp_pH_4 <- MARSS(plankton,
  model = list(m = 4, R = best_R),
  z.score = TRUE, 
  form = "dfa", 
  control = cntl.list, 
  covariates = rbind(Temp, pH), 
  method = "BFGS"
)
```
```{r, echo=FALSE}
models_df <- data.frame(
  model = c("m=6", "m=5", "m=4"),
  AICc = c(Temp_pH$AICc, Temp_pH_5$AICc, Temp_pH_4$AICc)
)
print(models_df)

#Model #AICc
#m=6	1543.220			
#m=5	1545.452			
#m=4	1558.606	

### The temp_ph model with 5 states performs about the same as 6 states, and is more parsimonious so this is our best and most parsimonious model

```

The most parsimonious model includes 5 states with diagonal and unequal R structure and temperature and pH as covariates.

# Matrix form
DFA model with covariates has the following structure:
$$
\mathbf{y}_t = \mathbf{Z} \mathbf{x}_t + \mathbf{a} + \mathbf{D} \mathbf{d}_t + \mathbf{v}_t \quad \text{where } \mathbf{v}_t \sim \text{MVN}(0, \mathbf{R})
$$

$$
\mathbf{x}_t = \mathbf{x}_{t-1} + \mathbf{w}_t \quad \text{where } \mathbf{w}_t \sim \text{MVN}(0, \mathbf{Q})
$$

Using the constraints, the observation equation for the DFA model above becomes:
$$
\begin{bmatrix}
y_1 \\
y_2 \\
y_3 \\
\vdots \\
y_{108}
\end{bmatrix}_t
=
\begin{bmatrix}
z_{1,1} & 0 & 0 & 0 & 0 \\
z_{2,1} & z_{2,2} & 0 & 0 & 0 \\
z_{3,1} & z_{3,2} & z_{3,3} & 0 & 0 \\
\vdots & \vdots & \vdots & \vdots & \vdots \\
z_{108,1} & z_{108,2} & z_{108,3} & z_{108,4} & z_{108,5}
\end{bmatrix}
\begin{bmatrix}
x_1 \\
x_2 \\
x_3 \\
x_4 \\
x_5
\end{bmatrix}_t
+
\begin{bmatrix}
0 \\
0 \\
0 \\
\vdots \\
0
\end{bmatrix}
+
\begin{bmatrix}
d_{1,temp} & d_{1,pH} \\
d_{2,temp} & d_{2,pH} \\
d_{3,temp} & d_{3,pH} \\
\vdots & \vdots \\
d_{108,temp} & d_{108,pH}
\end{bmatrix}
\begin{bmatrix}
\text{temp} \\
\text{pH} 
\end{bmatrix}_t
+
\begin{bmatrix}
v_1 \\
v_2 \\
v_3 \\
\vdots \\
v_{108}
\end{bmatrix}_t
$$ 

where $\textbf{v}_t \sim \mathcal{N}(0,\textbf{R})$ and:
$$
\textbf{R} = 
\begin{bmatrix}
r_1 & 0 & 0 & \cdots & 0 \\
0 & r_2 & 0 & \cdots & 0 \\
0 & 0 & r_3 & 0 & 0 \\
\vdots & \vdots & \vdots & \ddots & \vdots \\
0 & 0 & 0 & 0 & r_{108}
\end{bmatrix}
$$

and the process equation becomes:
$$
\begin{bmatrix}
x_1 \\
x_2 \\
x_3 \\
x_4 \\
x_5
\end{bmatrix}_t
=
\begin{bmatrix}
1 & 0 & 0 & 0 & 0 \\
0 & 1 & 0 & 0 & 0 \\
0 & 0 & 1 & 0 & 0 \\
0 & 0 & 0 & 1 & 0 \\
0 & 0 & 0 & 0 & 1
\end{bmatrix}
\begin{bmatrix}
x_1 \\
x_2 \\
x_3 \\
x_4 \\
x_5
\end{bmatrix}_{t-1}
+
\begin{bmatrix}
w_1 \\
w_2 \\
w_3 \\
w_4 \\
w_5
\end{bmatrix}_t \; \text{where} \: \textbf{w}_t \sim \mathcal{N}(0,\textbf{Q})
$$

where $\mathbf{Q}$ is set equal to the identity matrix $\mathbf{I}_m$.

# Plot trends and individual loadings 
```{r, echo=FALSE}
Z_est <- coef(Temp_pH_5, type = "matrix")$Z
mm<-5
## get the inverse of the rotation matrix
H_inv <- varimax(Z_est)$rotmat

Z_rot = Z_est %*% H_inv   

## rotate processes
proc_rot = solve(H_inv) %*% Temp_pH_5$states

w_ts <- seq(dim(plankton)[2])
par(mai = c(0.5, 0.5, 0.5, 0.1), omi = c(0, 0, 0, 0))

## plot the processes
for(i in 1:mm) {
  ylm <- c(-1, 1) * max(abs(proc_rot[i,]))
  ## set up plot area
    plot(w_ts,proc_rot[i,], type = "n", bty = "L",
         ylim = ylm, xlab = "", ylab = "", xaxt = "n")
    ## draw zero-line
    abline(h = 0, col = "gray")
    ## plot trend line
    lines(w_ts, proc_rot[i,], lwd = 2)
    lines(w_ts, proc_rot[i,], lwd = 2)
    ## add panel labels
    mtext(paste("State",i), side = 3, line = 0.5)
    axis(1, 12 * (0:dim(plankton)[2]) + 1, yr_frst + 0:dim(plankton)[2])
}
## plot the loadings
minZ <- 0
ylm <- c(-1, 1) * max(abs(Z_rot))
for(i in 1:mm) {
  plot(x = c(1:N.ts)[abs(Z_rot[,i])>minZ],
       y = as.vector(Z_rot[abs(Z_rot[,i])>minZ,i]),
       type = "h",
       lwd = 2, xlab = "", ylab = "", xaxt = "n", ylim = ylm,
       xlim = c(0.5, N.ts + 0.5), col = c("black"))
  for(j in 1:N.ts) {
    if(Z_rot[j,i] > minZ) {text(j, -0.03, spp[j], srt = 90, adj = 1, cex = 1.2, col = color[j])}
    if(Z_rot[j,i] < -minZ) {text(j, 0.03, spp[j], srt = 90, adj = 0, cex = 1.2, col = color[j])}
    abline(h = 0, lwd = 1.5, col = "gray")
  } 
  mtext(paste("Factor loadings on state", i), side = 3, line = 0.5)
}

## set up plotting area
par(mai = c(0.9,0.9,0.1,0.1))

## plot CCF's
ccf(proc_rot[1,],proc_rot[2,], lag.max = 12, main="")

```

# Plot Model fit and uncertainty 
Note: When we used the function from the lab materials to calculate model fit and uncertainty, it returned NaN values and so we could not include the upper and lower confidence intervals in the plot. Mark looked at the code during lab on 5/1, but could not figure out the cause of the error. 

```{r}
#### When we use this function we cannot get the confidence intervals, but we can get the means
# Mark says not to worry about it
# Mention this in the write up, describe the error- we were getting NaN's when trying to calculate the high and low confidence intervals.
 get_DFA_fits <- function(MLEobj, dd = NULL, alpha = 0.05) {
  ## empty list for results
  fits <- list()
  ## extra stuff for var() calcs
  Ey <- MARSS:::MARSShatyt(MLEobj)
  ## model params
    ZZ <- coef(MLEobj, type="matrix")$Z
    ## number of obs ts
    nn <- dim(Ey$ytT)[1]
    ## number of time steps
    TT <- dim(Ey$ytT)[2]
    ## get the inverse of the rotation matrix
    H_inv <- varimax(ZZ)$rotmat
    ## check for covars
    if(!is.null(dd)) {
      DD <- coef(MLEobj, type = "matrix")$D
      ## model expectation
      fits$ex <- ZZ %*% H_inv %*% MLEobj$states + DD %*% dd
    } else {
      ## model expectation
      fits$ex <- ZZ %*% H_inv %*% MLEobj$states
    }
    ## Var in model fits
    VtT <- MARSSkfss(MLEobj)$VtT
    VV <- NULL
    for(tt in 1:TT) {
      RZVZ <- coef(MLEobj, type = "matrix")$R - ZZ %*% VtT[,,tt] %*% t(ZZ)
      SS <- Ey$yxtT[,,tt] - Ey$ytT[,tt,drop = FALSE] %*% t(MLEobj$states[,tt,drop = FALSE])
      VV <- cbind(VV, diag(RZVZ + SS %*% t(ZZ) + ZZ %*% t(SS)))
      }
    SE <- sqrt(VV)
    ## upper & lower (1-alpha)% CI
    fits$up <- qnorm(1-alpha/2)*SE + fits$ex
    fits$lo <- qnorm(alpha/2)*SE + fits$ex
    return(fits)
}

## get model fits & CI's
mod_fit <- get_DFA_fits(Temp_pH_5,dd = rbind(Temp, pH), alpha = 0.05)

## set up plotting area
par(mai = c(0.5, 0.7, 0.1, 0.1), omi = c(0, 0, 0, 0))

## plot the fits
for(i in 1:N.ts) {
  #up <- mod_fit$up[i,]
  mn <- mod_fit$ex[i,]
  #lo <- mod_fit$lo[i,]
  plot(w_ts, mn, type = "n", #ylim = c(min(lo), max(up)),
       cex.lab = 1.2,
       xlab = "", ylab = spp[i], xaxt = "n")
  axis(1, 12 * (0:dim(plankton)[2]) + 1, yr_frst + 0:dim(plankton)[2])
  points(w_ts,plankton[i,], pch = 16, col = color[i])
  #lines(w_ts, up, col = "darkgray")
  lines(w_ts, mn, col = "black", lwd = 2)
  #lines(w_ts, lo, col = "darkgray")
}
```


##################
# Methods
##################

The nutrient composition and, subsequently, zooplankton and phytoplankton community changed dramatically following both the rapid increase in human development on its shores and the beginning of sewage treatment by Seattle metro in the early 60's. Its well documented that this change resulted in a decrease in blue-green algae and an increase in daphnia from nearly undetectable numbers to modern value. However, our group was curious to explore what continued trends have been occuring in phytoplankton communities in more recently years. To do so , we chose the most recent 10 year period (1984 - 1995) and selected all zooplankton types other than those with significant gaps, presumably due to low numbers. As a result, we analyzed trends in "Conochilus", "Cyclops", "Daphnia", "Diaptomus", "Epischura","Non daphnid cladocerans", and "Non colonial rotifers".

We tested several combinations of models including those with different number of states (1-7), different numbers and combinations of covariates, and different structures for the observation error matrix. The matrix forms are written throughout the script. We also monitored convergence and log-likelihood diagnostics from MARSS output to verify successful model fitting.

##################
# Results
##################

## Most parsimonious model

In models which compared a seasonal dummy variable comprised of sin and consine functions with a model containing all covariates. The model with all covariates was a better fit in terms of AICc. Keeping this value constant, we found that a model with 6 states had the best support, and that the best model included a "diagonal and unequal" observation variance structure, meaning each species has a different level of observation error and they were not correlated among species. Finally, pH and temperature had the best support as the combination of covariates with the lowest AICc. Once we determined what our best structure for observation error and covariates was, we check the AICc of the same model but with fewer states (5 as opposed to 6) and found that the model fit was the same. Therefore, this model was a more parsimonious fit. 
In models which compared a seasonal dummy variable comprised of sin and consine functions with a model containing all covariates. The model with all covariates was a better fit in terms of AICc. Keep this value constant, we found that a model with 6 states had the best support, and that the best model included a "diagonal and unequal" observation variance structure, meaning each species has a different level of observation error and they were not correlated among species. Finally, pH and temperature had the best support as the combination of covariates with the lowest AICc. Once we determined what our best structure for observation error and covariates was, we check the AICc of the same model but with fewer states (5 as opposed to 6) and found that the model fit was the same. Therefore, this model was a more parsimonious fit. 

## Loadings and interpretation 
The varimax-rotated factor loadings for the selected 5-trend DFA model revealed distinct associations between taxa and latent trends, suggesting different ecological groupings and temporal responses:

- Trend 1 (State 1) was most strongly associated with Non-colonial rotifers and Conochilus, both showing high positive loadings (~0.5), indicating they co-vary with the dominant pattern captured by this trend. Daphnia, Diaptomus, and Non-daphnid cladocerans also showed moderate positive loadings, while Epischura had a near-zero or slightly negative loading, suggesting it did not follow this dominant temporal signal.
- Trend 2 (State 2) was dominated by Cyclops, which had a distinctly high positive loading (>0.6), marking it as the primary taxon associated with this temporal pattern. Daphnia and Non-colonial rotifers also had moderate positive loadings, while Diaptomus and Epischura showed low or slightly negative loadings, indicating some divergence in community response.
- Trend 3 (State 3) showed strong negative loadings for both Epischura and Diaptomus, suggesting a shared response to a latent driver that opposes the pattern observed in other taxa. Most other taxa showed weak to neutral associations with this trend, indicating it may represent a factor more specific to copepod dynamics.
- Trend 4 (State 4) had uniformly weak loadings across all taxa, with most values near zero or slightly negative. This suggests Trend 4 may represent low-variance background variation or unstructured noise not strongly tied to any particular group.
- Trend 5 (State 5) highlighted a strong positive loading for Non-daphnid cladocerans and moderate contributions from Daphnia and Non-colonial rotifers, indicating that this trend may reflect dynamics specific to cladocerans and rotifers. In contrast, Diaptomus and Epischura had slight negative loadings, further distinguishing copepods from other taxa in their temporal behavior.

Overall, the loading structure revealed coherent groupings: rotifers and small-bodied cladocerans frequently tracked together, while copepods, particularly Epischura and Diaptomus, often displayed opposing or unique patterns. 

# Describe the effects of environmental or dummy variables on patterns in the data 
The best-supported covariate structure included both temperature and pH, outperforming seasonal dummy variables and other environmental combinations in terms of AICc. Daphnia (0.64), Epischura (0.64), and Non-daphnid cladocerans (0.58) had a strong positive response to temperature and Non-colonial rotifers (0.61), Cyclops (0.55), and Conochilus (0.42) had a strong positive response to pH.

##################
# Discussion
##################

Our dynamic factor analysis (DFA) identified distinct latent temporal patterns in zooplankton community dynamics, with several trends closely aligned with environmental drivers. The model that included temperature and pH as covariates had the strongest support based on AICc. 
Temperature and pH were key predictors of variation across taxa: temperature had a strong positive effect on Daphnia, Epischura, and non-daphnid cladocerans, indicating that warmer conditions favored their abundance. pH also played a significant role, particularly influencing Cyclops, Conochilus, and non-colonial rotifers, which may be more sensitive to acidity or benefit from conditions associated with elevated pH levels.



##################
# Team contributions
##################

All group members brought their own initial model fitting attempts to the first meeting. The group chose to build upon Ben’s version. As a result, Ben wrote the majority of the code for model fitting and analysis, while Nicole and Lilac contributed to key decisions, including model selection. Nicole wrote the code for plots of model fits and factor loadings. Ben and Lilac contributed to writing the methods, results, and discussion sections of the report. 
