---
title: Team X - Lab 2
subtitle: MARSS models
author: "Ben Makhlouf" 
date: April XX, 2025
output: 
  html_document:
    code-folding: true
    toc: true
    toc_float: true
---

# Data

Describe what data set you will work with and any subsetting you decided on. For example, you may have decided to look only at a section of the ESU that your team was assigned.

Load the data.
```{r}
load(here::here("Lab-2", "Data_Images", "esa-salmon.rda"))
library(tidyr)
```

Wrangle the data.

```{r}
library(dplyr)
# Our group was assigned only Steelhead. 
SteelheadAll<- esa.salmon %>%
  filter(species=="Steelhead")


# Plot all Summer Steelhead faceted by ESA group and log transformed value 
library(ggplot2)

Stl_plot<- SteelheadAll %>%
  mutate(log_value=log(value)) %>%
  ggplot(aes(x=spawningyear, y=log_value, color = run)) +
  geom_point() +
  facet_wrap(~esapopname) +
  theme_bw() +
  labs(title="Summer Steelhead by ESA group", x="Year", y="Log Abundance")


print(Stl_plot)

```

```{r}

# Step 1: Start with your full dataset
Stl <- SteelheadAll

# Step 2: Filter for Summer if you want
# Stl <- Stl %>% filter(run == "Summer") # optional

# Step 3: Create region and run labels BEFORE pivoting
Stl <- Stl %>%
  mutate(
    region = case_when(
      grepl("Lower", esapopname) ~ "Lower",
      grepl("Middle", esapopname) ~ "Middle",
      grepl("Upper", esapopname) ~ "Upper",
      TRUE ~ "Unknown"
    ),
    run_timing = case_when(
      grepl("summer", esapopname, ignore.case = TRUE) ~ "Summer",
      grepl("winter", esapopname, ignore.case = TRUE) ~ "Winter",
      TRUE ~ "Unknown"
    )
  )

# Step 4: Clean up the names (simplify for pivoting)
Stl$esapopname <- gsub("Steelhead", "", Stl$esapopname)
Stl$esapopname <- gsub("- summer", "", Stl$esapopname)
Stl$esapopname <- gsub(" - winter", "", Stl$esapopname)


# Step 5: Store lookup table for region/run AFTER cleaning names
meta_info <- Stl %>%
  select(esapopname, region, run_timing) %>%
  distinct()

# Step 6: Pivot wider to population x year
Stl_wide <- Stl %>%
  select(spawningyear, esapopname, value) %>%
  pivot_wider(names_from = esapopname, values_from = value)

# Step 7: Save years and transpose
Years <- Stl_wide$spawningyear
Stl_mat <- Stl_wide %>%
  select(-spawningyear) %>%
  as.matrix() %>%
  t()

# Step 8: Assign years as column names and reorder by year
colnames(Stl_mat) <- Years
Stl_mat <- Stl_mat[, order(as.numeric(colnames(Stl_mat)))]
#Make sure all NaN are NA 
Stl_mat[is.nan(Stl_mat)] <- NA

# Step 9: Create vectors for regions and run timings
region_vec <- meta_info$region[match(rownames(Stl_mat), meta_info$esapopname)]
run_vec <- meta_info$run_timing[match(rownames(Stl_mat), meta_info$esapopname)]

# Now you can subset like:
# Stl_mat[region_vec == "Lower", ]
# Stl_mat[run_vec == "Summer", ]


```


Models 

```{r, echo = FALSE, results = "hide"}

# ==========================================
# MODEL ASSUMPTIONS IN MATRIX FORM
# ==========================================
# General MARSS model:
# Process equation: x_t = B x_{t-1} + u + w_t  where w_t ~ N(0,Q)
# Observation equation: y_t = Z x_t + a + v_t  where v_t ~ N(0,R)
#
# Model 1: All Populations Same (H1)
# - One hidden state for all populations
# - Z = [1, 1, 1, ..., 1]^T (n_pops × 1 matrix of 1s)
# - U = [u] (1 × 1 scalar)
# - Q = [q] (1 × 1 scalar)
# - B = [1] (1 × 1 scalar)
# - R = diag(r, r, ..., r) (n_pops × n_pops diagonal matrix with equal values)
#
# Model 2: Region-based (H2)
# - One hidden state per region (Lower, Middle, Upper)
# - Z = [1 0 0]   (for each Lower region population)
#       [0 1 0]   (for each Middle region population)
#       [0 0 1]   (for each Upper region population)
# - U = [u_1, u_2, u_3]^T (different trend for each region)
# - Q = [q  0  0]
#       [0  q  0]
#       [0  0  q]
# - B = [b  0  0]
#       [0  b  0]
#       [0  0  b]
# - R = diag(r, r, ..., r) (diagonal matrix with equal values)
#
# Model 3: Run-based (H3)
# - One hidden state per run timing (Summer, Winter, Unknown)
# - Z = [1 0 0]   (for each Summer run population)
#       [0 1 0]   (for each Winter run population)
#       [0 0 1]   (for each Unknown run population)
# - U = [u_1, u_2, u_3]^T (different trend for each run timing)
# - Q = [q  0  0]
#       [0  q  0]
#       [0  0  q]
# - B = [b  0  0]
#       [0  b  0]
#       [0  0  b]
# - R = diag(r, r, ..., r) (diagonal matrix with equal values)
#
# Model 4: Region + Unequal Q
# - Same as Model 2 except:
# - Q = [q_1  0    0  ]
#       [0    q_2  0  ]
#       [0    0    q_3]
#
# Model 5: Run + Unequal Q
# - Same as Model 3 except:
# - Q = [q_1  0    0  ]
#       [0    q_2  0  ]
#       [0    0    q_3]

# ==========================================
# 1. All populations same (H1)
# ==========================================
mod1 <- list(
  B = matrix(1),
  U = matrix("u"),
  Q = matrix("q"),
  Z = matrix(1, n_pops, 1),
  A = "scaling",
  R = "diagonal and equal"
)

# Fit model 1
fit1 <- MARSS(Data, model = mod1, method = "BFGS", control = list(maxit = 1000))

# ==========================================
# 2. Region-based (H2)
# ==========================================
mod2 <- list(
  B = "diagonal and equal",
  U = "unequal",
  Q = "diagonal and equal",
  Z = Z2,
  A = "scaling",
  R = "diagonal and equal"
)

# Fit model 2
fit2 <- MARSS(Data, model = mod2, method = "BFGS", control = list(maxit = 1000))

# ==========================================
# 3. Run-based (H3)
# ==========================================
mod3 <- list(
  B = "diagonal and equal",
  U = "unequal",
  Q = "diagonal and equal",
  Z = Z3,
  A = "scaling",
  R = "diagonal and equal"
)

# Fit model 3
fit3 <- MARSS(Data, model = mod3, method = "BFGS", control = list(maxit = 1000))

# ==========================================
# 4. Region with unequal Q
# ==========================================
mod4 <- list(
  B = "diagonal and equal",
  U = "unequal",
  Q = "diagonal and unequal",
  Z = Z2,
  A = "scaling",
  R = "diagonal and equal"
)

# Fit model 4
fit4 <- MARSS(Data, model = mod4, method = "BFGS", control = list(maxit = 1000))

# ==========================================
# 5. Run with unequal Q
# ==========================================
mod5 <- list(
  B = "diagonal and equal",
  U = "unequal",
  Q = "diagonal and unequal",
  Z = Z3,
  A = "scaling",
  R = "diagonal and equal"
)

# Fit model 5
fit5 <- MARSS(Data, model = mod5, method = "BFGS", control = list(maxit = 1000))

```

```{r}
# ==========================================
# Compare via AIC and AICc
# ==========================================

# Collect all AICc values
model_names <- c("1. All populations same (H1)",
                "2. Region-based (H2)",
                "3. Run-based (H3)",
                "4. Region + Unequal Q",
                "5. Run + Unequal Q")

AICc_values <- c(fit1$AICc, fit2$AICc, fit3$AICc, fit4$AICc, fit5$AICc)
min_AICc <- min(AICc_values)
delta_AICc <- AICc_values - min_AICc

# Create and print the simple comparison table
comparison_table <- data.frame(
  Model = model_names,
  AICc = round(AICc_values, 2),
  delta_AICc = round(delta_AICc, 2)
)

# Sort by AICc
comparison_table <- comparison_table[order(comparison_table$AICc), ]

# Print the simple comparison table
cat("\n============ MODEL COMPARISON TABLE ============\n")
print(comparison_table, row.names = FALSE)
```





esa.salmon

### Get the steelhead data 


### Fill in the missing years


### Look to see if there appears to be trends. 
# could be grouped by Season 
# could be grouped by Region 
##### Upper , lower 
##### Distance Based 


# General Questions

Each group has the same general tasks, but you will adapt them as you work on the data.

1. Create estimates of spawner abundance for all missing years and provide estimates of the decline from the historical abundance.

2. Evaluate support for the major population groups. Are the populations in the groups more correlated than outside the groups?

3. Evaluate the evidence of cycling in the data. *We will talk about how to do this on the Tuesday after lab.*


# Methods

Address the following in your methods

* Describe your assumptions about the x and how the data time series are related to x.

   - How are the x and y (data) related? 1 x for 1 y or will you assume 1 x for all y or 1 x for each major population group? How will you choose? 
   - What will you assume about the U for the x's?
   - What will you assume about the Q matrix?
   
* Write out your assumptions as different models **in matrix form**, fit each and then compare these with AIC or AICc.

* Do your estimates differ depending on the assumptions you make about the structure of the data, i.e. you assumptions about the x's, Q, and U.

### Tips

**Simplify**

If your ESU has many populations, start with a smaller set of 4-7 populations.

**Assumptions**

You can assume that `R="diagonal and equal"` and `A="scaling"`. Assume that "historical" means the earliest years available for your group.

**States**

Your abundance estimate is the "x" or "state" estimates. You can get this from
```
fit$states
```
or 
```
tsSmooth(fit)
```
where `fit` is from `fit <- MARSS()`

**plotting**

Estimate of the mean of the spawner counts based on your x model.
```
autoplot(fit, plot.type="fitted.ytT")
```

**diagnostics**
```
autoplot(fit, plot.type="residuals")
```

# Results

# Discussion

# Description of each team member's contributions

Example: "All team members helped decide on the goal and ran the analyses for the individual regions. Team members 2 & 3 wrote most of the code for the analysis of the regions. Team member 4 researched approaches for measuring accuracy of forecasts in [Hyndman & Athanasopoulos[OTexts.com/fpp2] and team member 2 added code for that to the methods. Team member 4 also researched tests for stationarity and worked with team member 2 to code that up. Team member 1 worked on the plotting section of the report using and adapting code that team member 3 wrote. All team members helped edit the report and wrote the discussion together."

