---
title: Team X - Lab 2
subtitle: MARSS models
author: "Ben Makhlouf" 
date: April XX, 2025
output: 
  html_document:
    code-folding: true
    toc: true
    toc_float: true
---

# Data

Describe what data set you will work with and any subsetting you decided on. For example, you may have decided to look only at a section of the ESU that your team was assigned.

Load the data.
```{r}
load(here::here("Lab-2", "Data_Images", "esa-salmon.rda"))
library(tidyr)
```

Wrangle the data.

```{r}
library(dplyr)
# Our group was assigned only Steelhead. 
SteelheadAll<- esa.salmon %>%
  filter(species=="Steelhead")


# Plot all Summer Steelhead faceted by ESA group and log transformed value 
library(ggplot2)

Stl_plot<- SteelheadAll %>%
  mutate(log_value=log(value)) %>%
  ggplot(aes(x=spawningyear, y=log_value, color = run)) +
  geom_point() +
  facet_wrap(~esapopname) +
  theme_bw() +
  labs(title="Summer Steelhead by ESA group", x="Year", y="Log Abundance")


print(Stl_plot)

```

```{r}

# Step 1: Start with your full dataset
Stl <- SteelheadAll

# Step 2: Filter for Summer if you want
# Stl <- Stl %>% filter(run == "Summer") # optional

# Step 3: Create region and run labels BEFORE pivoting
Stl <- Stl %>%
  mutate(
    region = case_when(
      grepl("Lower", esapopname) ~ "Lower",
      grepl("Middle", esapopname) ~ "Middle",
      grepl("Upper", esapopname) ~ "Upper",
      TRUE ~ "Unknown"
    ),
    run_timing = case_when(
      grepl("summer", esapopname, ignore.case = TRUE) ~ "Summer",
      grepl("winter", esapopname, ignore.case = TRUE) ~ "Winter",
      TRUE ~ "Unknown"
    )
  )

# Step 4: Clean up the names (simplify for pivoting)
Stl$esapopname <- gsub("Steelhead", "", Stl$esapopname)
Stl$esapopname <- gsub("- summer", "", Stl$esapopname)
Stl$esapopname <- gsub(" - winter", "", Stl$esapopname)


# Step 5: Store lookup table for region/run AFTER cleaning names
meta_info <- Stl %>%
  select(esapopname, region, run_timing) %>%
  distinct()

# Step 6: Pivot wider to population x year
Stl_wide <- Stl %>%
  select(spawningyear, esapopname, value) %>%
  pivot_wider(names_from = esapopname, values_from = value)

# Step 7: Save years and transpose
Years <- Stl_wide$spawningyear
Stl_mat <- Stl_wide %>%
  select(-spawningyear) %>%
  as.matrix() %>%
  t()

# Step 8: Assign years as column names and reorder by year
colnames(Stl_mat) <- Years
Stl_mat <- Stl_mat[, order(as.numeric(colnames(Stl_mat)))]
#Make sure all NaN are NA 
Stl_mat[is.nan(Stl_mat)] <- NA

# Step 9: Create vectors for regions and run timings
region_vec <- meta_info$region[match(rownames(Stl_mat), meta_info$esapopname)]
run_vec <- meta_info$run_timing[match(rownames(Stl_mat), meta_info$esapopname)]

# Now you can subset like:
# Stl_mat[region_vec == "Lower", ]
# Stl_mat[run_vec == "Summer", ]


```







Models 

# Hypotheses to test 

#################
### Hypothesis 1 


2. Evaluate support for the major population groups. Are the populations in the groups more correlated than outside the groups?

```{r}
# Hypotheses to test 
#################
### Hypothesis 1 
# Evaluate support for the major population groups. Are the populations in the groups more correlated than outside the groups?

# Load required packages
library(MARSS)

# First let's examine our data structure after wrangling
dim(Stl)
pop_names <- rownames(Stl)
n_pops <- length(pop_names)
print(pop_names)

# Let's log-transform the data since we're working with population abundances
# Log transformation stabilizes variance and makes data more normally distributed
Stl_log <- log(Stl)
# Replace -Inf values (from log(0)) with NA
Stl_log[is.infinite(Stl_log)] <- NA

# Identify the major population groups based on the ESA designations
# Extract group information from the population names
# We'll create a function to identify the group for each population
get_group <- function(pop_name) {
  if(grepl("Middle Columbia", pop_name)) return("Middle Columbia")
  if(grepl("Upper Columbia", pop_name)) return("Upper Columbia")
  if(grepl("Snake River", pop_name)) return("Snake River")
  if(grepl("Lower Columbia", pop_name)) return("Lower Columbia")
  return("Other")
}

# Apply the function to get group for each population
pop_groups <- sapply(pop_names, get_group)
unique_groups <- unique(pop_groups)
n_groups <- length(unique_groups)

# Create a table showing the number of populations in each group
table(pop_groups)

# Create a matrix to define the population-to-group mapping for MARSS
Z_groups <- matrix(0, nrow = n_pops, ncol = n_groups)
colnames(Z_groups) <- unique_groups
rownames(Z_groups) <- pop_names

# Fill in the group membership matrix
for (i in 1:n_pops) {
  group_idx <- which(unique_groups == pop_groups[i])
  Z_groups[i, group_idx] <- 1
}

# Print the group matrix to verify
print(Z_groups)

# We'll now define and fit several competing models to test our hypothesis
# For each model, we'll provide detailed explanation of parameter choices

#------------------------------------------------------------------------------
# MODEL 1: Fully independent populations (null model)
#------------------------------------------------------------------------------
# In this model, each population has its own trend parameter
mod1 <- list(
  Z = "identity",           # Each population has its own trend
  R = "diagonal and unequal", # Each population has its own observation variance
  Q = "diagonal and unequal", # Each population has its own process variance
  U = "unequal",            # Each population has its own mean growth rate
  x0 = "unequal"            # Each population has its own initial state
)

# Parameter choice explanations:
# Z = "identity": One-to-one mapping between states and observations
#   - This assumes each population follows its own independent trajectory
#   - No shared dynamics between populations
# 
# R = "diagonal and unequal":
#   - Diagonal: Observation errors are independent across populations
#   - Unequal: Each population has its own observation error variance
#   - This is realistic as monitoring methods and precision may differ by population
# 
# Q = "diagonal and unequal":
#   - Diagonal: Process errors (environmental stochasticity) are independent
#   - Unequal: Each population experiences different environmental variation
#   - This allows for different magnitudes of year-to-year fluctuations
# 
# U = "unequal":
#   - Each population has its own growth trend
#   - This allows populations to have different long-term trajectories
# 
# x0 = "unequal":
#   - Each population starts from a different initial state
#   - Reflects different initial abundances and historical conditions

# Fit model 1 (this may take some time)
fit1 <- MARSS(Stl_log, model = mod1, silent = TRUE, control = list(maxit = 1000))

#------------------------------------------------------------------------------
# MODEL 2: Group-based trends (our hypothesis model)
#------------------------------------------------------------------------------
# In this model, populations within the same ESA group share a trend
mod2 <- list(
  Z = Z_groups,           # Populations map to their respective group trends
  R = "diagonal and unequal", # Each population has its own observation variance
  Q = "diagonal and equal",   # Each group has the same process variance
  U = "unequal",            # Each group has its own mean growth rate
  x0 = "unequal"            # Each group has its own initial state
)

# Parameter choice explanations:
# Z = Z_groups:
#   - Custom matrix that maps populations to their ESA groups
#   - Populations in the same group share the same underlying state (trend)
#   - This directly tests our hypothesis that populations within groups are more correlated
# 
# R = "diagonal and unequal":
#   - Observation errors remain independent between populations
#   - Allows for different monitoring precision for each population
# 
# Q = "diagonal and equal":
#   - Process errors are independent between groups
#   - Equal: All groups experience the same magnitude of environmental stochasticity
#   - This simplification reduces parameters and assumes similar environmental drivers
# 
# U = "unequal":
#   - Each group has its own mean growth rate
#   - Different ESA groups may have different overall trends
# 
# x0 = "unequal":
#   - Each group starts from a different initial state
#   - Different groups may have different historical abundances

# Fit model 2
fit2 <- MARSS(Stl_log, model = mod2, silent = TRUE, control = list(maxit = 1000))

#------------------------------------------------------------------------------
# MODEL 3: Shared trend with group-specific loadings
#------------------------------------------------------------------------------
# In this model, all populations follow one shared trend but with different loadings
mod3 <- list(
  Z = "unconstrained",     # Each population loads differently on the shared trend
  R = "diagonal and unequal", # Each population has its own observation variance
  Q = "identity",          # Single process variance for the shared trend
  U = "equal",             # Single mean growth rate
  x0 = "equal"             # Single initial state
)

# Parameter choice explanations:
# Z = "unconstrained":
#   - Allows each population to load differently on a single shared trend
#   - Some populations may be more responsive to the shared drivers than others
#   - The sign can also differ (some may increase while others decrease)
# 
# R = "diagonal and unequal":
#   - Consistent with previous models
# 
# Q = "identity":
#   - Single process variance as we have only one underlying trend
# 
# U = "equal":
#   - Single mean growth rate for the shared trend
# 
# x0 = "equal":
#   - Single initial state for the shared trend

# Fit model 3
fit3 <- MARSS(Stl_log, model = mod3, silent = TRUE, control = list(maxit = 1000, safe = TRUE))

#------------------------------------------------------------------------------
# MODEL 4: Correlated trends within groups
#------------------------------------------------------------------------------
# In this model, populations have individual trends but are correlated within groups
# First, create a block-diagonal Q matrix structure
Q_blocks <- list()
for (g in unique_groups) {
  # Find which populations are in this group
  group_pops <- which(pop_groups == g)
  n_in_group <- length(group_pops)
  
  # Create a matrix of 1s (will be converted to "unconstrained")
  if (n_in_group > 1) {
    Q_blocks[[g]] <- matrix("unconstrained", n_in_group, n_in_group)
    diag(Q_blocks[[g]]) <- paste0("q", g) # Diagonal elements share the same parameter
  } else if (n_in_group == 1) {
    Q_blocks[[g]] <- matrix(paste0("q", g), 1, 1)
  }
}

# Construct the block diagonal Q matrix
Q_matrix <- MARSS:::build.block.diagonal(Q_blocks)

mod4 <- list(
  Z = "identity",           # Each population has its own trend
  R = "diagonal and unequal", # Each population has its own observation variance
  Q = Q_matrix,             # Block diagonal - correlated within groups
  U = "unequal",            # Each population has its own mean growth rate
  x0 = "unequal"            # Each population has its own initial state
)

# Parameter choice explanations:
# Z = "identity":
#   - Each population still has its own state
#   - Unlike model 2, populations don't share states, but they are correlated
# 
# R = "diagonal and unequal":
#   - Consistent with previous models
# 
# Q = Q_matrix:
#   - Block diagonal structure where populations in the same group have correlated process errors
#   - This allows for within-group correlation of environmental responses
#   - Between-group processes remain independent
#   - This directly tests if populations within the same group respond similarly to environmental variation
# 
# U = "unequal":
#   - Each population maintains its own growth trend
# 
# x0 = "unequal":
#   - Each population has its own initial state

# Fit model 4 (this is more complex and might take longer)
fit4 <- MARSS(Stl_log, model = mod4, silent = TRUE, control = list(maxit = 1000, safe = TRUE))

#------------------------------------------------------------------------------
# Compare all models
#------------------------------------------------------------------------------
# Use AIC to compare model fit
model_table <- data.frame(
  Model = c("Independent", "Shared Group Trends", "Single Trend", "Correlated Group Trends"),
  AICc = c(fit1$AICc, fit2$AICc, fit3$AICc, fit4$AICc),
  num_params = c(fit1$num.params, fit2$num.params, fit3$num.params, fit4$num.params),
  logLik = c(fit1$logLik, fit2$logLik, fit3$logLik, fit4$logLik)
)

# Calculate delta AICc and weights
model_table$delta_AICc <- model_table$AICc - min(model_table$AICc)
model_table$AICc_weight <- exp(-0.5 * model_table$delta_AICc) / sum(exp(-0.5 * model_table$delta_AICc))

# Sort by AICc
model_table <- model_table[order(model_table$AICc), ]

# Print the table
print(model_table)

# Plot the results for the best model
best_model_idx <- which.min(c(fit1$AICc, fit2$AICc, fit3$AICc, fit4$AICc))
best_fits <- list(fit1, fit2, fit3, fit4)[[best_model_idx]]

# Plot the fitted values vs observed
# We'll need to autoplot for this
autoplot(best_fits, plot.type = "fitted.ytT")

# If the group-based model (Model 2) is best, we can examine group-specific trends
if (best_model_idx == 2) {
  # Extract the estimated states (trends) for each group
  states <- best_fits$states
  state_CIs <- best_fits$states.se * 1.96 # For 95% CI
  
  # Create a data frame for plotting
  states_df <- data.frame(
    Year = as.numeric(colnames(Stl_log)),
    t(states)
  )
  colnames(states_df)[-1] <- unique_groups
  
  # Plot the group trends
  states_long <- pivot_longer(states_df, -Year, names_to = "Group", values_to = "Trend")
  
  ggplot(states_long, aes(x = Year, y = Trend, color = Group)) +
    geom_line() +
    theme_bw() +
    labs(title = "Estimated population trends by ESA group",
         x = "Year",
         y = "Log Abundance")
}

# Examine the degree of correlation within vs between groups if Model 4 is best
if (best_model_idx == 4) {
  # Extract the Q matrix which contains the covariance structure
  Q_est <- coef(best_fits, type = "matrix")$Q
  
  # Convert to correlation matrix
  Q_sds <- sqrt(diag(Q_est))
  Q_cor <- Q_est / (Q_sds %*% t(Q_sds))
  
  # Create a heatmap of the correlation matrix
  Q_cor_df <- as.data.frame(Q_cor)
  rownames(Q_cor_df) <- pop_names
  
  # Add group information for sorting
  Q_cor_df$Group <- pop_groups
  Q_cor_df <- Q_cor_df[order(Q_cor_df$Group), ]
  Q_cor_df$Group <- NULL
  
  # Reorder the columns the same way
  Q_cor_df <- Q_cor_df[, rownames(Q_cor_df)]
  
  # Plot the correlation heatmap
  library(reshape2)
  Q_cor_melt <- melt(as.matrix(Q_cor_df))
  colnames(Q_cor_melt) <- c("Population1", "Population2", "Correlation")
  
  ggplot(Q_cor_melt, aes(x = Population1, y = Population2, fill = Correlation)) +
    geom_tile() +
    scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    labs(title = "Correlation structure between populations",
         x = "", y = "")
}

```







esa.salmon

### Get the steelhead data 


### Fill in the missing years


### Look to see if there appears to be trends. 
# could be grouped by Season 
# could be grouped by Region 
##### Upper , lower 
##### Distance Based 


# General Questions

Each group has the same general tasks, but you will adapt them as you work on the data.

1. Create estimates of spawner abundance for all missing years and provide estimates of the decline from the historical abundance.

2. Evaluate support for the major population groups. Are the populations in the groups more correlated than outside the groups?

3. Evaluate the evidence of cycling in the data. *We will talk about how to do this on the Tuesday after lab.*


# Methods

Address the following in your methods

* Describe your assumptions about the x and how the data time series are related to x.

   - How are the x and y (data) related? 1 x for 1 y or will you assume 1 x for all y or 1 x for each major population group? How will you choose? 
   - What will you assume about the U for the x's?
   - What will you assume about the Q matrix?
   
* Write out your assumptions as different models **in matrix form**, fit each and then compare these with AIC or AICc.

* Do your estimates differ depending on the assumptions you make about the structure of the data, i.e. you assumptions about the x's, Q, and U.

### Tips

**Simplify**

If your ESU has many populations, start with a smaller set of 4-7 populations.

**Assumptions**

You can assume that `R="diagonal and equal"` and `A="scaling"`. Assume that "historical" means the earliest years available for your group.

**States**

Your abundance estimate is the "x" or "state" estimates. You can get this from
```
fit$states
```
or 
```
tsSmooth(fit)
```
where `fit` is from `fit <- MARSS()`

**plotting**

Estimate of the mean of the spawner counts based on your x model.
```
autoplot(fit, plot.type="fitted.ytT")
```

**diagnostics**
```
autoplot(fit, plot.type="residuals")
```

# Results

# Discussion

# Description of each team member's contributions

Example: "All team members helped decide on the goal and ran the analyses for the individual regions. Team members 2 & 3 wrote most of the code for the analysis of the regions. Team member 4 researched approaches for measuring accuracy of forecasts in [Hyndman & Athanasopoulos[OTexts.com/fpp2] and team member 2 added code for that to the methods. Team member 4 also researched tests for stationarity and worked with team member 2 to code that up. Team member 1 worked on the plotting section of the report using and adapting code that team member 3 wrote. All team members helped edit the report and wrote the discussion together."

