---
title: Team X - Lab 2
subtitle: MARSS models
author: "Ben Makhlouf" 
date: April XX, 2025
output: 
  html_document:
    code-folding: true
    toc: true
    toc_float: true
---

# Data

Describe what data set you will work with and any subsetting you decided on. For example, you may have decided to look only at a section of the ESU that your team was assigned.

Load the data.
```{r}
load(here::here("Lab-2", "Data_Images", "esa-salmon.rda"))
library(tidyr)
library(MARSS)
```

Wrangle the data.

```{r}
library(dplyr)
# Our group was assigned only Steelhead. 
SteelheadAll<- esa.salmon %>%
  filter(species=="Steelhead")


# Plot all Summer Steelhead faceted by ESA group and log transformed value 
library(ggplot2)

Stl_plot<- SteelheadAll %>%
  mutate(log_value=log(value)) %>%
  ggplot(aes(x=spawningyear, y=log_value, color = run)) +
  geom_point() +
  facet_wrap(~esapopname) +
  theme_bw() +
  labs(title="Summer Steelhead by ESA group", x="Year", y="Log Abundance")


print(Stl_plot)

```

```{r}
# Step 1: Start with your full dataset
Stl <- SteelheadAll

# Step 2: Create region and run labels BEFORE pivoting
Stl <- Stl %>%
  mutate(
    region = case_when(
      grepl("Lower", esapopname) ~ "Lower",
      grepl("Middle", esapopname) ~ "Middle",
      grepl("Upper", esapopname) ~ "Upper",
      TRUE ~ "Unknown"
    ),
    run_timing = case_when(
      grepl("summer", esapopname, ignore.case = TRUE) ~ "Summer",
      grepl("winter", esapopname, ignore.case = TRUE) ~ "Winter",
      TRUE ~ "Unknown"
    )
  )

# Filter out Unknown categories 
Stl <- Stl %>%
  filter(region != "Unknown" & run_timing != "Unknown")

# Clean up the names (simplify for pivoting)
Stl$esapopname <- gsub("Steelhead", "", Stl$esapopname)
Stl$esapopname <- gsub("- summer", "", Stl$esapopname)
Stl$esapopname <- gsub(" - winter", "", Stl$esapopname)


# Store lookup table for region/run AFTER cleaning names
meta_info <- Stl %>%
  select(esapopname, region, run_timing) %>%
  distinct()

# Pivot wider to population x year
Stl_wide <- Stl %>%
  select(spawningyear, esapopname, value) %>%
  pivot_wider(names_from = esapopname, values_from = value)

# Save years and transpose
Years <- Stl_wide$spawningyear
Stl_mat <- Stl_wide %>%
  select(-spawningyear) %>%
  as.matrix() %>%
  t()

# Assign years as column names and reorder by year
colnames(Stl_mat) <- Years
Stl_mat <- Stl_mat[, order(as.numeric(colnames(Stl_mat)))]
#Make sure all NaN are NA 
Stl_mat[is.nan(Stl_mat)] <- NA

Data<- Stl_mat

#Create vectors for regions and run timings
region_vec <- meta_info$region[match(rownames(Stl_mat), meta_info$esapopname)]
run_vec <- meta_info$run_timing[match(rownames(Stl_mat), meta_info$esapopname)]

n_pops <- nrow(Stl_mat)


# For Region-based model
regions <- unique(region_vec)
n_regions <- length(regions)
Z2 <- matrix(0, n_pops, n_regions)
for (i in 1:n_pops) {
  region_index <- which(regions == region_vec[i])
  Z2[i, region_index] <- 1
}

# For Run-based model
runs <- unique(run_vec)
n_runs <- length(runs)
Z3 <- matrix(0, n_pops, n_runs)
for (i in 1:n_pops) {
  run_index <- which(runs == run_vec[i])
  Z3[i, run_index] <- 1
}


```


Models 

```{r, echo = FALSE, results = "hide"}
# ==========================================
# MODEL ASSUMPTIONS IN MATRIX FORM
# ==========================================
# General MARSS model:
# Process equation: x_t = B x_{t-1} + u + w_t  where w_t ~ N(0,Q)
# Observation equation: y_t = Z x_t + a + v_t  where v_t ~ N(0,R)
#
# Model 1: All Populations Same (H1)
# - One hidden state for all populations
# - Z = [1, 1, 1, ..., 1]^T (n_pops × 1 matrix of 1s)
# - U = [u] (1 × 1 scalar)
# - Q = [q] (1 × 1 scalar)
# - B = [1] (1 × 1 scalar)
# - R = diag(r, r, ..., r) (n_pops × n_pops diagonal matrix with equal values)
#
# Model 2: Region-based (H2)
# - One hidden state per region (Lower, Middle, Upper)
# - Z = [1 0 0]   (for each Lower region population)
#       [0 1 0]   (for each Middle region population)
#       [0 0 1]   (for each Upper region population)
# - U = [u_1, u_2, u_3]^T (different trend for each region)
# - Q = [q  0  0]
#       [0  q  0]
#       [0  0  q]
# - B = [b  0  0]
#       [0  b  0]
#       [0  0  b]
# - R = diag(r, r, ..., r) (diagonal matrix with equal values)
#
# Model 3: Run-based (H3)
# - One hidden state per run timing (Summer, Winter, Unknown)
# - Z = [1 0 0]   (for each Summer run population)
#       [0 1 0]   (for each Winter run population)
#       [0 0 1]   (for each Unknown run population)
# - U = [u_1, u_2, u_3]^T (different trend for each run timing)
# - Q = [q  0  0]
#       [0  q  0]
#       [0  0  q]
# - B = [b  0  0]
#       [0  b  0]
#       [0  0  b]
# - R = diag(r, r, ..., r) (diagonal matrix with equal values)
#
# Model 4: Region + Unequal Q
# - Same as Model 2 except:
# - Q = [q_1  0    0  ]
#       [0    q_2  0  ]
#       [0    0    q_3]
#
# Model 5: Run + Unequal Q
# - Same as Model 3 except:
# - Q = [q_1  0    0  ]
#       [0    q_2  0  ]
#       [0    0    q_3]

# ==========================================
# 1. All populations same (H1)
# ==========================================
mod1 <- list(
  B = matrix(1),
  U = matrix("u"),
  Q = matrix("q"),
  Z = matrix(1, n_pops, 1),
  A = "scaling",
  R = "diagonal and equal"
)

# Fit model 1
fit1 <- MARSS(Data, model = mod1, method = "BFGS", control = list(maxit = 1000))

# ==========================================
# 2. Region-based (H2)
# ==========================================
mod2 <- list(
  B = "diagonal and equal",
  U = "unequal",
  Q = "diagonal and equal",
  Z = Z2,
  A = "scaling",
  R = "diagonal and equal"
)

# Fit model 2
fit2 <- MARSS(Data, model = mod2, method = "BFGS", control = list(maxit = 1000))

# ==========================================
# 3. Run-based (H3)
# ==========================================
mod3 <- list(
  B = "diagonal and equal",
  U = "unequal",
  Q = "diagonal and equal",
  Z = Z3,
  A = "scaling",
  R = "diagonal and equal"
)

# Fit model 3
fit3 <- MARSS(Data, model = mod3, method = "BFGS", control = list(maxit = 1000))

# ==========================================
# 4. Region with unequal Q
# ==========================================
mod4 <- list(
  B = "diagonal and equal",
  U = "unequal",
  Q = "diagonal and unequal",
  Z = Z2,
  A = "scaling",
  R = "diagonal and equal"
)

# Fit model 4
fit4 <- MARSS(Data, model = mod4, method = "BFGS", control = list(maxit = 1000))

# ==========================================
# 5. Run with unequal Q
# ==========================================
mod5 <- list(
  B = "diagonal and equal",
  U = "unequal",
  Q = "diagonal and unequal",
  Z = Z3,
  A = "scaling",
  R = "diagonal and equal"
)

# Fit model 5
fit5 <- MARSS(Data, model = mod5, method = "BFGS", control = list(maxit = 1000))

```



```{r}
# ==========================================
# Compare via AIC and AICc
# ==========================================

# Collect all AICc values
model_names <- c("1. All populations same (H1)",
                "2. Region-based (H2)",
                "3. Run-based (H3)",
                "4. Region + Unequal Q",
                "5. Run + Unequal Q")

AICc_values <- c(fit1$AICc, fit2$AICc, fit3$AICc, fit4$AICc, fit5$AICc)
min_AICc <- min(AICc_values)
delta_AICc <- AICc_values - min_AICc

# Create and print the simple comparison table
comparison_table <- data.frame(
  Model = model_names,
  AICc = round(AICc_values, 2),
  delta_AICc = round(delta_AICc, 2)
)

# Sort by AICc
comparison_table <- comparison_table[order(comparison_table$AICc), ]

# Print the simple comparison table
cat("\n============ MODEL COMPARISON TABLE ============\n")
print(comparison_table, row.names = FALSE)
```



# DISCUSSION 

I employed several MARSS models to analyze population dynamics of Steelhead salmon across different regions and run timings. Five candidate models were developed to test competing hypotheses about the underlying structure of population dynamics. Our results provide strong support for run timing as the primary factor structuring Steelhead populations, with the run-based model that incorporated heterogeneous process variance (Model 5) outperforming all other models. The substantial difference in AICc values (ΔAICc > 21) between this model and the next best alternative provides compelling evidence that Summer and Winter run populations experience different environmental influences and respond differently to ecosystem changes.
The models differed in their assumptions about population structure (Z matrix) and process error variance (Q matrix). Model 1 represented the simplest case, treating all populations as reflecting a single underlying state with uniform process variance. Models 2 and 3 introduced structure based on either geographic regions or run timing, respectively, while maintaining equal process variance across groups. Models 4 and 5 expanded upon the previous models by allowing for heterogeneous process variance between groups. The superior performance of Model 5 indicates that not only do Summer and Winter steelhead populations follow distinct trends, but they also experience different levels of environmental stochasticity. 





