---
title: Team X - Lab 2
subtitle: MARSS models
author: "Ben Makhlouf" 
date: April XX, 2025
output: 
  html_document:
    code-folding: true
    toc: true
    toc_float: true
---

# Data

Describe what data set you will work with and any subsetting you decided on. For example, you may have decided to look only at a section of the ESU that your team was assigned.

Load the data.
```{r}
load(here::here("Lab-2", "Data_Images", "esa-salmon.rda"))
library(tidyr)
```

Wrangle the data.

```{r}
library(dplyr)
# Our group was assigned only Steelhead. 
SteelheadAll<- esa.salmon %>%
  filter(species=="Steelhead")


# Plot all Summer Steelhead faceted by ESA group and log transformed value 
library(ggplot2)

Stl_plot<- SteelheadAll %>%
  mutate(log_value=log(value)) %>%
  ggplot(aes(x=spawningyear, y=log_value, color = run)) +
  geom_point() +
  facet_wrap(~esapopname) +
  theme_bw() +
  labs(title="Summer Steelhead by ESA group", x="Year", y="Log Abundance")


print(Stl_plot)

```

```{r}

# Step 1: Start with your full dataset
Stl <- SteelheadAll

# Step 2: Filter for Summer if you want
# Stl <- Stl %>% filter(run == "Summer") # optional

# Step 3: Create region and run labels BEFORE pivoting
Stl <- Stl %>%
  mutate(
    region = case_when(
      grepl("Lower", esapopname) ~ "Lower",
      grepl("Middle", esapopname) ~ "Middle",
      grepl("Upper", esapopname) ~ "Upper",
      TRUE ~ "Unknown"
    ),
    run_timing = case_when(
      grepl("summer", esapopname, ignore.case = TRUE) ~ "Summer",
      grepl("winter", esapopname, ignore.case = TRUE) ~ "Winter",
      TRUE ~ "Unknown"
    )
  )

# Step 4: Clean up the names (simplify for pivoting)
Stl$esapopname <- gsub("Steelhead", "", Stl$esapopname)
Stl$esapopname <- gsub("- summer", "", Stl$esapopname)
Stl$esapopname <- gsub(" - winter", "", Stl$esapopname)


# Step 5: Store lookup table for region/run AFTER cleaning names
meta_info <- Stl %>%
  select(esapopname, region, run_timing) %>%
  distinct()

# Step 6: Pivot wider to population x year
Stl_wide <- Stl %>%
  select(spawningyear, esapopname, value) %>%
  pivot_wider(names_from = esapopname, values_from = value)

# Step 7: Save years and transpose
Years <- Stl_wide$spawningyear
Stl_mat <- Stl_wide %>%
  select(-spawningyear) %>%
  as.matrix() %>%
  t()

# Step 8: Assign years as column names and reorder by year
colnames(Stl_mat) <- Years
Stl_mat <- Stl_mat[, order(as.numeric(colnames(Stl_mat)))]
#Make sure all NaN are NA 
Stl_mat[is.nan(Stl_mat)] <- NA

# Step 9: Create vectors for regions and run timings
region_vec <- meta_info$region[match(rownames(Stl_mat), meta_info$esapopname)]
run_vec <- meta_info$run_timing[match(rownames(Stl_mat), meta_info$esapopname)]

# Now you can subset like:
# Stl_mat[region_vec == "Lower", ]
# Stl_mat[run_vec == "Summer", ]


```







Models 

2. Evaluate support for the major population groups. Are the populations in the groups more correlated than outside the groups?

```{r}
# Hypotheses to test 
#################
### Hypotheses
# H1: All populations act the same 
# H2: Each "Region" Lower, Middle, "Upper" acts differently
# H3: Each "Run" Summer, Winter acts differently

# Load required packages
library(MARSS)

# First let's log-transform the data for analysis
Stl_log <- log(Stl_mat)
# Replace -Inf values (from log(0)) with NA
Stl_log[is.infinite(Stl_log)] <- NA

# Examine our data structure
dim(Stl_log)
n_pops <- nrow(Stl_log)
n_years <- ncol(Stl_log)

# Print the region and run timing vectors to verify
print(table(region_vec))
print(table(run_vec))

#------------------------------------------------------------------------------
# MODEL 1: Single trend (H1: All populations act the same)
#------------------------------------------------------------------------------
# In this model, all populations follow one shared trend with different loadings
mod1 <- list(
  Z = "unconstrained",     # Each population loads differently on a single shared trend
  R = "diagonal and unequal", # Each population has its own observation variance
  Q = "identity",          # Single process variance for the shared trend
  U = "identity",          # Single mean growth rate
  x0 = "identity"          # Single initial state
)

# Parameter choice explanations:
# Z = "unconstrained":
#   - There is only one underlying trend that all populations follow
#   - Each population can have a different relationship (loading) with this trend
#   - "unconstrained" means loadings can be any value (+ or -) and different for each population
# 
# R = "diagonal and unequal":
#   - Diagonal: Observation errors are independent across populations
#   - Unequal: Each population has its own observation error variance
#   - This accounts for different monitoring methods/accuracy between populations
# 
# Q = "identity":
#   - Single process variance as we have only one underlying trend
# 
# U = "identity":
#   - Single mean growth rate for the shared trend
# 
# x0 = "identity":
#   - Single initial state for the shared trend

# Fit model 1
fit1 <- MARSS(Stl_log, model = mod1, silent = TRUE, control = list(maxit = 1000))

#------------------------------------------------------------------------------
# MODEL 2: Region-based trends (H2: Each Region acts differently)
#------------------------------------------------------------------------------
# In this model, populations within the same region share a trend

# Create region-based groups mapping matrix
unique_regions <- unique(region_vec)
n_regions <- length(unique_regions)

# Create a region membership matrix for MARSS
Z_regions <- matrix(0, nrow = n_pops, ncol = n_regions)
colnames(Z_regions) <- unique_regions
rownames(Z_regions) <- rownames(Stl_log)

# Fill in the region membership matrix
for (i in 1:n_pops) {
  region_idx <- which(unique_regions == region_vec[i])
  Z_regions[i, region_idx] <- 1
}

mod2 <- list(
  Z = Z_regions,           # Populations map to their respective region trends
  R = "diagonal and unequal", # Each population has its own observation variance
  Q = "diagonal and equal",   # Each region has the same process variance
  U = "unequal",            # Each region has its own mean growth rate
  x0 = "unequal"            # Each region has its own initial state
)

# Parameter choice explanations:
# Z = Z_regions:
#   - Custom matrix that maps populations to their geographical regions
#   - Populations in the same region share the same underlying state (trend)
# 
# R = "diagonal and unequal":
#   - Each population has its own observation error variance
# 
# Q = "diagonal and equal":
#   - Process errors are independent between regions
#   - Equal: All regions experience the same magnitude of environmental stochasticity
# 
# U = "unequal":
#   - Each region has its own mean growth rate
# 
# x0 = "unequal":
#   - Each region starts from a different initial state

# Fit model 2
fit2 <- MARSS(Stl_log, model = mod2, silent = TRUE, control = list(maxit = 1000))

#------------------------------------------------------------------------------
# MODEL 3: Run-timing based trends (H3: Each Run acts differently)
#------------------------------------------------------------------------------
# In this model, populations with the same run timing share a trend

# Create run-timing based groups mapping matrix
unique_runs <- unique(run_vec)
n_runs <- length(unique_runs)

# Create a run timing membership matrix for MARSS
Z_runs <- matrix(0, nrow = n_pops, ncol = n_runs)
colnames(Z_runs) <- unique_runs
rownames(Z_runs) <- rownames(Stl_log)

# Fill in the run timing membership matrix
for (i in 1:n_pops) {
  run_idx <- which(unique_runs == run_vec[i])
  Z_runs[i, run_idx] <- 1
}

mod3 <- list(
  Z = Z_runs,              # Populations map to their respective run timing trends
  R = "diagonal and unequal", # Each population has its own observation variance
  Q = "diagonal and equal",   # Each run timing has the same process variance
  U = "unequal",            # Each run timing has its own mean growth rate
  x0 = "unequal"            # Each run timing has its own initial state
)

# Parameter choice explanations:
# Z = Z_runs:
#   - Custom matrix that maps populations to their run timing groups (Summer, Winter)
#   - Populations with the same run timing share the same underlying state (trend)
# 
# R = "diagonal and unequal":
#   - Each population has its own observation error variance
# 
# Q = "diagonal and equal":
#   - Process errors are independent between run timing groups
#   - All run timing groups have the same environmental stochasticity magnitude
# 
# U = "unequal":
#   - Each run timing group has its own mean growth rate
# 
# x0 = "unequal":
#   - Each run timing group starts from a different initial state

# Fit model 3
fit3 <- MARSS(Stl_log, model = mod3, silent = TRUE, control = list(maxit = 1000))

#------------------------------------------------------------------------------
# Compare all models using AIC
#------------------------------------------------------------------------------
# Calculate AIC for each model
AIC1 <- fit1$AIC
AIC2 <- fit2$AIC
AIC3 <- fit3$AIC

# Calculate AICc (corrected AIC) for each model
AICc1 <- fit1$AICc
AICc2 <- fit2$AICc
AICc3 <- fit3$AICc

# Create a comparison table
model_table <- data.frame(
  Model = c("H1: All populations act the same", 
            "H2: Each Region acts differently", 
            "H3: Each Run acts differently"),
  AIC = c(AIC1, AIC2, AIC3),
  AICc = c(AICc1, AICc2, AICc3),
  num_params = c(fit1$num.params, fit2$num.params, fit3$num.params),
  logLik = c(fit1$logLik, fit2$logLik, fit3$logLik)
)

# Calculate delta AICc and weights
model_table$delta_AICc <- model_table$AICc - min(model_table$AICc)
model_table$AICc_weight <- exp(-0.5 * model_table$delta_AICc) / sum(exp(-0.5 * model_table$delta_AICc))

# Sort by AICc
model_table <- model_table[order(model_table$AICc), ]

# Print the table
print(model_table)

# Identify best model
best_model_idx <- which.min(c(AICc1, AICc2, AICc3))
best_model_name <- c("H1: All populations act the same", 
                     "H2: Each Region acts differently", 
                     "H3: Each Run acts differently")[best_model_idx]

cat("\nThe best model according to AICc is:", best_model_name, "\n")

# Calculate AICc differences for interpretation
cat("\nInterpretation guideline for delta_AICc:\n")
cat("delta_AICc < 2: Models have substantial support\n")
cat("delta_AICc 4-7: Models have considerably less support\n")
cat("delta_AICc > 10: Models have essentially no support\n")


```







esa.salmon

### Get the steelhead data 


### Fill in the missing years


### Look to see if there appears to be trends. 
# could be grouped by Season 
# could be grouped by Region 
##### Upper , lower 
##### Distance Based 


# General Questions

Each group has the same general tasks, but you will adapt them as you work on the data.

1. Create estimates of spawner abundance for all missing years and provide estimates of the decline from the historical abundance.

2. Evaluate support for the major population groups. Are the populations in the groups more correlated than outside the groups?

3. Evaluate the evidence of cycling in the data. *We will talk about how to do this on the Tuesday after lab.*


# Methods

Address the following in your methods

* Describe your assumptions about the x and how the data time series are related to x.

   - How are the x and y (data) related? 1 x for 1 y or will you assume 1 x for all y or 1 x for each major population group? How will you choose? 
   - What will you assume about the U for the x's?
   - What will you assume about the Q matrix?
   
* Write out your assumptions as different models **in matrix form**, fit each and then compare these with AIC or AICc.

* Do your estimates differ depending on the assumptions you make about the structure of the data, i.e. you assumptions about the x's, Q, and U.

### Tips

**Simplify**

If your ESU has many populations, start with a smaller set of 4-7 populations.

**Assumptions**

You can assume that `R="diagonal and equal"` and `A="scaling"`. Assume that "historical" means the earliest years available for your group.

**States**

Your abundance estimate is the "x" or "state" estimates. You can get this from
```
fit$states
```
or 
```
tsSmooth(fit)
```
where `fit` is from `fit <- MARSS()`

**plotting**

Estimate of the mean of the spawner counts based on your x model.
```
autoplot(fit, plot.type="fitted.ytT")
```

**diagnostics**
```
autoplot(fit, plot.type="residuals")
```

# Results

# Discussion

# Description of each team member's contributions

Example: "All team members helped decide on the goal and ran the analyses for the individual regions. Team members 2 & 3 wrote most of the code for the analysis of the regions. Team member 4 researched approaches for measuring accuracy of forecasts in [Hyndman & Athanasopoulos[OTexts.com/fpp2] and team member 2 added code for that to the methods. Team member 4 also researched tests for stationarity and worked with team member 2 to code that up. Team member 1 worked on the plotting section of the report using and adapting code that team member 3 wrote. All team members helped edit the report and wrote the discussion together."

